---
title: "Day 2: SDTM Domain Structure & Tidyverse Mastery"
subtitle: "Understanding SDTM Domain Classes Through Code"


format:
  html:
    theme:
      light: cosmo

    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css
    
editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

<div style="margin-bottom:12px;">< a target="_blank" href="../index.html" class="resource-link">← Back to Roadmap</a></div>

## Learning Objectives

By the end of Day 2, you will be able to:

1.  Understand the four SDTM domain classes and their purposes
2.  Recognize and interpret SDTM variable naming conventions
3.  Apply advanced tidyverse operations to clinical data
4.  Create meaningful summaries of clinical trial data
5.  Work fluently with `dplyr` and `tidyr` on pharmaverse datasets

------------------------------------------------------------------------

## Introduction

Day 2 builds on the foundations from Day 1. Now that you can load and explore SDTM datasets, we'll dive deeper into:

-   **SDTM Domain Classes**: Understanding why domains are organized the way they are
-   **Variable Naming Conventions**: Decoding the patterns like `--TESTCD`, `--ORRES`, `--DY`
-   **Tidyverse Mastery**: This is 50% of what you'll do daily as a clinical programmer

The pharmaverse packages are built on tidyverse principles, so mastering `dplyr` and `tidyr` is essential for working with `admiral`, `sdtm.oak`, and other clinical programming tools.

------------------------------------------------------------------------

## Package Installation & Loading

### Required Packages

| Package           | Purpose                                          |
|-------------------|--------------------------------------------------|
| `dplyr`           | Data manipulation (filter, select, mutate, etc.) |
| `tidyr`           | Data reshaping (pivot_wider, pivot_longer, etc.) |
| `pharmaversesdtm` | Example SDTM datasets for learning               |

### Load Packages

```{r}
#| label: load-packages

# Load required packages
library(dplyr)
library(tidyr)
library(pharmaversesdtm)
```

### Load SDTM Datasets

We'll work with multiple domains today to understand their structures and relationships.

```{r}
#| label: load-data

# Load all domains we'll use today
data("dm", package = "pharmaversesdtm")
data("ae", package = "pharmaversesdtm")
data("ex", package = "pharmaversesdtm")
data("lb", package = "pharmaversesdtm")
data("vs", package = "pharmaversesdtm")
data("cm", package = "pharmaversesdtm")
data("ds", package = "pharmaversesdtm")
```

------------------------------------------------------------------------

## SDTM Domain Classes

SDTM organizes clinical data into **four domain classes**, each designed for a specific type of information. Understanding these classes helps you know what to expect from each domain.

### The Four Domain Classes

| Class | Purpose | Example Domains | Key Characteristics |
|----|----|----|----|
| **Special Purpose** | Subject-level information | DM, CO, SE, SV | Usually one row per subject |
| **Events** | Discrete occurrences | AE, DS, MH, DV | Multiple rows per subject, start/end |
| **Interventions** | Treatments administered | EX, CM, EC, SU | Dosing info, frequencies, routes |
| **Findings** | Collected measurements/tests | LB, VS, EG, PE | Test codes, results, units, standards |

### Special Purpose Class: DM (Demographics)

The Demographics domain is the **anchor** of all SDTM data. It contains one row per subject and provides the reference information that other domains link to.

```{r}
#| label: explore-dm

# DM structure - note the one row per subject
dm %>%
  dplyr::select(STUDYID, USUBJID, SUBJID, RFSTDTC, RFENDTC, ARM, ACTARM, AGE, SEX, RACE) %>%
  utils::head(5)
```

::: {.callout-note collapse="true"}
## Key DM Variables

-   **STUDYID**: Study Identifier
-   **USUBJID**: Unique Subject Identifier (links all domains)
-   **RFSTDTC**: Reference Start Date/Time (usually first dose)
-   **RFENDTC**: Reference End Date/Time
-   **ARM/ACTARM**: Planned/Actual Treatment Arm
:::

### Events Class: AE (Adverse Events)

Events domains capture discrete occurrences that happen to subjects. Each event has a start and potentially an end.

```{r}
#| label: explore-ae

# AE structure - multiple rows per subject
ae %>%
  dplyr::select(USUBJID, AETERM, AEDECOD, AEBODSYS, AESEV, AESER, AESTDTC, AEENDTC) %>%
  utils::head(10)
```

### Interventions Class: EX (Exposure)

Interventions domains track treatments given to subjects. They include dosing information, routes, and frequencies.

```{r}
#| label: explore-ex

# EX structure - dosing records
ex %>%
  dplyr::select(USUBJID, EXTRT, EXDOSE, EXDOSU, EXROUTE, EXSTDTC, EXENDTC) %>%
  utils::head(10)
```

### Findings Class: LB (Laboratory)

Findings domains contain collected measurements. They have a standardized structure with test codes, results, units, and reference ranges.

```{r}
#| label: explore-lb

# LB structure - lab results with standardization
lb %>%
  dplyr::select(USUBJID, LBTESTCD, LBTEST, LBORRES, LBORRESU, LBSTRESN, LBSTRESU) %>%
  utils::head(10)
```

------------------------------------------------------------------------

## SDTM Variable Naming Conventions

SDTM uses consistent naming patterns across all domains. Understanding these patterns helps you quickly identify variable purposes without documentation.

### The Two-Character Domain Prefix

Every domain-specific variable starts with a **two-character prefix** that identifies the domain:

| Prefix | Domain           | Example Variables           |
|--------|------------------|-----------------------------|
| DM     | Demographics     | DMDTC, DMDY                 |
| AE     | Adverse Events   | AESTDTC, AEENDTC, AESEV     |
| EX     | Exposure         | EXSTDTC, EXDOSE, EXROUTE    |
| LB     | Laboratory       | LBTESTCD, LBORRES, LBSTRESN |
| VS     | Vital Signs      | VSTESTCD, VSORRES, VSSTRESN |
| CM     | Concomitant Meds | CMTRT, CMDOSE, CMSTDTC      |

### Common Variable Suffixes

The suffix tells you what type of information the variable contains:

| Suffix     | Meaning                        | Examples                  |
|------------|--------------------------------|---------------------------|
| `--TERM`   | Reported term (verbatim)       | AETERM, CMTRT             |
| `--DECOD`  | Decoded/standardized term      | AEDECOD, CMDECOD          |
| `--TESTCD` | Test code (short)              | LBTESTCD, VSTESTCD        |
| `--TEST`   | Test name (full)               | LBTEST, VSTEST            |
| `--ORRES`  | Original result (as collected) | LBORRES, VSORRES          |
| `--ORRESU` | Original result units          | LBORRESU, VSORRESU        |
| `--STRESN` | Standardized result (numeric)  | LBSTRESN, VSSTRESN        |
| `--STRESU` | Standardized result units      | LBSTRESU, VSSTRESU        |
| `--STDTC`  | Start date/time                | AESTDTC, EXSTDTC, CMSTDTC |
| `--ENDTC`  | End date/time                  | AEENDTC, EXENDTC, CMENDTC |
| `--DY`     | Study day                      | AESTDY, AEENDY, EXSTDY    |
| `--SEV`    | Severity                       | AESEV                     |
| `--SER`    | Serious (Y/N)                  | AESER                     |
| `--REL`    | Relationship/Causality         | AEREL                     |

### Understanding the Pattern: Findings Class

Let's examine how the naming convention works in the Findings class (LB, VS):

```{r}
#| label: naming-pattern-lb

# The Findings class pattern in LB
lb %>%
  dplyr::select(
    USUBJID,
    LBTESTCD,   # Test code (short identifier)
    LBTEST,     # Test name (full description)
    LBORRES,    # Original result (as collected)
    LBORRESU,   # Original units
    LBSTRESC,   # Standardized result (character)
    LBSTRESN,   # Standardized result (numeric)
    LBSTRESU    # Standardized units
  ) %>%
  dplyr::filter(LBTESTCD == "ALT") %>%
  utils::head(5)
```

::: {.callout-tip collapse="true"}
## Why Two Result Variables?

-   **ORRES/ORRESU**: Original result as collected (might be "\< 5", "\>100", or character values)
-   **STRESN/STRESU**: Standardized numeric result for analysis (handles unit conversions)

This separation preserves the original data while providing analysis-ready values.
:::

------------------------------------------------------------------------

## Tidyverse Mastery: 10 Essential Transformations

The deliverable for Day 2 is to demonstrate 10 tidyverse transformations. Let's work through them systematically.

### Transformation 1: `filter()` with Multiple Conditions

Filter AEs that are both serious AND severe.

```{r}
#| label: transform-1-filter

# Complex filtering: Serious AND Severe AEs
serious_severe_ae <- ae %>%
  dplyr::filter(AESER == "Y" & AESEV == "SEVERE")

cat("Total AEs:", nrow(ae), "\n")
cat("Serious + Severe AEs:", nrow(serious_severe_ae), "\n")

# Preview
serious_severe_ae %>%
  dplyr::select(USUBJID, AEDECOD, AESEV, AESER) %>%
  utils::head()
```

### Transformation 2: `filter()` with `%in%` for Multiple Values

Filter specific test codes from the lab data.

```{r}
#| label: transform-2-filter-in

# Filter for liver function tests
liver_tests <- c("ALT", "AST", "BILI", "ALP")

lft_data <- lb %>%
  dplyr::filter(LBTESTCD %in% liver_tests)

cat("Total LB records:", nrow(lb), "\n")
cat("Liver function tests:", nrow(lft_data), "\n")

# Count by test
lft_data %>%
  dplyr::count(LBTESTCD, LBTEST)
```

### Transformation 3: `select()` with Helper Functions

Use select helpers to choose columns by pattern.

```{r}
#| label: transform-3-select-helpers

# Select all date columns from AE
ae_dates <- ae %>%
  dplyr::select(USUBJID, dplyr::ends_with("DTC"))

names(ae_dates)
```

```{r}
#| label: transform-3b-select-contains

# Select all result columns from LB
lb_results <- lb %>%
  dplyr::select(USUBJID, LBTESTCD, dplyr::contains("RES"))

names(lb_results)
```

### Transformation 4: `mutate()` with `case_when()`

Create a severity score based on AE severity.

```{r}
#| label: transform-4-case-when

# Create numeric severity score
ae_scored <- ae %>%
  dplyr::mutate(
    AESEV_N = dplyr::case_when(
      AESEV == "MILD"     ~ 1L,
      AESEV == "MODERATE" ~ 2L,
      AESEV == "SEVERE"   ~ 3L,
      TRUE                ~ NA_integer_
    )
  )

# Check the mapping
ae_scored %>%
  dplyr::distinct(AESEV, AESEV_N) %>%
  dplyr::arrange(AESEV_N)
```

### Transformation 5: `mutate()` with `across()`

Apply transformations to multiple columns at once.

```{r}
#| label: transform-5-across

# Convert all character date columns to uppercase (demonstration)
ae_upper <- ae %>%
  dplyr::mutate(
    dplyr::across(
      dplyr::where(is.character),
      ~ toupper(.x)
    )
  )

# This is useful for standardizing text fields
ae_upper %>%
  dplyr::select(USUBJID, AETERM, AEDECOD) %>%
  utils::head(3)
```

### Transformation 6: `group_by()` + `summarise()` - AE Summary by SOC

This is the key deliverable: Create an AE summary by System Organ Class.

```{r}
#| label: transform-6-ae-summary-soc

# AE Summary by System Organ Class (SOC)
ae_soc_summary <- ae %>%
  dplyr::group_by(AEBODSYS) %>%
  dplyr::summarise(
    n_events = dplyr::n(),
    n_subjects = dplyr::n_distinct(USUBJID),
    n_serious = sum(AESER == "Y", na.rm = TRUE),
    n_severe = sum(AESEV == "SEVERE", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::arrange(dplyr::desc(n_events))

# Display the summary
ae_soc_summary
```

### Transformation 7: `group_by()` + `summarise()` - By Treatment Arm

Summarize by treatment arm using joined data.

```{r}
#| label: transform-7-by-arm

# First, join AE with DM to get treatment arm
ae_with_arm <- ae %>%
  dplyr::inner_join(
    dm %>% dplyr::select(USUBJID, ARM),
    by = "USUBJID"
  )

# Now summarize by treatment arm
ae_arm_summary <- ae_with_arm %>%
  dplyr::group_by(ARM) %>%
  dplyr::summarise(
    n_subjects_with_ae = dplyr::n_distinct(USUBJID),
    total_aes = dplyr::n(),
    mean_aes_per_subject = round(dplyr::n() / dplyr::n_distinct(USUBJID), 1),
    .groups = "drop"
  )

ae_arm_summary
```

### Transformation 8: `pivot_wider()` - Reshape Lab Data

Convert long lab data to wide format (one row per subject/visit with columns for each test).

```{r}
#| label: transform-8-pivot-wider

# Pivot LB data to wide format
lb_wide <- lb %>%
  dplyr::filter(LBTESTCD %in% c("ALT", "AST", "BILI")) %>%
  dplyr::select(USUBJID, VISIT, LBTESTCD, LBSTRESN) %>%
  tidyr::pivot_wider(
    names_from = LBTESTCD,
    values_from = LBSTRESN,
    values_fn = mean  # Handle multiple values per cell
  )

utils::head(lb_wide, 10)
```

### Transformation 9: `pivot_longer()` - Reshape Wide to Long

Convert wide data back to long format (useful for plotting).

```{r}
#| label: transform-9-pivot-longer

# Pivot back to long format
lb_long <- lb_wide %>%
  tidyr::pivot_longer(
    cols = c(ALT, AST, BILI),
    names_to = "LBTESTCD",
    values_to = "LBSTRESN"
  ) %>%
  dplyr::filter(!is.na(LBSTRESN))

utils::head(lb_long, 10)
```

### Transformation 10: `left_join()` + `anti_join()` - Data Quality Check

Find subjects with exposure but no adverse events (data quality check).

```{r}
#| label: transform-10-joins

# Subjects who were dosed (in EX)
subjects_dosed <- ex %>%
  dplyr::distinct(USUBJID)

# Subjects with AEs
subjects_with_ae <- ae %>%
  dplyr::distinct(USUBJID)

# Find dosed subjects with NO AEs
dosed_no_ae <- subjects_dosed %>%
  dplyr::anti_join(subjects_with_ae, by = "USUBJID")

cat("Subjects dosed:", nrow(subjects_dosed), "\n")
cat("Subjects with AEs:", nrow(subjects_with_ae), "\n")
cat("Dosed subjects with NO AEs:", nrow(dosed_no_ae), "\n")
```

------------------------------------------------------------------------

## Advanced: Combining Multiple Operations

Real clinical programming often chains many operations together. Here's a complete example:

```{r}
#| label: combined-example

# Complete analysis: AE summary by SOC and Treatment Arm
complete_ae_summary <- ae %>%
  # Join with demographics
  dplyr::inner_join(
    dm %>% dplyr::select(USUBJID, ARM, AGE, SEX),
    by = "USUBJID"
  ) %>%
  # Create severity score
  dplyr::mutate(
    AESEV_N = dplyr::case_when(
      AESEV == "MILD"     ~ 1L,
      AESEV == "MODERATE" ~ 2L,
      AESEV == "SEVERE"   ~ 3L,
      TRUE                ~ NA_integer_
    )
  ) %>%
  # Group and summarize
  dplyr::group_by(ARM, AEBODSYS) %>%
  dplyr::summarise(
    n_events = dplyr::n(),
    n_subjects = dplyr::n_distinct(USUBJID),
    n_serious = sum(AESER == "Y", na.rm = TRUE),
    mean_severity = round(mean(AESEV_N, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  # Filter to most common SOCs
  dplyr::filter(n_events >= 5) %>%
  # Sort
dplyr::arrange(ARM, dplyr::desc(n_events))

# Display top results
utils::head(complete_ae_summary, 15)
```

------------------------------------------------------------------------

## Day 2 Deliverable Summary

Today you completed the following:

| Task                                                | Status |
|-----------------------------------------------------|--------|
| Understood SDTM domain classes                      | Done   |
| Learned variable naming conventions                 | Done   |
| Transformation 1: filter() with multiple conditions | Done   |
| Transformation 2: filter() with %in%                | Done   |
| Transformation 3: select() with helper functions    | Done   |
| Transformation 4: mutate() with case_when()         | Done   |
| Transformation 5: mutate() with across()            | Done   |
| Transformation 6: group_by + summarise (AE by SOC)  | Done   |
| Transformation 7: Summarize by treatment arm        | Done   |
| Transformation 8: pivot_wider()                     | Done   |
| Transformation 9: pivot_longer()                    | Done   |
| Transformation 10: anti_join() for data QC          | Done   |

------------------------------------------------------------------------

## Key Takeaways

1.  **SDTM has four domain classes**: Special Purpose, Events, Interventions, Findings
2.  **Variable naming is systematic**: Prefix (domain) + suffix (content type)
3.  **`--TESTCD` vs `--TEST`**: Short code vs full name
4.  **`--ORRES` vs `--STRESN`**: Original vs standardized results
5.  **Tidyverse is foundational**: 50% of clinical programming is dplyr/tidyr operations

------------------------------------------------------------------------

## Resources

-   [CDISC SDTM Implementation Guide v3.4](https://www.cdisc.org/standards/foundational/sdtm) - Official SDTM documentation
-   [tidyverse.org](https://www.tidyverse.org/) - Complete tidyverse documentation
-   [pharmaverse blog: 'The pharmaverse (hi)story'](https://pharmaverse.github.io/blog/) - Background on pharmaverse development
-   [dplyr cheatsheet](https://rstudio.github.io/cheatsheets/data-transformation.pdf) - Quick reference for dplyr

------------------------------------------------------------------------

## What's Next?

In **Day 3**, we'll explore:

-   Controlled Terminology and why it matters for submissions
-   MedDRA coding and creating lookup tables
-   CDISC CT concepts: AESEV, AEREL, AESER values
-   How metacore + metatools automate CT validation

< a target="_blank" href="https://visitorbadge.io/status?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse"><img src="https://api.visitorbadge.io/api/visitors?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse&label=Visits%20till%20date&countColor=%23263759&style=plastic&labelStyle=none" /></a>
