---
title: "Day 3: Controlled Terminology & MedDRA Coding"
subtitle: "Building AE Codelists in R"


format:
  html:
    theme:
      light: cosmo

    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css
    
editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

<div style="margin-bottom:12px;"><a href="../index.html" class="resource-link">← Back to Roadmap</a></div>

## Learning Objectives

By the end of Day 3, you will be able to:

1.  Understand the importance of **Controlled Terminology (CT)** in regulatory submissions
2.  Create **lookup tables** (codelists) in R for standardized coding
3.  Perform **MedDRA coding** operations (mapping Preferred Terms to System Organ Classes)
4.  Use `left_join()` effectively to apply standardized codes to clinical data
5.  Validate that all terms have been correctly coded

------------------------------------------------------------------------

## Introduction

### Why Controlled Terminology Matters

In clinical trials, consistency is key. Regulatory agencies like the FDA need data to be standardized so they can analyze it across different studies. **Controlled Terminology (CT)** ensures that everyone uses the same words to describe the same things.

For example, without CT, severity might be recorded as:

-   "Mild", "Mod", "Severe"
-   "1", "2", "3"
-   "Light", "Moderate", "Heavy"

With CDISC CT, we must use: **"MILD", "MODERATE", "SEVERE"**.

### The Role of MedDRA

**MedDRA** (Medical Dictionary for Regulatory Activities) is the standard for coding medical terms. It organizes adverse events into a hierarchy:

```         
┌─────────────────────────────────────────────────────────────────┐
│  SOC  │ System Organ Class (e.g., "Gastrointestinal disorders") │
├───────┴─────────────────────────────────────────────────────────┤
│  HLGT │ High Level Group Term                                   │
├─────────────────────────────────────────────────────────────────┤
│  HLT  │ High Level Term                                         │
├─────────────────────────────────────────────────────────────────┤
│  PT   │ Preferred Term (e.g., "Nausea") → Used in AEDECOD       │
├─────────────────────────────────────────────────────────────────┤
│  LLT  │ Lowest Level Term (e.g., "Feeling queasy")              │
└─────────────────────────────────────────────────────────────────┘
```

In SDTM:

-   **AEDECOD** = Preferred Term (PT)
-   **AEBODSYS** = System Organ Class (SOC)

------------------------------------------------------------------------

## Package Installation & Loading

### Required Packages

| Package           | Purpose                              |
|-------------------|--------------------------------------|
| `dplyr`           | Data manipulation (joins, mutations) |
| `pharmaversesdtm` | Source of AE data                    |

### Load Packages

```{r}
#| label: load-packages

library(dplyr)
library(pharmaversesdtm)
```

### Load AE Data

```{r}
#| label: load-data

data("ae", package = "pharmaversesdtm")

# Preview the variables involved in coding
ae %>%
  select(USUBJID, AETERM, AEDECOD, AEBODSYS) %>%
  head()
```

------------------------------------------------------------------------

## Part 1: Building a Complete MedDRA Lookup Table

In a real environment, you would have access to the full MedDRA dictionary files. For this exercise, we will create a **complete** lookup table covering all terms in our dataset.

### Step 1: Identify All Unique Preferred Terms

First, let's see **all** unique Preferred Terms (AEDECOD) and their corresponding SOCs in our dataset.

```{r}
#| label: unique-terms

# Get all unique PT-SOC combinations from the source data
unique_terms <- ae %>%
  distinct(AEDECOD, AEBODSYS) %>%
  arrange(AEBODSYS, AEDECOD)

cat("Total unique Preferred Terms:", nrow(unique_terms), "\n\n")
print(unique_terms)
```

### Step 2: Create the Complete Lookup Table

Now we create a comprehensive lookup table that maps **all** Preferred Terms to their SOCs. This is the dictionary we would use in production.

```{r}
#| label: create-lookup

# Create a complete MedDRA-style lookup table
# In production, this comes from official MedDRA files
meddra_lookup <- tribble(
  ~AEDECOD,                            ~SOC_CODED,
  # CARDIAC DISORDERS
  "ATRIAL FIBRILLATION",               "CARDIAC DISORDERS",
  "ATRIAL FLUTTER",                    "CARDIAC DISORDERS",
  "BUNDLE BRANCH BLOCK LEFT",          "CARDIAC DISORDERS",
  "BUNDLE BRANCH BLOCK RIGHT",         "CARDIAC DISORDERS",
  "MYOCARDIAL INFARCTION",             "CARDIAC DISORDERS",
  "SINUS BRADYCARDIA",                 "CARDIAC DISORDERS",
  "SINUS TACHYCARDIA",                 "CARDIAC DISORDERS",
  
  # GASTROINTESTINAL DISORDERS
  "ABDOMINAL DISCOMFORT",              "GASTROINTESTINAL DISORDERS",
  "ABDOMINAL PAIN",                    "GASTROINTESTINAL DISORDERS",
  "CONSTIPATION",                      "GASTROINTESTINAL DISORDERS",
  "DIARRHOEA",                         "GASTROINTESTINAL DISORDERS",
  "DYSPEPSIA",                         "GASTROINTESTINAL DISORDERS",
  "FLATULENCE",                        "GASTROINTESTINAL DISORDERS",
  "NAUSEA",                            "GASTROINTESTINAL DISORDERS",
  "VOMITING",                          "GASTROINTESTINAL DISORDERS",
  
  # GENERAL DISORDERS AND ADMINISTRATION SITE CONDITIONS
  "APPLICATION SITE DERMATITIS",       "GENERAL DISORDERS AND ADMINISTRATION SITE CONDITIONS",
  "APPLICATION SITE ERYTHEMA",         "GENERAL DISORDERS AND ADMINISTRATION SITE CONDITIONS",
  "APPLICATION SITE IRRITATION",       "GENERAL DISORDERS AND ADMINISTRATION SITE CONDITIONS",
  "APPLICATION SITE PRURITUS",         "GENERAL DISORDERS AND ADMINISTRATION SITE CONDITIONS",
  "FATIGUE",                           "GENERAL DISORDERS AND ADMINISTRATION SITE CONDITIONS",
  "OEDEMA PERIPHERAL",                 "GENERAL DISORDERS AND ADMINISTRATION SITE CONDITIONS",
  "PYREXIA",                           "GENERAL DISORDERS AND ADMINISTRATION SITE CONDITIONS",
  
  # INFECTIONS AND INFESTATIONS
  "BRONCHITIS",                        "INFECTIONS AND INFESTATIONS",
  "GASTROENTERITIS",                   "INFECTIONS AND INFESTATIONS",
  "NASOPHARYNGITIS",                   "INFECTIONS AND INFESTATIONS",
  "UPPER RESPIRATORY TRACT INFECTION", "INFECTIONS AND INFESTATIONS",
  "URINARY TRACT INFECTION",           "INFECTIONS AND INFESTATIONS",
  
  # METABOLISM AND NUTRITION DISORDERS
  "DECREASED APPETITE",                "METABOLISM AND NUTRITION DISORDERS",
  "HYPERCHOLESTEROLAEMIA",             "METABOLISM AND NUTRITION DISORDERS",
  "HYPERGLYCAEMIA",                    "METABOLISM AND NUTRITION DISORDERS",
  "HYPOGLYCAEMIA",                     "METABOLISM AND NUTRITION DISORDERS",
  
  # MUSCULOSKELETAL AND CONNECTIVE TISSUE DISORDERS
  "ARTHRALGIA",                        "MUSCULOSKELETAL AND CONNECTIVE TISSUE DISORDERS",
  "BACK PAIN",                         "MUSCULOSKELETAL AND CONNECTIVE TISSUE DISORDERS",
  "MUSCLE SPASMS",                     "MUSCULOSKELETAL AND CONNECTIVE TISSUE DISORDERS",
  "MYALGIA",                           "MUSCULOSKELETAL AND CONNECTIVE TISSUE DISORDERS",
  "PAIN IN EXTREMITY",                 "MUSCULOSKELETAL AND CONNECTIVE TISSUE DISORDERS",
  
  # NERVOUS SYSTEM DISORDERS
  "DIZZINESS",                         "NERVOUS SYSTEM DISORDERS",
  "HEADACHE",                          "NERVOUS SYSTEM DISORDERS",
  "LETHARGY",                          "NERVOUS SYSTEM DISORDERS",
  "MIGRAINE",                          "NERVOUS SYSTEM DISORDERS",
  "PARAESTHESIA",                      "NERVOUS SYSTEM DISORDERS",
  "SYNCOPE",                           "NERVOUS SYSTEM DISORDERS",
  "TREMOR",                            "NERVOUS SYSTEM DISORDERS",
  
  # PSYCHIATRIC DISORDERS
  "ANXIETY",                           "PSYCHIATRIC DISORDERS",
  "DEPRESSION",                        "PSYCHIATRIC DISORDERS",
  "INSOMNIA",                          "PSYCHIATRIC DISORDERS",
  
  # RESPIRATORY, THORACIC AND MEDIASTINAL DISORDERS
  "COUGH",                             "RESPIRATORY, THORACIC AND MEDIASTINAL DISORDERS",
  "DYSPNOEA",                          "RESPIRATORY, THORACIC AND MEDIASTINAL DISORDERS",
  "NASAL CONGESTION",                  "RESPIRATORY, THORACIC AND MEDIASTINAL DISORDERS",
  "OROPHARYNGEAL PAIN",                "RESPIRATORY, THORACIC AND MEDIASTINAL DISORDERS",
  
  # SKIN AND SUBCUTANEOUS TISSUE DISORDERS
  "DERMATITIS CONTACT",                "SKIN AND SUBCUTANEOUS TISSUE DISORDERS",
  "ERYTHEMA",                          "SKIN AND SUBCUTANEOUS TISSUE DISORDERS",
  "HYPERHIDROSIS",                     "SKIN AND SUBCUTANEOUS TISSUE DISORDERS",
  "PRURITUS",                          "SKIN AND SUBCUTANEOUS TISSUE DISORDERS",
  "RASH",                              "SKIN AND SUBCUTANEOUS TISSUE DISORDERS",
  "URTICARIA",                         "SKIN AND SUBCUTANEOUS TISSUE DISORDERS",
  
  # VASCULAR DISORDERS
  "HYPERTENSION",                      "VASCULAR DISORDERS",
  "HYPOTENSION",                       "VASCULAR DISORDERS",
  "HOT FLUSH",                         "VASCULAR DISORDERS"
)

cat("Lookup table contains", nrow(meddra_lookup), "term mappings\n")
```

------------------------------------------------------------------------

## Part 2: Applying Codes with Joins

Now, we will simulate a scenario where we have the terms but *missing* the Body System (SOC). We will use our lookup table to fill it in.

### Step 3: Simulate Uncoded Data

Let's create a dataset that has `AEDECOD` but is missing `AEBODSYS`.

```{r}
#| label: simulate-uncoded

# Create uncoded dataset by dropping AEBODSYS
ae_uncoded <- ae %>%
  select(USUBJID, AESEQ, AETERM, AEDECOD, AESEV, AEREL, AESER)

cat("Uncoded dataset:", nrow(ae_uncoded), "records\n")
head(ae_uncoded)
```

### Step 4: Apply Coding via Left Join

We use `left_join()` to attach the SOC information from our lookup table to the clinical data based on the matching Preferred Term (`AEDECOD`).

```{r}
#| label: apply-coding

# Join the lookup table to the data
ae_coded <- ae_uncoded %>%
  left_join(meddra_lookup, by = "AEDECOD")

# Check the results
ae_coded %>%
  select(USUBJID, AEDECOD, SOC_CODED) %>%
  head(10)
```

::: {.callout-important collapse="true"}
## Why Left Join?

We use `left_join` because we want to keep **all** rows from our clinical data (`ae_uncoded`), even if a matching code isn't found in the dictionary. If we used `inner_join`, we would lose data for any term not in our lookup table!
:::

------------------------------------------------------------------------

## Part 3: Validating the Coding

After applying codes, it is critical to verify that:

1.  Every term got a code (no `NA` values in the new column).
2.  The coding is accurate.

### Step 5: Generate Validation Report

Let's create a comprehensive validation report.

```{r}
#| label: validate-coding

# Count coded vs uncoded
validation_summary <- ae_coded %>%
  summarise(
    Total_Records = n(),
    Coded = sum(!is.na(SOC_CODED)),
    Uncoded = sum(is.na(SOC_CODED)),
    Pct_Coded = round(100 * Coded / Total_Records, 1)
  )

print(validation_summary)
```

```{r}
#| label: find-uncoded

# Identify which terms are still uncoded
uncoded_terms <- ae_coded %>%
  filter(is.na(SOC_CODED)) %>%
  distinct(AEDECOD) %>%
  arrange(AEDECOD)

if (nrow(uncoded_terms) > 0) {
  cat("⚠️ WARNING: The following", nrow(uncoded_terms), "terms need to be added to the lookup table:\n\n")
  print(uncoded_terms)
} else {
  cat("✅ SUCCESS: All terms were successfully coded!\n")
}
```

### Step 6: Compare Against Original

Let's verify our coding matches the original data.

```{r}
#| label: verify-accuracy

# Join original SOC for comparison
coding_check <- ae %>%
  select(USUBJID, AESEQ, AEDECOD, AEBODSYS_ORIG = AEBODSYS) %>%
  inner_join(
    ae_coded %>% select(USUBJID, AESEQ, SOC_CODED),
    by = c("USUBJID", "AESEQ")
  ) %>%
  mutate(
    Match = AEBODSYS_ORIG == SOC_CODED
  )

# Summary
cat("Coding accuracy check:\n")
cat("Matching:", sum(coding_check$Match, na.rm = TRUE), "\n")
cat("Mismatched:", sum(!coding_check$Match, na.rm = TRUE), "\n")
```

------------------------------------------------------------------------

## Part 4: Other CDISC Controlled Terminology

Beyond MedDRA, we must ensure other variables match CDISC standards.

### Severity (AESEV)

**Valid CDISC CT values**: "MILD", "MODERATE", "SEVERE"

```{r}
#| label: check-severity

# Check for non-standard values
ae %>%
  count(AESEV, name = "Count") %>%
  mutate(
    Is_Standard = AESEV %in% c("MILD", "MODERATE", "SEVERE"),
    Status = ifelse(Is_Standard, "✓ Valid", "✗ Invalid")
  )
```

### Seriousness (AESER)

**Valid CDISC CT values**: "Y", "N"

```{r}
#| label: check-serious

ae %>%
  count(AESER, name = "Count") %>%
  mutate(
    Is_Standard = AESER %in% c("Y", "N"),
    Status = ifelse(Is_Standard, "✓ Valid", "✗ Invalid")
  )
```

### Causality (AEREL)

Causality can vary by study, but common values include:

```{r}
#| label: check-rel

ae %>%
  count(AEREL, name = "Count") %>%
  arrange(desc(Count))
```

------------------------------------------------------------------------

## 🎯 Practice Exercise

### Your Turn: Complete the Missing Mappings

If the validation above showed uncoded terms, add them to the lookup table below and re-run the validation.

```{r}
#| label: practice-exercise
#| eval: false

# TODO: Add any missing terms from the validation report
additional_terms <- tribble(
  ~AEDECOD,         ~SOC_CODED,
  # Add your mappings here:
  # "TERM_NAME",    "SYSTEM ORGAN CLASS",
)

# Combine with original lookup
complete_lookup <- bind_rows(meddra_lookup, additional_terms)

# Re-apply coding
ae_recoded <- ae_uncoded %>%
  left_join(complete_lookup, by = "AEDECOD")

# Check for remaining uncoded
ae_recoded %>%
  filter(is.na(SOC_CODED)) %>%
  distinct(AEDECOD)
```

------------------------------------------------------------------------

## Deliverable Summary

Today you completed the following:

| Task                                           | Status |
|------------------------------------------------|--------|
| Created MedDRA lookup table with 50+ terms     | ✓ Done |
| Applied coding using `left_join()`             | ✓ Done |
| Generated validation report (coded vs uncoded) | ✓ Done |
| Verified coding accuracy against source        | ✓ Done |
| Checked CDISC CT for AESEV, AESER, AEREL       | ✓ Done |

------------------------------------------------------------------------

## Key Takeaways

1.  **Consistency is Mandatory**: Regulatory submissions require strict adherence to Controlled Terminology (CT).
2.  **MedDRA Hierarchy**: PT (Preferred Term) → SOC (System Organ Class) is the key mapping for AE data.
3.  **Joins for Coding**: `left_join()` is the primary tool for applying codes from a dictionary to a dataset.
4.  **Validation is Critical**: Always check for `NA` values after a join to identify terms that didn't match.
5.  **Document Everything**: Keep track of which version of MedDRA you're using.

------------------------------------------------------------------------

## Resources

-   [CDISC Controlled Terminology Browser](https://www.cancer.gov/research/resources/terminology/cdisc) - Browse official values
-   [MedDRA Hierarchy Guide](https://www.meddra.org/how-to-use/basics/hierarchy) - Understanding the coding levels
-   [Metacore Documentation](https://atorus-research.github.io/metacore/) - R package for managing metadata and CT

------------------------------------------------------------------------

## What's Next?

In **Day 4**, we will focus on **Clinical Date Derivations**. You will learn:

-   How to parse ISO 8601 dates with `lubridate`
-   Calculating Study Days (`--DY`) with the "No Day 0" rule
-   Handling partial dates (e.g., "2023-05")
-   Creating reusable derivation functions

___
