---
title: "Day 4: Clinical Date Derivations with lubridate"
subtitle: "Study Day, Duration, and Partial Date Imputation"
author: "Indraneel Chakraborty"
date: today
format:
  html:
    theme:
      light: cosmo

    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css
    
editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

<div style="margin-bottom:12px;"><a href="../index.html" class="resource-link">← Back to Roadmap</a></div>

## Learning Objectives

By the end of Day 4, you will be able to:

1.  Parse ISO 8601 clinical dates using **`lubridate`**
2.  Calculate **Study Days** (`--DY` variables) adhering to CDISC rules (no Day 0)
3.  Calculate **Durations** (`--DUR`) between start and end dates
4.  Handle **Partial Dates** (e.g., "2023-05") using imputation strategies
5.  Create **reusable functions** for date derivations

------------------------------------------------------------------------

## Introduction

### Dates in CDISC (ISO 8601)

CDISC standards use the **ISO 8601** format for dates and times: `YYYY-MM-DDThh:mm:ss`.

| Format          | Example            | Description           |
|-----------------|--------------------|-----------------------|
| Complete date   | `2024-01-15`       | Full date             |
| Date + time     | `2024-01-15T09:30` | Date with time        |
| Year-Month only | `2024-01`          | Partial (missing day) |
| Year only       | `2024`             | Partial (month + day) |

### The "No Day 0" Rule

CDISC has a specific rule for calculating Study Days:

-   **Day 1** = Reference date (usually first dose)
-   **Day -1** = The day before reference date
-   **There is NO Day 0**

### The Role of `lubridate`

The `lubridate` package (part of tidyverse) makes working with dates in R easy. It handles parsing, arithmetic, and formatting much better than base R functions.

------------------------------------------------------------------------

## Package Installation & Loading

### Required Packages

| Package           | Purpose                |
|-------------------|------------------------|
| `dplyr`           | Data manipulation      |
| `lubridate`       | Date/time manipulation |
| `pharmaversesdtm` | Source data            |

### Load Packages

```{r}
#| label: load-packages

library(dplyr)
library(lubridate)
library(pharmaversesdtm)

# Load the data we'll use
data("ae", package = "pharmaversesdtm")
data("dm", package = "pharmaversesdtm")
```

------------------------------------------------------------------------

## Part 1: Basic Date Parsing

First, let's learn how to convert character strings (which SDTM dates always are) into actual Date objects in R.

### Parsing Complete Dates

```{r}
#| label: parsing-dates

# Use ymd() for Year-Month-Day strings
date_str <- "2024-01-15"
date_obj <- ymd(date_str)

cat("Original string:", date_str, "\n")
cat("Parsed date:", as.character(date_obj), "\n")
cat("Class:", class(date_obj), "\n")
```

### Parsing Dates with Times

```{r}
#| label: parsing-datetime

# ymd_hms() for full datetime
datetime_full <- "2024-01-15T14:30:00"
ymd_hms(datetime_full)

# ymd_hm() when seconds are missing
datetime_short <- "2024-01-15T14:30"
ymd_hm(datetime_short)
```

### Extracting Date Components

```{r}
#| label: extract-components

sample_date <- ymd("2024-03-15")

cat("Year:", year(sample_date), "\n")
cat("Month:", month(sample_date), "\n")
cat("Day:", day(sample_date), "\n")
cat("Weekday:", wday(sample_date, label = TRUE), "\n")
```

------------------------------------------------------------------------

## Part 2: Calculating Study Day (`--DY`)

The Study Day (`--DY`) indicates the number of days from the Reference Start Date (`RFSTDTC`).

### The Formula

```         
If event_date >= reference_date:
    Study Day = (event_date - reference_date) + 1

If event_date < reference_date:
    Study Day = (event_date - reference_date)
```

### Creating a Reusable Function

This function can be reused across all your SDTM and ADaM programs.

```{r}
#| label: derive-study-day-function

#' Derive Study Day following CDISC "No Day 0" convention
#' 
#' @param event_date Date of the event
#' @param reference_date Reference start date (usually RFSTDTC)
#' @return Integer study day (positive = on/after ref, negative = before ref)
derive_study_day <- function(event_date, reference_date) {
  # Calculate raw difference in days
  diff_days <- as.numeric(event_date - reference_date)
  
  # Apply "No Day 0" rule
  study_day <- ifelse(diff_days >= 0, diff_days + 1, diff_days)
  
  return(as.integer(study_day))
}

# Test the function
test_ref <- ymd("2024-01-01")
test_dates <- c("2023-12-30", "2023-12-31", "2024-01-01", "2024-01-02", "2024-01-05")

data.frame(
  Event_Date = test_dates,
  Reference = "2024-01-01",
  Study_Day = derive_study_day(ymd(test_dates), test_ref)
)
```

::: callout-note
## Notice the "No Day 0"

Look at the output above: Day -2, Day -1, **Day 1**, Day 2, Day 5. There is no Day 0!
:::

### Applying to Real Data: Derive `AESTDY`

```{r}
#| label: calc-aestdy

# Merge RFSTDTC from DM into AE
ae_with_ref <- ae %>%
  select(USUBJID, AESEQ, AESTDTC, AEENDTC) %>%
  inner_join(
    dm %>% select(USUBJID, RFSTDTC),
    by = "USUBJID"
  ) %>%
  # Filter to complete dates only (length >= 10)
  filter(nchar(AESTDTC) >= 10 & nchar(RFSTDTC) >= 10)

# Apply our function
ae_study_days <- ae_with_ref %>%
  mutate(
    AE_DT = ymd(substring(AESTDTC, 1, 10)),
    REF_DT = ymd(substring(RFSTDTC, 1, 10)),
    AESTDY = derive_study_day(AE_DT, REF_DT)
  )

# Preview
ae_study_days %>%
  select(USUBJID, AESTDTC, RFSTDTC, AESTDY) %>%
  head(10)
```

------------------------------------------------------------------------

## Part 3: Calculating Duration (`--DUR`)

Duration is typically: $EndDate - StartDate + 1$ (inclusive of both days).

### Duration in Days

```{r}
#| label: calc-duration

# Filter for records with both start and end dates
ae_dur <- ae %>%
  select(USUBJID, AESEQ, AESTDTC, AEENDTC) %>%
  filter(nchar(AESTDTC) >= 10 & nchar(AEENDTC) >= 10) %>%
  mutate(
    START_DT = ymd(substring(AESTDTC, 1, 10)),
    END_DT = ymd(substring(AEENDTC, 1, 10)),
    
    # Duration in days (inclusive)
    AEDUR = as.integer(END_DT - START_DT) + 1L
  )

ae_dur %>%
  select(USUBJID, AESTDTC, AEENDTC, AEDUR) %>%
  head(10)
```

### Duration Statistics

```{r}
#| label: dur-stats

ae_dur %>%
  summarise(
    N = n(),
    Mean_Duration = round(mean(AEDUR, na.rm = TRUE), 1),
    Median_Duration = median(AEDUR, na.rm = TRUE),
    Min = min(AEDUR, na.rm = TRUE),
    Max = max(AEDUR, na.rm = TRUE)
  )
```

------------------------------------------------------------------------

## Part 4: Handling Partial Dates

In clinical data, dates are often partial (e.g., "2023-06" or just "2023"). We cannot calculate durations or study days without **imputing** these to full dates.

### Common Imputation Strategies

| Date Type  | Missing Parts | Start Date Imputation | End Date Imputation |
|------------|---------------|-----------------------|---------------------|
| Year-Month | Day           | 01 (first)            | Last day of month   |
| Year only  | Month + Day   | 01-01                 | 12-31               |

### Reusable Imputation Functions

```{r}
#| label: impute-functions

#' Impute partial START dates (conservative: use earliest possible date)
#' 
#' @param date_str Character date string in ISO 8601 format
#' @return Character string with complete date
impute_start_date <- function(date_str) {
  dplyr::case_when(
    is.na(date_str) | date_str == "" ~ NA_character_,
    # Only Year (YYYY) -> YYYY-01-01
    nchar(date_str) == 4 ~ paste0(date_str, "-01-01"),
    # Year-Month (YYYY-MM) -> YYYY-MM-01
    nchar(date_str) == 7 ~ paste0(date_str, "-01"),
    # Full date -> Keep first 10 characters
    TRUE ~ substring(date_str, 1, 10)
  )
}

#' Impute partial END dates (conservative: use latest possible date)
#' 
#' @param date_str Character date string in ISO 8601 format
#' @return Character string with complete date
impute_end_date <- function(date_str) {
  dplyr::case_when(
    is.na(date_str) | date_str == "" ~ NA_character_,
    # Only Year (YYYY) -> YYYY-12-31
    nchar(date_str) == 4 ~ paste0(date_str, "-12-31"),
    # Year-Month (YYYY-MM) -> Last day of that month
    nchar(date_str) == 7 ~ {
      # Parse as first of month, then get last day
      first_of_month <- ymd(paste0(date_str, "-01"))
      last_of_month <- ceiling_date(first_of_month, "month") - days(1)
      as.character(last_of_month)
    },
    # Full date -> Keep first 10 characters
    TRUE ~ substring(date_str, 1, 10)
  )
}
```

### Testing the Imputation Functions

```{r}
#| label: test-impute

# Test cases with various partial dates
test_dates <- c("2023-05-12", "2023-05", "2023", "2023-02", NA)

data.frame(
  Original = test_dates,
  Imputed_Start = impute_start_date(test_dates),
  Imputed_End = impute_end_date(test_dates)
)
```

::: callout-tip
## Why Different Imputation Rules?

-   For **start dates**, we impute to the **earliest** possible date to be conservative about when something started.
-   For **end dates**, we impute to the **latest** possible date to be conservative about when something ended.
-   This approach ensures we don't underestimate durations or miss events.
:::

------------------------------------------------------------------------

## Part 5: Complete Exercise - Derive AESTDY and AEENDY

Let's bring everything together with a complete derivation.

```{r}
#| label: complete-derivation

# Full AE date derivation workflow
ae_dates_complete <- ae %>%
  select(USUBJID, AESEQ, AETERM, AESTDTC, AEENDTC) %>%
  # Join reference date
  inner_join(
    dm %>% select(USUBJID, RFSTDTC),
    by = "USUBJID"
  ) %>%
  mutate(
    # Impute partial dates
    AESTDTC_IMP = impute_start_date(AESTDTC),
    AEENDTC_IMP = impute_end_date(AEENDTC),
    RFSTDTC_IMP = impute_start_date(RFSTDTC),
    
    # Create imputation flags (for traceability)
    AESTDTF = ifelse(nchar(AESTDTC) < 10, "D", NA_character_),
    AEENDTF = ifelse(nchar(AEENDTC) < 10, "D", NA_character_),
    
    # Parse dates
    ASTDT = ymd(AESTDTC_IMP),
    AENDT = ymd(AEENDTC_IMP),
    RFSTDT = ymd(RFSTDTC_IMP),
    
    # Derive study days
    AESTDY = derive_study_day(ASTDT, RFSTDT),
    AEENDY = derive_study_day(AENDT, RFSTDT),
    
    # Derive duration
    AEDUR = ifelse(!is.na(ASTDT) & !is.na(AENDT),
                   as.integer(AENDT - ASTDT) + 1L,
                   NA_integer_)
  )

# Preview key results
ae_dates_complete %>%
  select(USUBJID, AETERM, AESTDY, AEENDY, AEDUR, AESTDTF) %>%
  head(15)
```

### Summary Statistics

```{r}
#| label: date-summary

ae_dates_complete %>%
  summarise(
    Total_Records = n(),
    Records_with_AESTDY = sum(!is.na(AESTDY)),
    Records_with_AEENDY = sum(!is.na(AEENDY)),
    Imputed_Start = sum(!is.na(AESTDTF)),
    Mean_Duration = round(mean(AEDUR, na.rm = TRUE), 1)
  )
```

------------------------------------------------------------------------

## 🎯 Practice Exercise

### Your Turn: Apply to Exposure (EX) Domain

Using the functions we created, derive study days for the Exposure domain.

```{r}
#| label: practice-exercise
#| eval: false

data("ex", package = "pharmaversesdtm")

# TODO: Complete the following
ex_dates <- ex %>%
  select(USUBJID, EXSEQ, EXSTDTC, EXENDTC) %>%
  inner_join(
    dm %>% select(USUBJID, RFSTDTC),
    by = "USUBJID"
  ) %>%
  mutate(
    # Step 1: Impute dates
    # EXSTDTC_IMP = ???
    # EXENDTC_IMP = ???
    
    # Step 2: Parse dates
    # EXSTDT = ???
    # EXENDT = ???
    
    # Step 3: Derive study days
    # EXSTDY = ???
    # EXENDY = ???
  )

# Check your results
head(ex_dates)
```

------------------------------------------------------------------------

## Deliverable Summary

Today you completed the following:

| Task                                   | Status |
|----------------------------------------|--------|
| Parsed ISO 8601 dates with `lubridate` | ✓ Done |
| Created `derive_study_day()` function  | ✓ Done |
| Created `impute_start_date()` function | ✓ Done |
| Created `impute_end_date()` function   | ✓ Done |
| Derived AESTDY, AEENDY, AEDUR from AE  | ✓ Done |

------------------------------------------------------------------------

## Key Takeaways

1.  **ISO 8601** (`YYYY-MM-DD`) is the standard format.
2.  **No Day 0**: Day -1 is immediately followed by Day 1.
3.  **Reusable Functions**: Create functions for derivations you'll use repeatedly.
4.  **Imputation Flags**: Use `--DTF` variables to indicate imputed dates.
5.  **Documentation**: Comment your logic for traceability.

------------------------------------------------------------------------

## Resources

-   [lubridate Cheatsheet](https://rawgit.com/rstudio/cheatsheets/main/lubridate.pdf)
-   [CDISC SDTM Date Conventions](https://www.cdisc.org/standards)
-   [admiral derive_vars_dt() Vignette](https://pharmaverse.github.io/admiral/)

------------------------------------------------------------------------

## What's Next?

In **Day 5**, we will move to **Advanced Tidyverse** operations:

-   Reshaping data with `pivot_wider()` and `pivot_longer()`
-   Creating analysis-ready datasets
-   Performing complex joins for QC
