---
title: "Day 5: Advanced Tidyverse: Pivoting & Joining"
subtitle: "Reshaping SDTM Data for Analysis"


format:
  html:
    theme:
      light: cosmo

    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css
    
editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

<div style="margin-bottom:12px;">< a target="_blank" href="../index.html" class="resource-link">← Back to Roadmap</a></div>

## Learning Objectives

By the end of Day 5, you will be able to:

1.  Reshape datasets using `pivot_wider()` and `pivot_longer()`
2.  Understand when to use **Wide vs. Long** formats in clinical data
3.  Perform advanced joins (`anti_join()`) for Data Quality Checks
4.  Create **3 analysis-ready datasets** as specified in the roadmap

------------------------------------------------------------------------

## Introduction

### Wide vs. Long Data

| Format | Structure | Use Case |
|----|----|----|
| **Long** | One measurement per row | Storage, filtering, SDTM standard |
| **Wide** | All measurements in columns | Calculations (BMI, Change from Baseline) |

SDTM stores data in **long format** (one row per observation). For analysis, we often need to reshape to **wide format** (one row per subject/visit).

------------------------------------------------------------------------

## Package Installation & Loading

```{r}
#| label: load-packages

library(dplyr)
library(tidyr)
library(pharmaversesdtm)

# Load all domains we'll use today
data("vs", package = "pharmaversesdtm")
data("dm", package = "pharmaversesdtm")
data("ae", package = "pharmaversesdtm")
data("ex", package = "pharmaversesdtm")
data("ds", package = "pharmaversesdtm")
```

------------------------------------------------------------------------

## Part 1: Pivoting Data

### From Long to Wide: `pivot_wider()`

Vital Signs (VS) comes in long format. Each test (Height, Weight, BP) is a separate row. Let's reshape it so each test has its own column.

```{r}
#| label: examine-vs-long

# Current long format
vs %>%
  filter(USUBJID == "01-701-1015" & VISIT == "BASELINE") %>%
  select(USUBJID, VISIT, VSTESTCD, VSORRES, VSORRESU) %>%
  head(10)
```

```{r}
#| label: pivot-wider-basic

# First, see what tests are available
vs %>%
  distinct(VSTESTCD) %>%
  print()

# Pivot to wide format
vs_wide <- vs %>%
  filter(VISIT == "BASELINE") %>%
  select(USUBJID, VSTESTCD, VSSTRESN) %>%
  pivot_wider(
    names_from = VSTESTCD,
    values_from = VSSTRESN,
    values_fn = mean  # Handle duplicates
  )

# Show available columns
cat("Columns after pivot:", paste(names(vs_wide), collapse = ", "), "\n\n")
head(vs_wide)
```

### Real Example: Calculate BMI

Now that we have Height and Weight side-by-side, we can easily calculate BMI.

```{r}
#| label: calc-bmi

# Check if HEIGHT and WEIGHT columns exist
if (all(c("HEIGHT", "WEIGHT") %in% names(vs_wide))) {
  vs_bmi <- vs_wide %>%
    filter(!is.na(HEIGHT) & !is.na(WEIGHT)) %>%
    mutate(
      # BMI = Weight (kg) / Height (m)^2
      HEIGHT_M = HEIGHT / 100,
      BMI = round(WEIGHT / (HEIGHT_M^2), 1)
    ) %>%
    select(USUBJID, HEIGHT, WEIGHT, BMI)
  
  head(vs_bmi)
  
  # Summary statistics
  vs_bmi %>%
    summarise(
      N = n(),
      Mean_BMI = round(mean(BMI), 1),
      Min_BMI = min(BMI),
      Max_BMI = max(BMI)
    )
} else {
  # Fallback: Calculate Blood Pressure ratio as alternative example
  cat("Note: HEIGHT/WEIGHT not available at BASELINE. Using BP data instead.\n\n")
  
  vs_bp <- vs_wide %>%
    filter(!is.na(SYSBP) & !is.na(DIABP)) %>%
    mutate(
      # Pulse Pressure = Systolic - Diastolic
      PULSE_PRESSURE = SYSBP - DIABP,
      # Mean Arterial Pressure ≈ DBP + 1/3(SBP - DBP)
      MAP = round(DIABP + (PULSE_PRESSURE / 3), 1)
    ) %>%
    select(USUBJID, SYSBP, DIABP, PULSE_PRESSURE, MAP)
  
  head(vs_bp)
  
  # Summary
  vs_bp %>%
    summarise(
      N = n(),
      Mean_MAP = round(mean(MAP, na.rm = TRUE), 1),
      Mean_PP = round(mean(PULSE_PRESSURE, na.rm = TRUE), 1)
    )
}
```

### From Wide to Long: `pivot_longer()`

Sometimes we need to convert wide data back to long format (often for plotting).

```{r}
#| label: pivot-longer

vs_long_again <- vs_wide %>%
  select(USUBJID, DIABP, SYSBP, PULSE) %>%
  pivot_longer(
    cols = c(DIABP, SYSBP, PULSE),
    names_to = "VSTESTCD",
    values_to = "VSSTRESN"
  ) %>%
  filter(!is.na(VSSTRESN))

head(vs_long_again, 10)
```

------------------------------------------------------------------------

## Part 2: Advanced Joins for QC

### Using `anti_join()` for Quality Checks

`anti_join(x, y)` returns rows in x that are **NOT** in y. This is essential for finding missing or orphaned records.

```{r}
#| label: anti-join-intro

# Subjects in DM (everyone enrolled)
dm_subjects <- dm %>% distinct(USUBJID)

# Subjects in AE (those with adverse events)
ae_subjects <- ae %>% distinct(USUBJID)

# Find subjects with NO adverse events
subjects_no_ae <- dm_subjects %>%
  anti_join(ae_subjects, by = "USUBJID")

cat("Total subjects:", nrow(dm_subjects), "\n")
cat("Subjects with AEs:", nrow(ae_subjects), "\n")
cat("Subjects with NO AEs:", nrow(subjects_no_ae), "\n")
```

### Finding Screen Failures

Screen failures are subjects in DM who:

-   Have disposition (DS) record showing SCREEN FAILURE
-   Often have no exposure (EX) records

```{r}
#| label: screen-failures

# Subjects with exposure
ex_subjects <- ex %>% distinct(USUBJID)

# Screen failures: in DM but NOT in EX
potential_screen_fail <- dm_subjects %>%
  anti_join(ex_subjects, by = "USUBJID")

cat("Subjects without exposure (potential screen failures):", nrow(potential_screen_fail), "\n")

# Verify against DS
screen_fail_ds <- ds %>%
  filter(grepl("SCREEN", DSDECOD, ignore.case = TRUE)) %>%
  distinct(USUBJID)

cat("Screen failures per DS:", nrow(screen_fail_ds), "\n")
```

------------------------------------------------------------------------

## Part 3: Creating Analysis-Ready Datasets

Per the roadmap, we need to create **3 analysis-ready datasets**.

### Dataset 1: Wide VS with One Row per Subject/Visit

This dataset has vital signs organized as columns, useful for visit-level analysis.

```{r}
#| label: dataset-1

# Dataset 1: Wide VS by Subject and Visit
analysis_vs_wide <- vs %>%
  # Select key columns
  select(USUBJID, VISIT, VISITNUM, VSTESTCD, VSSTRESN, VSSTRESU) %>%
  # Keep standard tests
  filter(VSTESTCD %in% c("SYSBP", "DIABP", "PULSE", "HEIGHT", "WEIGHT", "TEMP")) %>%
  # Pivot to wide
  pivot_wider(
    id_cols = c(USUBJID, VISIT, VISITNUM),
    names_from = VSTESTCD,
    values_from = VSSTRESN,
    values_fn = mean
  ) %>%
  # Sort
  arrange(USUBJID, VISITNUM)

cat("Dataset 1: Wide VS\n")
cat("Rows:", nrow(analysis_vs_wide), "\n")
cat("Columns:", ncol(analysis_vs_wide), "\n\n")

head(analysis_vs_wide, 10)
```

### Dataset 2: DM + AE with Demographic Covariates

This dataset merges adverse events with demographic information for safety analysis.

```{r}
#| label: dataset-2

# Dataset 2: AE with demographic covariates
analysis_ae_demo <- ae %>%
  # Select key AE variables
  select(USUBJID, AESEQ, AETERM, AEDECOD, AEBODSYS, AESEV, AESER, AESTDTC) %>%
  # Join demographics
  inner_join(
    dm %>% select(USUBJID, AGE, AGEU, SEX, RACE, ARM, ACTARM, RFSTDTC),
    by = "USUBJID"
  ) %>%
  # Add derived columns
  mutate(
    # Age group
    AGEGR1 = case_when(
      AGE < 40 ~ "<40",
      AGE >= 40 & AGE < 65 ~ "40-64",
      AGE >= 65 ~ ">=65"
    ),
    # Serious flag as factor
    AESER_F = factor(AESER, levels = c("N", "Y"), labels = c("Non-Serious", "Serious"))
  )

cat("Dataset 2: AE with Demographics\n")
cat("Rows:", nrow(analysis_ae_demo), "\n")
cat("Columns:", ncol(analysis_ae_demo), "\n\n")

# Summary by age group and treatment
analysis_ae_demo %>%
  group_by(ARM, AGEGR1) %>%
  summarise(
    N_Events = n(),
    N_Subjects = n_distinct(USUBJID),
    .groups = "drop"
  ) %>%
  arrange(ARM, AGEGR1)
```

### Dataset 3: Screen Failure Report using Anti-Join

This dataset identifies subjects who were screened but never received treatment.

```{r}
#| label: dataset-3

# Dataset 3: Screen Failure Report
screen_failure_report <- dm %>%
  # Get basic demographics
  select(USUBJID, SITEID, AGE, SEX, RACE, RFSTDTC, ARM) %>%
  # Anti-join: subjects NOT in EX
  anti_join(
    ex %>% distinct(USUBJID),
    by = "USUBJID"
  ) %>%
  # Add disposition reason if available
  left_join(
    ds %>%
      filter(DSCAT == "DISPOSITION EVENT") %>%
      select(USUBJID, DSDECOD, DSTERM),
    by = "USUBJID"
  ) %>%
  # Sort by site
  arrange(SITEID, USUBJID)

cat("Dataset 3: Screen Failure Report\n")
cat("Rows (Screen Failures):", nrow(screen_failure_report), "\n\n")

# Summary by site
screen_failure_report %>%
  group_by(SITEID) %>%
  summarise(
    N = n(),
    Reasons = paste(unique(na.omit(DSDECOD)), collapse = ", ")
  ) %>%
  head(10)
```

------------------------------------------------------------------------

## Part 4: Multi-Domain Analysis

Let's combine data from multiple domains to answer a clinical question:

**Question**: For subjects with Serious Adverse Events, what was their total drug exposure?

```{r}
#| label: multi-domain

# Subjects with Serious AEs
sae_subjects <- ae %>%
  filter(AESER == "Y") %>%
  distinct(USUBJID, .keep_all = TRUE) %>%
  select(USUBJID, AEDECOD)

# Add demographics
sae_analysis <- sae_subjects %>%
  inner_join(
    dm %>% select(USUBJID, AGE, SEX, ARM),
    by = "USUBJID"
  ) %>%
  # Add exposure summary
  left_join(
    ex %>%
      group_by(USUBJID) %>%
      summarise(
        TOTAL_DOSE = sum(EXDOSE, na.rm = TRUE),
        N_DOSES = n(),
        FIRST_DOSE = min(EXSTDTC),
        LAST_DOSE = max(EXENDTC),
        .groups = "drop"
      ),
    by = "USUBJID"
  )

cat("Serious AE Analysis Dataset\n")
cat("Subjects with SAEs:", nrow(sae_analysis), "\n\n")

# Display
sae_analysis %>%
  select(USUBJID, AEDECOD, ARM, TOTAL_DOSE, N_DOSES) %>%
  head(10)
```

------------------------------------------------------------------------

## 🎯 Practice Exercise

### Your Turn: Create a Lab Analysis Dataset

Create a wide dataset from LB (Laboratory) data with:

-   One row per subject and visit
-   Columns for ALT, AST, BILI (Liver Function Tests)
-   Joined with DM for treatment arm

```{r}
#| label: practice-exercise
#| eval: false

data("lb", package = "pharmaversesdtm")

# TODO: Complete the following
lb_wide <- lb %>%
  # Step 1: Filter for liver function tests
  filter(LBTESTCD %in% c("ALT", "AST", "BILI")) %>%
  # Step 2: Select key columns
  select(USUBJID, VISIT, VISITNUM, LBTESTCD, LBSTRESN) %>%
  # Step 3: Pivot wider
  pivot_wider(
    # Fill in parameters...
  ) %>%
  # Step 4: Join with DM to get ARM
  inner_join(
    # Fill in...
  )

# Check your results
head(lb_wide)
```

------------------------------------------------------------------------

## Deliverable Summary

Today you completed the following:

| Task                                 | Status |
|--------------------------------------|--------|
| Pivoted VS from long to wide format  | ✓ Done |
| Calculated BMI using pivoted data    | ✓ Done |
| Dataset 1: Wide VS per Subject/Visit | ✓ Done |
| Dataset 2: DM + AE with covariates   | ✓ Done |
| Dataset 3: Screen Failure Report     | ✓ Done |
| Multi-domain SAE analysis            | ✓ Done |

------------------------------------------------------------------------

## Key Takeaways

1.  **Pivot Wider**: Use when you need variables side-by-side for calculations.
2.  **Pivot Longer**: Use when you need to group by a variable for plotting.
3.  **Anti-joins**: Essential for finding missing, orphaned, or excluded records.
4.  **Multi-domain Analysis**: Real analysis requires combining multiple SDTM domains.

------------------------------------------------------------------------

## Resources

-   [tidyr pivot vignette](https://tidyr.tidyverse.org/articles/pivot.html)
-   [dplyr two-table verbs](https://dplyr.tidyverse.org/articles/two-table.html)
-   [pharmaverse examples site](https://pharmaverse.github.io/examples/)

------------------------------------------------------------------------

## What's Next?

In **Day 6**, we will introduce **`sdtm.oak`** concepts for transforming raw data to SDTM, including:

-   Simulating raw EDC data
-   Algorithm-based transformations
-   Creating a complete SDTM LB domain with unit standardization

< a target="_blank" href="https://visitorbadge.io/status?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse"><img src="https://api.visitorbadge.io/api/visitors?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse&label=Visits%20till%20date&countColor=%23263759&style=plastic&labelStyle=none" /></a>
