---
title: "Day 6: Introduction to sdtm.oak"
subtitle: "EDC-to-SDTM Transformation Patterns"


format:
  html:
    theme:
      light: cosmo

    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css
    
editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

<div style="margin-bottom:12px;"><a href="../index.html" class="resource-link">← Back to Roadmap</a></div>

## Learning Objectives

By the end of Day 6, you will be able to:

1.  Understand the philosophy of **`sdtm.oak`** for SDTM generation
2.  Simulate "Raw" EDC datasets for both VS and LB
3.  Apply algorithm-based transformations (mapping, pivoting, hardcoding)
4.  Perform **unit standardization** (e.g., mg/dL to mmol/L)
5.  Create a complete **SDTM LB domain** with sequence numbers

------------------------------------------------------------------------

## Introduction

### What is `sdtm.oak`?

`sdtm.oak` is a pharmaverse package that provides a standardized, algorithm-based approach to creating SDTM datasets. Key concepts:

-   **Algorithm-based**: Instead of custom code, use reusable algorithms
-   **Metadata-driven**: Transformations defined in specifications
-   **Modular**: Small, testable functions that can be chained

### The sdtm.oak Philosophy

| Algorithm     | Description                    | Example Use                |
|---------------|--------------------------------|----------------------------|
| **Assign**    | Copy source to target          | Subject ID → USUBJID       |
| **Hardcode**  | Set a constant value           | DOMAIN = "LB"              |
| **Condition** | Apply logic based on condition | If severity \>= 3 then "Y" |
| **Assign CT** | Map to Controlled Terminology  | "Male" → "M"               |

*Note: We'll simulate sdtm.oak patterns using dplyr/tidyr to understand the concepts.*

------------------------------------------------------------------------

## Package Loading

```{r}
#| label: load-packages

library(dplyr)
library(tidyr)
library(tibble)
library(lubridate)
```

------------------------------------------------------------------------

## Part 1: Simulating Raw EDC Data

### Raw Vital Signs Data

First, let's create data that looks like it came from an Electronic Data Capture (EDC) system like Rave or Veeva.

```{r}
#| label: raw-vs

# Simulated Raw Vital Signs (EDC export format)
raw_vs <- tribble(
  ~SubjectID,  ~Site, ~Visit,       ~Date,        ~SysBP, ~DiaBP, ~Pulse, ~Temp_C,
  "001",       "101", "Screening",  "2024-01-01", 120,    80,     72,     36.5,
  "001",       "101", "Baseline",   "2024-01-15", 118,    78,     70,     36.8,
  "001",       "101", "Week 4",     "2024-02-12", 115,    76,     68,     36.6,
  "002",       "101", "Screening",  "2024-01-02", 130,    85,     88,     37.0,
  "002",       "101", "Baseline",   "2024-01-16", 128,    82,     85,     36.7,
  "003",       "102", "Screening",  "2024-01-03", 145,    92,     95,     36.9,
  "003",       "102", "Baseline",   "2024-01-17", 140,    88,     90,     36.5
)

print(raw_vs)
```

### Raw Laboratory Data

Now let's create lab data with values in different units that need standardization.

```{r}
#| label: raw-lb

# Simulated Raw Lab Data (with different units per site)
raw_lb <- tribble(
  ~SubjectID,  ~Site, ~Visit,      ~Date,        ~Test,        ~Result, ~Unit,
  # Site 101: Uses US units
  "001",       "101", "Screening", "2024-01-01", "Glucose",    95,      "mg/dL",
  "001",       "101", "Screening", "2024-01-01", "Cholesterol", 180,    "mg/dL",
  "001",       "101", "Screening", "2024-01-01", "ALT",        25,      "U/L",
  "001",       "101", "Baseline",  "2024-01-15", "Glucose",    92,      "mg/dL",
  "001",       "101", "Baseline",  "2024-01-15", "Cholesterol", 175,    "mg/dL",
  "001",       "101", "Baseline",  "2024-01-15", "ALT",        28,      "U/L",
  "002",       "101", "Screening", "2024-01-02", "Glucose",    110,     "mg/dL",
  "002",       "101", "Screening", "2024-01-02", "Cholesterol", 220,    "mg/dL",
  "002",       "101", "Screening", "2024-01-02", "ALT",        45,      "U/L",
  # Site 102: Uses SI units
  "003",       "102", "Screening", "2024-01-03", "Glucose",    5.5,     "mmol/L",
  "003",       "102", "Screening", "2024-01-03", "Cholesterol", 4.8,    "mmol/L",
  "003",       "102", "Screening", "2024-01-03", "ALT",        30,      "U/L",
  "003",       "102", "Baseline",  "2024-01-17", "Glucose",    5.2,     "mmol/L",
  "003",       "102", "Baseline",  "2024-01-17", "Cholesterol", 4.5,    "mmol/L",
  "003",       "102", "Baseline",  "2024-01-17", "ALT",        32,      "U/L"
)

print(raw_lb)
```

------------------------------------------------------------------------

## Part 2: Creating SDTM VS Domain

### Step 1: Hardcode Standard Variables

```{r}
#| label: vs-step1

# Algorithm: Hardcode
vs_step1 <- raw_vs %>%
  mutate(
    STUDYID = "DEMO-001",
    DOMAIN = "VS",
    # Algorithm: Assign (concatenate)
    USUBJID = paste(STUDYID, Site, SubjectID, sep = "-")
  )

head(vs_step1)
```

### Step 2: Pivot to Long Format (Findings Algorithm)

SDTM Findings domains are **long** (one row per test per visit).

```{r}
#| label: vs-step2

# Algorithm: Transpose/Pivot
vs_step2 <- vs_step1 %>%
  pivot_longer(
    cols = c(SysBP, DiaBP, Pulse, Temp_C),
    names_to = "RAW_TEST",
    values_to = "VSORRES_NUM"
  )

head(vs_step2, 10)
```

### Step 3: Map to Controlled Terminology

```{r}
#| label: vs-step3

# Algorithm: Assign CT (Map test codes)
vs_step3 <- vs_step2 %>%
  mutate(
    # Controlled Terminology mapping
    VSTESTCD = case_when(
      RAW_TEST == "SysBP"  ~ "SYSBP",
      RAW_TEST == "DiaBP"  ~ "DIABP",
      RAW_TEST == "Pulse"  ~ "PULSE",
      RAW_TEST == "Temp_C" ~ "TEMP"
    ),
    VSTEST = case_when(
      VSTESTCD == "SYSBP" ~ "Systolic Blood Pressure",
      VSTESTCD == "DIABP" ~ "Diastolic Blood Pressure",
      VSTESTCD == "PULSE" ~ "Pulse Rate",
      VSTESTCD == "TEMP"  ~ "Temperature"
    ),
    VSORRESU = case_when(
      VSTESTCD %in% c("SYSBP", "DIABP") ~ "mmHg",
      VSTESTCD == "PULSE" ~ "BEATS/MIN",
      VSTESTCD == "TEMP"  ~ "C"
    )
  )

head(vs_step3)
```

### Step 4: Add Sequence Number

Every SDTM record needs a unique sequence number (`--SEQ`) within subject.

```{r}
#| label: vs-step4

# Algorithm: Derive sequence
vs_step4 <- vs_step3 %>%
  arrange(USUBJID, Date, VSTESTCD) %>%
  group_by(USUBJID) %>%
  mutate(VSSEQ = row_number()) %>%
  ungroup()

vs_step4 %>%
  select(USUBJID, VSSEQ, VSTESTCD, Visit) %>%
  head(10)
```

### Step 5: Final SDTM VS Domain

```{r}
#| label: vs-final

sdtm_vs <- vs_step4 %>%
  mutate(
    VSORRES = as.character(VSORRES_NUM),
    VSSTRESN = VSORRES_NUM,
    VSSTRESU = VSORRESU,
    VSDTC = Date,
    VISIT = Visit
  ) %>%
  select(
    STUDYID, DOMAIN, USUBJID, VSSEQ, VSTESTCD, VSTEST,
    VSORRES, VSORRESU, VSSTRESN, VSSTRESU, VSDTC, VISIT
  )

cat("SDTM VS Domain:\n")
cat("Records:", nrow(sdtm_vs), "\n\n")
head(sdtm_vs, 10)
```

------------------------------------------------------------------------

## Part 3: Creating SDTM LB Domain with Unit Standardization

The LB domain is more complex because we need to **standardize units** across sites.

### Unit Conversion Reference

| Test        | Original Unit | Standard Unit | Conversion Factor |
|-------------|---------------|---------------|-------------------|
| Glucose     | mg/dL         | mmol/L        | ÷ 18.0182         |
| Cholesterol | mg/dL         | mmol/L        | ÷ 38.67           |
| ALT         | U/L           | U/L           | None (already SI) |

### Step 1: Initial Mapping

```{r}
#| label: lb-step1

lb_step1 <- raw_lb %>%
  mutate(
    STUDYID = "DEMO-001",
    DOMAIN = "LB",
    USUBJID = paste(STUDYID, Site, SubjectID, sep = "-"),
    
    # Map test codes
    LBTESTCD = case_when(
      Test == "Glucose"     ~ "GLUC",
      Test == "Cholesterol" ~ "CHOL",
      Test == "ALT"         ~ "ALT"
    ),
    LBTEST = case_when(
      LBTESTCD == "GLUC" ~ "Glucose",
      LBTESTCD == "CHOL" ~ "Cholesterol",
      LBTESTCD == "ALT"  ~ "Alanine Aminotransferase"
    ),
    
    # Original results
    LBORRES = as.character(Result),
    LBORRESU = Unit
  )

head(lb_step1)
```

### Step 2: Unit Standardization

This is where the real work happens - converting all values to standard SI units.

```{r}
#| label: lb-step2

lb_step2 <- lb_step1 %>%
  mutate(
    # Standard unit is SI (mmol/L for glucose/cholesterol, U/L for ALT)
    LBSTRESU = case_when(
      LBTESTCD %in% c("GLUC", "CHOL") ~ "mmol/L",
      LBTESTCD == "ALT" ~ "U/L"
    ),
    
    # Convert to standard units
    LBSTRESN = case_when(
      # Glucose: mg/dL to mmol/L
      LBTESTCD == "GLUC" & LBORRESU == "mg/dL" ~ round(Result / 18.0182, 2),
      LBTESTCD == "GLUC" & LBORRESU == "mmol/L" ~ Result,
      
      # Cholesterol: mg/dL to mmol/L
      LBTESTCD == "CHOL" & LBORRESU == "mg/dL" ~ round(Result / 38.67, 2),
      LBTESTCD == "CHOL" & LBORRESU == "mmol/L" ~ Result,
      
      # ALT: Already in U/L
      LBTESTCD == "ALT" ~ Result
    ),
    
    # Character version of standardized result
    LBSTRESC = as.character(LBSTRESN)
  )

# Show the conversion
lb_step2 %>%
  select(USUBJID, LBTESTCD, LBORRES, LBORRESU, LBSTRESN, LBSTRESU) %>%
  head(10)
```

::: callout-note
## Unit Standardization in Practice

Notice how subjects from Site 101 (US units) and Site 102 (SI units) now have comparable values in `LBSTRESN`. This is essential for cross-site analysis!
:::

### Step 3: Add Sequence and Reference Ranges

```{r}
#| label: lb-step3

lb_step3 <- lb_step2 %>%
  arrange(USUBJID, Date, LBTESTCD) %>%
  group_by(USUBJID) %>%
  mutate(LBSEQ = row_number()) %>%
  ungroup() %>%
  # Add reference ranges (in standard units)
  mutate(
    LBSTNRLO = case_when(
      LBTESTCD == "GLUC" ~ 3.9,
      LBTESTCD == "CHOL" ~ 0.0,
      LBTESTCD == "ALT"  ~ 7.0
    ),
    LBSTNRHI = case_when(
      LBTESTCD == "GLUC" ~ 5.6,
      LBTESTCD == "CHOL" ~ 5.2,
      LBTESTCD == "ALT"  ~ 56.0
    ),
    # Normal range indicator
    LBNRIND = case_when(
      LBSTRESN < LBSTNRLO ~ "LOW",
      LBSTRESN > LBSTNRHI ~ "HIGH",
      TRUE ~ "NORMAL"
    )
  )

lb_step3 %>%
  select(USUBJID, LBTESTCD, LBSTRESN, LBSTNRLO, LBSTNRHI, LBNRIND) %>%
  head(10)
```

### Step 4: Final SDTM LB Domain

```{r}
#| label: lb-final

sdtm_lb <- lb_step3 %>%
  mutate(
    LBDTC = Date,
    VISIT = Visit
  ) %>%
  select(
    STUDYID, DOMAIN, USUBJID, LBSEQ, LBTESTCD, LBTEST,
    LBORRES, LBORRESU, LBSTRESC, LBSTRESN, LBSTRESU,
    LBSTNRLO, LBSTNRHI, LBNRIND, LBDTC, VISIT
  )

cat("SDTM LB Domain:\n")
cat("Records:", nrow(sdtm_lb), "\n")
cat("Variables:", ncol(sdtm_lb), "\n\n")

head(sdtm_lb, 10)
```

### Summary Statistics

```{r}
#| label: lb-summary

sdtm_lb %>%
  group_by(LBTESTCD, LBTEST) %>%
  summarise(
    N = n(),
    Mean = round(mean(LBSTRESN, na.rm = TRUE), 2),
    SD = round(sd(LBSTRESN, na.rm = TRUE), 2),
    Low = sum(LBNRIND == "LOW"),
    Normal = sum(LBNRIND == "NORMAL"),
    High = sum(LBNRIND == "HIGH"),
    .groups = "drop"
  )
```

------------------------------------------------------------------------

## 🎯 Practice Exercise

### Your Turn: Add LOINC Codes

LOINC codes are standard lab test identifiers. Add `LBLOINC` to the LB domain.

```{r}
#| label: practice-exercise
#| eval: false

# LOINC Reference
loinc_lookup <- tribble(
  ~LBTESTCD, ~LBLOINC,
  "GLUC",    "2345-7",
  "CHOL",    "2093-3",
  "ALT",     "1742-6"
)

# TODO: Join the LOINC codes to sdtm_lb
sdtm_lb_loinc <- sdtm_lb %>%
  # Your code here...

head(sdtm_lb_loinc)
```

------------------------------------------------------------------------

## Deliverable Summary

Today you completed the following:

| Task                                            | Status |
|-------------------------------------------------|--------|
| Created simulated Raw VS and LB data            | ✓ Done |
| Built SDTM VS with sequence numbers             | ✓ Done |
| Performed unit standardization (mg/dL → mmol/L) | ✓ Done |
| Added reference ranges and normal indicators    | ✓ Done |
| Created complete SDTM LB domain                 | ✓ Done |

------------------------------------------------------------------------

## Key Takeaways

1.  **sdtm.oak Philosophy**: Algorithm-based, modular, traceable.
2.  **Pivoting**: Raw data is wide; SDTM Findings are long.
3.  **Unit Standardization**: Critical for multi-site studies.
4.  **Sequence Numbers**: Every record needs a unique `--SEQ`.
5.  **Reference Ranges**: `LBSTNRLO`, `LBSTNRHI`, `LBNRIND` are key for flagging abnormals.

------------------------------------------------------------------------

## Resources

-   [sdtm.oak GitHub Repository](https://github.com/pharmaverse/sdtm.oak)
-   [CDISC SDTM Implementation Guide](https://www.cdisc.org/standards)
-   [LOINC Database](https://loinc.org/)

------------------------------------------------------------------------

## What's Next?

In **Day 7**, we will complete the **Week 1 Capstone**:

-   Build DM, AE, EX domains from scratch (20+ subjects)
-   Apply all concepts learned this week
-   Export to submission-ready `.xpt` files
