---
title: "Day 7: Week 1 Capstone: End-to-End SDTM Script"
subtitle: "Build DM, AE, EX from Scratch with xportr"


format:
  html:
    theme:
      light: cosmo

    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css
    
editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

<div style="margin-bottom:12px;"><a href="../index.html" class="resource-link">← Back to Roadmap</a></div>

## Learning Objectives

By the end of Day 7, you will be able to:

1.  Integrate all skills from Days 1-6 into a single workflow
2.  Produce valid SDTM domains (DM, AE, EX) from simulated raw data
3.  Work with **20+ subjects** across multiple sites
4.  Apply metadata (Labels, Types) using **`xportr`**
5.  Export submission-ready SAS Transport files (`.xpt`)

------------------------------------------------------------------------

## Capstone Project Overview

**Goal**: Create three SDTM domains from simulated raw clinical trial data:

-   **DM** (Demographics) - One row per subject
-   **AE** (Adverse Events) - Multiple rows per subject
-   **EX** (Exposure) - Dosing records

### The Pipeline

```         
Raw Data → Transform → Validate → Format → Export (.xpt)
```

------------------------------------------------------------------------

## Part 1: Generate Simulated Raw Data (20+ Subjects)

### Raw Demographics

```{r}
#| label: generate-raw-dm

library(dplyr)
library(tidyr)
library(lubridate)
library(xportr)

set.seed(42)  # For reproducibility

# Generate 25 subjects across 3 sites
n_subjects <- 25

raw_dm <- tibble(
  Site = sample(c("101", "102", "103"), n_subjects, replace = TRUE, prob = c(0.4, 0.35, 0.25)),
  Subject = sprintf("%03d", 1:n_subjects)
) %>%
  mutate(
    # Demographics
    Sex = sample(c("Male", "Female"), n_subjects, replace = TRUE),
    Race = sample(c("White", "Black", "Asian", "Other"), n_subjects, replace = TRUE, 
                  prob = c(0.6, 0.2, 0.15, 0.05)),
    Ethnicity = sample(c("Hispanic", "Not Hispanic"), n_subjects, replace = TRUE, 
                       prob = c(0.15, 0.85)),
    
    # Age 25-75
    BirthYear = sample(1950:2000, n_subjects, replace = TRUE),
    BirthDate = paste0(BirthYear, "-", 
                       sprintf("%02d", sample(1:12, n_subjects, replace = TRUE)), "-",
                       sprintf("%02d", sample(1:28, n_subjects, replace = TRUE))),
    
    # Treatment assignment
    Arm = sample(c("Placebo", "Active 10mg", "Active 20mg"), n_subjects, replace = TRUE,
                 prob = c(0.33, 0.33, 0.34)),
    
    # First dose date (Jan-Mar 2024)
    FirstDoseDate = as.Date("2024-01-01") + sample(0:89, n_subjects, replace = TRUE)
  ) %>%
  select(-BirthYear)

cat("Raw Demographics: ", nrow(raw_dm), "subjects\n\n")
head(raw_dm, 10)
```

### Raw Adverse Events

```{r}
#| label: generate-raw-ae

# Common adverse events with their SOC mappings
ae_terms <- tribble(
  ~Term,                   ~SOC,
  "Headache",              "Nervous system disorders",
  "Nausea",                "Gastrointestinal disorders",
  "Fatigue",               "General disorders",
  "Dizziness",             "Nervous system disorders",
  "Diarrhea",              "Gastrointestinal disorders",
  "Rash",                  "Skin disorders",
  "Back pain",             "Musculoskeletal disorders",
  "Insomnia",              "Psychiatric disorders",
  "Cough",                 "Respiratory disorders",
  "Upper respiratory infection", "Infections"
)

# Generate 2-5 AEs per subject (not all subjects have AEs)
subjects_with_ae <- sample(raw_dm$Subject, size = round(n_subjects * 0.8))

raw_ae <- tibble(
  Subject = rep(subjects_with_ae, each = 3)
) %>%
  slice_sample(n = 60) %>%  # ~60 AE records total
  mutate(
    Site = raw_dm$Site[match(Subject, raw_dm$Subject)],
    AE_Num = row_number(),
    Term = sample(ae_terms$Term, n(), replace = TRUE),
    Severity = sample(c("Mild", "Moderate", "Severe"), n(), replace = TRUE, 
                      prob = c(0.6, 0.3, 0.1)),
    Serious = sample(c("No", "Yes"), n(), replace = TRUE, prob = c(0.92, 0.08)),
    Related = sample(c("Not Related", "Possibly Related", "Probably Related"), n(), 
                     replace = TRUE, prob = c(0.5, 0.35, 0.15)),
    # Start 1-60 days after first dose
    StartDay = sample(1:60, n(), replace = TRUE)
  ) %>%
  left_join(
    raw_dm %>% select(Subject, FirstDoseDate),
    by = "Subject"
  ) %>%
  mutate(
    StartDate = as.character(FirstDoseDate + StartDay),
    # End date: 1-14 days after start, or ongoing
    EndDate = ifelse(
      sample(c(TRUE, FALSE), n(), replace = TRUE, prob = c(0.85, 0.15)),
      as.character(as.Date(StartDate) + sample(1:14, n(), replace = TRUE)),
      NA_character_
    )
  ) %>%
  select(Site, Subject, Term, StartDate, EndDate, Severity, Serious, Related)

cat("Raw Adverse Events:", nrow(raw_ae), "records\n\n")
head(raw_ae, 10)
```

### Raw Exposure

```{r}
#| label: generate-raw-ex

# Generate exposure records (weekly dosing)
raw_ex <- raw_dm %>%
  filter(!is.na(FirstDoseDate)) %>%
  select(Site, Subject, Arm, FirstDoseDate) %>%
  # Create 8 weeks of dosing per subject
  crossing(Week = 0:7) %>%
  mutate(
    DoseDate = FirstDoseDate + (Week * 7),
    Dose = case_when(
      Arm == "Placebo" ~ 0,
      Arm == "Active 10mg" ~ 10,
      Arm == "Active 20mg" ~ 20
    ),
    DoseUnit = "mg",
    Route = "ORAL"
  ) %>%
  # Some subjects discontinue early
  group_by(Subject) %>%
  filter(row_number() <= sample(4:8, 1)) %>%
  ungroup() %>%
  select(Site, Subject, DoseDate, Dose, DoseUnit, Route)

cat("Raw Exposure:", nrow(raw_ex), "records\n\n")
head(raw_ex, 10)
```

------------------------------------------------------------------------

## Part 2: Create SDTM DM Domain

```{r}
#| label: build-dm

# SDTM DM: Demographics (one row per subject)
sdtm_dm <- raw_dm %>%
  mutate(
    # Standard variables - HARDCODE algorithm
    STUDYID = "CAPSTONE-01",
    DOMAIN = "DM",
    
    # Subject identifiers - ASSIGN algorithm
    USUBJID = paste(STUDYID, Site, Subject, sep = "-"),
    SUBJID = Subject,
    SITEID = Site,
    
    # Demographics - MAP to CT
    SEX = case_when(
      Sex == "Male" ~ "M",
      Sex == "Female" ~ "F"
    ),
    RACE = toupper(Race),
    ETHNIC = case_when(
      Ethnicity == "Hispanic" ~ "HISPANIC OR LATINO",
      TRUE ~ "NOT HISPANIC OR LATINO"
    ),
    
    # Dates - DERIVE
    RFSTDTC = as.character(FirstDoseDate),
    RFENDTC = NA_character_,  # Would come from disposition
    BRTHDTC = BirthDate,
    
    # Age calculation
    AGE = as.integer(floor(interval(ymd(BirthDate), ymd(FirstDoseDate)) / years(1))),
    AGEU = "YEARS",
    
    # Treatment
    ARM = Arm,
    ACTARM = Arm,  # Assuming planned = actual
    ARMCD = case_when(
      Arm == "Placebo" ~ "PBO",
      Arm == "Active 10mg" ~ "ACT10",
      Arm == "Active 20mg" ~ "ACT20"
    ),
    ACTARMCD = ARMCD
  ) %>%
  select(
    STUDYID, DOMAIN, USUBJID, SUBJID, SITEID,
    RFSTDTC, RFENDTC, BRTHDTC,
    AGE, AGEU, SEX, RACE, ETHNIC,
    ARM, ARMCD, ACTARM, ACTARMCD
  )

cat("SDTM DM Domain\n")
cat("Subjects:", nrow(sdtm_dm), "\n")
cat("Variables:", ncol(sdtm_dm), "\n\n")

head(sdtm_dm, 10)
```

------------------------------------------------------------------------

## Part 3: Create SDTM AE Domain

```{r}
#| label: build-ae

# Helper: derive study day (from Day 4)
derive_study_day <- function(event_date, reference_date) {
  diff_days <- as.numeric(event_date - reference_date)
  ifelse(diff_days >= 0, diff_days + 1, diff_days)
}

# SDTM AE: Adverse Events
sdtm_ae <- raw_ae %>%
  mutate(
    STUDYID = "CAPSTONE-01",
    DOMAIN = "AE",
    USUBJID = paste(STUDYID, Site, Subject, sep = "-")
  ) %>%
  # Add reference date from DM
  left_join(
    sdtm_dm %>% select(USUBJID, RFSTDTC),
    by = "USUBJID"
  ) %>%
  # Derive AE variables
  mutate(
    AETERM = Term,
    AEDECOD = toupper(Term),
    AEBODSYS = toupper(case_when(
      grepl("Headache|Dizziness", Term) ~ "NERVOUS SYSTEM DISORDERS",
      grepl("Nausea|Diarrhea", Term) ~ "GASTROINTESTINAL DISORDERS",
      grepl("Fatigue", Term) ~ "GENERAL DISORDERS AND ADMINISTRATION SITE CONDITIONS",
      grepl("Rash", Term) ~ "SKIN AND SUBCUTANEOUS TISSUE DISORDERS",
      grepl("Back", Term) ~ "MUSCULOSKELETAL AND CONNECTIVE TISSUE DISORDERS",
      grepl("Insomnia", Term) ~ "PSYCHIATRIC DISORDERS",
      grepl("Cough", Term) ~ "RESPIRATORY, THORACIC AND MEDIASTINAL DISORDERS",
      grepl("infection", Term, ignore.case = TRUE) ~ "INFECTIONS AND INFESTATIONS",
      TRUE ~ "OTHER"
    )),
    
    # Dates
    AESTDTC = StartDate,
    AEENDTC = EndDate,
    
    # Study days
    AESTDY = derive_study_day(ymd(StartDate), ymd(RFSTDTC)),
    AEENDY = ifelse(!is.na(EndDate), 
                    derive_study_day(ymd(EndDate), ymd(RFSTDTC)),
                    NA_integer_),
    
    # CT mapping
    AESEV = toupper(Severity),
    AESER = ifelse(Serious == "Yes", "Y", "N"),
    AEREL = toupper(Related),
    
    # Outcome
    AEOUT = case_when(
      is.na(EndDate) ~ "NOT RECOVERED/NOT RESOLVED",
      TRUE ~ "RECOVERED/RESOLVED"
    ),
    
    # Action taken
    AEACN = "DOSE NOT CHANGED"
  ) %>%
  # Add sequence
  arrange(USUBJID, AESTDTC) %>%
  group_by(USUBJID) %>%
  mutate(AESEQ = row_number()) %>%
  ungroup() %>%
  select(
    STUDYID, DOMAIN, USUBJID, AESEQ,
    AETERM, AEDECOD, AEBODSYS,
    AESTDTC, AEENDTC, AESTDY, AEENDY,
    AESEV, AESER, AEREL, AEOUT, AEACN
  )

cat("SDTM AE Domain\n")
cat("Records:", nrow(sdtm_ae), "\n")
cat("Subjects:", n_distinct(sdtm_ae$USUBJID), "\n")
cat("Variables:", ncol(sdtm_ae), "\n\n")

head(sdtm_ae, 10)
```

------------------------------------------------------------------------

## Part 4: Create SDTM EX Domain

```{r}
#| label: build-ex

# SDTM EX: Exposure
sdtm_ex <- raw_ex %>%
  mutate(
    STUDYID = "CAPSTONE-01",
    DOMAIN = "EX",
    USUBJID = paste(STUDYID, Site, Subject, sep = "-"),
    
    # Treatment
    EXTRT = case_when(
      Dose == 0 ~ "PLACEBO",
      Dose == 10 ~ "STUDY DRUG 10 MG",
      Dose == 20 ~ "STUDY DRUG 20 MG"
    ),
    
    # Dose info
    EXDOSE = Dose,
    EXDOSU = "mg",
    EXDOSFRM = "TABLET",
    EXROUTE = "ORAL",
    
    # Dates (single-day dosing)
    EXSTDTC = as.character(DoseDate),
    EXENDTC = as.character(DoseDate)
  ) %>%
  # Add reference date from DM
  left_join(
    sdtm_dm %>% select(USUBJID, RFSTDTC),
    by = "USUBJID"
  ) %>%
  mutate(
    EXSTDY = derive_study_day(ymd(DoseDate), ymd(RFSTDTC)),
    EXENDY = EXSTDY
  ) %>%
  # Add sequence
  arrange(USUBJID, EXSTDTC) %>%
  group_by(USUBJID) %>%
  mutate(EXSEQ = row_number()) %>%
  ungroup() %>%
  select(
    STUDYID, DOMAIN, USUBJID, EXSEQ,
    EXTRT, EXDOSE, EXDOSU, EXDOSFRM, EXROUTE,
    EXSTDTC, EXENDTC, EXSTDY, EXENDY
  )

cat("SDTM EX Domain\n")
cat("Records:", nrow(sdtm_ex), "\n")
cat("Subjects:", n_distinct(sdtm_ex$USUBJID), "\n")
cat("Variables:", ncol(sdtm_ex), "\n\n")

head(sdtm_ex, 10)
```

------------------------------------------------------------------------

## Part 5: Validation Checks

Before exporting, let's validate our datasets.

```{r}
#| label: validation

# Check 1: Subject counts consistency
cat("=== Subject Count Validation ===\n")
cat("DM subjects:", nrow(sdtm_dm), "\n")
cat("EX subjects:", n_distinct(sdtm_ex$USUBJID), "\n")
cat("AE subjects:", n_distinct(sdtm_ae$USUBJID), "\n\n")

# Check 2: All EX subjects are in DM
ex_not_in_dm <- sdtm_ex %>%
  anti_join(sdtm_dm, by = "USUBJID") %>%
  nrow()
cat("EX subjects missing from DM:", ex_not_in_dm, "\n")

# Check 3: All AE subjects are in DM
ae_not_in_dm <- sdtm_ae %>%
  anti_join(sdtm_dm, by = "USUBJID") %>%
  nrow()
cat("AE subjects missing from DM:", ae_not_in_dm, "\n\n")

# Check 4: Variable types
cat("=== Variable Type Check ===\n")
cat("DM - AGE is numeric:", is.numeric(sdtm_dm$AGE), "\n")
cat("AE - AESEQ is integer:", is.integer(sdtm_ae$AESEQ), "\n")
cat("EX - EXDOSE is numeric:", is.numeric(sdtm_ex$EXDOSE), "\n")
```

------------------------------------------------------------------------

## Part 6: Export with xportr

```{r}
#| label: export

# Create output directory in project (relative path)
output_dir <- "output"
if (!dir.exists(output_dir)) {
  dir.create(output_dir)
}

# Export DM
xportr_write(sdtm_dm, file.path(output_dir, "dm.xpt"))
cat("✓ Exported: output/dm.xpt\n")

# Export AE
xportr_write(sdtm_ae, file.path(output_dir, "ae.xpt"))
cat("✓ Exported: output/ae.xpt\n")

# Export EX
xportr_write(sdtm_ex, file.path(output_dir, "ex.xpt"))
cat("✓ Exported: output/ex.xpt\n")

# Verify files exist
cat("\n=== Export Verification ===\n")
cat("dm.xpt:", file.size(file.path(output_dir, "dm.xpt")), "bytes\n")
cat("ae.xpt:", file.size(file.path(output_dir, "ae.xpt")), "bytes\n")
cat("ex.xpt:", file.size(file.path(output_dir, "ex.xpt")), "bytes\n")
```

------------------------------------------------------------------------

## Summary Statistics

```{r}
#| label: summary-stats

# Treatment arm distribution
cat("=== Treatment Distribution (DM) ===\n")
sdtm_dm %>%
  count(ARM, ARMCD) %>%
  print()

cat("\n=== AE Severity Distribution ===\n")
sdtm_ae %>%
  count(AESEV) %>%
  print()

cat("\n=== Exposure Summary ===\n")
sdtm_ex %>%
  group_by(EXTRT) %>%
  summarise(
    N_Subjects = n_distinct(USUBJID),
    N_Doses = n(),
    Total_Dose = sum(EXDOSE)
  ) %>%
  print()
```

------------------------------------------------------------------------

## 🎉 Congratulations! Week 1 Complete!

You have successfully completed the Week 1 Capstone and demonstrated:

| Skill                  | Application                                       |
|------------------------|---------------------------------------------------|
| Data Generation        | Created 25 subjects with realistic demographics   |
| SDTM Structure         | Built DM (1 row/subject), AE & EX (multiple rows) |
| Controlled Terminology | Mapped Sex, Race, Severity, SOC to CT             |
| Date Derivations       | Calculated study days using "No Day 0" rule       |
| Sequence Numbers       | Added AESEQ, EXSEQ per subject                    |
| Validation             | Cross-domain consistency checks                   |
| Export                 | Created FDA-ready .xpt files                      |

------------------------------------------------------------------------

## Key Takeaways from Week 1

1.  **CDISC Structure**: SDTM organizes clinical data into standardized domains
2.  **Controlled Terminology**: Consistency across studies is mandatory
3.  **Derivations**: Study days, durations, and flags follow specific rules
4.  **Traceability**: Every derived variable should trace back to source
5.  **Validation**: Check data before export

------------------------------------------------------------------------

## What's Next?

**Week 2: Production SDTM & Validation**

-   Complex domains: LB (Labs), VS (Vital Signs), DS (Disposition)
-   Industry-grade validation with `sdtmchecks`
-   Metadata management with `metacore`
-   Specification-driven workflows
