---
title: "Day 12: Data Cuts with datacutr"
subtitle: "Applying Clinical Cutoff Dates for Interim & Final Analyses"


format:
  html:
    theme:
      light: cosmo

    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css
    
editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

::: {style="margin-bottom:12px;"}
<a href="../index.html" class="resource-link">← Back to Roadmap</a>
:::

## Learning Objectives

By the end of Day 12, you will be able to:

1.  Explain what a **clinical data cut** is and why it's essential for submissions
2.  Understand the difference between **patient-level** and **record-level** data cuts
3.  Use the **`datacutr`** package to apply cutoff dates across multiple SDTM domains
4.  Handle edge cases: **ongoing AEs**, **partial dates**, and **records at the cutoff boundary**
5.  Build a reproducible data cut pipeline that works across DM, AE, LB, and VS domains

------------------------------------------------------------------------

## What is a Data Cut?

### The Problem

Clinical trials collect data **continuously** over months or years. But at some point, you need to **freeze** the data for analysis - this is the **data cutoff**.

```         
Timeline of a clinical trial:
────────────────────────────────────────────────────────────────────────
  First Subject In          Interim Analysis         Final Analysis
       │                         │                        │
       ▼                         ▼                        ▼
  ────────────────────────────────────────────────────────────────────
  ██████████████████████████│ DATA CUT │████████████████│ DATA CUT │
                             (Cutoff 1)                  (Cutoff 2)
                                 │                           │
                                 ▼                           ▼
                        Only data BEFORE           Only data BEFORE
                        this date included         this date included
```

::: {.callout-important collapse="true"}
## Why Data Cuts Matter

-   **Regulatory submissions** require data frozen at a specific cutoff date
-   **Interim analyses** (e.g., for a Data Safety Monitoring Board) need a clean snapshot
-   **Data integrity** - analyses must be reproducible based on the same data
-   **Blinding** - in ongoing studies, only data up to the cutoff should be analyzed
:::

### Types of Data Cuts

| Cut Type | What It Means | Example |
|----|----|----|
| **Patient-level cut** | Exclude entire subjects enrolled after cutoff | Subject consented on 2024-07-01, cutoff is 2024-06-15 → exclude entire subject |
| **Record-level cut** | Keep the subject but remove records after cutoff | Subject enrolled 2024-01-01, AE started 2024-07-01, cutoff 2024-06-15 → keep subject, remove this AE |
| **Hybrid** | Patient-level for some domains, record-level for others | Common in practice |

------------------------------------------------------------------------

## Package Installation & Loading

```{r}
#| label: install-and-load

if (!requireNamespace("dplyr", quietly = TRUE)) suppressMessages(install.packages("dplyr"))
if (!requireNamespace("lubridate", quietly = TRUE)) suppressMessages(install.packages("lubridate"))
if (!requireNamespace("tidyr", quietly = TRUE)) suppressMessages(install.packages("tidyr"))
if (!requireNamespace("stringr", quietly = TRUE)) suppressMessages(install.packages("stringr"))

# datacutr is the Pharmaverse package for data cuts
if (!requireNamespace("datacutr", quietly = TRUE)) suppressMessages(install.packages("datacutr"))

library(dplyr)
library(lubridate)
library(tidyr)
library(stringr)
library(datacutr)
```

------------------------------------------------------------------------

## Understanding datacutr

### What is datacutr?

**`datacutr`** is a Pharmaverse package designed to apply clinical data cutoff dates to SDTM datasets in a standardized, reproducible way. It handles the complex logic of:

-   Identifying which date variable to use for each domain
-   Applying patient-level vs. record-level cuts
-   Handling special cases (ongoing events, missing dates)

### How datacutr Works

The package uses a **datacut metadata** approach:

1.  You define a **patient-level cutoff date** for each subject
2.  You specify **which date variable** to cut on for each domain
3.  `datacutr` applies the cut rules and returns clean datasets

```{r}
#| label: datacutr-functions

# What functions does datacutr provide?
cat("Key datacutr functions:\n\n")
cat("1. create_dcut()        - Create the patient-level datacut dataset\n")
cat("2. date_cut()           - Flag records for removal (adds DCUT_TEMP_REMOVE)\n")
cat("3. apply_cut()          - Remove flagged records\n")
cat("4. special_dm_cut()     - Special handling for DM domain\n")
cat("5. impute_sdtm()        - Impute partial dates before cutting\n")
cat("6. impute_dcutdtc()     - Impute partial datacut dates\n")
cat("7. pt_cut()             - Patient-level cut\n")
cat("8. process_cut()        - Wrapper function for entire workflow\n")
```

------------------------------------------------------------------------

## Setting Up Sample Data for Data Cuts

Let's create a realistic multi-domain SDTM dataset to practice data cuts on:

```{r}
#| label: create-sample-data

set.seed(42)

# ---- Study Parameters ----
STUDYID <- "CDISC01"
cutoff_date <- as.Date("2024-06-15")

n_subjects <- 10
subjects <- paste0("CDISC01-001-", sprintf("%03d", 1:n_subjects))

# ---- DM Domain ----
dm_sample <- tibble(
  STUDYID  = STUDYID,
  DOMAIN   = "DM",
  USUBJID  = subjects,
  RFSTDTC  = as.character(as.Date("2024-01-01") + sample(0:120, n_subjects, replace = TRUE)),
  RFENDTC  = NA_character_,
  DTHDTC   = NA_character_,  # Required by special_dm_cut()
  ARM      = sample(c("Placebo", "Active 10mg", "Active 20mg"), n_subjects, replace = TRUE),
  SEX      = sample(c("M", "F"), n_subjects, replace = TRUE),
  AGE      = sample(30:70, n_subjects, replace = TRUE),
  AGEU     = "YEARS"
) %>%
  mutate(
    # Some subjects started AFTER the cutoff (to test patient-level cuts)
    RFSTDTC = case_when(
      row_number() == 9 ~ as.character(cutoff_date + 5),
      row_number() == 10 ~ as.character(cutoff_date + 15),
      TRUE ~ RFSTDTC
    )
  )

cat("DM domain:", nrow(dm_sample), "subjects\n")
cat("Cutoff date:", as.character(cutoff_date), "\n\n")

# Which subjects started after cutoff?
dm_sample %>%
  mutate(AFTER_CUTOFF = ymd(RFSTDTC) > cutoff_date) %>%
  select(USUBJID, RFSTDTC, ARM, AFTER_CUTOFF)
```

```{r}
#| label: create-ae-data

# ---- AE Domain ----
ae_sample <- tibble(
  STUDYID = STUDYID,
  DOMAIN  = "AE",
  USUBJID = sample(subjects[1:8], 20, replace = TRUE),
  AETERM  = sample(c("Headache", "Nausea", "Fatigue", "Dizziness", "Rash"), 20, replace = TRUE),
  AESEV   = sample(c("MILD", "MODERATE", "SEVERE"), 20, replace = TRUE, prob = c(0.6, 0.3, 0.1)),
  AESER   = sample(c("Y", "N"), 20, replace = TRUE, prob = c(0.1, 0.9))
) %>%
  mutate(AEDECOD = AETERM)

# Add dates - some AEs span across or start after cutoff
ae_sample <- ae_sample %>%
  left_join(dm_sample %>% select(USUBJID, RFSTDTC), by = "USUBJID") %>%
  rowwise() %>%
  mutate(
    start_offset = sample(1:180, 1),
    ae_start = ymd(RFSTDTC) + start_offset,
    AESTDTC = as.character(ae_start),
    # Some AEs are ongoing (no end date), some end after cutoff
    ae_end_offset = sample(c(NA, 3, 7, 14, 30, 60, 90), 1),
    AEENDTC = if_else(!is.na(ae_end_offset),
                      as.character(ae_start + ae_end_offset),
                      NA_character_)
  ) %>%
  ungroup() %>%
  group_by(USUBJID) %>%
  mutate(AESEQ = row_number()) %>%
  ungroup() %>%
  select(STUDYID, DOMAIN, USUBJID, AESEQ, AETERM, AEDECOD, AESEV, AESER,
         AESTDTC, AEENDTC)

cat("AE domain:", nrow(ae_sample), "records\n\n")

# Show which AEs cross the cutoff boundary
ae_sample %>%
  mutate(
    START_VS_CUT = case_when(
      ymd(AESTDTC) > cutoff_date ~ "AFTER cutoff",
      ymd(AESTDTC) <= cutoff_date ~ "BEFORE cutoff",
      TRUE ~ "UNKNOWN"
    ),
    END_VS_CUT = case_when(
      is.na(AEENDTC) ~ "ONGOING",
      ymd(AEENDTC) > cutoff_date ~ "AFTER cutoff",
      ymd(AEENDTC) <= cutoff_date ~ "BEFORE cutoff",
      TRUE ~ "UNKNOWN"
    )
  ) %>%
  count(START_VS_CUT, END_VS_CUT, name = "Count")
```

```{r}
#| label: create-lb-data

# ---- LB Domain (simplified) ----
visits <- tibble(
  VISITNUM = c(1, 2, 3, 4, 5),
  VISIT    = c("BASELINE", "WEEK 4", "WEEK 8", "WEEK 12", "WEEK 16")
)

lb_sample <- expand_grid(
  USUBJID = subjects[1:8],
  tibble(
    LBTESTCD = c("ALT", "CREAT", "GLUC"),
    LBTEST   = c("Alanine Aminotransferase", "Creatinine", "Glucose")
  ),
  visits
) %>%
  left_join(dm_sample %>% select(USUBJID, RFSTDTC), by = "USUBJID") %>%
  mutate(
    STUDYID  = STUDYID,
    DOMAIN   = "LB",
    visit_offset = case_when(
      VISITNUM == 1 ~ 0,
      VISITNUM == 2 ~ 28,
      VISITNUM == 3 ~ 56,
      VISITNUM == 4 ~ 84,
      VISITNUM == 5 ~ 112
    ),
    LBDTC = as.character(ymd(RFSTDTC) + visit_offset),
    LBSTRESN = round(rnorm(n(), 50, 15), 1),
    LBSTRESU = "U/L"
  ) %>%
  group_by(USUBJID) %>%
  mutate(LBSEQ = row_number()) %>%
  ungroup() %>%
  select(STUDYID, DOMAIN, USUBJID, LBSEQ, LBTESTCD, LBTEST,
         LBSTRESN, LBSTRESU, VISITNUM, VISIT, LBDTC)

cat("LB domain:", nrow(lb_sample), "records\n")

# How many LB records are after cutoff?
cat("LB records after cutoff:", 
    sum(ymd(lb_sample$LBDTC) > cutoff_date, na.rm = TRUE), "\n")
```

------------------------------------------------------------------------

## Applying Data Cuts Manually (Understanding the Logic)

Before using `datacutr`, let's understand the logic by implementing cuts manually:

### Patient-Level Cut

```{r}
#| label: manual-patient-cut

# ---- Patient-Level Cut ----
# Remove subjects whose first dose date is AFTER the cutoff

# Step 1: Identify subjects to keep
subjects_to_keep <- dm_sample %>%
  filter(ymd(RFSTDTC) <= cutoff_date) %>%
  pull(USUBJID)

cat("Patient-level cut:\n")
cat("  Subjects before cut:", n_distinct(dm_sample$USUBJID), "\n")
cat("  Subjects after cut:", length(subjects_to_keep), "\n")
cat("  Subjects removed:", n_distinct(dm_sample$USUBJID) - length(subjects_to_keep), "\n\n")

# Step 2: Apply to all domains
dm_cut <- dm_sample %>% filter(USUBJID %in% subjects_to_keep)
ae_patient_cut <- ae_sample %>% filter(USUBJID %in% subjects_to_keep)
lb_patient_cut <- lb_sample %>% filter(USUBJID %in% subjects_to_keep)

cat("Records after patient-level cut:\n")
cat("  DM:", nrow(dm_cut), "\n")
cat("  AE:", nrow(ae_patient_cut), "(from", nrow(ae_sample), ")\n")
cat("  LB:", nrow(lb_patient_cut), "(from", nrow(lb_sample), ")\n")
```

### Record-Level Cut

```{r}
#| label: manual-record-cut

# ---- Record-Level Cut ----
# Keep subjects, but remove individual records after cutoff

# AE: Remove AEs that STARTED after cutoff
ae_record_cut <- ae_patient_cut %>%
  filter(ymd(AESTDTC) <= cutoff_date)

cat("AE Record-level cut:\n")
cat("  Records before:", nrow(ae_patient_cut), "\n")
cat("  Records after:", nrow(ae_record_cut), "\n")
cat("  Records removed:", nrow(ae_patient_cut) - nrow(ae_record_cut), "\n\n")

# LB: Remove lab results collected after cutoff
lb_record_cut <- lb_patient_cut %>%
  filter(ymd(LBDTC) <= cutoff_date)

cat("LB Record-level cut:\n")
cat("  Records before:", nrow(lb_patient_cut), "\n")
cat("  Records after:", nrow(lb_record_cut), "\n")
cat("  Records removed:", nrow(lb_patient_cut) - nrow(lb_record_cut), "\n")
```

::: {.callout-warning collapse="true"}
## What About Ongoing AEs at Cutoff?

An AE that **started before** the cutoff but has **no end date** (ongoing) needs special handling:

-   **Keep the AE record** (it started before cutoff)
-   **Set AEENDTC to the cutoff date** (or leave it blank depending on convention)
-   **Flag it** as "ongoing at data cutoff"

This is where `datacutr` really helps - it handles these edge cases automatically.
:::

### Handling Ongoing AEs at Cutoff

```{r}
#| label: ongoing-ae-handling

# Handle AEs that span the cutoff boundary
ae_final_cut <- ae_patient_cut %>%
  mutate(
    ae_start = ymd(AESTDTC),
    ae_end   = ymd(AEENDTC)
  ) %>%
  # Keep AEs that started on or before cutoff
  filter(ae_start <= cutoff_date) %>%
  mutate(
    # If AE ends AFTER cutoff, truncate the end date to cutoff
    AEENDTC_ORIG = AEENDTC,
    AEENDTC = case_when(
      # AE ends after cutoff → truncate
      !is.na(ae_end) & ae_end > cutoff_date ~ as.character(cutoff_date),
      # AE is ongoing → leave as is (still ongoing at cutoff)
      is.na(ae_end) ~ NA_character_,
      # AE ended before cutoff → keep original
      TRUE ~ AEENDTC
    ),
    # Flag for records modified by the cut
    DCUT_TEMP_REMOVE = "N",
    DCUT_TEMP_DTHCHANGE = if_else(
      !is.na(AEENDTC_ORIG) & AEENDTC != AEENDTC_ORIG,
      "Y", "N"
    )
  ) %>%
  select(-ae_start, -ae_end, -AEENDTC_ORIG)

cat("AE handling at cutoff boundary:\n")
cat("  Records kept:", nrow(ae_final_cut), "\n")
cat("  Records with modified end dates:", 
    sum(ae_final_cut$DCUT_TEMP_DTHCHANGE == "Y", na.rm = TRUE), "\n")
```

------------------------------------------------------------------------

## Using datacutr

Now let's use `datacutr` to do this properly:

### Step 1: Create the Datacut Dataset

```{r}
#| label: datacutr-create-dcut

# The datacut dataset defines per-patient cutoff dates
# In simple cases, all subjects have the same cutoff date
# In complex cases (e.g., rolling enrollment), dates may differ

# For special_dm_cut, we need RFSTDTC and DTHDTC in the dcut dataset
dcut <- dm_sample %>%
  select(USUBJID, RFSTDTC, DTHDTC) %>%
  mutate(
    DCUTDTC = as.character(cutoff_date),  # Same cutoff for all
    DCUTDTM = as.POSIXct(paste0(cutoff_date, " 23:59:59"))
  )

cat("Datacut metadata:\n")
print(dcut)
```

### Step 2: Define the Cut Strategy

```{r}
#| label: datacutr-strategy

# datacutr uses a two-step process:
# 1. date_cut() flags records (adds DCUT_TEMP_REMOVE)
# 2. apply_cut() removes flagged records

# For DM: special_dm_cut() handles patient-level cut based on RFSTDTC
cat("Cut strategy:\n")
cat("  DM: Patient-level cut on RFSTDTC (first dose date)\n")
cat("  AE: Record-level cut on AESTDTC (AE start date)\n")
cat("  LB: Record-level cut on LBDTC (lab collection date)\n\n")
```

### Step 3: Apply the Cut Using datacutr Functions

```{r}
#| label: apply-datacutr

# ---- Step 3a: DM Domain ----
# special_dm_cut handles the patient-level cut for DM
# It compares RFSTDTC (from dcut) against DCUTDTC
dm_after_cut <- special_dm_cut(
  dataset_dm = dm_sample,
  dataset_cut = dcut
)

cat("DM after cut:", nrow(dm_after_cut), "subjects (from", nrow(dm_sample), ")\n")
```

```{r}
#| label: apply-datacutr-ae

# ---- Step 3b: AE Domain ----
# date_cut handles record-level cuts and adds DCUT_TEMP_REMOVE flag
# First, we need to know which subjects passed the DM cut
subjects_after_dm_cut <- dm_after_cut$USUBJID

ae_for_cut <- ae_sample %>%
  filter(USUBJID %in% subjects_after_dm_cut)

ae_after_cut_temp <- date_cut(
  dataset_sdtm = ae_for_cut,
  sdtm_date_var = AESTDTC,
  dataset_cut = dcut,
  cut_var = DCUTDTM
)

# Apply the cut using apply_cut which removes flagged records
ae_after_cut <- apply_cut(
  dsin = ae_after_cut_temp,
  dcutvar = DCUT_TEMP_REMOVE,
  dthchangevar = DCUT_TEMP_DTHCHANGE
)

cat("AE after cut:", nrow(ae_after_cut), "records (from", nrow(ae_sample), ")\n")
```

```{r}
#| label: apply-datacutr-lb

# ---- Step 3c: LB Domain ----
lb_for_cut <- lb_sample %>%
  filter(USUBJID %in% subjects_after_dm_cut)

lb_after_cut_temp <- date_cut(
  dataset_sdtm = lb_for_cut,
  sdtm_date_var = LBDTC,
  dataset_cut = dcut,
  cut_var = DCUTDTM
)

# Apply the cut using apply_cut which removes flagged records
lb_after_cut <- apply_cut(
  dsin = lb_after_cut_temp,
  dcutvar = DCUT_TEMP_REMOVE,
  dthchangevar = DCUT_TEMP_DTHCHANGE
)

cat("LB after cut:", nrow(lb_after_cut), "records (from", nrow(lb_sample), ")\n")
```

### Step 4: Summary of Data Cut

```{r}
#| label: cut-summary

cat("=== DATA CUT SUMMARY ===\n")
cat("Cutoff date:", as.character(cutoff_date), "\n\n")

summary_table <- tibble::tribble(
  ~Domain, ~Before_Cut, ~After_Cut, ~Removed,
  "DM", nrow(dm_sample), nrow(dm_after_cut), nrow(dm_sample) - nrow(dm_after_cut),
  "AE", nrow(ae_sample), nrow(ae_after_cut), nrow(ae_sample) - nrow(ae_after_cut),
  "LB", nrow(lb_sample), nrow(lb_after_cut), nrow(lb_sample) - nrow(lb_after_cut)
)

print(summary_table)
```

------------------------------------------------------------------------

## Real-World Data Cut Considerations

### Partial Dates

```{r}
#| label: partial-dates

# In real data, you'll encounter partial dates
cat("Common partial date scenarios:\n\n")

partial_examples <- tibble::tribble(
  ~AESTDTC,       ~Problem,                          ~Solution,
  "2024-06",       "Day missing",                    "Impute to 1st or last of month",
  "2024",          "Month and day missing",          "Impute based on SAP rules",
  "",              "Completely missing",             "Flag for medical review",
  "2024-06-15",    "Complete date",                  "No imputation needed",
  "2024-06-UN",    "Day unknown (UN notation)",      "Impute per convention"
)

print(partial_examples)
```

::: {.callout-tip collapse="true"}
## datacutr and Partial Dates

The `datacutr::impute_sdtm()` function can handle partial date imputation before applying the cut. Common imputation strategies:

-   **Conservative (for safety)**: Impute missing day to the **1st of the month** - this maximizes the chance of including the record (more conservative for safety data)
-   **Conservative (for efficacy)**: Impute missing day to the **last of the month** - interpretation depends on the analysis
-   **SAP-driven**: Always follow the Statistical Analysis Plan's imputation rules
:::

### Multiple Cutoff Dates

```{r}
#| label: multiple-cutoffs

# In some studies, cutoff dates differ by subject (rolling enrollment)
# or by analysis (interim vs final)

cat("Example: Different cutoffs per analysis:\n\n")

analysis_cuts <- tibble::tribble(
  ~Analysis,          ~Cutoff_Date,     ~Purpose,
  "DSMB Review 1",    "2024-03-31",     "Safety review at 50% enrollment",
  "Interim Analysis",  "2024-06-15",     "Pre-specified interim for efficacy",
  "Final Analysis",    "2024-12-31",     "Primary analysis for submission"
)

print(analysis_cuts)
```

------------------------------------------------------------------------

## Best Practices for Data Cuts

### DO

```{r}
#| label: best-practices

cat("=== DATA CUT BEST PRACTICES ===\n\n")

best_practices <- tibble::tribble(
  ~Rule,                                              ~Why,
  "Document the cutoff date in TS domain",            "Traceability for regulators",
  "Apply cuts BEFORE creating ADaM datasets",         "ADaM should be based on cut data",
  "Keep both pre-cut and post-cut datasets",          "Audit trail and reproducibility",
  "Handle ongoing events explicitly",                 "AEs without end dates need special logic",
  "Use datacutr for consistency",                     "Avoids manual errors across domains",
  "Log all records removed by the cut",               "Quality control and transparency",
  "Validate record counts before and after",          "Ensure no data is accidentally lost"
)

print(best_practices)
```

### The Data Cut in the Submission Package

```{r}
#| label: submission-context

cat("Where the data cut fits in the submission process:\n\n")
cat("1. Raw data collected continuously\n")
cat("2. DATABASE LOCK (data cleaning complete)\n")
cat("3. DATA CUT applied (datacutr)\n")
cat("4. SDTM datasets created from cut data\n")
cat("5. ADaM datasets derived from SDTM\n")
cat("6. TFLs (Tables, Figures, Listings) generated from ADaM\n")
cat("7. Clinical Study Report written\n")
cat("8. Submission package assembled (eCTD)\n\n")

cat("The cutoff date is recorded in:\n")
cat("  - TS domain (TSPARMCD = 'DCUTDTC')\n")
cat("  - Study report\n")
cat("  - define.xml\n")
```

------------------------------------------------------------------------

## Deliverable Summary

Today you completed the following:

| Task | Status |
|----|----|
| Understood what data cuts are and why they matter | ✓ Done |
| Implemented patient-level and record-level cuts manually | ✓ Done |
| Handled ongoing AEs at the cutoff boundary | ✓ Done |
| Used `datacutr` functions: `special_dm_cut()`, `date_cut()`, `apply_cut()` | ✓ Done |
| Learned about partial date imputation | ✓ Done |
| Reviewed best practices for data cuts | ✓ Done |

------------------------------------------------------------------------

## Key Takeaways

1.  **Data cuts freeze the data** - Only data before the cutoff is included in the analysis
2.  **Patient-level cuts** exclude entire subjects; **record-level cuts** exclude individual records
3.  **Ongoing events need special handling** - AEs that span the cutoff require careful logic
4.  **`datacutr` standardizes the process** - Avoids manual errors and ensures reproducibility
5.  **Cuts happen BEFORE SDTM/ADaM** - The cut data is the foundation for all downstream datasets
6.  **Document everything** - The cutoff date must be recorded in TS and the study report

------------------------------------------------------------------------

## Resources

-   [datacutr Documentation](https://pharmaverse.github.io/datacutr/) - Official datacutr package documentation
-   [datacutr GitHub](https://github.com/pharmaverse/datacutr) - Source code and examples
-   [CDISC Implementation Guide - Data Cuts](https://www.cdisc.org/standards/foundational/sdtm) - SDTM guidance
-   [Pharmaverse.org](https://pharmaverse.org) - R packages for clinical data
-   [ICH E9(R1)](https://database.ich.org/sites/default/files/E9-R1_Step4_Guideline_2019_1203.pdf) - Estimands and data handling

------------------------------------------------------------------------

## What's Next?

In **Day 13**, we will focus on **SDTM Validation with sdtmchecks**:

-   Why validation is essential before creating ADaM datasets
-   Running FDA business rules against your SDTM domains
-   Understanding validation reports and resolving findings
-   Common SDTM issues and how to fix them

___
