---
title: "Day 11: Disposition (DS) & Trial Design Domains"
subtitle: "Screen Failures, Completers, and Study Structure"


format:
  html:
    theme:
      light: cosmo

    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css
    
editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

<div style="margin-bottom:12px;"><a href="../index.html" class="resource-link">← Back to Roadmap</a></div>

## Learning Objectives

By the end of Day 11, you will be able to:

1.  Build the **DS (Disposition) domain** - tracking every subject's journey through the trial
2.  Understand **EPOCH** and how it relates to study periods (SCREENING, TREATMENT, FOLLOW-UP)
3.  Derive key disposition variables: **DSDECOD**, **DSSTDTC**, **DSCAT**, and completion flags
4.  Survey the **Trial Design domains** (TA, TE, TV, TI, TS) and understand what each contains
5.  Create a **subject disposition flow** - the essential "patient accounting" table
6.  Understand how DS feeds into the **ADaM ADSL** dataset

------------------------------------------------------------------------

## Why Disposition Matters

The DS domain answers the most fundamental question in a clinical trial: **What happened to each subject?**

::: callout-important
## Regulatory Requirement

The FDA requires a complete accounting of **every subject** who entered the study:

-   Did they complete the study? → `COMPLETED`
-   Did they withdraw early? → Why? (adverse event, lost to follow-up, withdrawal by subject, etc.)
-   Were they a screen failure? → Why? (did not meet inclusion criteria, etc.)

This information feeds into **CONSORT diagrams**, disposition tables, and the ADaM `ADSL` dataset - all critical for regulatory submissions.
:::

------------------------------------------------------------------------

## Package Installation & Loading

```{r}
#| label: install-and-load

if (!requireNamespace("dplyr", quietly = TRUE)) suppressMessages(install.packages("dplyr"))
if (!requireNamespace("lubridate", quietly = TRUE)) suppressMessages(install.packages("lubridate"))
if (!requireNamespace("tidyr", quietly = TRUE)) suppressMessages(install.packages("tidyr"))
if (!requireNamespace("pharmaversesdtm", quietly = TRUE)) suppressMessages(install.packages("pharmaversesdtm"))

library(dplyr)
library(lubridate)
library(tidyr)
library(pharmaversesdtm)
```

------------------------------------------------------------------------

## The DS Domain Structure

### Key Variables

```
┌──────────────────────────────────────────────────────────────────────────────┐
│                        DS DOMAIN - KEY VARIABLES                             │
├──────────────────────────────────────────────────────────────────────────────┤
│  IDENTIFIER VARIABLES                                                        │
│    STUDYID    = Study identifier                                             │
│    USUBJID    = Unique subject identifier                                    │
│    DSSEQ      = Sequence number within subject                               │
├──────────────────────────────────────────────────────────────────────────────┤
│  TOPIC / RESULT VARIABLES                                                    │
│    DSTERM     = Reported disposition term (as collected)                      │
│    DSDECOD    = Standardized disposition term (CT)                            │
│    DSCAT      = Category: "DISPOSITION EVENT" or "PROTOCOL MILESTONE"        │
│    DSSCAT     = Subcategory: e.g., "STUDY PARTICIPATION", "TREATMENT"        │
├──────────────────────────────────────────────────────────────────────────────┤
│  TIMING VARIABLES                                                            │
│    DSSTDTC    = Date/time of disposition event (ISO 8601)                     │
│    DSDY       = Study day of disposition event                               │
│    EPOCH      = Epoch (SCREENING, TREATMENT, FOLLOW-UP)                      │
│    VISITNUM   = Visit number                                                 │
│    VISIT      = Visit name                                                   │
└──────────────────────────────────────────────────────────────────────────────┘
```

### Understanding DSCAT and DSSCAT

These categorization variables are critical:

| DSCAT | DSSCAT | DSDECOD Examples |
|-------|--------|-----------------|
| `DISPOSITION EVENT` | `STUDY PARTICIPATION` | COMPLETED, SCREEN FAILURE |
| `DISPOSITION EVENT` | `TREATMENT` | COMPLETED, ADVERSE EVENT, WITHDRAWAL BY SUBJECT |
| `PROTOCOL MILESTONE` | `INFORMED CONSENT` | INFORMED CONSENT OBTAINED |
| `PROTOCOL MILESTONE` | `RANDOMIZATION` | RANDOMIZED |

::: callout-note
## DSCAT vs DSSCAT

Think of it this way:

-   **DSCAT** = The broad type of event (disposition event vs. protocol milestone)
-   **DSSCAT** = Which part of the study this relates to (study participation vs. treatment)

A subject can have **multiple DS records**: one for each protocol milestone and each disposition decision point.
:::

------------------------------------------------------------------------

## Exploring DS Data from pharmaversesdtm

```{r}
#| label: load-ds-data

# Load DS domain
data("ds", package = "pharmaversesdtm")
data("dm", package = "pharmaversesdtm")

cat("DS domain dimensions:", nrow(ds), "rows x", ncol(ds), "columns\n")
cat("Number of unique subjects:", n_distinct(ds$USUBJID), "\n\n")

cat("Variables available:\n")
cat(paste(names(ds), collapse = ", "), "\n")
```

```{r}
#| label: glimpse-ds

dplyr::glimpse(ds)
```

### Understanding the DS Records

```{r}
#| label: explore-ds-categories

# What categories exist?
cat("DS Categories:\n")
# Use only columns that exist in the dataset
cat_vars <- intersect(c("DSCAT", "DSSCAT"), names(ds))
ds %>%
  count(across(all_of(cat_vars)), name = "Count") %>%
  arrange(across(all_of(cat_vars)))
```

```{r}
#| label: explore-ds-terms

# What disposition terms exist?
cat("Disposition terms (DSDECOD):\n")
term_vars <- intersect(c("DSCAT", "DSSCAT", "DSDECOD"), names(ds))
ds %>%
  count(across(all_of(term_vars)), name = "Count") %>%
  arrange(across(all_of(term_vars[term_vars != "DSDECOD"])), desc(Count))
```

::: callout-tip
## Reading DS Data

Each subject typically has **multiple records** in DS:

1.  **Informed consent record** - When they signed the ICF
2.  **Randomization record** - When they were randomized (if applicable)
3.  **Treatment disposition** - Did they complete treatment? If not, why?
4.  **Study disposition** - Did they complete the study? If not, why?

These records tell the full story of each subject's journey.
:::

------------------------------------------------------------------------

## Understanding EPOCH

### What is EPOCH?

**EPOCH** is a variable that indicates which period of the study the data belongs to. It divides the trial into distinct phases:

```
Timeline:
────────────────────────────────────────────────────────────────────
  SCREENING    │   TREATMENT (Run-in)   │   TREATMENT    │   FOLLOW-UP
  (Pre-study)  │   (Placebo lead-in)    │   (Active)     │   (Post-treatment)
────────────────────────────────────────────────────────────────────
```

```{r}
#| label: epoch-exploration

# Check EPOCH values in DM (EPOCH is often in DM as well)
if ("EPOCH" %in% names(ds)) {
  cat("EPOCH values in DS:\n")
  ds %>%
    count(EPOCH, name = "Count") %>%
    arrange(desc(Count))
} else {
  cat("EPOCH not found in DS. Checking DM...\n")
  if ("EPOCH" %in% names(dm)) {
    cat("EPOCH values in DM:\n")
    dm %>% count(EPOCH) %>% print()
  }
}
```

### EPOCH in the DM Domain

The DM domain has **EPOCH-related variables** that indicate the last known state of the subject:

```{r}
#| label: dm-epoch-vars

# Key epoch/disposition-related variables in DM
dm_disp_vars <- intersect(
  c("USUBJID", "ARM", "ACTARM", "RFSTDTC", "RFENDTC", "DTHDTC", "DTHFL"),
  names(dm)
)
dm %>%
  select(all_of(dm_disp_vars)) %>%
  head(10)
```

::: callout-note
## Important DM Variables Related to DS

| Variable | Description |
|----------|-------------|
| `RFSTDTC` | Reference start date (first exposure) |
| `RFENDTC` | Reference end date (last exposure) |
| `RFXSTDTC` | Date/time of first study treatment |
| `RFXENDTC` | Date/time of last study treatment |
| `DTHDTC` | Date of death |
| `DTHFL` | Death flag (Y/N) |
| `DSDECOD` | Disposition decoded term (in some implementations) |

These DM variables are **derived from** the DS and EX domain data.
:::

------------------------------------------------------------------------

## Building a DS Domain from Scratch

Let's simulate raw disposition data and build a complete DS domain:

### Step 1: Create Raw Disposition Data

```{r}
#| label: raw-disposition

set.seed(42)

n_subjects <- 30

# Generate subject disposition decisions
raw_disp <- tibble(
  STUDYID = "CDISC01",
  USUBJID = paste0("CDISC01-001-", sprintf("%03d", 1:n_subjects)),
  
  # Informed consent date
  CONSENT_DATE = as.Date("2024-01-01") + sample(0:30, n_subjects, replace = TRUE),
  
  # Screening result
  SCREEN_RESULT = sample(
    c("PASSED", "FAILED"),
    n_subjects, replace = TRUE, prob = c(0.85, 0.15)
  ),
  SCREEN_REASON = if_else(
    SCREEN_RESULT == "FAILED",
    sample(c("DID NOT MEET INCLUSION CRITERIA",
             "EXCLUSION CRITERIA MET",
             "WITHDREW CONSENT"),
           n_subjects, replace = TRUE, prob = c(0.5, 0.3, 0.2)),
    NA_character_
  ),
  
  # Randomization date (only for those who passed screening)
  RAND_DATE = if_else(
    SCREEN_RESULT == "PASSED",
    CONSENT_DATE + sample(7:21, n_subjects, replace = TRUE),
    as.Date(NA)
  ),
  
  # Treatment arm (only for randomized subjects)
  ARMCD = if_else(
    SCREEN_RESULT == "PASSED",
    sample(c("PBO", "ACT10", "ACT20"), n_subjects, replace = TRUE),
    NA_character_
  )
) %>%
  # Add first dose date
  mutate(
    FIRST_DOSE = if_else(!is.na(RAND_DATE), RAND_DATE + sample(0:3, n(), replace = TRUE), as.Date(NA))
  )

# Determine treatment disposition for randomized subjects
raw_disp <- raw_disp %>%
  mutate(
    TRT_RESULT = case_when(
      SCREEN_RESULT == "FAILED" ~ NA_character_,
      TRUE ~ sample(c("COMPLETED", "ADVERSE EVENT", "WITHDRAWAL BY SUBJECT",
                       "LOST TO FOLLOW-UP", "PHYSICIAN DECISION"),
                    n(), replace = TRUE,
                    prob = c(0.65, 0.12, 0.10, 0.08, 0.05))
    ),
    # End of treatment date
    TRT_END_DATE = case_when(
      is.na(FIRST_DOSE) ~ as.Date(NA),
      TRT_RESULT == "COMPLETED" ~ FIRST_DOSE + 84,  # 12-week study
      TRUE ~ FIRST_DOSE + sample(7:60, n(), replace = TRUE)  # Early termination
    )
  )

cat("Raw disposition data:\n")
cat("Total subjects:", nrow(raw_disp), "\n")
cat("Screen failures:", sum(raw_disp$SCREEN_RESULT == "FAILED"), "\n")
cat("Randomized:", sum(raw_disp$SCREEN_RESULT == "PASSED"), "\n\n")

raw_disp %>%
  select(USUBJID, SCREEN_RESULT, TRT_RESULT) %>%
  head(10)
```

### Step 2: Transform to SDTM DS

```{r}
#| label: build-ds-domain

# Create DS records for each disposition event
# A subject may have MULTIPLE DS records

# Record 1: Informed Consent
ds_consent <- raw_disp %>%
  mutate(
    DOMAIN   = "DS",
    DSSEQ    = 1L,
    DSTERM   = "INFORMED CONSENT OBTAINED",
    DSDECOD  = "INFORMED CONSENT OBTAINED",
    DSCAT    = "PROTOCOL MILESTONE",
    DSSCAT   = "INFORMED CONSENT",
    DSSTDTC  = as.character(CONSENT_DATE),
    EPOCH    = "SCREENING"
  ) %>%
  select(STUDYID, DOMAIN, USUBJID, DSSEQ, DSTERM, DSDECOD, DSCAT, DSSCAT, DSSTDTC, EPOCH)

# Record 2: Randomization (only for those who passed screening)
ds_random <- raw_disp %>%
  filter(SCREEN_RESULT == "PASSED") %>%
  mutate(
    DOMAIN   = "DS",
    DSSEQ    = 2L,
    DSTERM   = "RANDOMIZED",
    DSDECOD  = "RANDOMIZED",
    DSCAT    = "PROTOCOL MILESTONE",
    DSSCAT   = "RANDOMIZATION",
    DSSTDTC  = as.character(RAND_DATE),
    EPOCH    = "SCREENING"
  ) %>%
  select(STUDYID, DOMAIN, USUBJID, DSSEQ, DSTERM, DSDECOD, DSCAT, DSSCAT, DSSTDTC, EPOCH)

# Record 3: Treatment Disposition
ds_treatment <- raw_disp %>%
  filter(!is.na(TRT_RESULT)) %>%
  mutate(
    DOMAIN   = "DS",
    DSSEQ    = 3L,
    DSTERM   = TRT_RESULT,
    DSDECOD  = TRT_RESULT,
    DSCAT    = "DISPOSITION EVENT",
    DSSCAT   = "TREATMENT",
    DSSTDTC  = as.character(TRT_END_DATE),
    EPOCH    = "TREATMENT"
  ) %>%
  select(STUDYID, DOMAIN, USUBJID, DSSEQ, DSTERM, DSDECOD, DSCAT, DSSCAT, DSSTDTC, EPOCH)

# Record 4: Screen Failure disposition
ds_screenfail <- raw_disp %>%
  filter(SCREEN_RESULT == "FAILED") %>%
  mutate(
    DOMAIN   = "DS",
    DSSEQ    = 2L,
    DSTERM   = SCREEN_REASON,
    DSDECOD  = "SCREEN FAILURE",
    DSCAT    = "DISPOSITION EVENT",
    DSSCAT   = "STUDY PARTICIPATION",
    DSSTDTC  = as.character(CONSENT_DATE + sample(1:14, n(), replace = TRUE)),
    EPOCH    = "SCREENING"
  ) %>%
  select(STUDYID, DOMAIN, USUBJID, DSSEQ, DSTERM, DSDECOD, DSCAT, DSSCAT, DSSTDTC, EPOCH)

# Combine all DS records
sdtm_ds <- bind_rows(ds_consent, ds_random, ds_treatment, ds_screenfail) %>%
  # Re-sequence within each subject
  arrange(STUDYID, USUBJID, DSSTDTC) %>%
  group_by(USUBJID) %>%
  mutate(DSSEQ = row_number()) %>%
  ungroup()

cat("SDTM DS domain created:\n")
cat("Total records:", nrow(sdtm_ds), "\n")
cat("Unique subjects:", n_distinct(sdtm_ds$USUBJID), "\n\n")

sdtm_ds %>%
  filter(USUBJID == first(sdtm_ds$USUBJID)) %>%
  print()
```

### Step 3: Validate DS Domain

```{r}
#| label: validate-ds

cat("=== DS Domain Validation ===\n\n")

# Check 1: Every subject has at least one record
cat("Subjects with DS records:", n_distinct(sdtm_ds$USUBJID), "\n")

# Check 2: Every subject has informed consent
consent_check <- sdtm_ds %>%
  filter(DSDECOD == "INFORMED CONSENT OBTAINED") %>%
  n_distinct(.$USUBJID)
cat("Subjects with informed consent:", consent_check, "\n\n")

# Check 3: Disposition summary
cat("Treatment Disposition Summary:\n")
sdtm_ds %>%
  filter(DSCAT == "DISPOSITION EVENT", DSSCAT == "TREATMENT") %>%
  count(DSDECOD, name = "Count") %>%
  mutate(Percent = round(100 * Count / sum(Count), 1)) %>%
  arrange(desc(Count))
```

------------------------------------------------------------------------

## Subject Disposition Flow

This is one of the most important outputs in a clinical study report:

```{r}
#| label: disposition-flow

# Create a CONSORT-style flow summary
cat("=== SUBJECT DISPOSITION FLOW ===\n\n")

n_consented <- n_distinct(sdtm_ds$USUBJID)
cat("Subjects consented:         ", n_consented, "\n")

n_screened_out <- sdtm_ds %>%
  filter(DSDECOD == "SCREEN FAILURE") %>%
  n_distinct(.$USUBJID)
cat("  Screen failures:           ", n_screened_out, "\n")

# Screen failure reasons
if (n_screened_out > 0) {
  sdtm_ds %>%
    filter(DSDECOD == "SCREEN FAILURE") %>%
    count(DSTERM, name = "N") %>%
    arrange(desc(N)) %>%
    mutate(line = paste0("    - ", DSTERM, ": ", N)) %>%
    pull(line) %>%
    cat(sep = "\n")
  cat("\n")
}

n_randomized <- sdtm_ds %>%
  filter(DSDECOD == "RANDOMIZED") %>%
  n_distinct(.$USUBJID)
cat("Subjects randomized:        ", n_randomized, "\n")

n_completed <- sdtm_ds %>%
  filter(DSCAT == "DISPOSITION EVENT", DSSCAT == "TREATMENT", DSDECOD == "COMPLETED") %>%
  n_distinct(.$USUBJID)
cat("  Completed treatment:       ", n_completed, "\n")

n_discontinued <- n_randomized - n_completed
cat("  Discontinued treatment:    ", n_discontinued, "\n")

# Discontinuation reasons
if (n_discontinued > 0) {
  sdtm_ds %>%
    filter(DSCAT == "DISPOSITION EVENT", DSSCAT == "TREATMENT", DSDECOD != "COMPLETED") %>%
    count(DSDECOD, name = "N") %>%
    arrange(desc(N)) %>%
    mutate(line = paste0("    - ", DSDECOD, ": ", N)) %>%
    pull(line) %>%
    cat(sep = "\n")
  cat("\n")
}

cat("\nCompletion rate:", round(100 * n_completed / n_randomized, 1), "%\n")
```

------------------------------------------------------------------------

## Trial Design Domains Overview

### What Are Trial Design Domains?

Trial Design domains describe the **structure of the study itself**, not individual subject data. Think of them as the **blueprint** of the clinical trial.

| Domain | Name | Purpose |
|--------|------|---------|
| **TA** | Trial Arms | Describes each treatment arm and its sequence of elements |
| **TE** | Trial Elements | Describes the building blocks of each arm (e.g., washout, treatment) |
| **TV** | Trial Visits | Lists all planned visits and their timing windows |
| **TI** | Trial Inclusion/Exclusion | Documents the inclusion/exclusion criteria |
| **TS** | Trial Summary | Key parameters about the study (phase, indication, design) |

::: callout-note
## When Do You Need These?

Trial Design domains are **required** for FDA submissions. They provide the **context** for understanding the subject-level data. An FDA reviewer needs to know:

-   What arms existed? (TA)
-   What was the study schedule? (TV)
-   What were the eligibility criteria? (TI)
-   What type of study was this? (TS)
:::

### TA - Trial Arms

```{r}
#| label: trial-arms

# Create a Trial Arms (TA) domain example
ta <- tibble::tribble(
  ~STUDYID,  ~DOMAIN, ~ARMCD,  ~ARM,           ~TAESSION, ~TATRANS,       ~EPOCH,       ~ELEMENT,
  "CDISC01", "TA",    "PBO",   "Placebo",      1L,        "No criteria",  "SCREENING",  "Screen",
  "CDISC01", "TA",    "PBO",   "Placebo",      2L,        "Randomized",   "TREATMENT",  "Placebo",
  "CDISC01", "TA",    "PBO",   "Placebo",      3L,        "End treatment","FOLLOW-UP",  "Follow-up",
  "CDISC01", "TA",    "ACT10", "Active 10mg",  1L,        "No criteria",  "SCREENING",  "Screen",
  "CDISC01", "TA",    "ACT10", "Active 10mg",  2L,        "Randomized",   "TREATMENT",  "Active 10mg",
  "CDISC01", "TA",    "ACT10", "Active 10mg",  3L,        "End treatment","FOLLOW-UP",  "Follow-up",
  "CDISC01", "TA",    "ACT20", "Active 20mg",  1L,        "No criteria",  "SCREENING",  "Screen",
  "CDISC01", "TA",    "ACT20", "Active 20mg",  2L,        "Randomized",   "TREATMENT",  "Active 20mg",
  "CDISC01", "TA",    "ACT20", "Active 20mg",  3L,        "End treatment","FOLLOW-UP",  "Follow-up"
)

cat("Trial Arms (TA):\n")
print(ta)
```

### TV - Trial Visits

```{r}
#| label: trial-visits

# Create a Trial Visits (TV) domain example
tv <- tibble::tribble(
  ~STUDYID,  ~DOMAIN, ~VISITNUM, ~VISIT,       ~TVSTRL,     ~TVENRL,
  "CDISC01", "TV",    -1,        "SCREENING",  "Day -21",   "Day -1",
  "CDISC01", "TV",     1,        "BASELINE",   "Day 1",     "Day 1",
  "CDISC01", "TV",     2,        "WEEK 2",     "Day 8",     "Day 18",
  "CDISC01", "TV",     3,        "WEEK 4",     "Day 22",    "Day 32",
  "CDISC01", "TV",     4,        "WEEK 8",     "Day 50",    "Day 64",
  "CDISC01", "TV",     5,        "WEEK 12",    "Day 78",    "Day 92",
  "CDISC01", "TV",     6,        "END OF STUDY","Day 92",   "Day 106"
)

cat("Trial Visits (TV) - Study Schedule:\n")
print(tv)
```

::: callout-tip
## Visit Windows

Notice `TVSTRL` and `TVENRL` - these define the **visit window**. In practice:

-   A "WEEK 2" visit might be acceptable if it occurs between Day 8 and Day 18
-   Programming logic must map actual visit dates to planned visits based on these windows
-   Visits outside windows may be flagged as **unscheduled visits**
:::

### TS - Trial Summary

```{r}
#| label: trial-summary

# Create a Trial Summary (TS) domain example
ts <- tibble::tribble(
  ~STUDYID,  ~DOMAIN, ~TSSEQ, ~TSPARMCD,  ~TSPARM,                    ~TSVAL,
  "CDISC01", "TS",    1L,     "ADDON",     "Added on to Existing Treatments", "N",
  "CDISC01", "TS",    2L,     "SSTDTC",    "Study Start Date",         "2024-01-01",
  "CDISC01", "TS",    3L,     "SENDTC",    "Study End Date",           "2024-12-31",
  "CDISC01", "TS",    4L,     "INDIC",     "Trial Indication",        "Type 2 Diabetes Mellitus",
  "CDISC01", "TS",    5L,     "TRT",       "Investigational Therapy", "Study Drug XYZ",
  "CDISC01", "TS",    6L,     "TPHASE",    "Trial Phase",             "PHASE III",
  "CDISC01", "TS",    7L,     "STYPE",     "Study Type",              "INTERVENTIONAL",
  "CDISC01", "TS",    8L,     "RANDOM",    "Trial is Randomized",     "Y",
  "CDISC01", "TS",    9L,     "STOTEFFC",  "Study to Evaluate",       "EFFICACY AND SAFETY",
  "CDISC01", "TS",    10L,    "PCLAS",     "Pharmacological Class",   "SGLT2 INHIBITOR"
)

cat("Trial Summary (TS) - Key Study Parameters:\n")
print(ts)
```

::: callout-note
## TS Parameters

The TS domain uses a **parameter-value** structure where each row is one study-level parameter. Common parameters include:

-   **TPHASE** - Phase I, II, III, IV
-   **INDIC** - What disease is being studied
-   **RANDOM** - Is this a randomized trial
-   **STYPE** - Interventional vs. Observational
-   **REGID** - Regulatory registration number (e.g., IND number)

The full list of valid TS parameters is in the [CDISC CT Package](https://www.cancer.gov/research/resources/terminology/cdisc).
:::

------------------------------------------------------------------------

## How DS Feeds Into ADSL

The DS domain is a primary source for the **ADaM ADSL** (Subject-Level Analysis Dataset). Here's the connection:

```{r}
#| label: ds-to-adsl-preview

# Preview: Deriving ADSL disposition variables from DS
adsl_disp <- raw_disp %>%
  mutate(
    # Core ADSL variables derived from DS logic
    RANDFL = if_else(SCREEN_RESULT == "PASSED", "Y", "N"),
    RANDDT = RAND_DATE,
    
    # Treatment completion
    EOSSTT = case_when(
      SCREEN_RESULT == "FAILED" ~ "SCREEN FAILURE",
      TRT_RESULT == "COMPLETED" ~ "COMPLETED",
      TRUE ~ "DISCONTINUED"
    ),
    
    # Reason for discontinuation
    DCSREAS = case_when(
      EOSSTT == "COMPLETED" ~ NA_character_,
      EOSSTT == "SCREEN FAILURE" ~ SCREEN_REASON,
      TRUE ~ TRT_RESULT
    ),
    
    # Safety population flag (received at least one dose)
    SAFFL = if_else(!is.na(FIRST_DOSE), "Y", "N"),
    
    # ITT population flag (randomized)
    ITTFL = RANDFL,
    
    # Treatment duration (days)
    TRTDURD = if_else(
      !is.na(FIRST_DOSE) & !is.na(TRT_END_DATE),
      as.numeric(TRT_END_DATE - FIRST_DOSE) + 1,
      NA_real_
    )
  )

cat("ADSL Disposition Variables Preview:\n")
adsl_disp %>%
  select(USUBJID, RANDFL, EOSSTT, DCSREAS, SAFFL, ITTFL, TRTDURD) %>%
  head(10)
```

```{r}
#| label: adsl-population-summary

# Population flags summary
cat("\n=== Analysis Population Summary ===\n")
cat("Screened:", nrow(adsl_disp), "\n")
cat("Randomized (ITT):", sum(adsl_disp$ITTFL == "Y"), "\n")
cat("Safety Population:", sum(adsl_disp$SAFFL == "Y"), "\n\n")

cat("End of Study Status:\n")
adsl_disp %>%
  count(EOSSTT, name = "N") %>%
  mutate(Percent = round(100 * N / sum(N), 1)) %>%
  arrange(desc(N))
```

::: callout-important
## Key ADSL Variables from DS

| ADSL Variable | Source | Description |
|---------------|--------|-------------|
| RANDFL | DS (RANDOMIZED record) | Randomized flag (Y/N) |
| RANDDT | DS (DSSTDTC where DSDECOD = RANDOMIZED) | Randomization date |
| EOSSTT | DS (treatment disposition) | End of study status |
| DCSREAS | DS (DSTERM for discontinuation) | Reason for discontinuation |
| SAFFL | EX (received at least one dose) | Safety population flag |
| ITTFL | DS (randomized) | ITT population flag |
| TRTDURD | Derived from EX dates | Treatment duration |

Understanding these mappings is essential for ADaM work in Week 3.
:::

------------------------------------------------------------------------

## Deliverable Summary

Today you completed the following:

| Task | Status |
|------|--------|
| Understood DS domain structure and purpose | ✓ Done |
| Built a DS domain with consent, randomization, and treatment disposition | ✓ Done |
| Explored EPOCH and study period concepts | ✓ Done |
| Surveyed Trial Design domains (TA, TV, TS) | ✓ Done |
| Created subject disposition flow (CONSORT-style) | ✓ Done |
| Previewed how DS feeds into ADSL | ✓ Done |

------------------------------------------------------------------------

## Key Takeaways

1.  **DS is an Events class domain** - Each record is a milestone or disposition event
2.  **Multiple records per subject** - Consent, randomization, treatment disposition, study disposition
3.  **DSCAT/DSSCAT distinguish event types** - Milestones vs. disposition decisions
4.  **EPOCH defines study periods** - Screening, Treatment, Follow-up
5.  **Trial Design domains are required** - TA, TE, TV, TI, TS describe the study blueprint
6.  **DS drives ADSL** - Randomization, completion, discontinuation all come from DS

------------------------------------------------------------------------

## Resources

-   [CDISC SDTM Implementation Guide - DS Domain](https://www.cdisc.org/standards/foundational/sdtm) - Official DS specification
-   [Trial Design Domains Guide](https://www.cdisc.org/standards/foundational/sdtm) - TA, TE, TV, TI, TS specifications
-   [CONSORT Statement](http://www.consort-statement.org/) - Reporting standards for clinical trials
-   [Admiral ADSL Vignette](https://pharmaverse.github.io/admiral/articles/adsl.html) - Building ADSL from DS
-   [Pharmaverse.org](https://pharmaverse.org) - R packages for clinical data

------------------------------------------------------------------------

## What's Next?

In **Day 12**, we will focus on **Data Cuts with datacutr**:

-   What clinical data cuts are and why they matter for interim analyses
-   Using the `datacutr` package to apply cutoff dates across domains
-   Handling ongoing AEs, partial dates, and records at the cutoff boundary
-   Patient-level vs. record-level cut logic
