---
title: "Day 10: AE Domain Mastery & SAE Logic"
subtitle: "Deep Dive into Severity, Causality, and Outcomes"


format:
  html:
    theme:
      light: cosmo

    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css
    
editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

::: {style="margin-bottom:12px;"}
<a href="../index.html" class="resource-link">← Back to Roadmap</a>
:::

## Learning Objectives

By the end of Day 10, you will be able to:

1.  Understand the **AE domain** at a production level - beyond the basics covered in Day 7
2.  Distinguish between **severity grading** (MILD/MODERATE/SEVERE) and **toxicity grading** (CTCAE Grade 1-5)
3.  Work with all **SAE-related variables**: AESER, AESDTH, AESLIFE, AESHOSP, AESDISAB, AESCONG, AESMIE
4.  Derive **treatment-emergent adverse events** (TEAEs) - one of the most critical derivations in clinical programming
5.  Calculate **AE duration** and handle **ongoing AEs** correctly
6.  Understand how AE data maps to the **ADaM ADAE** dataset

------------------------------------------------------------------------

## Why This Day Matters

On Day 7 we built a basic AE domain as part of the capstone. Today we go **much deeper** - because in production clinical programming, AE data is arguably the **most scrutinized dataset** by regulatory agencies. Getting AE logic wrong can delay a submission or trigger FDA queries.

::: callout-important
## The Stakes Are High

Adverse event data directly impacts:

-   **Patient safety decisions** - Should the trial continue?
-   **Drug labeling** - What warnings go on the label?
-   **Regulatory approval** - Is the risk-benefit profile acceptable?
-   **Post-marketing surveillance** - What to watch for after approval

Every variable, every flag, every derivation matters.
:::

------------------------------------------------------------------------

## Package Installation & Loading

### Required Packages

| Package           | Purpose                                   |
|-------------------|-------------------------------------------|
| `dplyr`           | Data manipulation (filter, mutate, joins) |
| `lubridate`       | Date/time arithmetic for AE durations     |
| `pharmaversesdtm` | Example SDTM datasets including AE and DM |

### Install & Load

```{r}
#| label: install-and-load

if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("lubridate", quietly = TRUE)) install.packages("lubridate")
if (!requireNamespace("pharmaversesdtm", quietly = TRUE)) install.packages("pharmaversesdtm")

library(dplyr)
library(lubridate)
library(pharmaversesdtm)
```

------------------------------------------------------------------------

## Exploring the AE Domain from pharmaversesdtm

### Load and Inspect

```{r}
#| label: load-ae-data

# Load AE and DM domains
data("ae", package = "pharmaversesdtm")
data("dm", package = "pharmaversesdtm")

cat("AE domain dimensions:", nrow(ae), "rows x", ncol(ae), "columns\n")
cat("Number of unique subjects:", n_distinct(ae$USUBJID), "\n")
cat("Variables available:\n")
cat(paste(names(ae), collapse = ", "), "\n")
```

### Understanding What Each AE Variable Means

Let's look at the actual data - every column tells a story:

```{r}
#| label: glimpse-ae

dplyr::glimpse(ae)
```

::: callout-note
## AE Variables Quick Reference

Here's what the key variables represent in plain English:

| Variable   | What it means                        | Example                    |
|------------|--------------------------------------|----------------------------|
| `AETERM`   | AE as reported by investigator       | "Headache"                 |
| `AEDECOD`  | Standardized preferred term (MedDRA) | "HEADACHE"                 |
| `AEBODSYS` | Body system (MedDRA SOC)             | "NERVOUS SYSTEM DISORDERS" |
| `AESEV`    | Severity: MILD, MODERATE, SEVERE     | "MODERATE"                 |
| `AESER`    | Is it serious? Y/N                   | "N"                        |
| `AEREL`    | Related to study drug?               | "POSSIBLY RELATED"         |
| `AEACN`    | Action taken with study drug         | "DOSE NOT CHANGED"         |
| `AEOUT`    | Outcome                              | "RECOVERED/RESOLVED"       |
| `AESTDTC`  | Start date (ISO 8601)                | "2014-01-03"               |
| `AEENDTC`  | End date (ISO 8601)                  | "2014-01-12"               |
:::

------------------------------------------------------------------------

## Severity vs. Toxicity Grading

This is a concept that confuses many programmers. Let's clarify it with code.

### Severity Grading (AESEV)

**Severity** describes the *intensity* of the adverse event. It's a clinical judgment:

-   **MILD**: Awareness of event but easily tolerated
-   **MODERATE**: Discomfort causing interference with usual activity
-   **SEVERE**: Incapacitating; unable to do usual activities

```{r}
#| label: severity-distribution

# What severity values exist in our data?
ae %>%
  count(AESEV, name = "Count") %>%
  mutate(Percent = round(100 * Count / sum(Count), 1)) %>%
  arrange(match(AESEV, c("MILD", "MODERATE", "SEVERE")))
```

### Toxicity Grading (AETOXGR - CTCAE Scale)

**Toxicity grading** uses the **CTCAE (Common Terminology Criteria for Adverse Events)** scale and is more granular:

| Grade | Description                                                      |
|-------|------------------------------------------------------------------|
| 1     | Mild; asymptomatic or mild symptoms                              |
| 2     | Moderate; minimal, local, or non-invasive intervention indicated |
| 3     | Severe or medically significant; hospitalization indicated       |
| 4     | Life-threatening; urgent intervention indicated                  |
| 5     | Death related to AE                                              |

```{r}
#| label: toxicity-vs-severity

# Check if AETOXGR exists in our data
if ("AETOXGR" %in% names(ae)) {
  ae %>%
    count(AETOXGR, AESEV) %>%
    arrange(AETOXGR)
} else {
  cat("AETOXGR is not present in the pharmaversesdtm AE dataset.\n")
  cat("This is common - not all studies use CTCAE grading.\n\n")
  
  cat("When AETOXGR IS available, it goes in the AE domain as:\n")
  cat("  AETOXGR = Toxicity grade (1-5)\n")
  cat("  AETOXGRS = Toxicity grade from source\n")
}
```

::: callout-important
## Severity ≠ Seriousness

This is the **#1 most common confusion** in clinical programming:

-   **AESEV (Severity)** = How intense is the event? (MILD/MODERATE/SEVERE)
-   **AESER (Seriousness)** = Does it meet SAE criteria? (Y/N)

A **MILD rash** could be an SAE if it requires hospitalization. A **SEVERE headache** may NOT be an SAE if it resolves quickly with OTC medication.

**Severity describes intensity. Seriousness describes regulatory significance.**
:::

------------------------------------------------------------------------

## Serious Adverse Events (SAEs)

### What Makes an AE "Serious"?

An adverse event is classified as **serious** (AESER = "Y") if it meets **any** of the following criteria:

```         
┌─────────────────────────────────────────────────────────────────────┐
│                   SAE CRITERIA VARIABLES                            │
├─────────────────────────────────────────────────────────────────────┤
│  AESER    = "Y" if ANY of the following are "Y":                    │
│                                                                     │
│  AESDTH   = Results in Death                                        │
│  AESLIFE  = Is Life-Threatening                                     │
│  AESHOSP  = Requires or Prolongs Hospitalization                    │
│  AESDISAB = Results in Persistent/Significant Disability            │
│  AESCONG  = Congenital Anomaly/Birth Defect                         │
│  AESMIE   = Other Medically Important Event                         │
└─────────────────────────────────────────────────────────────────────┘
```

### Exploring SAE Data

```{r}
#| label: sae-exploration

# How many SAEs in our data?
ae %>%
  count(AESER, name = "Count") %>%
  mutate(Percent = round(100 * Count / sum(Count), 1))
```

### SAE Criteria Breakdown

```{r}
#| label: sae-criteria

# Check which SAE criteria variables are available
sae_vars <- c("AESDTH", "AESLIFE", "AESHOSP", "AESDISAB", "AESCONG", "AESMIE")
available_sae_vars <- sae_vars[sae_vars %in% names(ae)]

cat("SAE criteria variables available:", paste(available_sae_vars, collapse = ", "), "\n\n")

if (length(available_sae_vars) > 0) {
  # Show SAE details
  ae %>%
    filter(AESER == "Y") %>%
    select(USUBJID, AEDECOD, AESEV, AESER, any_of(sae_vars)) %>%
    head(10)
} else {
  cat("No SAE criteria sub-variables found in this dataset.\n")
  cat("In production, you would create them from raw CRF data.\n")
}
```

### Simulating SAE Logic from Scratch

Since the practice dataset may not have all SAE sub-variables, let's build the logic ourselves. This is exactly what you'd do on a real study:

```{r}
#| label: sae-derivation

# Create simulated AE data with SAE criteria
set.seed(42)

ae_sample <- tibble(
  USUBJID  = rep(paste0("CDISC01-001-00", 1:5), each = 4),
  AESEQ    = rep(1:4, 5),
  AEDECOD  = sample(c("HEADACHE", "NAUSEA", "RASH", "FALL", "PNEUMONIA",
                       "MYOCARDIAL INFARCTION", "SEIZURE", "ANEMIA"), 20, replace = TRUE),
  AESEV    = sample(c("MILD", "MODERATE", "SEVERE"), 20, replace = TRUE, 
                    prob = c(0.5, 0.35, 0.15)),
  # SAE criteria - simulate realistic probabilities
  AESDTH   = sample(c("Y", "N"), 20, replace = TRUE, prob = c(0.02, 0.98)),
  AESLIFE  = sample(c("Y", "N"), 20, replace = TRUE, prob = c(0.05, 0.95)),
  AESHOSP  = sample(c("Y", "N"), 20, replace = TRUE, prob = c(0.10, 0.90)),
  AESDISAB = sample(c("Y", "N"), 20, replace = TRUE, prob = c(0.03, 0.97)),
  AESCONG  = "N",  # Very rare in adult trials
  AESMIE   = sample(c("Y", "N"), 20, replace = TRUE, prob = c(0.05, 0.95))
)

# Derive AESER: "Y" if ANY criterion is "Y"
ae_with_ser <- ae_sample %>%
  mutate(
    AESER = case_when(
      AESDTH   == "Y" ~ "Y",
      AESLIFE  == "Y" ~ "Y",
      AESHOSP  == "Y" ~ "Y",
      AESDISAB == "Y" ~ "Y",
      AESCONG  == "Y" ~ "Y",
      AESMIE   == "Y" ~ "Y",
      TRUE ~ "N"
    )
  )

cat("SAE derivation results:\n")
ae_with_ser %>%
  count(AESER, name = "Count") %>%
  mutate(Percent = round(100 * Count / sum(Count), 1))
```

```{r}
#| label: sae-detail-view

# View SAE records with their criteria
cat("\nSAE records detail:\n")
ae_with_ser %>%
  filter(AESER == "Y") %>%
  select(USUBJID, AEDECOD, AESEV, AESER, AESDTH, AESLIFE, AESHOSP, AESDISAB, AESMIE)
```

::: callout-tip
## Read the Logic Carefully

The `case_when()` above uses a **waterfall** approach - it checks each criterion in sequence and returns `"Y"` at the first match. This works well for deriving `AESER`, but in production you'd typically use a more explicit approach:

``` r
AESER = if_else(
  AESDTH == "Y" | AESLIFE == "Y" | AESHOSP == "Y" | 
  AESDISAB == "Y" | AESCONG == "Y" | AESMIE == "Y",
  "Y", "N"
)
```

Both approaches give the same result. The `if_else()` version makes the OR-logic more explicit.
:::

------------------------------------------------------------------------

## Treatment-Emergent Adverse Events (TEAEs)

### What is a TEAE?

A **treatment-emergent adverse event** is an AE that:

1.  **Started on or after the first dose** of study drug, OR
2.  **Was present before treatment but worsened** after the first dose

This is one of the most important derivations in clinical programming because **most safety analyses focus exclusively on TEAEs**.

### The TEAE Decision Tree

```         
┌─────────────────────────────────────────────────────────────────────┐
│                      TEAE DECISION LOGIC                            │
├─────────────────────────────────────────────────────────────────────┤
│                                                                     │
│  Is AE Start Date (AESTDTC) >= First Dose Date (RFSTDTC)?          │
│    YES → TEAE = "Y"                                                │
│    NO  → Was severity worse than pre-treatment?                     │
│           YES → TEAE = "Y"                                         │
│           NO  → TEAE = "N" (Pre-treatment AE, not worsened)        │
│                                                                     │
│  Is AE Start Date missing?                                          │
│    → Compare AE End Date with First Dose Date                       │
│    → If AEENDTC >= RFSTDTC, treat as potentially TEAE              │
│    → Flag for medical review                                        │
└─────────────────────────────────────────────────────────────────────┘
```

### Deriving TEAEs in Code

```{r}
#| label: teae-derivation

# Load DM for reference dates
data("dm", package = "pharmaversesdtm")
data("ae", package = "pharmaversesdtm")

# Get reference start date per subject
ref_dates <- dm %>%
  select(USUBJID, RFSTDTC) %>%
  filter(!is.na(RFSTDTC))

cat("Reference dates (first dose) for sample subjects:\n")
head(ref_dates)
```

```{r}
#| label: teae-join-and-derive

# Join AE with reference dates and derive TEAE flag
ae_with_teae <- ae %>%
  left_join(ref_dates, by = "USUBJID") %>%
  mutate(
    # Parse dates - handle potential partial dates
    ae_start = ymd(AESTDTC),
    ref_start = ymd(RFSTDTC),
    
    # Core TEAE derivation
    TRTEMFL = case_when(
      # Case 1: AE starts on or after first dose
      !is.na(ae_start) & !is.na(ref_start) & ae_start >= ref_start ~ "Y",
      
      # Case 2: AE start date is missing - conservative approach
      is.na(ae_start) & !is.na(AEENDTC) & !is.na(ref_start) &
        ymd(AEENDTC) >= ref_start ~ "Y",
      
      # Case 3: Both dates available, AE started before treatment
      !is.na(ae_start) & !is.na(ref_start) & ae_start < ref_start ~ "N",
      
      # Case 4: Cannot determine
      TRUE ~ NA_character_
    )
  )

# Summary
cat("TEAE derivation results:\n")
ae_with_teae %>%
  count(TRTEMFL, name = "Count") %>%
  mutate(Percent = round(100 * Count / sum(Count), 1))
```

```{r}
#| label: teae-examples

# Show examples of TEAE vs non-TEAE
cat("\nSample of TEAE records (started on/after first dose):\n")
ae_with_teae %>%
  filter(TRTEMFL == "Y") %>%
  select(USUBJID, AEDECOD, AESTDTC, RFSTDTC, TRTEMFL) %>%
  head(5)

cat("\nSample of non-TEAE records (started before first dose):\n")
ae_with_teae %>%
  filter(TRTEMFL == "N") %>%
  select(USUBJID, AEDECOD, AESTDTC, RFSTDTC, TRTEMFL) %>%
  head(5)
```

::: callout-warning
## Partial Dates Are Common

In real clinical data, you will frequently encounter **partial dates** like:

-   `"2014-01"` (month known, day unknown)
-   `"2014"` (only year known)
-   `""` or `NA` (completely missing)

The `lubridate::ymd()` function will return `NA` for partial dates. In production, you'd implement **date imputation rules** as specified in the Statistics Analysis Plan (SAP). Common approaches:

-   **Conservative**: Impute to the latest possible date (e.g., first of the month)
-   **Non-conservative**: Impute to the earliest possible date
-   **Rule-based**: Use other available information to make the best guess
:::

------------------------------------------------------------------------

## AE Duration Calculations

### Computing Duration in Days

```{r}
#| label: ae-duration

# Calculate AE duration
ae_duration <- ae %>%
  mutate(
    ae_start = ymd(AESTDTC),
    ae_end   = ymd(AEENDTC),
    
    # Duration = end - start + 1 (inclusive of start and end day)
    AEDUR = as.numeric(ae_end - ae_start) + 1,
    
    # Flag ongoing AEs (no end date)
    AEONGO = if_else(is.na(ae_end), "Y", "N")
  )

# Summary of durations
cat("AE Duration Summary (days):\n")
ae_duration %>%
  filter(!is.na(AEDUR)) %>%
  summarise(
    N = n(),
    Mean = round(mean(AEDUR), 1),
    Median = median(AEDUR),
    Min = min(AEDUR),
    Max = max(AEDUR),
    SD = round(sd(AEDUR), 1)
  )
```

```{r}
#| label: ongoing-ae-count

# Ongoing AEs
cat("\nOngoing AEs (no end date):\n")
ae_duration %>%
  count(AEONGO, name = "Count") %>%
  mutate(Percent = round(100 * Count / sum(Count), 1))
```

::: callout-note
## The "+1" Rule

Notice we calculate duration as `(end - start) + 1`. This is the standard clinical convention:

-   An AE that starts on Day 5 and ends on Day 5 lasted **1 day** (not 0)
-   An AE that starts on Day 5 and ends on Day 7 lasted **3 days** (not 2)

This is the same "inclusive" counting rule used for study day calculations.
:::

------------------------------------------------------------------------

## Causality Assessment

### How Causality Is Determined

Causality (AEREL) indicates whether the AE is related to the study drug. Common categories:

| AEREL Value   | Description                               |
|---------------|-------------------------------------------|
| `NOT RELATED` | No reasonable possibility of relationship |
| `UNLIKELY`    | Doubtful relationship                     |
| `POSSIBLE`    | Cannot rule out relationship              |
| `PROBABLE`    | Likely related                            |
| `DEFINITE`    | Clearly related                           |

```{r}
#| label: causality-analysis

# Causality distribution
cat("Causality (AEREL) distribution:\n")
ae %>%
  count(AEREL, name = "Count") %>%
  mutate(Percent = round(100 * Count / sum(Count), 1)) %>%
  arrange(desc(Count))
```

### Deriving a Binary Relatedness Flag

For many analyses, we simplify causality into a binary flag:

```{r}
#| label: relatedness-flag

# In ADaM, we often create a binary related flag
ae_related <- ae %>%
  mutate(
    # RELFL = "Y" if possibly, probably, or definitely related
    RELFL = case_when(
      AEREL %in% c("POSSIBLE", "PROBABLE", "DEFINITE") ~ "Y",
      AEREL %in% c("NOT RELATED", "UNLIKELY") ~ "N",
      # Handle variations in terminology
      grepl("RELAT", AEREL, ignore.case = TRUE) ~ "Y",
      TRUE ~ "N"
    )
  )

cat("Binary relatedness flag:\n")
ae_related %>%
  count(AEREL, RELFL) %>%
  arrange(RELFL, desc(n))
```

------------------------------------------------------------------------

## Action Taken and Outcome

### AEACN - Action Taken with Study Drug

The action taken in response to the AE is captured in `AEACN`:

```{r}
#| label: action-taken

ae %>%
  count(AEACN, name = "Count") %>%
  mutate(Percent = round(100 * Count / sum(Count), 1)) %>%
  arrange(desc(Count))
```

### AEOUT - Outcome of the AE

```{r}
#| label: ae-outcome

ae %>%
  count(AEOUT, name = "Count") %>%
  mutate(Percent = round(100 * Count / sum(Count), 1)) %>%
  arrange(desc(Count))
```

::: callout-tip
## Controlled Terminology Matters

All of these variables - `AEACN`, `AEOUT`, `AESEV`, `AESER` - use **CDISC Controlled Terminology**. You can't invent your own values. The allowed values are specified in the [CDISC CT Package](https://www.cancer.gov/research/resources/terminology/cdisc).

For example, valid `AEOUT` values are: - RECOVERED/RESOLVED - RECOVERING/RESOLVING - NOT RECOVERED/NOT RESOLVED - RECOVERED/RESOLVED WITH SEQUELAE - FATAL - UNKNOWN
:::

------------------------------------------------------------------------

## Complete Example: Production AE Processing

Let's put everything together into a production-quality AE processing pipeline:

```{r}
#| label: production-ae-pipeline

# ---- Production AE Processing Pipeline ----

# Step 1: Start with raw AE data
data("ae", package = "pharmaversesdtm")
data("dm", package = "pharmaversesdtm")

# Step 2: Get reference dates
ref <- dm %>%
  select(STUDYID, USUBJID, RFSTDTC, RFENDTC) %>%
  mutate(
    ref_start = ymd(RFSTDTC),
    ref_end   = ymd(RFENDTC)
  )

# Step 3: Enhance AE data with all derivations
ae_production <- ae %>%
  # Join reference dates
  left_join(ref %>% select(USUBJID, ref_start, ref_end, RFSTDTC), by = "USUBJID") %>%
  mutate(
    # Parse dates
    ae_start = ymd(AESTDTC),
    ae_end   = ymd(AEENDTC),
    
    # ---- TEAE Flag ----
    TRTEMFL = case_when(
      !is.na(ae_start) & !is.na(ref_start) & ae_start >= ref_start ~ "Y",
      is.na(ae_start) & !is.na(ae_end) & !is.na(ref_start) & ae_end >= ref_start ~ "Y",
      !is.na(ae_start) & !is.na(ref_start) & ae_start < ref_start ~ "N",
      TRUE ~ NA_character_
    ),
    
    # ---- Study Days ----
    AESTDY = case_when(
      !is.na(ae_start) & !is.na(ref_start) & ae_start >= ref_start ~ 
        as.numeric(ae_start - ref_start) + 1,
      !is.na(ae_start) & !is.na(ref_start) & ae_start < ref_start ~ 
        as.numeric(ae_start - ref_start),
      TRUE ~ NA_real_
    ),
    AEENDY = case_when(
      !is.na(ae_end) & !is.na(ref_start) & ae_end >= ref_start ~ 
        as.numeric(ae_end - ref_start) + 1,
      !is.na(ae_end) & !is.na(ref_start) & ae_end < ref_start ~ 
        as.numeric(ae_end - ref_start),
      TRUE ~ NA_real_
    ),
    
    # ---- Duration ----
    AEDUR = if_else(!is.na(ae_start) & !is.na(ae_end),
                    as.numeric(ae_end - ae_start) + 1,
                    NA_real_),
    
    # ---- Ongoing Flag ----
    AEONGO = if_else(is.na(ae_end), "Y", "N"),
    
    # ---- Binary Relatedness ----
    RELFL = case_when(
      AEREL %in% c("POSSIBLE", "PROBABLE", "DEFINITE") ~ "Y",
      grepl("RELAT", AEREL, ignore.case = TRUE) ~ "Y",
      TRUE ~ "N"
    )
  ) %>%
  # Clean up helper columns
  select(-ae_start, -ae_end, -ref_start, -ref_end)

cat("Production AE dataset:\n")
cat("Total AE records:", nrow(ae_production), "\n")
cat("TEAEs:", sum(ae_production$TRTEMFL == "Y", na.rm = TRUE), "\n")
cat("SAEs:", sum(ae_production$AESER == "Y", na.rm = TRUE), "\n")
cat("Drug-related:", sum(ae_production$RELFL == "Y", na.rm = TRUE), "\n\n")
```

```{r}
#| label: production-ae-preview

# Preview the enhanced dataset
ae_production %>%
  select(USUBJID, AEDECOD, AESEV, AESER, TRTEMFL, AESTDY, AEDUR, AEONGO, RELFL) %>%
  head(15)
```

------------------------------------------------------------------------

## Key AE Counts for Safety Reporting

In clinical study reports, AE tables almost always include these counts:

```{r}
#| label: ae-safety-summary

# ---- Summary Table: AE Incidence ----
cat("=== AE Incidence Summary ===\n\n")

# Total subjects
n_total <- n_distinct(dm$USUBJID)
n_ae <- n_distinct(ae_production$USUBJID)

cat("Total subjects enrolled:", n_total, "\n")
cat("Subjects with any AE:", n_ae, 
    sprintf("(%.1f%%)", 100 * n_ae / n_total), "\n\n")

# TEAE summary
teae_data <- ae_production %>% filter(TRTEMFL == "Y")
n_teae <- n_distinct(teae_data$USUBJID)
cat("Subjects with any TEAE:", n_teae,
    sprintf("(%.1f%%)", 100 * n_teae / n_total), "\n")

# SAE summary
sae_data <- ae_production %>% filter(AESER == "Y")
n_sae <- n_distinct(sae_data$USUBJID)
cat("Subjects with any SAE:", n_sae,
    sprintf("(%.1f%%)", 100 * n_sae / n_total), "\n")

# Drug-related TEAE
rel_teae <- ae_production %>% filter(TRTEMFL == "Y", RELFL == "Y")
n_rel <- n_distinct(rel_teae$USUBJID)
cat("Subjects with drug-related TEAE:", n_rel,
    sprintf("(%.1f%%)", 100 * n_rel / n_total), "\n")
```

```{r}
#| label: ae-by-severity

# TEAE by maximum severity
cat("\nTEAE by Maximum Severity per Subject:\n")
teae_data %>%
  mutate(SEV_NUM = case_when(
    AESEV == "MILD" ~ 1,
    AESEV == "MODERATE" ~ 2,
    AESEV == "SEVERE" ~ 3
  )) %>%
  group_by(USUBJID) %>%
  summarise(MAX_SEV = max(SEV_NUM, na.rm = TRUE), .groups = "drop") %>%
  mutate(MAX_AESEV = case_when(
    MAX_SEV == 1 ~ "MILD",
    MAX_SEV == 2 ~ "MODERATE",
    MAX_SEV == 3 ~ "SEVERE"
  )) %>%
  count(MAX_AESEV, name = "N_Subjects") %>%
  mutate(Percent = round(100 * N_Subjects / n_total, 1)) %>%
  arrange(match(MAX_AESEV, c("MILD", "MODERATE", "SEVERE")))
```

------------------------------------------------------------------------

## Preview: From AE to ADAE

The SDTM AE domain feeds into the ADaM **ADAE** dataset. Here's a preview of the key mappings:

| SDTM AE          | ADaM ADAE | Description                            |
|------------------|-----------|----------------------------------------|
| AEDECOD          | AEDECOD   | Preferred term (carried forward)       |
| AEBODSYS         | AEBODSYS  | Body system (carried forward)          |
| AESEV            | AESEV     | Severity (carried forward)             |
| AESTDTC → parsed | ASTDT     | Analysis start date (numeric)          |
| AEENDTC → parsed | AENDT     | Analysis end date (numeric)            |
| *derived*        | TRTEMFL   | Treatment-emergent flag                |
| *derived*        | AESEQ_GR  | Worst event selection per subject/term |
| from DM          | TRT01A    | Actual treatment (from ADSL)           |

::: callout-note
## Looking Ahead

In Week 3, when we build ADaM datasets with `admiral`, the `derive_var_trtemfl()` function will handle TEAE derivation automatically - but understanding the logic behind it (as we've done today) is essential for debugging and validation.
:::

------------------------------------------------------------------------

## Deliverable Summary

Today you completed the following:

| Task                                                      | Status |
|-----------------------------------------------------------|--------|
| Understood severity vs. toxicity grading                  | ✓ Done |
| Explored all SAE criteria variables (AESER, AESDTH, etc.) | ✓ Done |
| Derived SAE flag from sub-criteria using OR-logic         | ✓ Done |
| Derived treatment-emergent AE (TEAE) flag                 | ✓ Done |
| Calculated AE duration and identified ongoing AEs         | ✓ Done |
| Analyzed causality, action taken, and outcome             | ✓ Done |
| Built a production AE processing pipeline                 | ✓ Done |
| Generated safety summary counts                           | ✓ Done |

------------------------------------------------------------------------

## Key Takeaways

1.  **Severity ≠ Seriousness** - A mild AE can be serious; a severe AE may not be serious
2.  **SAE criteria are additive** - AESER = "Y" if ANY sub-criterion is "Y"
3.  **TEAEs are the focus** - Most safety analyses exclude pre-treatment AEs
4.  **Duration uses the +1 rule** - Inclusive of both start and end day
5.  **Causality is simplified** - Binary related/not-related flags are common in ADaM
6.  **Controlled Terminology is mandatory** - Use only CDISC-approved values

------------------------------------------------------------------------

## Resources

-   [CDISC SDTM Implementation Guide - AE Domain](https://www.cdisc.org/standards/foundational/sdtm) - Official AE specification
-   [MedDRA Terminology](https://www.meddra.org/) - Medical Dictionary for Regulatory Activities
-   [CTCAE v5.0](https://ctep.cancer.gov/protocoldevelopment/electronic_applications/ctc.htm) - Common Terminology Criteria for Adverse Events
-   [ICH E2A Guidelines](https://www.ich.org/page/safety-guidelines) - Clinical Safety Data Management
-   [Admiral ADAE Vignette](https://pharmaverse.github.io/admiral/articles/adae.html) - Building ADAE with admiral

------------------------------------------------------------------------

## What's Next?

In **Day 11**, we will focus on **Disposition (DS) & Trial Design Domains**:

-   Understanding the DS domain for screen failures, completers, early terminators
-   Trial Design domains: TA, TE, TV, TI, TS
-   Working with EPOCH and milestone variables
-   Subject flow and disposition summaries

<a href="https://visitorbadge.io/status?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse"><img src="https://api.visitorbadge.io/api/visitors?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse&label=Visits%20till%20date&countColor=%23263759&style=plastic&labelStyle=none" /></a>
