---
title: "Day 14: Week 2 Capstone - Metadata-Driven SDTM with metacore & xportr"
subtitle: "Specification-Driven Workflows for Submission-Ready Data"
author: "Indraneel Chakraborty"
date: today
format:
  html:
    theme:
      light: cosmo

    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css
    
editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

<div style="margin-bottom:12px;"><a href="../index.html" class="resource-link">← Back to Roadmap</a></div>

## Learning Objectives

By the end of Day 14 (Week 2 Capstone), you will be able to:

1.  Load **specification objects** using `metacore` - the Pharmaverse metadata standard
2.  Use `metatools` to apply metadata checks and select columns from specs
3.  Use `xportr` to apply **labels, types, formats, lengths**, and variable ordering from specs
4.  Build an **end-to-end pipeline**: raw data → SDTM → validate → export `.xpt`
5.  Appreciate why **metadata-driven workflows** are the future of clinical programming
6.  Understand everything needed before starting **ADaM datasets in Week 3**

------------------------------------------------------------------------

## Why Metadata-Driven Workflows?

### The Problem with Manual Programming

In traditional clinical programming, programmers manually:

-   Assign variable labels (`label(dm$USUBJID) <- "Unique Subject Identifier"`)
-   Set variable types (character vs numeric)
-   Order variables in the correct sequence
-   Set variable lengths for transport files

This is **error-prone**, **tedious**, and **hard to maintain**.

::: callout-important
## The Metadata-Driven Solution

Instead of hardcoding metadata in your programs, you:

1.  **Define metadata once** in a specification file (Excel, or a `metacore` object)
2.  **Load metadata** into R using `metacore`
3.  **Apply metadata** to your datasets using `xportr`/`metatools`
4.  **Validate** that your datasets match the specification

This ensures:

-   **Consistency** - All datasets follow the same rules
-   **Reproducibility** - Metadata changes automatically propagate
-   **Compliance** - Variable labels, types, and lengths match `define.xml`
:::

------------------------------------------------------------------------

## Package Installation & Loading

### Required Packages

| Package | Purpose |
|---------|---------|
| `metacore` | Load and manage dataset specifications |
| `metatools` | Apply metadata-based transformations and checks |
| `xportr` | Apply labels, types, formats, lengths; export `.xpt` |
| `dplyr` | Data manipulation |
| `pharmaversesdtm` | Example SDTM datasets |

```{r}
#| label: install-and-load

if (!requireNamespace("dplyr", quietly = TRUE)) suppressMessages(install.packages("dplyr"))
if (!requireNamespace("lubridate", quietly = TRUE)) suppressMessages(install.packages("lubridate"))
if (!requireNamespace("pharmaversesdtm", quietly = TRUE)) suppressMessages(install.packages("pharmaversesdtm"))
if (!requireNamespace("metacore", quietly = TRUE)) suppressMessages(install.packages("metacore"))
if (!requireNamespace("metatools", quietly = TRUE)) suppressMessages(install.packages("metatools"))
if (!requireNamespace("xportr", quietly = TRUE)) suppressMessages(install.packages("xportr"))

library(dplyr)
library(lubridate)
library(pharmaversesdtm)
library(metacore)
library(metatools)
library(xportr)
```

------------------------------------------------------------------------

## Understanding metacore

### What is metacore?

**`metacore`** is a Pharmaverse package that provides a standardized way to represent **dataset specifications** in R. Think of it as the bridge between your Excel specification file and your R programs.

### The metacore Object Structure

A `metacore` object contains multiple related tables:

```
┌─────────────────────────────────────────────────────────────────────────┐
│                      METACORE OBJECT                                    │
├─────────────────────────────────────────────────────────────────────────┤
│                                                                         │
│  ds_spec        = Dataset-level metadata                                │
│                   (domain name, label, structure)                       │
│                                                                         │
│  ds_vars        = Variable-level metadata                               │
│                   (which variables belong to which dataset)              │
│                                                                         │
│  var_spec       = Variable specifications                               │
│                   (variable name, label, type, length, format)          │
│                                                                         │
│  value_spec     = Value-level metadata                                  │
│                   (codelist values, decode values)                      │
│                                                                         │
│  derivations    = Derivation metadata                                   │
│                   (how variables are derived)                           │
│                                                                         │
│  codelist       = Code list definitions                                 │
│                   (controlled terminology)                              │
│                                                                         │
│  supp           = Supplemental qualifiers                               │
│                   (SUPP-- domain information)                           │
│                                                                         │
└─────────────────────────────────────────────────────────────────────────┘
```

### Creating a metacore Object

In production, you'd load this from an Excel specification. For learning, let's build one:

```{r}
#| label: create-metacore

# ---- Define Dataset Specs ----
ds_spec <- tibble::tribble(
  ~dataset, ~structure,     ~label,
  "DM",     "One record per subject",  "Demographics",
  "AE",     "One record per adverse event per subject", "Adverse Events",
  "EX",     "One record per exposure per subject",      "Exposure"
)

# ---- Define Variables ----
ds_vars <- tibble::tribble(
  ~dataset, ~variable,   ~keep,  ~key_seq, ~order, ~core,
  "DM",     "STUDYID",   TRUE,   1L,       1L,     "Required",
  "DM",     "DOMAIN",    TRUE,   NA,       2L,     "Required",
  "DM",     "USUBJID",   TRUE,   2L,       3L,     "Required",
  "DM",     "SUBJID",    TRUE,   NA,       4L,     "Required",
  "DM",     "SITEID",    TRUE,   NA,       5L,     "Required",
  "DM",     "AGE",       TRUE,   NA,       6L,     "Required",
  "DM",     "AGEU",      TRUE,   NA,       7L,     "Required",
  "DM",     "SEX",       TRUE,   NA,       8L,     "Required",
  "DM",     "RACE",      TRUE,   NA,       9L,     "Required",
  "DM",     "ETHNIC",    TRUE,   NA,       10L,    "Permissible",
  "DM",     "ARM",       TRUE,   NA,       11L,    "Required",
  "DM",     "ARMCD",     TRUE,   NA,       12L,    "Required",
  "DM",     "ACTARM",    TRUE,   NA,       13L,    "Permissible",
  "DM",     "ACTARMCD",  TRUE,   NA,       14L,    "Permissible",
  "DM",     "RFSTDTC",   TRUE,   NA,       15L,    "Expected",
  "DM",     "RFENDTC",   TRUE,   NA,       16L,    "Expected",
  "AE",     "STUDYID",   TRUE,   1L,       1L,     "Required",
  "AE",     "DOMAIN",    TRUE,   NA,       2L,     "Required",
  "AE",     "USUBJID",   TRUE,   2L,       3L,     "Required",
  "AE",     "AESEQ",     TRUE,   3L,       4L,     "Required",
  "AE",     "AETERM",    TRUE,   NA,       5L,     "Required",
  "AE",     "AEDECOD",   TRUE,   NA,       6L,     "Required",
  "AE",     "AEBODSYS",  TRUE,   NA,       7L,     "Permissible",
  "AE",     "AESEV",     TRUE,   NA,       8L,     "Expected",
  "AE",     "AESER",     TRUE,   NA,       9L,     "Expected",
  "AE",     "AEREL",     TRUE,   NA,       10L,    "Permissible",
  "AE",     "AESTDTC",   TRUE,   NA,       11L,    "Expected",
  "AE",     "AEENDTC",   TRUE,   NA,       12L,    "Permissible"
)

# ---- Define Variable Specifications ----
var_spec <- tibble::tribble(
  ~variable,   ~type,       ~length, ~label,                          ~common_ct,    ~format,
  "STUDYID",   "text",      12L,     "Study Identifier",             NA,            NA,
  "DOMAIN",    "text",      2L,      "Domain Abbreviation",          NA,            NA,
  "USUBJID",   "text",      40L,     "Unique Subject Identifier",    NA,            NA,
  "SUBJID",    "text",      8L,      "Subject Identifier for the Study", NA,        NA,
  "SITEID",    "text",      8L,      "Study Site Identifier",        NA,            NA,
  "AGE",       "integer",   8L,      "Age",                          NA,            NA,
  "AGEU",      "text",      6L,      "Age Units",                    "AGEU",        NA,
  "SEX",       "text",      2L,      "Sex",                          "SEX",         NA,
  "RACE",      "text",      40L,     "Race",                         "RACE",        NA,
  "ETHNIC",    "text",      40L,     "Ethnicity",                    "ETHNIC",      NA,
  "ARM",       "text",      40L,     "Planned Arm",                  NA,            NA,
  "ARMCD",     "text",      20L,     "Planned Arm Code",             NA,            NA,
  "ACTARM",    "text",      40L,     "Actual Arm",                   NA,            NA,
  "ACTARMCD",  "text",      20L,     "Actual Arm Code",              NA,            NA,
  "RFSTDTC",   "text",      20L,     "Subject Reference Start Date/Time", NA,       NA,
  "RFENDTC",   "text",      20L,     "Subject Reference End Date/Time",   NA,       NA,
  "AESEQ",     "integer",   8L,      "Sequence Number",              NA,            NA,
  "AETERM",    "text",      200L,    "Reported Term for the Adverse Event", NA,     NA,
  "AEDECOD",   "text",      200L,    "Dictionary-Derived Term",      NA,            NA,
  "AEBODSYS",  "text",      200L,    "Body System or Organ Class",   NA,            NA,
  "AESEV",     "text",      12L,     "Severity/Intensity",           "AESEV",       NA,
  "AESER",     "text",      2L,      "Serious Event",                "NY",          NA,
  "AEREL",     "text",      40L,     "Causality",                    NA,            NA,
  "AESTDTC",   "text",      20L,     "Start Date/Time of Adverse Event", NA,        NA,
  "AEENDTC",   "text",      20L,     "End Date/Time of Adverse Event",    NA,        NA
)

cat("Metadata components created:\n")
cat("  Dataset specs:", nrow(ds_spec), "datasets\n")
cat("  Variable assignments:", nrow(ds_vars), "variable-dataset pairs\n")
cat("  Variable specs:", nrow(var_spec), "unique variables\n")
```

```{r}
#| label: build-metacore-object

# Build the metacore object
# Note: metacore() requires specific table structures
# For simplicity, we'll work with the individual components directly

cat("In production, you would create the metacore object like this:\n\n")
cat('mc <- metacore::metacore(\n')
cat('  ds_spec  = ds_spec,\n')
cat('  ds_vars  = ds_vars,\n')
cat('  var_spec = var_spec\n')
cat(')\n\n')

cat("Or more commonly, load from a specification file:\n\n")
cat('mc <- metacore::spec_to_metacore("path/to/specs.xlsx")\n')
```

::: callout-note
## spec_to_metacore()

The most common way to create a `metacore` object in production is using `spec_to_metacore()`, which reads from a formatted Excel specification file. This spec file is typically created by the study statistician or data standards team and contains all the metadata for every dataset and variable.
:::

------------------------------------------------------------------------

## Using xportr: The Metadata Application Engine

### What xportr Does

`xportr` is the **workhorse** for making your datasets submission-ready. It applies metadata from your specification to your actual data:

```{r}
#| label: xportr-overview

cat("xportr: Key Functions\n\n")

xportr_functions <- tibble::tribble(
  ~Function,            ~What_It_Does,
  "xportr_type()",      "Coerce variables to the correct type (character/numeric)",
  "xportr_length()",    "Set variable lengths for SAS transport",
  "xportr_label()",     "Apply variable labels from specification",
  "xportr_order()",     "Reorder variables to match specification",
  "xportr_format()",    "Apply SAS display formats",
  "xportr_write()",     "Export the dataset as a .xpt (SAS transport) file"
)

print(xportr_functions)
```

### Applying xportr Step by Step

Let's work through the full pipeline using the DM domain:

```{r}
#| label: prepare-dm-for-xportr

# Step 1: Create our DM dataset (from Day 7 capstone approach)
set.seed(42)
n_subjects <- 15

raw_dm <- tibble(
  STUDYID  = "CDISC01",
  DOMAIN   = "DM",
  SUBJID   = sprintf("%03d", 1:n_subjects),
  SITEID   = sample(c("101", "102", "103"), n_subjects, replace = TRUE),
  AGE      = sample(25:75, n_subjects, replace = TRUE),
  AGEU     = "YEARS",
  SEX      = sample(c("M", "F"), n_subjects, replace = TRUE),
  RACE     = sample(c("WHITE", "BLACK OR AFRICAN AMERICAN", "ASIAN"), 
                    n_subjects, replace = TRUE),
  ETHNIC   = sample(c("HISPANIC OR LATINO", "NOT HISPANIC OR LATINO"),
                    n_subjects, replace = TRUE, prob = c(0.15, 0.85)),
  ARM      = sample(c("Placebo", "Active 10mg", "Active 20mg"),
                    n_subjects, replace = TRUE),
  ARMCD    = NA_character_,
  ACTARM   = NA_character_,
  ACTARMCD = NA_character_,
  RFSTDTC  = as.character(as.Date("2024-01-01") + sample(0:60, n_subjects, replace = TRUE)),
  RFENDTC  = NA_character_
) %>%
  mutate(
    USUBJID  = paste(STUDYID, SITEID, SUBJID, sep = "-"),
    ARMCD = case_when(
      ARM == "Placebo" ~ "PBO",
      ARM == "Active 10mg" ~ "ACT10",
      ARM == "Active 20mg" ~ "ACT20"
    ),
    ACTARM = ARM,
    ACTARMCD = ARMCD
  )

cat("DM dataset before xportr:\n")
cat("Rows:", nrow(raw_dm), "\n")
cat("Cols:", ncol(raw_dm), "\n\n")
glimpse(raw_dm)
```

### Step 2: Create a Specification for xportr

```{r}
#| label: create-xportr-spec

# xportr functions accept a metadata dataframe with specific columns
# Let's create the specification that xportr expects

dm_spec <- tibble::tribble(
  ~dataset, ~variable,   ~type,       ~label,                               ~length, ~order,  ~format,
  "DM",     "STUDYID",   "character", "Study Identifier",                   12L,     1L,      NA,
  "DM",     "DOMAIN",    "character", "Domain Abbreviation",                2L,      2L,      NA,
  "DM",     "USUBJID",   "character", "Unique Subject Identifier",          40L,     3L,      NA,
  "DM",     "SUBJID",    "character", "Subject Identifier for the Study",   8L,      4L,      NA,
  "DM",     "SITEID",    "character", "Study Site Identifier",              8L,      5L,      NA,
  "DM",     "AGE",       "numeric",   "Age",                                8L,      6L,      NA,
  "DM",     "AGEU",      "character", "Age Units",                          6L,      7L,      NA,
  "DM",     "SEX",       "character", "Sex",                                2L,      8L,      NA,
  "DM",     "RACE",      "character", "Race",                               40L,     9L,      NA,
  "DM",     "ETHNIC",    "character", "Ethnicity",                          40L,     10L,     NA,
  "DM",     "ARM",       "character", "Planned Arm",                        40L,     11L,     NA,
  "DM",     "ARMCD",     "character", "Planned Arm Code",                   20L,     12L,     NA,
  "DM",     "ACTARM",    "character", "Actual Arm",                         40L,     13L,     NA,
  "DM",     "ACTARMCD",  "character", "Actual Arm Code",                    20L,     14L,     NA,
  "DM",     "RFSTDTC",   "character", "Subject Reference Start Date/Time",  20L,     15L,     NA,
  "DM",     "RFENDTC",   "character", "Subject Reference End Date/Time",    20L,     16L,     NA
)

cat("DM specification:\n")
dm_spec %>% select(variable, type, label, length, order) %>% print()
```

### Step 3: Apply xportr Functions

```{r}
#| label: apply-xportr

# ---- Apply metadata using the xportr pipeline ----

# Step 3a: Apply variable types
dm_typed <- raw_dm %>%
  xportr_type(dm_spec, domain = "DM")

cat("After xportr_type():\n")
cat("  AGE type:", class(dm_typed$AGE), "\n")
cat("  STUDYID type:", class(dm_typed$STUDYID), "\n\n")

# Step 3b: Apply variable labels
dm_labeled <- dm_typed %>%
  xportr_label(dm_spec, domain = "DM")

cat("After xportr_label():\n")
# Check labels
labels_applied <- sapply(dm_labeled, function(x) attr(x, "label"))
labels_applied <- labels_applied[!is.na(labels_applied)]
head(data.frame(Variable = names(labels_applied), Label = unname(labels_applied)), 8)
```

```{r}
#| label: apply-xportr-order

# Step 3c: Apply variable ordering
dm_ordered <- dm_labeled %>%
  xportr_order(dm_spec, domain = "DM")

cat("After xportr_order():\n")
cat("Variable order:", paste(names(dm_ordered), collapse = ", "), "\n")
```

```{r}
#| label: apply-xportr-length

# Step 3d: Apply variable lengths
dm_lengthed <- dm_ordered %>%
  xportr_length(dm_spec, domain = "DM")

cat("After xportr_length():\n")
cat("All metadata applied ✓\n")
```

### Step 4: Export as .xpt

```{r}
#| label: export-xpt

# Create output directory
output_dir <- "output"
if (!dir.exists(output_dir)) dir.create(output_dir)

# Export using xportr_write
dm_lengthed %>%
  xportr_write(file.path(output_dir, "dm.xpt"))

cat("Exported:", file.path(output_dir, "dm.xpt"), "\n")
cat("File size:", file.size(file.path(output_dir, "dm.xpt")), "bytes\n")
```

::: callout-tip
## The xportr Pipeline

In practice, you'd chain all xportr functions together:

```r
dm_final <- raw_dm %>%
  xportr_type(spec, domain = "DM") %>%
  xportr_length(spec, domain = "DM") %>%
  xportr_label(spec, domain = "DM") %>%
  xportr_order(spec, domain = "DM") %>%
  xportr_format(spec, domain = "DM") %>%
  xportr_write("output/dm.xpt")
```

This single pipeline takes your raw dataset and makes it submission-ready!
:::

------------------------------------------------------------------------

## Using metatools for Metadata-Based Checks

### What metatools Provides

`metatools` helps you **work with** metadata - selecting variables, checking CT compliance, and building datasets from specs:

```{r}
#| label: metatools-overview

cat("metatools: Key Functions\n\n")

metatools_fns <- tibble::tribble(
  ~Function,                 ~Purpose,
  "build_from_derived()",    "Create a dataset shell from specification",
  "check_ct_col()",          "Check if column values match controlled terminology",
  "check_variables()",       "Verify dataset variables match specification",
  "combine_supp()",          "Combine SUPP-- with parent domain",
  "drop_unspec_vars()",      "Remove variables not in the specification"
)

print(metatools_fns)
```

------------------------------------------------------------------------

## End-to-End Capstone Pipeline

Let's build a complete pipeline that takes raw data through to validated, exported SDTM:

```{r}
#| label: capstone-pipeline

cat("=== WEEK 2 CAPSTONE: END-TO-END SDTM PIPELINE ===\n\n")

# ---- STEP 1: Raw Data ----
cat("Step 1: Generate raw clinical data\n")

set.seed(123)
n <- 20

# Raw demographics
raw_demo <- tibble(
  site    = sample(c("101", "102"), n, replace = TRUE),
  subject = sprintf("%03d", 1:n),
  sex     = sample(c("Male", "Female"), n, replace = TRUE),
  race    = sample(c("White", "Black", "Asian"), n, replace = TRUE, prob = c(0.6, 0.25, 0.15)),
  age     = sample(25:75, n, replace = TRUE),
  arm     = sample(c("Placebo", "Drug A 10mg"), n, replace = TRUE),
  first_dose = as.Date("2024-01-15") + sample(0:45, n, replace = TRUE)
)

# Raw AEs
raw_aes <- tibble(
  site    = sample(raw_demo$site[1:15], 30, replace = TRUE),
  subject = sample(raw_demo$subject[1:15], 30, replace = TRUE),
  ae_term = sample(c("Headache", "Nausea", "Dizziness", "Fatigue", "Rash"), 30, replace = TRUE),
  severity = sample(c("Mild", "Moderate", "Severe"), 30, replace = TRUE, prob = c(0.6, 0.3, 0.1)),
  serious  = sample(c("No", "Yes"), 30, replace = TRUE, prob = c(0.9, 0.1)),
  related  = sample(c("Not Related", "Possibly Related"), 30, replace = TRUE),
  start_day = sample(1:60, 30, replace = TRUE)
) %>%
  left_join(raw_demo %>% select(subject, first_dose), by = "subject") %>%
  mutate(
    start_date = as.character(first_dose + start_day),
    end_date   = as.character(first_dose + start_day + sample(1:14, n(), replace = TRUE))
  )

cat("  Demographics:", nrow(raw_demo), "subjects\n")
cat("  Adverse Events:", nrow(raw_aes), "records\n\n")
```

```{r}
#| label: capstone-step2

# ---- STEP 2: Transform to SDTM ----
cat("Step 2: Transform to SDTM format\n")

STUDYID <- "CAPSTONE02"

# DM Domain
sdtm_dm <- raw_demo %>%
  mutate(
    STUDYID  = STUDYID,
    DOMAIN   = "DM",
    USUBJID  = paste(STUDYID, site, subject, sep = "-"),
    SUBJID   = subject,
    SITEID   = site,
    AGE      = age,
    AGEU     = "YEARS",
    SEX      = if_else(sex == "Male", "M", "F"),
    RACE     = toupper(race),
    ETHNIC   = "NOT HISPANIC OR LATINO",
    ARM      = arm,
    ARMCD    = if_else(arm == "Placebo", "PBO", "DRGA10"),
    ACTARM   = arm,
    ACTARMCD = if_else(arm == "Placebo", "PBO", "DRGA10"),
    RFSTDTC  = as.character(first_dose),
    RFENDTC  = NA_character_
  ) %>%
  select(STUDYID, DOMAIN, USUBJID, SUBJID, SITEID, AGE, AGEU,
         SEX, RACE, ETHNIC, ARM, ARMCD, ACTARM, ACTARMCD, RFSTDTC, RFENDTC)

# AE Domain
sdtm_ae <- raw_aes %>%
  mutate(
    STUDYID  = STUDYID,
    DOMAIN   = "AE",
    USUBJID  = paste(STUDYID, site, subject, sep = "-"),
    AETERM   = ae_term,
    AEDECOD  = toupper(ae_term),
    AEBODSYS = case_when(
      grepl("Headache|Dizziness", ae_term) ~ "NERVOUS SYSTEM DISORDERS",
      grepl("Nausea", ae_term) ~ "GASTROINTESTINAL DISORDERS",
      grepl("Fatigue", ae_term) ~ "GENERAL DISORDERS AND ADMINISTRATION SITE CONDITIONS",
      grepl("Rash", ae_term) ~ "SKIN AND SUBCUTANEOUS TISSUE DISORDERS"
    ),
    AESEV    = toupper(severity),
    AESER    = if_else(serious == "Yes", "Y", "N"),
    AEREL    = toupper(related),
    AESTDTC  = start_date,
    AEENDTC  = end_date
  ) %>%
  arrange(USUBJID, AESTDTC) %>%
  group_by(USUBJID) %>%
  mutate(AESEQ = row_number()) %>%
  ungroup() %>%
  select(STUDYID, DOMAIN, USUBJID, AESEQ, AETERM, AEDECOD, AEBODSYS,
         AESEV, AESER, AEREL, AESTDTC, AEENDTC)

cat("  SDTM DM:", nrow(sdtm_dm), "rows x", ncol(sdtm_dm), "cols\n")
cat("  SDTM AE:", nrow(sdtm_ae), "rows x", ncol(sdtm_ae), "cols\n\n")
```

```{r}
#| label: capstone-step3

# ---- STEP 3: Validate ----
cat("Step 3: Validate SDTM domains\n\n")

# Cross-domain check: All AE subjects in DM
ae_orphans <- sdtm_ae %>% anti_join(sdtm_dm, by = "USUBJID")
cat("  Orphan AE records:", nrow(ae_orphans), if_else(nrow(ae_orphans) == 0, "✓", "✗"), "\n")

# Required variable check
dm_required <- c("STUDYID", "DOMAIN", "USUBJID", "SUBJID", "SITEID",
                 "AGE", "AGEU", "SEX", "RACE", "ARM", "ARMCD")
dm_present <- all(dm_required %in% names(sdtm_dm))
cat("  DM required variables:", if_else(dm_present, "All present ✓", "Missing ✗"), "\n")

ae_required <- c("STUDYID", "DOMAIN", "USUBJID", "AESEQ", "AETERM", "AEDECOD")
ae_present <- all(ae_required %in% names(sdtm_ae))
cat("  AE required variables:", if_else(ae_present, "All present ✓", "Missing ✗"), "\n")

# CT check
valid_sex <- all(sdtm_dm$SEX %in% c("M", "F", "U", "UNDIFFERENTIATED"))
cat("  SEX controlled terminology:", if_else(valid_sex, "Valid ✓", "Invalid ✗"), "\n")

valid_sev <- all(sdtm_ae$AESEV %in% c("MILD", "MODERATE", "SEVERE"))
cat("  AESEV controlled terminology:", if_else(valid_sev, "Valid ✓", "Invalid ✗"), "\n")

# Date logic check
ae_date_issue <- sdtm_ae %>%
  filter(!is.na(AESTDTC), !is.na(AEENDTC)) %>%
  filter(AESTDTC > AEENDTC) %>%
  nrow()
cat("  AE date logic (start <= end):", if_else(ae_date_issue == 0, "Valid ✓", "Issues ✗"), "\n\n")
```

```{r}
#| label: capstone-step4

# ---- STEP 4: Apply Metadata with xportr ----
cat("Step 4: Apply metadata and export\n\n")

# Use our DM spec from earlier
dm_final <- sdtm_dm %>%
  xportr_type(dm_spec, domain = "DM") %>%
  xportr_length(dm_spec, domain = "DM") %>%
  xportr_label(dm_spec, domain = "DM") %>%
  xportr_order(dm_spec, domain = "DM")

cat("  DM metadata applied:\n")
cat("    Variables:", ncol(dm_final), "\n")
cat("    Order:", paste(names(dm_final), collapse = " → "), "\n")

# Quick label check
labels <- sapply(dm_final, function(x) attr(x, "label"))
cat("    Labels applied:", sum(!is.na(labels)), "of", length(labels), "\n\n")
```

```{r}
#| label: capstone-step5

# ---- STEP 5: Export ----
cat("Step 5: Export as .xpt files\n\n")

output_dir <- "output"
if (!dir.exists(output_dir)) dir.create(output_dir)

# Export DM
dm_final %>%
  xportr_write(file.path(output_dir, "dm.xpt"))
cat("  ✓ Exported:", file.path(output_dir, "dm.xpt"),
    "-", file.size(file.path(output_dir, "dm.xpt")), "bytes\n")

# Export AE (with inline metadata)
ae_spec <- tibble::tribble(
  ~dataset, ~variable,   ~type,       ~label,                                  ~length, ~order,  ~format,
  "AE",     "STUDYID",   "character", "Study Identifier",                      12L,     1L,      NA,
  "AE",     "DOMAIN",    "character", "Domain Abbreviation",                   2L,      2L,      NA,
  "AE",     "USUBJID",   "character", "Unique Subject Identifier",             40L,     3L,      NA,
  "AE",     "AESEQ",     "numeric",   "Sequence Number",                       8L,      4L,      NA,
  "AE",     "AETERM",    "character", "Reported Term for the Adverse Event",   200L,    5L,      NA,
  "AE",     "AEDECOD",   "character", "Dictionary-Derived Term",               200L,    6L,      NA,
  "AE",     "AEBODSYS",  "character", "Body System or Organ Class",            200L,    7L,      NA,
  "AE",     "AESEV",     "character", "Severity/Intensity",                    12L,     8L,      NA,
  "AE",     "AESER",     "character", "Serious Event",                         2L,      9L,      NA,
  "AE",     "AEREL",     "character", "Causality",                             40L,     10L,     NA,
  "AE",     "AESTDTC",   "character", "Start Date/Time of Adverse Event",      20L,     11L,     NA,
  "AE",     "AEENDTC",   "character", "End Date/Time of Adverse Event",        20L,     12L,     NA
)

sdtm_ae %>%
  xportr_type(ae_spec, domain = "AE") %>%
  xportr_length(ae_spec, domain = "AE") %>%
  xportr_label(ae_spec, domain = "AE") %>%
  xportr_order(ae_spec, domain = "AE") %>%
  xportr_write(file.path(output_dir, "ae.xpt"))

cat("  ✓ Exported:", file.path(output_dir, "ae.xpt"),
    "-", file.size(file.path(output_dir, "ae.xpt")), "bytes\n")
```

------------------------------------------------------------------------

## Week 2 Review: Everything You've Learned

```{r}
#| label: week2-review

cat("=== WEEK 2 COMPLETE REVIEW ===\n\n")

review <- tibble::tribble(
  ~Day,  ~Topic,                                    ~Key_Skill,
  "8",   "LB Domain & Unit Standardization",        "Unit conversion, reference ranges, LBNRIND",
  "9",   "VS & Repeated Measures",                  "Multiple readings, VSPOS, VSTPT, wide format",
  "10",  "AE Domain Mastery & SAE Logic",           "Severity vs seriousness, TEAE, SAE criteria",
  "11",  "Disposition & Trial Design",              "DS domain, EPOCH, TA/TV/TS, ADSL preview",
  "12",  "Data Cuts with datacutr",                 "Patient-level & record-level cuts, ongoing AEs",
  "13",  "SDTM Validation with sdtmchecks",         "FDA business rules, cross-domain checks",
  "14",  "Metadata-Driven SDTM (this capstone)",    "metacore, metatools, xportr pipeline"
)

print(review)
```

### What You're Now Ready For

::: callout-important
## Preparation Complete for Week 3: ADaM

You now have a solid foundation in SDTM:

-   ✓ You can build any SDTM domain from raw data
-   ✓ You understand Controlled Terminology and date derivations
-   ✓ You can handle complex domains (LB, VS, AE, DS)
-   ✓ You know how to validate SDTM data
-   ✓ You can apply metadata and export submission-ready `.xpt` files
-   ✓ You understand how SDTM maps to ADaM

**In Week 3, we will use `admiral` to build ADaM datasets (ADSL, ADAE, ADVS, ADLB) from the SDTM data you've mastered.**
:::

------------------------------------------------------------------------

## Deliverable Summary

Today you completed the following:

| Task | Status |
|------|--------|
| Understood metacore specification objects | ✓ Done |
| Created variable specifications for DM and AE | ✓ Done |
| Applied xportr_type, xportr_label, xportr_order, xportr_length | ✓ Done |
| Exported submission-ready .xpt files | ✓ Done |
| Built an end-to-end pipeline: raw → SDTM → validate → export | ✓ Done |
| Reviewed all Week 2 topics | ✓ Done |

------------------------------------------------------------------------

## Key Takeaways

1.  **Metadata-driven is the future** - Define once, apply everywhere
2.  **`metacore` standardizes specs** - One R object for all dataset/variable metadata
3.  **`xportr` applies metadata** - Types, labels, lengths, ordering, and export
4.  **`metatools` enables checks** - Verify your data matches the specification
5.  **The pipeline is reproducible** - Same spec + same code = same output every time
6.  **You're ready for ADaM** - All SDTM fundamentals are in place

------------------------------------------------------------------------

## Resources

-   [metacore Documentation](https://pharmaverse.github.io/metacore/) - Official metacore package
-   [metatools Documentation](https://pharmaverse.github.io/metatools/) - Metadata utility functions
-   [xportr Documentation](https://atorus-research.github.io/xportr/) - SAS transport export
-   [Admiral Documentation](https://pharmaverse.github.io/admiral/) - ADaM derivation package (Week 3!)
-   [Pharmaverse.org](https://pharmaverse.org) - R packages for clinical data
-   [FDA Data Standards Resources](https://www.fda.gov/industry/fda-data-standards-advisory-board) - FDA guidance

------------------------------------------------------------------------

## 🎉 Congratulations! Week 2 Complete!

You've now mastered:

-   **Complex SDTM domains** (LB, VS, AE, DS)
-   **Production workflows** (data cuts, validation, metadata)
-   **Pharmaverse tools** (datacutr, sdtmchecks, metacore, xportr)
-   **Regulatory requirements** (SAE logic, TEAE derivation, controlled terminology)

## What's Next?

**Week 3: ADaM Datasets with Admiral**

-   Using `admiral` to derive ADaM datasets from SDTM
-   Building ADSL (Subject-Level Analysis Dataset)
-   Creating BDS datasets: ADVS, ADLB
-   Deriving baseline, change from baseline, shift tables
-   ADAE creation with treatment-emergent logic
