---
title: "Day 8: Complex SDTM Domains - LB (Lab Results)"
subtitle: "Findings Class with Unit Standardization"


format:
  html:
    theme:
      light: cosmo

    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css
    
editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

<div style="margin-bottom:12px;">< a target="_blank" href="../index.html" class="resource-link">← Back to Roadmap</a></div>

## Learning Objectives

By the end of Day 8, you will be able to:

1.  Understand the **LB (Laboratory) domain** structure and its role as a **Findings class** domain
2.  Identify and explain the key LB variables: **LBTESTCD, LBORRES, LBORRESU, LBSTRESC, LBSTRESN, LBSTRESU**
3.  Perform **unit conversions** (e.g., mg/dL → mmol/L) for standardized results
4.  Derive **reference range flags** (LBNRIND) and understand normal range variables
5.  Create a complete LB domain from simulated raw data using `dplyr` and `tidyr`

------------------------------------------------------------------------

## Introduction to the LB Domain

### What is the LB Domain?

The **LB (Laboratory) domain** is one of the most complex and data-rich domains in SDTM. It belongs to the **Findings** class of SDTM domains, which are designed to capture observations or measurements collected during a study.

::: {.callout-note collapse="true"}
## Findings Class Characteristics

Findings domains share a common structure:

-   **--TESTCD**: Short name for the test (e.g., "ALT", "GLUC")
-   **--TEST**: Full descriptive name (e.g., "Alanine Aminotransferase", "Glucose")
-   **--ORRES**: Original result as collected (character)
-   **--ORRESU**: Original unit of the result
-   **--STRESC**: Standardized result in character format
-   **--STRESN**: Standardized result in numeric format
-   **--STRESU**: Standardized unit
:::

### Why is LB Important?

Laboratory data is fundamental to clinical trials because:

-   It provides **objective measurements** of safety (liver enzymes, kidney function)
-   It helps assess **efficacy** (biomarkers, disease markers)
-   Regulatory agencies require **standardized lab data** for safety reviews
-   Lab abnormalities often determine eligibility and safety signals

### The LB Domain Structure

```
┌──────────────────────────────────────────────────────────────────────────────┐
│                        LB DOMAIN - KEY VARIABLES                             │
├──────────────────────────────────────────────────────────────────────────────┤
│  IDENTIFIER VARIABLES                                                        │
│    STUDYID    = Study identifier                                             │
│    USUBJID    = Unique subject identifier (links to DM)                      │
│    LBSEQ      = Sequence number within subject                               │
├──────────────────────────────────────────────────────────────────────────────┤
│  TOPIC VARIABLE                                                              │
│    LBTESTCD   = Lab test short name (e.g., "ALT", "BILI", "GLUC")            │
│    LBTEST     = Lab test full name (e.g., "Alanine Aminotransferase")        │
│    LBCAT      = Category (e.g., "CHEMISTRY", "HEMATOLOGY")                   │
├──────────────────────────────────────────────────────────────────────────────┤
│  RESULT VARIABLES (Original)                                                 │
│    LBORRES    = Result as originally collected (character)                   │
│    LBORRESU   = Unit of original result                                      │
├──────────────────────────────────────────────────────────────────────────────┤
│  RESULT VARIABLES (Standardized)                                             │
│    LBSTRESC   = Standardized result (character)                              │
│    LBSTRESN   = Standardized result (numeric)                                │
│    LBSTRESU   = Standardized unit                                            │
├──────────────────────────────────────────────────────────────────────────────┤
│  REFERENCE RANGE VARIABLES                                                   │
│    LBSTNRLO   = Standard normal range lower limit                            │
│    LBSTNRHI   = Standard normal range upper limit                            │
│    LBNRIND    = Normal range indicator (LOW/NORMAL/HIGH)                     │
├──────────────────────────────────────────────────────────────────────────────┤
│  TIMING VARIABLES                                                            │
│    VISITNUM   = Visit number                                                 │
│    VISIT      = Visit name (e.g., "SCREENING", "WEEK 2")                     │
│    LBDTC      = Date/time of specimen collection (ISO 8601)                  │
│    LBDY       = Study day of specimen collection                             │
└──────────────────────────────────────────────────────────────────────────────┘
```

------------------------------------------------------------------------

## Package Installation & Loading

### Required Packages

| Package           | Purpose                                            |
|-------------------|----------------------------------------------------|
| `dplyr`           | Data manipulation (filter, mutate, joins)          |
| `tidyr`           | Data reshaping (pivot operations)                  |
| `pharmaversesdtm` | Example SDTM datasets including LB                 |

### Install Packages (if needed)

```{r}
#| label: install-packages

# Install packages if not already installed
if (!requireNamespace("dplyr", quietly = TRUE)) suppressMessages(install.packages("dplyr"))
if (!requireNamespace("tidyr", quietly = TRUE)) suppressMessages(install.packages("tidyr"))
if (!requireNamespace("pharmaversesdtm", quietly = TRUE)) suppressMessages(install.packages("pharmaversesdtm"))
```

### Load Packages

```{r}
#| label: load-packages

library(dplyr)
library(tidyr)
library(pharmaversesdtm)
```

------------------------------------------------------------------------

## Exploring pharmaversesdtm LB Data

Let's start by loading and exploring the LB domain from `pharmaversesdtm`. This is the same data used by production teams in the pharmaverse community when developing the `admiral` package.

### Load the LB Domain

```{r}
#| label: load-lb-data

# Load LB domain from pharmaversesdtm
data("lb", package = "pharmaversesdtm")

# Quick overview
cat("LB domain dimensions:", nrow(lb), "rows x", ncol(lb), "columns\n")
cat("Number of unique subjects:", n_distinct(lb$USUBJID), "\n")
cat("Number of unique tests:", n_distinct(lb$LBTESTCD), "\n")
```

### Explore LB Structure

```{r}
#| label: glimpse-lb

# View the structure of the LB domain
dplyr::glimpse(lb)
```

### Key LB Variables Explained

Let's examine some critical LB variables:

```{r}
#| label: explore-lb-variables

# View unique lab tests
lb %>%
  distinct(LBTESTCD, LBTEST, LBCAT) %>%
  arrange(LBCAT, LBTESTCD) %>%
  head(15)
```

::: {.callout-tip collapse="true"}
## Understanding Test Codes

Notice how **LBTESTCD** is a short, standardized code while **LBTEST** is the full descriptive name. This follows CDISC conventions where:

-   **Short codes** (LBTESTCD) enable efficient data processing
-   **Full names** (LBTEST) provide human-readable descriptions
-   **Categories** (LBCAT) group related tests (CHEMISTRY, HEMATOLOGY, etc.)
:::

------------------------------------------------------------------------

## Understanding Original vs Standardized Results

One of the most important concepts in the LB domain is the distinction between **original** and **standardized** results.

### Why Standardization Matters

Labs from different sites may report the same test in different units:

-   Site A reports Glucose as **90 mg/dL**
-   Site B reports Glucose as **5.0 mmol/L**

Both values are the same measurement, just in different units! SDTM requires us to **standardize** these values so they can be analyzed together.

### The Result Variable Pairs

| Original Variable | Standardized Variable | Description |
|-------------------|----------------------|-------------|
| `LBORRES` | `LBSTRESC` | Result (character format) |
| - | `LBSTRESN` | Result (numeric format) |
| `LBORRESU` | `LBSTRESU` | Unit of measurement |

### Viewing Original vs Standardized Values

```{r}
#| label: compare-results

# Compare original and standardized values for a specific test
lb %>%
  filter(LBTESTCD == "GLUC") %>%
  select(USUBJID, VISIT, LBORRES, LBORRESU, LBSTRESC, LBSTRESN, LBSTRESU) %>%
  head(10)
```

------------------------------------------------------------------------

## Unit Conversion: A Hands-On Example

Let's create a practical example of unit conversion. This is one of the most common tasks when building an LB domain from raw lab data.

### Common Lab Unit Conversions

| Test | Original Unit | Standard Unit | Conversion Factor |
|------|---------------|---------------|-------------------|
| Glucose | mg/dL | mmol/L | ÷ 18.02 |
| Creatinine | mg/dL | µmol/L | × 88.42 |
| Bilirubin | mg/dL | µmol/L | × 17.10 |
| Cholesterol | mg/dL | mmol/L | ÷ 38.67 |

::: {.callout-note collapse="true"}
## Reference for Unit Conversions

For a comprehensive list of clinical lab unit conversions, see:

-   [National Institutes of Health - Unit Conversion Tables](https://www.ncbi.nlm.nih.gov/books/NBK535333/)
-   [CDISC SEND/SDTM Unit Standards](https://www.cdisc.org/standards/foundational/sdtm)
:::

### Step-by-Step: Creating Standardized Results

Let's simulate a scenario where we receive raw lab data and need to standardize it.

```{r}
#| label: create-raw-data

# Create simulated raw lab data with mixed units
raw_lab <- tibble::tribble(
  ~USUBJID,           ~LBTESTCD, ~LBTEST,              ~LBORRES, ~LBORRESU, ~VISITNUM, ~VISIT,
  "CDISC01-001-001",  "GLUC",    "Glucose",            "95",     "mg/dL",   1,         "BASELINE",
  "CDISC01-001-001",  "GLUC",    "Glucose",            "5.1",    "mmol/L",  2,         "WEEK 2",
  "CDISC01-001-001",  "CREAT",   "Creatinine",         "1.2",    "mg/dL",   1,         "BASELINE",
  "CDISC01-001-001",  "CREAT",   "Creatinine",         "98",     "umol/L",  2,         "WEEK 2",
  "CDISC01-001-002",  "GLUC",    "Glucose",            "102",    "mg/dL",   1,         "BASELINE",
  "CDISC01-001-002",  "GLUC",    "Glucose",            "88",     "mg/dL",   2,         "WEEK 2",
  "CDISC01-001-002",  "BILI",    "Bilirubin Total",    "0.8",    "mg/dL",   1,         "BASELINE",
  "CDISC01-001-002",  "BILI",    "Bilirubin Total",    "1.1",    "mg/dL",   2,         "WEEK 2"
)

cat("Raw lab data with mixed units:\n")
print(raw_lab)
```

### Create a Unit Conversion Function

```{r}
#| label: conversion-function

# Define a reusable conversion function
convert_lab_units <- function(testcd, value, from_unit, to_unit) {
  # Convert character value to numeric
  val <- as.numeric(value)
  
  # Define conversion factors (from original to standard)
  # Standard units: Glucose = mmol/L, Creatinine = umol/L, Bilirubin = umol/L
  
  result <- case_when(
    # Glucose: mg/dL to mmol/L
    testcd == "GLUC" & from_unit == "mg/dL" & to_unit == "mmol/L" ~ val / 18.02,
    
    # Glucose: already in mmol/L
    testcd == "GLUC" & from_unit == "mmol/L" & to_unit == "mmol/L" ~ val,
    
    # Creatinine: mg/dL to umol/L
    testcd == "CREAT" & from_unit == "mg/dL" & to_unit == "umol/L" ~ val * 88.42,
    
    # Creatinine: already in umol/L
    testcd == "CREAT" & from_unit == "umol/L" & to_unit == "umol/L" ~ val,
    
    # Bilirubin: mg/dL to umol/L
    testcd == "BILI" & from_unit == "mg/dL" & to_unit == "umol/L" ~ val * 17.10,
    
    # Default: no conversion
    TRUE ~ val
  )
  
  return(round(result, 2))
}
```

### Apply Unit Conversions

```{r}
#| label: apply-conversions

# Define standard units for each test
standard_units <- tibble::tribble(
  ~LBTESTCD, ~LBSTRESU,
  "GLUC",    "mmol/L",
  "CREAT",   "umol/L",
  "BILI",    "umol/L"
)

# Apply standardization
lab_standardized <- raw_lab %>%
  # Join to get target standard units
  left_join(standard_units, by = "LBTESTCD") %>%
  # Apply conversion
  mutate(
    LBSTRESN = convert_lab_units(LBTESTCD, LBORRES, LBORRESU, LBSTRESU),
    LBSTRESC = as.character(LBSTRESN)
  ) %>%
  # Reorder columns for clarity
  select(USUBJID, LBTESTCD, LBTEST, VISITNUM, VISIT, 
         LBORRES, LBORRESU, LBSTRESC, LBSTRESN, LBSTRESU)

cat("Standardized lab data:\n")
print(lab_standardized)
```

::: {.callout-important collapse="true"}
## Key Learning Point

Notice how the **original** values are preserved (`LBORRES`, `LBORRESU`) while **standardized** values are derived (`LBSTRESC`, `LBSTRESN`, `LBSTRESU`). This is a fundamental SDTM principle: **never lose the original data!**
:::

------------------------------------------------------------------------

## Reference Ranges and Normal Range Indicators

### What are Reference Ranges?

Reference ranges (also called normal ranges) define the expected values for a lab test in a healthy population. Values outside this range may indicate:

-   **LOW**: Value below the normal range (potentially abnormal)
-   **NORMAL**: Value within expected range
-   **HIGH**: Value above the normal range (potentially abnormal)

### Key Reference Range Variables

| Variable | Description |
|----------|-------------|
| `LBSTNRLO` | Standard normal range - lower limit |
| `LBSTNRHI` | Standard normal range - upper limit |
| `LBNRIND` | Normal range indicator (LOW/NORMAL/HIGH) |
| `LBORNRLO` | Original normal range - lower limit |
| `LBORNRHI` | Original normal range - upper limit |

### Deriving Reference Range Flags

```{r}
#| label: reference-ranges

# Define reference ranges (typically from lab vendor or specifications)
reference_ranges <- tibble::tribble(
  ~LBTESTCD, ~LBSTNRLO, ~LBSTNRHI, ~LBSTRESU,
  "GLUC",    3.9,       5.6,       "mmol/L",   # Fasting glucose
  "CREAT",   62,        106,       "umol/L",   # Creatinine (male)
  "BILI",    5.1,       17.0,      "umol/L"    # Total bilirubin
)

# Join reference ranges and derive LBNRIND
lab_with_ranges <- lab_standardized %>%
  left_join(
    reference_ranges %>% select(LBTESTCD, LBSTNRLO, LBSTNRHI),
    by = "LBTESTCD"
  ) %>%
  mutate(
    # Derive normal range indicator
    LBNRIND = case_when(
      is.na(LBSTRESN) ~ NA_character_,                    # Missing result
      LBSTRESN < LBSTNRLO ~ "LOW",                        # Below range
      LBSTRESN > LBSTNRHI ~ "HIGH",                       # Above range
      TRUE ~ "NORMAL"                                     # Within range
    )
  )

# Display results
lab_with_ranges %>%
  select(USUBJID, LBTESTCD, VISIT, LBSTRESN, LBSTRESU, 
         LBSTNRLO, LBSTNRHI, LBNRIND)
```

::: {.callout-tip collapse="true"}
## Clinical Significance

The `LBNRIND` variable is critical for safety analyses. It helps identify:

-   **Clinically significant abnormalities** that may require action
-   **Trends** in lab values over time (e.g., worsening liver function)
-   **Treatment-emergent abnormalities** (normal at baseline, abnormal post-treatment)
:::

------------------------------------------------------------------------

## LOINC Coding Concepts

### What is LOINC?

**LOINC** (Logical Observation Identifiers Names and Codes) is a universal standard for identifying medical laboratory observations. While not always required in SDTM, it enhances interoperability.

### How LOINC Integrates with SDTM

| SDTM Variable | LOINC Equivalent |
|---------------|------------------|
| `LBLOINC` | LOINC code for the test |
| `LBTESTCD` | Mapped to LOINC component |

### Example LOINC Codes

| Test | LBTESTCD | LOINC Code | Description |
|------|----------|------------|-------------|
| Glucose | GLUC | 2339-0 | Glucose [Mass/volume] in Blood |
| Creatinine | CREAT | 2160-0 | Creatinine [Mass/volume] in Serum or Plasma |
| ALT | ALT | 1742-6 | Alanine aminotransferase [Enzymatic activity/volume] in Serum or Plasma |

::: {.callout-note collapse="true"}
## Learn More About LOINC

-   [LOINC Official Website](https://loinc.org/)
-   [Search LOINC Codes](https://loinc.org/search/)
-   [LOINC and CDISC Integration Guide](https://www.cdisc.org/kb/articles/loinc-cdisc-lab-test-terminology)
:::

------------------------------------------------------------------------

## Complete Example: Building an LB Domain

Let's put everything together and build a more complete LB domain from simulated raw data.

### Step 1: Create Comprehensive Raw Data

```{r}
#| label: comprehensive-raw-data

# Create more comprehensive raw lab data
set.seed(42)

subjects <- c("CDISC01-001-001", "CDISC01-001-002", "CDISC01-001-003",
              "CDISC01-001-004", "CDISC01-001-005")
visits <- c("SCREENING", "BASELINE", "WEEK 2", "WEEK 4", "WEEK 8")
visit_nums <- c(-1, 1, 2, 3, 4)

# Generate raw lab data for 5 subjects across 5 visits
raw_lb_data <- expand_grid(
  USUBJID = subjects,
  VISIT_INFO = tibble(VISIT = visits, VISITNUM = visit_nums)
) %>%
  unnest(VISIT_INFO) %>%
  # Add lab tests
  crossing(
    tibble::tribble(
      ~LBTESTCD, ~LBTEST,                      ~LBCAT,      ~LBORRESU,
      "GLUC",    "Glucose",                    "CHEMISTRY", "mg/dL",
      "CREAT",   "Creatinine",                 "CHEMISTRY", "mg/dL",
      "ALT",     "Alanine Aminotransferase",   "CHEMISTRY", "U/L",
      "AST",     "Aspartate Aminotransferase", "CHEMISTRY", "U/L",
      "BILI",    "Bilirubin Total",            "CHEMISTRY", "mg/dL",
      "WBC",     "Leukocytes",                 "HEMATOLOGY","10^9/L",
      "RBC",     "Erythrocytes",               "HEMATOLOGY","10^12/L",
      "HGB",     "Hemoglobin",                 "HEMATOLOGY","g/dL",
      "PLT",     "Platelets",                  "HEMATOLOGY","10^9/L",
      "SODIUM",  "Sodium",                     "CHEMISTRY", "mmol/L"
    )
  ) %>%
  # Generate random values within typical ranges
  rowwise() %>%
  mutate(
    LBORRES = as.character(round(case_when(
      LBTESTCD == "GLUC"   ~ rnorm(1, 100, 15),
      LBTESTCD == "CREAT"  ~ rnorm(1, 1.0, 0.2),
      LBTESTCD == "ALT"    ~ rnorm(1, 30, 10),
      LBTESTCD == "AST"    ~ rnorm(1, 28, 8),
      LBTESTCD == "BILI"   ~ rnorm(1, 0.8, 0.3),
      LBTESTCD == "WBC"    ~ rnorm(1, 7.0, 2.0),
      LBTESTCD == "RBC"    ~ rnorm(1, 4.8, 0.5),
      LBTESTCD == "HGB"    ~ rnorm(1, 14.5, 1.5),
      LBTESTCD == "PLT"    ~ rnorm(1, 250, 50),
      LBTESTCD == "SODIUM" ~ rnorm(1, 140, 3),
      TRUE ~ NA_real_
    ), 2))
  ) %>%
  ungroup() %>%
  # Add STUDYID and sequence
  mutate(STUDYID = "CDISC01") %>%
  group_by(USUBJID) %>%
  mutate(LBSEQ = row_number()) %>%
  ungroup()

cat("Raw LB data created:", nrow(raw_lb_data), "records\n")
cat("Tests per subject per visit:", n_distinct(raw_lb_data$LBTESTCD), "\n")
```

### Step 2: Apply Standardization

```{r}
#| label: full-standardization

# Define standard units and reference ranges for all tests
lab_standards <- tibble::tribble(
  ~LBTESTCD, ~LBSTRESU,  ~LBSTNRLO, ~LBSTNRHI, ~CONV_FACTOR,
  "GLUC",    "mmol/L",   3.9,       5.6,       0.0555,    # mg/dL to mmol/L
  "CREAT",   "umol/L",   62,        106,       88.42,     # mg/dL to umol/L
  "ALT",     "U/L",      7,         56,        1.0,       # No conversion
  "AST",     "U/L",      10,        40,        1.0,       # No conversion
  "BILI",    "umol/L",   5.1,       17.0,      17.10,     # mg/dL to umol/L
  "WBC",     "10^9/L",   4.5,       11.0,      1.0,       # No conversion
  "RBC",     "10^12/L",  4.2,       5.4,       1.0,       # No conversion
  "HGB",     "g/dL",     12.0,      17.0,      1.0,       # No conversion
  "PLT",     "10^9/L",   150,       400,       1.0,       # No conversion
  "SODIUM",  "mmol/L",   136,       145,       1.0        # No conversion
)

# Create complete LB domain
lb_complete <- raw_lb_data %>%
  # Join standards
  left_join(lab_standards, by = "LBTESTCD") %>%
  # Calculate standardized numeric result
  mutate(
    LBSTRESN = round(as.numeric(LBORRES) * CONV_FACTOR, 2),
    LBSTRESC = as.character(LBSTRESN),
    # Derive normal range indicator
    LBNRIND = case_when(
      is.na(LBSTRESN) ~ NA_character_,
      LBSTRESN < LBSTNRLO ~ "LOW",
      LBSTRESN > LBSTNRHI ~ "HIGH",
      TRUE ~ "NORMAL"
    ),
    # Add specimen type (typically from raw data, simulated here)
    LBSPEC = "SERUM"
  ) %>%
  # Select and order columns per SDTM standard
  select(
    STUDYID, USUBJID, LBSEQ,
    LBTESTCD, LBTEST, LBCAT, LBSPEC,
    LBORRES, LBORRESU,
    LBSTRESC, LBSTRESN, LBSTRESU,
    LBSTNRLO, LBSTNRHI, LBNRIND,
    VISITNUM, VISIT
  ) %>%
  arrange(STUDYID, USUBJID, VISITNUM, LBTESTCD)

# Preview
cat("\nComplete LB domain preview:\n")
head(lb_complete, 15)
```

### Step 3: Validate the LB Domain

```{r}
#| label: validate-lb

# Validation checks
cat("=== LB Domain Validation ===\n\n")

# Check 1: All required variables present
required_vars <- c("STUDYID", "USUBJID", "LBSEQ", "LBTESTCD", "LBTEST", 
                   "LBORRES", "LBORRESU", "LBSTRESN", "LBSTRESU")
present <- required_vars %in% names(lb_complete)
cat("Required variables check:\n")
for (i in seq_along(required_vars)) {
  cat("  ", required_vars[i], ":", ifelse(present[i], "✓", "✗"), "\n")
}

# Check 2: No missing USUBJID
cat("\nMissing USUBJID:", sum(is.na(lb_complete$USUBJID)), "\n")

# Check 3: Distribution of LBNRIND
cat("\nNormal Range Indicator distribution:\n")
table(lb_complete$LBNRIND, useNA = "ifany")

# Check 4: Unique subjects and visits
cat("\nData coverage:\n")
cat("  Unique subjects:", n_distinct(lb_complete$USUBJID), "\n")
cat("  Unique visits:", n_distinct(lb_complete$VISIT), "\n")
cat("  Unique tests:", n_distinct(lb_complete$LBTESTCD), "\n")
cat("  Total records:", nrow(lb_complete), "\n")
```

------------------------------------------------------------------------

## Summary Statistics for LB Data

When working with lab data, it's often useful to generate summary statistics.

```{r}
#| label: lb-summaries

# Summary by test and visit
lb_summary <- lb_complete %>%
  group_by(LBTESTCD, LBTEST, VISIT, LBSTRESU) %>%
  summarise(
    N = n(),
    Mean = round(mean(LBSTRESN, na.rm = TRUE), 2),
    SD = round(sd(LBSTRESN, na.rm = TRUE), 2),
    Min = round(min(LBSTRESN, na.rm = TRUE), 2),
    Max = round(max(LBSTRESN, na.rm = TRUE), 2),
    N_Low = sum(LBNRIND == "LOW", na.rm = TRUE),
    N_High = sum(LBNRIND == "HIGH", na.rm = TRUE),
    .groups = "drop"
  )

# Display chemistry panel at BASELINE
cat("Chemistry Panel Summary at BASELINE:\n")
lb_summary %>%
  filter(VISIT == "BASELINE", LBTESTCD %in% c("GLUC", "CREAT", "ALT", "AST", "BILI")) %>%
  print()
```

------------------------------------------------------------------------

## Deliverable Summary

Today you completed the following:

| Task | Status |
|------|--------|
| Understood LB domain structure and Findings class | ✓ Done |
| Explored key LB variables (LBTESTCD, LBORRES, LBSTRESN, etc.) | ✓ Done |
| Performed unit conversions (mg/dL → mmol/L) | ✓ Done |
| Derived reference range flags (LBNRIND) | ✓ Done |
| Created a complete LB domain with 10+ tests | ✓ Done |
| Generated summary statistics | ✓ Done |

------------------------------------------------------------------------

## Key Takeaways

1.  **LB is a Findings class domain** - It follows the --TESTCD, --ORRES, --STRESC pattern shared by VS, LB, EG
2.  **Original vs Standardized** - Always preserve original results while creating standardized versions
3.  **Unit conversions are critical** - Labs from different sites may use different units
4.  **Reference ranges enable safety analysis** - LBNRIND flags abnormal values
5.  **LOINC provides universal identification** - Enhances data interoperability

------------------------------------------------------------------------

## Resources

-   [CDISC SDTM Implementation Guide - LB Domain](https://www.cdisc.org/standards/foundational/sdtm) - Official LB specification
-   [LOINC Official Website](https://loinc.org/) - Medical laboratory codes
-   [NIH Unit Conversion Tables](https://www.ncbi.nlm.nih.gov/books/NBK535333/) - Lab unit conversions
-   [Pharmaverse.org](https://pharmaverse.org) - R packages for clinical data
-   [sdtm.oak Documentation](https://pharmaverse.github.io/sdtm.oak/) - SDTM creation package

------------------------------------------------------------------------

## What's Next?

In **Day 9**, we will focus on **VS (Vital Signs) & Repeated Measures**:

-   Understanding visit-level data and multiple readings per timepoint
-   Working with positional variables (VSPOS, VSLOC)
-   Deriving statistics across multiple readings
-   Preparing for ADaM BDS structure (baseline, change from baseline)

< a target="_blank" href="https://visitorbadge.io/status?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse"><img src="https://api.visitorbadge.io/api/visitors?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse&label=Visits%20till%20date&countColor=%23263759&style=plastic&labelStyle=none" /></a>
