---
title: "Day 9: VS (Vital Signs) & Repeated Measures"
subtitle: "Visit-Level Data and Positional Readings"


format:
  html:
    theme:
      light: cosmo

    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css
    
editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

<div style="margin-bottom:12px;"><a href="../index.html" class="resource-link">← Back to Roadmap</a></div>

## Learning Objectives

By the end of Day 9, you will be able to:

1.  Understand the **VS (Vital Signs) domain** structure and its role as a **Findings class** domain
2.  Handle **repeated measures** - multiple readings per subject per visit per timepoint
3.  Work with **positional variables** like `VSPOS` (SITTING/STANDING) and `VSLOC` (ARM)
4.  Derive **summary statistics** (mean, min, max) across multiple readings
5.  Understand how VS data maps to the **ADaM BDS (Basic Data Structure)** for ADVS

------------------------------------------------------------------------

## Introduction to the VS Domain

### What is the VS Domain?

The **VS (Vital Signs) domain** captures measurements of a subject's vital functions during a clinical trial. Like the LB domain we covered on Day 8, VS belongs to the **Findings** class of SDTM domains.

Common vital signs measured include:

-   **Blood Pressure** (Systolic and Diastolic)
-   **Heart Rate** (Pulse)
-   **Temperature**
-   **Respiratory Rate**
-   **Weight and Height**
-   **Oxygen Saturation**

### Why is VS Special?

The VS domain introduces an important concept: **repeated measures at the same timepoint**. Unlike demographics (one row per subject) or adverse events (one row per event), vital signs are often measured multiple times:

-   **Multiple blood pressure readings** at the same visit (e.g., 3 readings taken 2 minutes apart)
-   **Position-dependent measurements** (sitting vs. standing blood pressure)
-   **Location-specific readings** (left arm vs. right arm)

::: {.callout-note collapse="true"}
## Repeated Measures Challenge

When analyzing vital signs, we need to decide:

-   Do we use the **first** reading?
-   Do we calculate the **mean** of all readings?
-   Do we report the **minimum** or **maximum**?

This is a key consideration when building ADaM datasets (ADVS) from SDTM VS data.
:::

### The VS Domain Structure

```
┌──────────────────────────────────────────────────────────────────────────────┐
│                        VS DOMAIN - KEY VARIABLES                             │
├──────────────────────────────────────────────────────────────────────────────┤
│  IDENTIFIER VARIABLES                                                        │
│    STUDYID    = Study identifier                                             │
│    USUBJID    = Unique subject identifier (links to DM)                      │
│    VSSEQ      = Sequence number within subject                               │
├──────────────────────────────────────────────────────────────────────────────┤
│  TOPIC VARIABLE                                                              │
│    VSTESTCD   = Vital sign test short name (e.g., "SYSBP", "DIABP", "HR")    │
│    VSTEST     = Vital sign test name (e.g., "Systolic Blood Pressure")       │
├──────────────────────────────────────────────────────────────────────────────┤
│  RESULT VARIABLES                                                            │
│    VSORRES    = Result as originally collected (character)                   │
│    VSORRESU   = Unit of original result                                      │
│    VSSTRESC   = Standardized result (character)                              │
│    VSSTRESN   = Standardized result (numeric)                                │
│    VSSTRESU   = Standardized unit                                            │
├──────────────────────────────────────────────────────────────────────────────┤
│  POSITIONAL VARIABLES (Unique to VS)                                         │
│    VSPOS      = Position of subject (SITTING, STANDING, SUPINE)              │
│    VSLOC      = Location on body (LEFT ARM, RIGHT ARM)                       │
├──────────────────────────────────────────────────────────────────────────────┤
│  TIMING VARIABLES                                                            │
│    VISITNUM   = Visit number                                                 │
│    VISIT      = Visit name (e.g., "SCREENING", "WEEK 2")                     │
│    VSDY       = Study day of measurement                                     │
│    VSTPT      = Planned time point name (PRE-DOSE, 1 HOUR POST-DOSE)         │
│    VSTPTNUM   = Planned time point number                                    │
│    VSDTC      = Date/time of measurement (ISO 8601)                          │
└──────────────────────────────────────────────────────────────────────────────┘
```

------------------------------------------------------------------------

## Package Installation & Loading

### Required Packages

| Package           | Purpose                                            |
|-------------------|----------------------------------------------------|
| `dplyr`           | Data manipulation (filter, mutate, summarise)      |
| `tidyr`           | Data reshaping (pivot operations)                  |
| `pharmaversesdtm` | Example SDTM datasets including VS                 |

### Install Packages (if needed)

```{r}
#| label: install-packages

# Install packages if not already installed
if (!requireNamespace("dplyr", quietly = TRUE)) suppressMessages(install.packages("dplyr"))
if (!requireNamespace("tidyr", quietly = TRUE)) suppressMessages(install.packages("tidyr"))
if (!requireNamespace("pharmaversesdtm", quietly = TRUE)) suppressMessages(install.packages("pharmaversesdtm"))
```

### Load Packages

```{r}
#| label: load-packages

library(dplyr)
library(tidyr)
library(pharmaversesdtm)
```

------------------------------------------------------------------------

## Exploring pharmaversesdtm VS Data

Let's start by loading and exploring the VS domain from `pharmaversesdtm`.

### Load the VS Domain

```{r}
#| label: load-vs-data

# Load VS domain from pharmaversesdtm
data("vs", package = "pharmaversesdtm")

# Quick overview
cat("VS domain dimensions:", nrow(vs), "rows x", ncol(vs), "columns\n")
cat("Number of unique subjects:", n_distinct(vs$USUBJID), "\n")
cat("Number of unique tests:", n_distinct(vs$VSTESTCD), "\n")
```

### Explore VS Structure

```{r}
#| label: glimpse-vs

# View the structure of the VS domain
dplyr::glimpse(vs)
```

### Key VS Variables

Let's examine the vital sign tests available in our data:

```{r}
#| label: explore-tests

# View unique vital sign tests
vs %>%
  distinct(VSTESTCD, VSTEST, VSSTRESU) %>%
  arrange(VSTESTCD)
```

::: {.callout-tip collapse="true"}
## Understanding VS Test Codes

Common CDISC vital sign test codes:

-   **SYSBP** = Systolic Blood Pressure
-   **DIABP** = Diastolic Blood Pressure
-   **PULSE** = Pulse Rate / Heart Rate
-   **TEMP** = Temperature
-   **RESP** = Respiratory Rate
-   **WEIGHT** = Weight
-   **HEIGHT** = Height
-   **BMI** = Body Mass Index (often derived)
:::

------------------------------------------------------------------------

## Understanding Visit Structure

### Visits and Timepoints

In clinical trials, vital signs are collected at specific **visits** (e.g., Screening, Week 2, Week 8) and sometimes at specific **timepoints within a visit** (e.g., pre-dose, 1 hour post-dose).

```{r}
#| label: explore-visits

# View visit structure
vs %>%
  distinct(VISITNUM, VISIT) %>%
  arrange(VISITNUM)
```

### SDTM Timing Variables Explained

| Variable | Description | Example |
|----------|-------------|---------|
| `VISITNUM` | Numeric visit identifier | 1, 2, 3 |
| `VISIT` | Visit name | "BASELINE", "WEEK 2" |
| `VSDY` | Study day of measurement | 1, 15, 29 |
| `VSTPT` | Planned time point name | "PRE-DOSE", "1HR POST-DOSE" |
| `VSTPTNUM` | Planned time point number | 1, 2, 3 |
| `VSDTC` | Date/time (ISO 8601) | "2023-03-15T09:30:00" |

```{r}
#| label: timing-example

# Example of timing variables
vs %>%
  filter(USUBJID == first(vs$USUBJID), VSTESTCD == "SYSBP") %>%
  select(USUBJID, VSTESTCD, VISITNUM, VISIT, VSDY, VSTPT, VSTPTNUM) %>%
  arrange(VISITNUM, VSTPTNUM) %>%
  head(10)
```

------------------------------------------------------------------------

## Positional Variables: VSPOS and VSLOC

### Why Position and Location Matter

Blood pressure measurements can vary significantly based on:

-   **Position**: Sitting, standing, or lying down (supine)
-   **Location**: Which arm the reading was taken from

Clinical protocols often specify:

> "Blood pressure should be measured after the patient has been seated for at least 5 minutes, using the right arm."

### VSPOS - Position of Subject

The `VSPOS` variable captures the subject's position during measurement:

| VSPOS Value | Description |
|-------------|-------------|
| `SITTING` | Subject is seated |
| `STANDING` | Subject is standing |
| `SUPINE` | Subject is lying flat |
| `RECLINING` | Subject is partially reclined |

```{r}
#| label: explore-position

# Check position values in our data
vs %>%
  filter(VSTESTCD %in% c("SYSBP", "DIABP")) %>%
  count(VSPOS, name = "Count") %>%
  arrange(desc(Count))
```

### VSLOC - Location on Body

The `VSLOC` variable specifies where on the body the measurement was taken:

| VSLOC Value | Description |
|-------------|-------------|
| `LEFT ARM` | Blood pressure cuff on left arm |
| `RIGHT ARM` | Blood pressure cuff on right arm |
| `ARM` | Arm (unspecified) |
| `ORAL` | Oral temperature |
| `AXILLARY` | Under the armpit (temperature) |

```{r}
#| label: explore-location

# Check location values
vs %>%
  filter(VSTESTCD %in% c("SYSBP", "DIABP", "TEMP")) %>%
  count(VSTESTCD, VSLOC, name = "Count") %>%
  arrange(VSTESTCD, desc(Count))
```

------------------------------------------------------------------------

## Handling Multiple Readings

### The Problem: Multiple Measurements

In many studies, vital signs are measured multiple times at the same visit to ensure accuracy. For blood pressure, it's common to take **3 readings** and use the **average**.

### Creating Sample Data with Multiple Readings

Let's create a dataset that demonstrates this scenario:

```{r}
#| label: create-multiple-readings

# Create sample data with 3 BP readings per visit
set.seed(42)

subjects <- c("CDISC01-001-001", "CDISC01-001-002", "CDISC01-001-003")
visits <- c("SCREENING", "BASELINE", "WEEK 2", "WEEK 4")
visit_nums <- c(-1, 1, 2, 3)

# Generate VS data with multiple readings per visit
vs_multiple <- expand_grid(
  USUBJID = subjects,
  VISIT_INFO = tibble(VISIT = visits, VISITNUM = visit_nums)
) %>%
  unnest(VISIT_INFO) %>%
  # Create 3 readings per vital sign per visit
  crossing(
    READING = 1:3,
    tibble::tribble(
      ~VSTESTCD, ~VSTEST,                     ~VSSTRESU, ~BASE_VALUE, ~SD_VALUE,
      "SYSBP",   "Systolic Blood Pressure",   "mmHg",    125,         10,
      "DIABP",   "Diastolic Blood Pressure",  "mmHg",    80,          8,
      "PULSE",   "Pulse Rate",                "beats/min", 72,        8
    )
  ) %>%
  # Generate values with some variation between readings
  rowwise() %>%
  mutate(
    VSORRES = as.character(round(rnorm(1, BASE_VALUE, SD_VALUE), 0)),
    VSSTRESN = as.numeric(VSORRES),
    VSSTRESC = VSORRES,
    VSORRESU = VSSTRESU,
    VSPOS = "SITTING",
    VSLOC = ifelse(VSTESTCD %in% c("SYSBP", "DIABP"), "RIGHT ARM", NA_character_)
  ) %>%
  ungroup() %>%
  # Add identifiers
  mutate(
    STUDYID = "CDISC01",
    VSTPT = paste("READING", READING),
    VSTPTNUM = READING
  ) %>%
  group_by(USUBJID) %>%
  mutate(VSSEQ = row_number()) %>%
  ungroup() %>%
  select(STUDYID, USUBJID, VSSEQ, VSTESTCD, VSTEST, 
         VSORRES, VSORRESU, VSSTRESC, VSSTRESN, VSSTRESU,
         VSPOS, VSLOC, VISITNUM, VISIT, VSTPT, VSTPTNUM)

cat("Sample VS data with multiple readings:\n")
cat("Total records:", nrow(vs_multiple), "\n")
cat("Subjects:", n_distinct(vs_multiple$USUBJID), "\n")
cat("Visits:", n_distinct(vs_multiple$VISIT), "\n\n")

# Show example for one subject at one visit
vs_multiple %>%
  filter(USUBJID == first(vs_multiple$USUBJID), VISIT == "BASELINE") %>%
  print()
```

::: {.callout-note collapse="true"}
## Understanding the Data Structure

Notice that for each subject at each visit, we have:

-   3 readings × 3 tests = **9 records per subject per visit**
-   Each reading is identified by `VSTPT` and `VSTPTNUM`
-   This granular data allows for various analysis approaches
:::

------------------------------------------------------------------------

## Deriving Statistics Across Readings

Now let's derive **mean**, **minimum**, and **maximum** values across the multiple readings.

### Calculate Summary Statistics

```{r}
#| label: derive-statistics

# Calculate statistics across readings for each subject/visit/test
vs_summary <- vs_multiple %>%
  group_by(STUDYID, USUBJID, VSTESTCD, VSTEST, VSSTRESU, 
           VSPOS, VSLOC, VISITNUM, VISIT) %>%
  summarise(
    N_READINGS = n(),
    VSSTRESN_MEAN = round(mean(VSSTRESN, na.rm = TRUE), 1),
    VSSTRESN_MIN = min(VSSTRESN, na.rm = TRUE),
    VSSTRESN_MAX = max(VSSTRESN, na.rm = TRUE),
    VSSTRESN_SD = round(sd(VSSTRESN, na.rm = TRUE), 2),
    .groups = "drop"
  )

cat("Summary statistics by subject/visit/test:\n")
vs_summary %>%
  filter(USUBJID == first(vs_summary$USUBJID)) %>%
  print()
```

### Create Analysis-Ready Dataset

When creating ADaM ADVS, we typically select **one representative value** from the multiple readings. Common approaches:

```{r}
#| label: analysis-ready

# Create analysis-ready dataset using MEAN of readings
vs_analysis <- vs_summary %>%
  mutate(
    AVAL = VSSTRESN_MEAN,
    AVALC = as.character(AVAL),
    DTYPE = "MEAN"  # Derivation type
  ) %>%
  select(STUDYID, USUBJID, VSTESTCD, VSTEST, 
         VISITNUM, VISIT, VSPOS, VSLOC,
         AVAL, AVALC, VSSTRESU, N_READINGS, DTYPE)

cat("Analysis-ready dataset (using mean values):\n")
head(vs_analysis, 12)
```

::: {.callout-important collapse="true"}
## ADaM Considerations

When building ADVS (Analysis Dataset for Vital Signs):

-   **AVAL** = Analysis Value (numeric)
-   **AVALC** = Analysis Value (character)
-   **DTYPE** = Derivation Type (documents how AVAL was calculated)

Common DTYPE values: "MEAN", "MEDIAN", "LAST", "FIRST", "WORST"
:::

------------------------------------------------------------------------

## Creating a Wide Format View

For some analyses, having one row per subject per visit (with BP columns side by side) is useful.

### Pivot to Wide Format

```{r}
#| label: pivot-wide

# Create wide format with one row per subject/visit
vs_wide <- vs_summary %>%
  select(USUBJID, VISITNUM, VISIT, VSTESTCD, VSSTRESN_MEAN) %>%
  pivot_wider(
    names_from = VSTESTCD,
    values_from = VSSTRESN_MEAN,
    names_prefix = "MEAN_"
  )

cat("Wide format - One row per subject per visit:\n")
print(vs_wide)
```

::: {.callout-tip collapse="true"}
## When to Use Wide Format

Wide format is useful for:

-   **Correlation analysis** (e.g., systolic vs. diastolic BP)
-   **Patient profiles** (view all vitals at once)
-   **Exploratory data analysis**

SDTM VS is always in **long format** (one row per measurement).
:::

------------------------------------------------------------------------

## Complete Example: Building a VS Domain

Let's build a complete VS domain from simulated raw data, similar to what you might receive from an EDC system.

### Step 1: Create Raw Vital Signs Data

```{r}
#| label: raw-vs-data

# Simulate raw vital signs data as it might come from EDC
set.seed(123)

subjects <- paste0("CDISC01-001-00", 1:5)
visits <- c("SCREENING", "BASELINE", "WEEK 2", "WEEK 4", "WEEK 8")
visit_nums <- c(-1, 1, 2, 3, 4)

# Generate comprehensive VS data
raw_vs <- expand_grid(
  USUBJID = subjects,
  VISIT_INFO = tibble(VISIT = visits, VISITNUM = visit_nums)
) %>%
  unnest(VISIT_INFO) %>%
  # Add vital sign tests with 3 readings each for BP, 1 for others
  crossing(
    tibble::tribble(
      ~TEST, ~TESTCD, ~UNIT,       ~MEAN_VAL, ~SD_VAL, ~N_READ,
      "Systolic Blood Pressure",  "SYSBP",  "mmHg",      125,     12,    3,
      "Diastolic Blood Pressure", "DIABP",  "mmHg",      78,       8,    3,
      "Pulse Rate",               "PULSE",  "beats/min", 72,       8,    1,
      "Temperature",              "TEMP",   "C",         36.8,   0.3,    1,
      "Respiratory Rate",         "RESP",   "breaths/min", 16,    2,    1,
      "Weight",                   "WEIGHT", "kg",        75,      10,    1,
      "Height",                   "HEIGHT", "cm",        170,     10,    1
    )
  ) %>%
  # Expand for number of readings
  rowwise() %>%
  mutate(data = list(tibble(READING = 1:N_READ))) %>%
  unnest(data) %>%
  ungroup() %>%
  # Generate values
  rowwise() %>%
  mutate(
    VALUE = round(rnorm(1, MEAN_VAL, SD_VAL), 1),
    # Height stays constant within subject
    VALUE = ifelse(TESTCD == "HEIGHT", round(rnorm(1, 170, 10), 0), VALUE),
    # Weight stays relatively constant
    VALUE = ifelse(TESTCD == "WEIGHT", round(rnorm(1, 75 + (VISITNUM * 0.5), 2), 1), VALUE)
  ) %>%
  ungroup() %>%
  # Clean up - HEIGHT only at SCREENING
  filter(!(TESTCD == "HEIGHT" & VISIT != "SCREENING")) %>%
  mutate(
    STUDYID = "CDISC01",
    VSPOS = case_when(
      TESTCD %in% c("SYSBP", "DIABP", "PULSE") ~ "SITTING",
      TRUE ~ NA_character_
    ),
    VSLOC = case_when(
      TESTCD %in% c("SYSBP", "DIABP") ~ "RIGHT ARM",
      TESTCD == "TEMP" ~ "ORAL",
      TRUE ~ NA_character_
    ),
    VSTPT = ifelse(N_READ > 1, paste("READING", READING), NA_character_),
    VSTPTNUM = ifelse(N_READ > 1, READING, NA_integer_)
  )

cat("Raw VS data generated:\n")
cat("Total records:", nrow(raw_vs), "\n")
```

### Step 2: Transform to SDTM VS Format

```{r}
#| label: create-sdtm-vs

# Transform to proper SDTM VS format
sdtm_vs <- raw_vs %>%
  # Add required SDTM variables
  mutate(
    DOMAIN = "VS",
    VSTESTCD = TESTCD,
    VSTEST = TEST,
    VSORRES = as.character(VALUE),
    VSORRESU = UNIT,
    VSSTRESC = VSORRES,
    VSSTRESN = VALUE,
    VSSTRESU = UNIT,
    VSDTC = "2023-03-15",  # Simplified for this example
    EPOCH = case_when(
      VISITNUM < 1 ~ "SCREENING",
      VISITNUM == 1 ~ "TREATMENT",
      TRUE ~ "TREATMENT"
    )
  ) %>%
  # Add sequence number
  group_by(USUBJID) %>%
  mutate(VSSEQ = row_number()) %>%
  ungroup() %>%
  # Select and order SDTM variables
  select(
    STUDYID, DOMAIN, USUBJID, VSSEQ,
    VSTESTCD, VSTEST,
    VSORRES, VSORRESU,
    VSSTRESC, VSSTRESN, VSSTRESU,
    VSPOS, VSLOC,
    VISITNUM, VISIT, VSTPT, VSTPTNUM,
    EPOCH, VSDTC
  ) %>%
  arrange(STUDYID, USUBJID, VISITNUM, VSTESTCD, VSTPTNUM)

cat("SDTM VS domain created:\n")
cat("Total records:", nrow(sdtm_vs), "\n\n")

# Preview
head(sdtm_vs, 15)
```

### Step 3: Validate the VS Domain

```{r}
#| label: validate-vs

# Validation checks
cat("=== VS Domain Validation ===\n\n")

# Check 1: Required variables
required_vars <- c("STUDYID", "DOMAIN", "USUBJID", "VSSEQ", 
                   "VSTESTCD", "VSTEST", "VSORRES", "VSORRESU")
present <- required_vars %in% names(sdtm_vs)
cat("Required variables check:\n")
for (i in seq_along(required_vars)) {
  cat("  ", required_vars[i], ":", ifelse(present[i], "✓", "✗"), "\n")
}

# Check 2: Data completeness
cat("\nData completeness:\n")
cat("  Missing VSSTRESN:", sum(is.na(sdtm_vs$VSSTRESN)), "\n")
cat("  Missing VSPOS (for BP/pulse):", 
    sum(is.na(sdtm_vs$VSPOS) & sdtm_vs$VSTESTCD %in% c("SYSBP", "DIABP", "PULSE")), "\n")

# Check 3: Value ranges
cat("\nValue ranges by test:\n")
sdtm_vs %>%
  group_by(VSTESTCD, VSTEST, VSSTRESU) %>%
  summarise(
    Min = min(VSSTRESN, na.rm = TRUE),
    Max = max(VSSTRESN, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  print()

# Check 4: Coverage
cat("\nData coverage:\n")
cat("  Subjects:", n_distinct(sdtm_vs$USUBJID), "\n")
cat("  Visits:", n_distinct(sdtm_vs$VISIT), "\n")
cat("  Tests:", n_distinct(sdtm_vs$VSTESTCD), "\n")
```

------------------------------------------------------------------------

## Preview: ADaM BDS Structure

The VS domain maps directly to the **ADaM ADVS** dataset using the **BDS (Basic Data Structure)** format. Here's a preview of how this transition works:

### Key ADaM Concepts for ADVS

| SDTM Variable | ADaM Variable | Description |
|---------------|---------------|-------------|
| VSSTRESN | AVAL | Analysis value (numeric) |
| VSSTRESC | AVALC | Analysis value (character) |
| VISIT | AVISIT | Analysis visit |
| VISITNUM | AVISITN | Analysis visit number |
| - | ABLFL | Baseline flag ("Y") |
| - | BASE | Baseline value |
| - | CHG | Change from baseline |
| - | PCHG | Percent change from baseline |

### Mock ADVS Creation

```{r}
#| label: mock-advs

# Create a simple ADVS preview
# First, aggregate to one value per subject/visit/test (using mean for BP)
advs_prep <- sdtm_vs %>%
  group_by(STUDYID, USUBJID, VSTESTCD, VSTEST, VSSTRESU, VISITNUM, VISIT) %>%
  summarise(
    AVAL = round(mean(VSSTRESN, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  rename(PARAMCD = VSTESTCD, PARAM = VSTEST, AVISITN = VISITNUM, AVISIT = VISIT)

# Add baseline flag and value
advs_preview <- advs_prep %>%
  group_by(USUBJID, PARAMCD) %>%
  mutate(
    # Flag baseline (VISITNUM = 1)
    ABLFL = ifelse(AVISITN == 1, "Y", NA_character_),
    # Get baseline value
    BASE = AVAL[AVISITN == 1][1],
    # Calculate change from baseline
    CHG = ifelse(AVISITN >= 1, round(AVAL - BASE, 1), NA_real_),
    # Calculate percent change
    PCHG = ifelse(AVISITN >= 1 & BASE != 0, 
                  round(100 * (AVAL - BASE) / BASE, 1), NA_real_)
  ) %>%
  ungroup()

cat("ADVS Preview (BDS Structure):\n")
advs_preview %>%
  filter(USUBJID == first(advs_preview$USUBJID), PARAMCD == "SYSBP") %>%
  select(USUBJID, PARAMCD, AVISITN, AVISIT, AVAL, ABLFL, BASE, CHG, PCHG) %>%
  print()
```

::: {.callout-note collapse="true"}
## Looking Ahead to Week 3

In Week 3 (Days 15-21), we'll use the `admiral` package to properly build ADaM datasets like ADVS. The key admiral functions include:

-   `derive_var_extreme_flag()` - Create baseline flags
-   `derive_var_base()` - Derive baseline values
-   `derive_vars_chg()` - Calculate change from baseline

Learn more at: [Admiral ADVS Vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html)
:::

------------------------------------------------------------------------

## Deliverable Summary

Today you completed the following:

| Task | Status |
|------|--------|
| Understood VS domain structure and repeated measures | ✓ Done |
| Explored positioning variables (VSPOS, VSLOC) | ✓ Done |
| Handled multiple readings per visit | ✓ Done |
| Derived summary statistics (mean, min, max) | ✓ Done |
| Created wide format view for analysis | ✓ Done |
| Built a complete VS domain with 5+ vital signs | ✓ Done |
| Previewed ADaM BDS structure (ADVS) | ✓ Done |

------------------------------------------------------------------------

## Key Takeaways

1.  **VS is a Findings class domain** - It shares structure with LB (--TESTCD, --ORRES, --STRESN)
2.  **Repeated measures are common** - Blood pressure often has 3 readings per visit
3.  **Position matters** - VSPOS (SITTING/STANDING) affects measurements significantly
4.  **Location matters** - VSLOC (LEFT ARM/RIGHT ARM) should be consistent per protocol
5.  **Summary statistics are essential** - Choose mean, min, max based on analysis needs
6.  **VS maps to ADVS** - BDS structure includes baseline, change, and percent change

------------------------------------------------------------------------

## Resources

-   [CDISC SDTM Implementation Guide - VS Domain](https://www.cdisc.org/standards/foundational/sdtm) - Official VS specification
-   [Admiral ADVS Vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html) - Building vital signs analysis datasets
-   [Pharmaverse.org](https://pharmaverse.org) - R packages for clinical data
-   [tidyr pivot documentation](https://tidyr.tidyverse.org/articles/pivot.html) - Wide/long format transformations
-   [CDISC ADaM BDS Structure](https://www.cdisc.org/standards/foundational/adam) - Basic Data Structure specification

------------------------------------------------------------------------

## What's Next?

In **Day 10**, we will focus on **AE Domain Mastery & SAE Logic**:

-   Deep dive into severity, causality, and outcomes
-   Understanding SAE criteria (AESER, AESDTH, AESHOSP)
-   Deriving AE duration and treatment-emergent flags
-   Preparing for ADaM ADAE creation

___
