---
title: "Day 29: ggsurvfit + gtsummary - Survival Plots and Clinical Figures"
subtitle: "Publication-ready TTE figures in pharmaverse"
format:
  html:
    theme:
      light: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css
editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---
::: {style="margin-bottom:12px;"}
<a href="../index.html" class="resource-link"> Back to Roadmap</a>
:::

## Overview

`visR` was archived from CRAN on 2024-06-22. The pharmaverse recommends two actively maintained replacements:

| Task                      | Package     | Status                       |
|---------------------------|-------------|------------------------------|
| KM curves + risk tables   | `ggsurvfit` | Active on CRAN + pharmaverse |
| Demographics / Cox tables | `gtsummary` | Active on CRAN + pharmaverse |

`ggsurvfit` is a proper **ggplot2 extension** -- every `add_*()` call is a real geom or stat, so any `ggplot2::theme()`, `labs()`, or scale function works directly without wrappers.

> **CNSR convention:** CDISC ADaM uses `CNSR = 1` for censored and `CNSR = 0` for event (opposite of base R). `Surv_CNSR(AVAL, CNSR)` from `ggsurvfit` handles this automatically.

> **Note:** `pharmaverseadam` does not ship an `adtte` dataset. This day derives Overall Survival from `pharmaverseadam::adsl` using `DTHDT`, `LSTALVDT`, `EOSDT`, and `TRTSDT` -- a standard clinical derivation.

------------------------------------------------------------------------

## Setup

```{r setup}
library(ggsurvfit)
library(gtsummary)
library(survival)
library(broom)
library(pharmaverseadam)
library(dplyr)
library(knitr)

adsl <- pharmaverseadam::adsl

# Inspect date and death columns confirmed in pharmaverseadam::adsl
date_cols <- grep("DT$|DTF$|ALV", names(adsl), value = TRUE)
cat("Date/death columns:", paste(date_cols, collapse = ", "), "\n")

# Derive Overall Survival ADTTE from ADSL:
#   AVAL = days from TRTSDT to DTHDT (event) or LSTALVDT / EOSDT (censored)
#   CNSR = 0 (death = event) | 1 (no death = censored)  [ADaM convention]
adtte <- adsl |>
  dplyr::filter(SAFFL == "Y") |>
  dplyr::mutate(
    TRTP = TRT01P,
    CNSR = dplyr::if_else(!is.na(DTHDT), 0L, 1L),
    AVAL = dplyr::case_when(
      !is.na(DTHDT)    ~ as.numeric(DTHDT    - TRTSDT),
      !is.na(LSTALVDT) ~ as.numeric(LSTALVDT - TRTSDT),
      !is.na(EOSDT)    ~ as.numeric(EOSDT    - TRTSDT),
      !is.na(TRTEDT)   ~ as.numeric(TRTEDT   - TRTSDT) + 1,
      TRUE             ~ NA_real_
    )
  ) |>
  dplyr::filter(!is.na(AVAL), AVAL >= 0)

cat("Derived ADTTE rows:", nrow(adtte), "\n")
cat("Events  (CNSR=0):", sum(adtte$CNSR == 0), "\n")
cat("Censored(CNSR=1):", sum(adtte$CNSR == 1), "\n")
cat("Median AVAL (days):", median(adtte$AVAL), "\n")
```

------------------------------------------------------------------------

## Part 1: Demographic Summary

### Step 1: gtsummary::tbl_summary()

```{r demo-table}
# tbl_summary(by = ...) stratifies by a column.
# add_overall() appends a combined Total column.
# as_kable() renders to a kable table compatible with Quarto HTML.
adtte |>
  dplyr::select(AGE, AGEGR1, SEX, RACE, TRTP) |>
  gtsummary::tbl_summary(
    by        = TRTP,
    label     = list(AGE ~ "Age (years)", AGEGR1 ~ "Age Group",
                     SEX ~ "Sex",         RACE   ~ "Race"),
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits  = all_continuous() ~ 1,
    missing = "no"
  ) |>
  gtsummary::add_overall() |>
  gtsummary::bold_labels() |>
  gtsummary::as_kable()
```

------------------------------------------------------------------------

## Part 2: Kaplan-Meier Plot

### Step 2: survfit2() + ggsurvfit()

```{r km-basic}
# survfit2() is ggsurvfit's survfit wrapper -- required for add_pvalue().
# It tracks the calling environment to correctly label plot elements.
# Surv_CNSR(AVAL, CNSR) converts ADaM CNSR coding to standard Surv() internally:
#   CNSR=1 (censored) -> status=0  |  CNSR=0 (event) -> status=1
# ggsurvfit uses '+' (ggplot2 extension), NOT '|>'.

km_fit <- survfit2(Surv_CNSR(AVAL, CNSR) ~ TRTP, data = adtte)
km_fit

km_fit |>
  ggsurvfit(linewidth = 1) +
  scale_ggsurvfit() +
  labs(
    title = "Kaplan-Meier Overall Survival",
    x     = "Time (Days)",
    y     = "Survival Probability"
  )
```

------------------------------------------------------------------------

### Step 3: add_confidence_interval() + add_censor_mark() + add_risktable()

```{r km-enhanced}
# add_risktable() attaches a risk table below the plot using patchwork.
# It MUST come last in the chain -- it wraps the ggplot and cannot be
# further modified by add_* calls after it.
# risktable_stats accepts: "n.risk", "cum.event", "cum.censor",
#   or glue-style strings like "{n.risk} ({cum.event})".
# stats_label renames the risk table row labels.

km_fit |>
  ggsurvfit(linewidth = 1) +
  add_confidence_interval() +
  add_censor_mark(shape = 3, size = 2) +
  scale_ggsurvfit() +
  labs(x = "Time (Days)", y = "Survival Probability",
       title = "Overall Survival with Risk Table") +
  add_risktable(
    risktable_stats = c("n.risk", "cum.event"),
    stats_label     = list(n.risk = "At Risk", cum.event = "Events")
  )
```

------------------------------------------------------------------------

## Part 3: Annotated KM Plot

### Step 4: add_pvalue() -- log-rank p-value on the figure

```{r km-pvalue}
# add_pvalue() calls survival::survdiff() log-rank test on km_fit.
# location = "annotation" places the p-value inside the plot panel.
# location = "caption"    places it in the figure caption below.
# survfit2() (not survfit()) is REQUIRED for add_pvalue() to work.

km_fit |>
  ggsurvfit(linewidth = 1) +
  add_confidence_interval() +
  add_censor_mark(shape = 3, size = 1.5) +
  add_pvalue(location = "annotation", prepend_p = TRUE) +
  scale_ggsurvfit() +
  labs(x = "Time (Days)", y = "Survival Probability",
       title = "Overall Survival with Log-Rank P-value")
```

------------------------------------------------------------------------

### Step 5: add_quantile() -- median survival line

```{r km-quantile}
# add_quantile(y_value = 0.5) draws lines at median survival (50th percentile).
# y_value is the survival probability level, NOT the time value.
# Must be called BEFORE add_risktable().

km_fit |>
  ggsurvfit(linewidth = 1) +
  add_confidence_interval() +
  add_censor_mark(shape = 3, size = 1.5) +
  add_quantile(
    y_value   = 0.5,
    color     = "grey40",
    linewidth = 0.75,
    linetype  = "dashed"
  ) +
  add_pvalue(location = "annotation", prepend_p = TRUE) +
  scale_ggsurvfit() +
  labs(x = "Time (Days)", y = "Survival Probability",
       title = "Median Survival with Annotation") +
  add_risktable(
    risktable_stats = "n.risk",
    stats_label     = list(n.risk = "At Risk")
  )
```

------------------------------------------------------------------------

## Part 4: KMunicate Theme

### Step 6: theme_ggsurvfit_KMunicate()

```{r km-kmunicate}
# theme_ggsurvfit_KMunicate() applies the KMunicate style recommended for
# transparent reporting of KM plots (Morris et al., BMJ Open 2019).
# It repositions risk table elements for integrated, publication-ready output.

km_fit |>
  ggsurvfit(linetype_aes = TRUE, linewidth = 0.9) +
  add_confidence_interval() +
  add_censor_mark(shape = 3, size = 1.5) +
  add_risktable(
    risktable_stats = c("n.risk", "cum.event"),
    stats_label     = list(n.risk = "At Risk", cum.event = "Cum. Events")
  ) +
  theme_ggsurvfit_KMunicate() +
  scale_ggsurvfit() +
  labs(x = "Time (Days)", y = "Survival Probability",
       title = "KMunicate-Style Overall Survival")
```

------------------------------------------------------------------------

## Part 5: Cox Proportional Hazards

### Step 7: coxph() + broom::tidy()

```{r cox-tidy}
# Surv_CNSR() works inside coxph() as well -- same ADaM convention handling.
# broom::tidy(exponentiate = TRUE) returns hazard ratios with 95% CI.

cox_fit <- survival::coxph(
  Surv_CNSR(AVAL, CNSR) ~ TRTP,
  data = adtte
)

broom::tidy(cox_fit, exponentiate = TRUE, conf.int = TRUE) |>
  dplyr::select(term, estimate, conf.low, conf.high, p.value) |>
  dplyr::rename(
    "Treatment" = term,
    "HR"        = estimate,
    "95% CI LB" = conf.low,
    "95% CI UB" = conf.high,
    "P-value"   = p.value
  ) |>
  knitr::kable(digits = 3, caption = "Cox PH - Hazard Ratios vs Reference")
```

------------------------------------------------------------------------

### Step 8: gtsummary::tbl_regression() -- formatted HR table

```{r cox-forest}
# tbl_regression(exponentiate = TRUE) renders a formatted HR table with CI.
# add_global_p() adds a global Wald test p-value for the treatment term.
# as_kable() converts to plain kable for Quarto HTML output.

cox_fit |>
  gtsummary::tbl_regression(
    exponentiate = TRUE,
    label        = list(TRTP ~ "Treatment")
  ) |>
  gtsummary::add_global_p() |>
  gtsummary::bold_p() |>
  gtsummary::bold_labels() |>
  gtsummary::as_kable()
```

------------------------------------------------------------------------

## Part 6: Accessing Survival Estimates

### Step 9: tidy_survfit() -- extract estimates as a data frame

```{r tidy-survfit}
# tidy_survfit() returns survival estimates as a tidy tibble.
# Useful for custom ggplot2 figures or programmatic summaries.

surv_df <- km_fit |> tidy_survfit()
cat("Columns:", paste(names(surv_df), collapse = ", "), "\n")
cat("Rows:", nrow(surv_df), "\n")

# First time each stratum crosses 50% survival (= median OS)
surv_df |>
  dplyr::group_by(strata) |>
  dplyr::filter(estimate <= 0.5) |>
  dplyr::slice_head(n = 1) |>
  dplyr::select(strata, time, estimate, conf.low, conf.high) |>
  knitr::kable(digits = 3, caption = "Median Overall Survival by Treatment")
```

------------------------------------------------------------------------

## Validation Checks

```{r validate}
cat("\n=== Day 29 Validation ===\n\n")
cat("Check 1 - adtte rows derived:", nrow(adtte), "\n")
cat("Check 2 - Events (CNSR=0):", sum(adtte$CNSR == 0), "\n")
cat("Check 3 - KM strata count:", length(km_fit$strata), "\n")
cat("Check 4 - tidy_survfit rows:", nrow(surv_df), "\n")
cat("Check 5 - Cox terms:", length(coef(cox_fit)), "\n")
cat("Check 6 - Log-rank p:", format(survfit2_p(km_fit), digits = 4), "\n")
cat("\nValidation complete\n")
```

------------------------------------------------------------------------

## Key Takeaways

1.  **`pharmaverseadam` has no `adtte`** -- derive Overall Survival from `adsl` using `DTHDT - TRTSDT` (event) and `LSTALVDT`/`EOSDT - TRTSDT` (censored)
2.  **`survfit2()` not `survfit()`** -- required for `add_pvalue()` to work; tracks the calling environment for the log-rank test
3.  **`Surv_CNSR(AVAL, CNSR)`** -- handles ADaM CNSR convention (1=censored, 0=event) automatically; works in both `survfit2()` and `coxph()`
4.  **`ggsurvfit` uses `+`** (ggplot2 extension), not `|>` -- any `ggplot2::theme()`, `labs()`, or scale function applies directly
5.  **`add_risktable()` uses `risktable_stats`** (not `statlist`) and `stats_label` (not `label`); must always be the last call in the chain
6.  **`add_quantile(y_value = 0.5)`** draws median survival lines; **`add_pvalue(location = "annotation")`** overlays the log-rank result
7.  **`tbl_summary()` + `tbl_regression(exponentiate = TRUE)`** from `gtsummary` replace `visR::tableone()` and Cox HR table wrappers
8.  **`tidy_survfit()`** exports the KM curve as a plain tibble for custom figures

------------------------------------------------------------------------

## Resources

-   ggsurvfit documentation: <https://www.danieldsjoberg.com/ggsurvfit/>
-   ggsurvfit gallery: <https://www.danieldsjoberg.com/ggsurvfit/articles/gallery.html>
-   gtsummary documentation: <https://www.danieldsjoberg.com/gtsummary/>
-   KMunicate style: Morris et al. (2019) BMJ Open <https://doi.org/10.1136/bmjopen-2019-030215>

------------------------------------------------------------------------

End of Day 29

<a href="https://visitorbadge.io/status?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse"><img src="https://api.visitorbadge.io/api/visitors?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse&label=Visits%20till%20date&countColor=%23263759&style=plastic&labelStyle=none" /></a>
