---
title: "Day 24: ARD-First Reporting with cards and cardx"
subtitle: "CDISC Analysis Results Data - computation decoupled from presentation"
format:
  html:
    theme:
      light: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css

editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

::: {style="margin-bottom:12px;"}
<a href="../index.html" class="resource-link"> Back to Roadmap</a>
:::

## Overview

The **Analysis Results Dataset (ARD)** is an emerging CDISC standard that stores statistical results in a tidy, machine-readable format - reusable across tables, figures, and reports without re-running analyses.

Today's packages:

-   `cards` - core ARD objects: continuous, categorical summaries
-   `cardx` - extended ARD: t-tests, chi-square, regression models
-   `pharmaverseadam` - reference ADaM datasets as input

------------------------------------------------------------------------

## Setup

```{r setup}
library(cards)
library(cardx)
library(pharmaverseadam)
library(dplyr)

# Safety population from ADSL
adsl_safe <- pharmaverseadam::adsl %>%
  dplyr::filter(SAFFL == "Y")

cat("Safety population N:", nrow(adsl_safe), "\n")
cat("Treatment arms:\n")
adsl_safe %>% dplyr::count(TRT01A)
```

------------------------------------------------------------------------

## Step 1: Confirm Available Variables

```{r check-vars}
# Always verify before use - BMIBL is NOT in pharmaverseadam::adsl
names(adsl_safe)
```

> **Note:** `pharmaverseadam::adsl` does not contain `BMIBL`. We use `AGE` as the only continuous variable.

------------------------------------------------------------------------

## Step 2: Continuous Summary ARD

```{r ard-continuous}
ard_cont <- cards::ard_continuous(
  adsl_safe,
  by        = TRT01A,
  variables = AGE,
  statistic = ~ cards::continuous_summary_fns(c("mean", "sd", "median", "p25", "p75", "min", "max"))
)

ard_cont %>%
  dplyr::select(group1, group1_level, variable, stat_name, stat_label, stat)
```

**ARD Structure:**

| Column         | Meaning                        |
|----------------|--------------------------------|
| `group1`       | Grouping variable (TRT01A)     |
| `group1_level` | Level of grouping variable     |
| `variable`     | Analysis variable (AGE)        |
| `stat_name`    | Statistic name (mean, sd, ...) |
| `stat_label`   | Human-readable label           |
| `stat`         | Computed value (list column)   |

------------------------------------------------------------------------

## Step 3: Categorical Summary ARD

```{r ard-categorical}
ard_cat <- cards::ard_categorical(
  adsl_safe,
  by        = TRT01A,
  variables = c(SEX, RACE, AGEGR1)
)

ard_cat %>%
  dplyr::select(group1, group1_level, variable, variable_level, stat_name, stat_label, stat) %>%
  head(12)
```

------------------------------------------------------------------------

## Step 4: Missing Values ARD

```{r ard-missing}
# Note: keep ard_missing separate - do not combine with ard_cont/cat
# (overlapping stat_names cause duplicates in bind_ard)
ard_miss <- cards::ard_missing(
  adsl_safe,
  by        = TRT01A,
  variables = c(AGE, SEX, RACE, AGEGR1, ETHNIC)
)

ard_miss %>%
  dplyr::select(group1, group1_level, variable, stat_name, stat_label, stat)
```

------------------------------------------------------------------------

## Step 5: T-test ARD (cardx)

```{r ard-ttest}
ard_t <- cardx::ard_stats_t_test(
  adsl_safe,
  by        = TRT01A,
  variables = AGE
)

ard_t %>%
  dplyr::select(group1, variable, stat_name, stat_label, stat)
```

------------------------------------------------------------------------

## Step 6: Chi-Square Test ARD (cardx)

```{r ard-chisq}
ard_chi <- cardx::ard_stats_chisq_test(
  adsl_safe,
  by        = TRT01A,
  variables = SEX
)

ard_chi %>%
  dplyr::select(group1, variable, stat_name, stat_label, stat)
```

------------------------------------------------------------------------

## Step 7: Wilcoxon Rank-Sum ARD (cardx)

```{r ard-wilcox}
ard_wilcox <- cardx::ard_stats_wilcox_test(
  adsl_safe,
  by        = TRT01A,
  variables = AGE
)

ard_wilcox %>%
  dplyr::select(group1, variable, stat_name, stat_label, stat)
```

------------------------------------------------------------------------

## Step 8: Combine Summary ARDs

```{r ard-combine}
# bind_ard() with .update = TRUE resolves duplicate stat_names across ARD objects
# Keep inferential (t-test, chisq, wilcox) separate from descriptive (cont, cat)
ard_descriptive <- cards::bind_ard(
  ard_cont,
  ard_cat,
  .update = TRUE
)

ard_inferential <- cards::bind_ard(
  ard_t,
  ard_chi,
  ard_wilcox,
  .update = TRUE
)

cat("Descriptive ARD rows:", nrow(ard_descriptive), "\n")
cat("Inferential ARD rows:", nrow(ard_inferential), "\n")

cat("\nDescriptive contexts:\n")
ard_descriptive %>% dplyr::count(context)

cat("\nInferential contexts:\n")
ard_inferential %>% dplyr::count(context)
```

------------------------------------------------------------------------

## Step 9: Inspect the Descriptive ARD

```{r ard-inspect}
ard_descriptive %>%
  dplyr::select(group1, group1_level, variable, context, stat_name, stat_label, stat) %>%
  head(20)
```

------------------------------------------------------------------------

## Step 10: Extract Specific Statistics

```{r ard-extract}
# Pull mean AGE per treatment arm
ard_descriptive %>%
  dplyr::filter(
    variable  == "AGE",
    stat_name == "mean",
    context   == "continuous"
  ) %>%
  dplyr::select(group1_level, stat_name, stat) %>%
  dplyr::mutate(stat = round(unlist(stat), 1))
```

------------------------------------------------------------------------

## Step 11: Format Statistics from ARD

```{r ard-format}
ard_formatted <- ard_descriptive %>%
  dplyr::filter(
    variable  == "AGE",
    context   == "continuous",
    stat_name %in% c("mean", "sd", "median", "p25", "p75")
  ) %>%
  cards::apply_fmt_fn()

ard_formatted %>%
  dplyr::select(group1_level, stat_name, stat_label, stat_fmt)
```

------------------------------------------------------------------------

## Step 12: ARD-Backed Summary Table

```{r ard-summary-table}
mean_age <- ard_descriptive %>%
  dplyr::filter(variable == "AGE", stat_name == "mean", context == "continuous") %>%
  dplyr::mutate(treatment = as.character(group1_level), mean_age = round(unlist(stat), 1)) %>%
  dplyr::select(treatment, mean_age)

sd_age <- ard_descriptive %>%
  dplyr::filter(variable == "AGE", stat_name == "sd", context == "continuous") %>%
  dplyr::mutate(treatment = as.character(group1_level), sd_age = round(unlist(stat), 1)) %>%
  dplyr::select(treatment, sd_age)

age_summary <- dplyr::left_join(mean_age, sd_age, by = "treatment") %>%
  dplyr::mutate(`Mean (SD)` = paste0(mean_age, " (", sd_age, ")")) %>%
  dplyr::select(Treatment = treatment, `Mean (SD)`)

age_summary
```

------------------------------------------------------------------------

## Step 13: ARD for Regression (cardx)

```{r ard-regression}
adsl_model <- adsl_safe %>%
  dplyr::mutate(SEX_BIN = dplyr::if_else(SEX == "M", 1L, 0L))

logit_mod <- glm(
  SEX_BIN ~ AGE + TRT01A,
  data   = adsl_model,
  family = binomial()
)

ard_logit <- cardx::ard_regression(
  logit_mod,
  exponentiate = TRUE
)

ard_logit %>%
  dplyr::select(variable, stat_name, stat_label, stat) %>%
  head(12)
```

------------------------------------------------------------------------

## Step 14: Validation Checks

```{r validate}
cat("\n=== ARD Validation ===\n\n")

# Check 1: No error records in descriptive ARD
check1 <- ard_descriptive %>%
  dplyr::filter(!sapply(error, is.null))
cat("Check 1 - Descriptive ARD error records:", nrow(check1), "\n")

# Check 2: AGE mean stat is numeric
mean_stat <- ard_descriptive %>%
  dplyr::filter(variable == "AGE", stat_name == "mean", context == "continuous") %>%
  dplyr::pull(stat)
check2 <- all(sapply(mean_stat, is.numeric))
cat("Check 2 - AGE mean is numeric:", check2, "\n")

# Check 3: SEX counts per treatment arm
cat_n <- ard_descriptive %>%
  dplyr::filter(variable == "SEX", stat_name == "n", context == "categorical") %>%
  dplyr::mutate(val = unlist(stat)) %>%
  dplyr::group_by(group1_level) %>%
  dplyr::summarise(total = sum(val), .groups = "drop")
cat("Check 3 - SEX counts per arm:\n")
print(cat_n)

cat("\n✓ ARD validation complete\n")
```

------------------------------------------------------------------------

## Summary

```{r summary}
cat("\n=== Day 24 Summary ===\n\n")

cat("Descriptive ARD contexts:\n")
ard_descriptive %>% dplyr::count(context) %>% print()

cat("\nInferential ARD contexts:\n")
ard_inferential %>% dplyr::count(context) %>% print()

cat("\nVariables covered: AGE, SEX, RACE, AGEGR1, ETHNIC\n")
cat("Tests: t-test, chi-square, Wilcoxon rank-sum\n")
cat("Regression: logistic (AGE + TRT01A)\n")
```

------------------------------------------------------------------------

## Export ARD as CSV

```{r export}
ard_export <- ard_descriptive %>%
  dplyr::mutate(
    stat_value     = sapply(stat, function(x) if (is.null(x)) NA_character_ else as.character(x)),
    group1_level   = as.character(group1_level),
    variable_level = sapply(variable_level, function(x) if (is.null(x)) NA_character_ else as.character(x))
  ) %>%
  dplyr::select(group1, group1_level, variable, variable_level,
                context, stat_name, stat_label, stat_value)

readr::write_csv(ard_export, "ard_day24.csv")
cat("✓ ard_day24.csv exported -", nrow(ard_export), "rows\n")
```

------------------------------------------------------------------------

## Key Takeaways

1.  `cards::ard_continuous()` and `ard_categorical()` build tidy ARDs from ADaM data
2.  `cardx` extends with inferential stats: t-test, chi-square, Wilcoxon, regression
3.  `bind_ard(.update = TRUE)` resolves duplicate stat_names across ARD objects
4.  Keep **descriptive** and **inferential** ARDs in separate `bind_ard()` calls to avoid duplicates
5.  ARD `stat` column is a **list column** - use `unlist()` or `apply_fmt_fn()` to extract values
6.  `pharmaverseadam::adsl` has no `BMIBL` - only `AGE` is available as continuous

------------------------------------------------------------------------

## Next Steps

**Day 25:** `gtsummary` + `tfrmt` - ARD-backed production tables **Day 26:** `flextable` + `officer` - Word/RTF clinical output **Day 27:** `r2rtf` - submission-ready RTF tables

------------------------------------------------------------------------

## Resources

-   cards CRAN: <https://cran.r-project.org/package=cards>
-   cardx documentation: <https://insightsengineering.github.io/cardx/>
-   CDISC ARS standard: <https://www.cdisc.org/standards/foundational/ars>
-   pharmaverse cardinal project: <https://github.com/pharmaverse/cardinal>

------------------------------------------------------------------------

**End of Day 24**
