---
title: "Day 26: flextable and officer - Word and RTF Clinical Tables"
subtitle: "Formatted TLF output to Word and RTF"


format:
  html:
    theme:
      light: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css

editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

::: {style="margin-bottom:12px;"}
<a href="../index.html" class="resource-link"> Back to Roadmap</a>
:::

## Overview

`flextable` is the most flexible R package for creating formatted tables in Word, RTF, HTML, and PowerPoint -- comparable in power to SAS PROC REPORT for clinical reporting. `officer` provides the Word document container.

Key rule: `body_add_flextable()` lives in the **flextable** package, not officer. Always call it as `flextable::body_add_flextable()`.

------------------------------------------------------------------------

## Setup

```{r setup}
library(flextable)
library(officer)
library(pharmaverseadam)
library(dplyr)
library(tidyr)

adsl <- pharmaverseadam::adsl
adae <- pharmaverseadam::adae
adlb <- pharmaverseadam::adlb

cat("ADSL columns:\n");       print(names(adsl))
cat("\nADAE columns (first 20):\n"); print(head(names(adae), 20))
cat("\nADLB columns (first 20):\n"); print(head(names(adlb), 20))
```

------------------------------------------------------------------------

## Table 1: Demographics

### Step 1: Build Summary Data

```{r demo-data}
adsl_safe <- adsl |> dplyr::filter(SAFFL == "Y")

demo_sum <- adsl_safe |>
  dplyr::group_by(TRT01A) |>
  dplyr::summarise(
    N          = dplyr::n(),
    AGE_MEAN   = sprintf("%.1f (%.2f)", mean(AGE, na.rm = TRUE), sd(AGE, na.rm = TRUE)),
    FEMALE_PCT = sprintf("%d (%.1f%%)", sum(SEX == "F", na.rm = TRUE),
                         mean(SEX == "F", na.rm = TRUE) * 100),
    .groups = "drop"
  )

print(demo_sum)
```

------------------------------------------------------------------------

### Step 2: Build flextable

```{r demo-ft}
ft_demo <- flextable::flextable(demo_sum) |>
  flextable::set_header_labels(
    TRT01A     = "Treatment",
    N          = "N",
    AGE_MEAN   = "Age, Mean (SD)",
    FEMALE_PCT = "Female, n (%)"
  ) |>
  flextable::theme_booktabs() |>
  flextable::align(align = "center", part = "all") |>
  flextable::align(j = 1, align = "left", part = "all") |>
  flextable::bold(part = "header") |>
  flextable::hline_top(border = officer::fp_border(width = 2), part = "header") |>
  flextable::hline_bottom(border = officer::fp_border(width = 2), part = "header") |>
  flextable::hline_bottom(border = officer::fp_border(width = 1.5), part = "body") |>
  flextable::add_header_lines("Table 14.1.1 -- Subject Demographics (Safety Population)") |>
  flextable::add_footer_lines("SD = standard deviation") |>
  flextable::autofit()

ft_demo
```

------------------------------------------------------------------------

### Step 3: Export Demographics to Word

```{r demo-docx}
# body_add_flextable() is in flextable, NOT officer
doc_demo <- officer::read_docx()
doc_demo <- officer::body_add_par(doc_demo,
                                   "Table 14.1.1 Demographics",
                                   style = "heading 1")
doc_demo <- flextable::body_add_flextable(doc_demo, value = ft_demo)
doc_demo <- officer::body_add_par(doc_demo, "", style = "Normal")
print(doc_demo, target = "table14_1_1_demographics.docx")
cat("Exported: table14_1_1_demographics.docx\n")
```

------------------------------------------------------------------------

### Step 4: Export Demographics to RTF

```{r demo-rtf}
flextable::save_as_rtf(ft_demo, path = "table14_1_1_demographics.rtf")
cat("Exported: table14_1_1_demographics.rtf\n")
```

------------------------------------------------------------------------

## Table 2: AE Frequency by SOC

### Step 5: Build AE Summary Data

```{r ae-data}
adae_te <- adae |>
  dplyr::filter(TRTEMFL == "Y", SAFFL == "Y")

n_per_arm <- adsl_safe |>
  dplyr::count(TRT01A, name = "ARM_N")

ae_sum <- adae_te |>
  dplyr::distinct(USUBJID, TRT01A, AEBODSYS, AEDECOD) |>
  dplyr::group_by(TRT01A, AEBODSYS, AEDECOD) |>
  dplyr::summarise(subj_n = dplyr::n_distinct(USUBJID), .groups = "drop") |>
  dplyr::left_join(n_per_arm, by = "TRT01A") |>
  dplyr::mutate(
    cell = sprintf("%d (%.1f%%)", subj_n, subj_n / ARM_N * 100)
  ) |>
  dplyr::select(AEBODSYS, AEDECOD, TRT01A, cell) |>
  tidyr::pivot_wider(names_from = TRT01A, values_from = cell,
                     values_fill = "0 (0.0%)") |>
  dplyr::arrange(AEBODSYS, AEDECOD) |>
  utils::head(15)

cat("AE summary rows:", nrow(ae_sum), "\n")
print(head(ae_sum, 3))
```

------------------------------------------------------------------------

### Step 6: Build AE flextable

```{r ae-ft}
ft_ae <- flextable::flextable(ae_sum) |>
  flextable::set_header_labels(
    AEBODSYS = "System Organ Class",
    AEDECOD  = "Preferred Term"
  ) |>
  flextable::theme_booktabs() |>
  flextable::bold(part = "header") |>
  flextable::align(align = "center", part = "all") |>
  flextable::align(j = c(1, 2), align = "left", part = "all") |>
  flextable::merge_v(j = "AEBODSYS") |>
  flextable::valign(j = "AEBODSYS", valign = "top", part = "body") |>
  flextable::hline_top(border = officer::fp_border(width = 2), part = "header") |>
  flextable::hline_bottom(border = officer::fp_border(width = 2), part = "header") |>
  flextable::hline_bottom(border = officer::fp_border(width = 1.5), part = "body") |>
  flextable::add_header_lines(
    "Table 14.3.1 -- Treatment-Emergent Adverse Events by SOC (Safety Population)"
  ) |>
  flextable::add_footer_lines(
    "n (%) = subjects with at least one event; TEAE = treatment-emergent adverse event"
  ) |>
  flextable::autofit()

ft_ae
```

------------------------------------------------------------------------

### Step 7: Export AE Table to Word

```{r ae-docx}
doc_ae <- officer::read_docx()
doc_ae <- officer::body_add_par(doc_ae, "Table 14.3.1 TEAEs by SOC",
                                 style = "heading 1")
doc_ae <- flextable::body_add_flextable(doc_ae, value = ft_ae)
doc_ae <- officer::body_add_par(doc_ae, "", style = "Normal")
print(doc_ae, target = "table14_3_1_ae.docx")
cat("Exported: table14_3_1_ae.docx\n")
```

------------------------------------------------------------------------

### Step 8: Export AE Table to RTF

```{r ae-rtf}
flextable::save_as_rtf(ft_ae, path = "table14_3_1_ae.rtf")
cat("Exported: table14_3_1_ae.rtf\n")
```

------------------------------------------------------------------------

## Table 3: Lab Shift Table

### Step 9: Build Lab Shift Data

```{r lab-data}
adlb_safe <- adlb |>
  dplyr::filter(SAFFL == "Y", PARAMCD == "ALT", !is.na(AVAL))

cat("ADLB indicator columns present:\n")
print(intersect(names(adlb_safe), c("BNRIND", "ANRIND", "ABLFL", "ANL01FL")))

shift_data <- adlb_safe |>
  dplyr::filter(!is.na(BNRIND), !is.na(ANRIND), ANL01FL == "Y") |>
  dplyr::count(TRT01A, BNRIND, ANRIND, name = "n") |>
  dplyr::mutate(
    BNRIND = factor(BNRIND, levels = c("LOW", "NORMAL", "HIGH")),
    ANRIND = factor(ANRIND, levels = c("LOW", "NORMAL", "HIGH"))
  ) |>
  tidyr::pivot_wider(
    names_from  = ANRIND,
    values_from = n,
    values_fill = 0L
  ) |>
  dplyr::arrange(TRT01A, BNRIND)

cat("Lab shift rows:", nrow(shift_data), "\n")
print(shift_data)
```

------------------------------------------------------------------------

### Step 10: Build Lab Shift flextable

```{r lab-ft}
ft_lab <- flextable::flextable(shift_data) |>
  flextable::set_header_labels(
    TRT01A = "Treatment",
    BNRIND = "Baseline"
  ) |>
  flextable::add_header_row(
    values     = c("", "", "Post-Baseline Reference Range"),
    colwidths  = c(1, 1, ncol(shift_data) - 2)
  ) |>
  flextable::theme_booktabs() |>
  flextable::bold(part = "header") |>
  flextable::align(align = "center", part = "all") |>
  flextable::align(j = c(1, 2), align = "left", part = "all") |>
  flextable::merge_v(j = "TRT01A") |>
  flextable::valign(j = "TRT01A", valign = "top", part = "body") |>
  flextable::hline_top(border = officer::fp_border(width = 2), part = "header") |>
  flextable::hline_bottom(border = officer::fp_border(width = 2), part = "header") |>
  flextable::hline_bottom(border = officer::fp_border(width = 1.5), part = "body") |>
  flextable::add_header_lines(
    "Table 14.4.1 -- ALT Shift Table: Baseline vs Post-Baseline (Safety Population)"
  ) |>
  flextable::add_footer_lines(
    "BNRIND = Baseline Reference Range Indicator; ANRIND = Post-Baseline Reference Range Indicator"
  ) |>
  flextable::autofit()

ft_lab
```

------------------------------------------------------------------------

### Step 11: Export Lab Shift to Word

```{r lab-docx}
doc_lab <- officer::read_docx()
doc_lab <- officer::body_add_par(doc_lab, "Table 14.4.1 ALT Shift Table",
                                  style = "heading 1")
doc_lab <- flextable::body_add_flextable(doc_lab, value = ft_lab)
doc_lab <- officer::body_add_par(doc_lab, "", style = "Normal")
print(doc_lab, target = "table14_4_1_lab_shift.docx")
cat("Exported: table14_4_1_lab_shift.docx\n")
```

------------------------------------------------------------------------

### Step 12: Export Lab Shift to RTF

```{r lab-rtf}
flextable::save_as_rtf(ft_lab, path = "table14_4_1_lab_shift.rtf")
cat("Exported: table14_4_1_lab_shift.rtf\n")
```

------------------------------------------------------------------------

## Step 13: Multi-Table Word Document

```{r multi-doc}
doc_all <- officer::read_docx()
doc_all <- officer::body_add_par(doc_all,
                                  "Table 14.1.1 Demographics",
                                  style = "heading 1")
doc_all <- flextable::body_add_flextable(doc_all, value = ft_demo)
doc_all <- officer::body_add_break(doc_all)
doc_all <- officer::body_add_par(doc_all,
                                  "Table 14.3.1 Treatment-Emergent Adverse Events",
                                  style = "heading 1")
doc_all <- flextable::body_add_flextable(doc_all, value = ft_ae)
doc_all <- officer::body_add_break(doc_all)
doc_all <- officer::body_add_par(doc_all,
                                  "Table 14.4.1 ALT Shift Table",
                                  style = "heading 1")
doc_all <- flextable::body_add_flextable(doc_all, value = ft_lab)
doc_all <- officer::body_add_par(doc_all, "", style = "Normal")
print(doc_all, target = "tlf_package_day26.docx")
cat("Exported: tlf_package_day26.docx (all 3 tables)\n")
```

------------------------------------------------------------------------

## Validation Checks

```{r validate}
cat("\n=== Day 26 Validation ===\n\n")

cat("Check 1 - Treatment arms in demo summary:\n")
print(demo_sum |> dplyr::select(TRT01A, N))

cat("\nCheck 2 - AE table columns:\n")
print(names(ae_sum))

cat("\nCheck 3 - Lab shift rows:", nrow(shift_data), "\n")

files_out <- c(
  "table14_1_1_demographics.docx",
  "table14_1_1_demographics.rtf",
  "table14_3_1_ae.docx",
  "table14_3_1_ae.rtf",
  "table14_4_1_lab_shift.docx",
  "table14_4_1_lab_shift.rtf",
  "tlf_package_day26.docx"
)

cat("\nCheck 4 - Output files:\n")
for (f in files_out) {
  cat(sprintf("  %s: %s\n", f,
              if (file.exists(f)) "EXISTS" else "MISSING"))
}

cat("\nValidation complete\n")
```

------------------------------------------------------------------------

## Key Takeaways

1.  `body_add_flextable()` is exported from **flextable**, not officer -- always use `flextable::body_add_flextable()`
2.  Use base assignment (`doc <- flextable::body_add_flextable(doc, ...)`) not pipe chains for officer documents
3.  `theme_booktabs()` produces the standard clinical table look
4.  `merge_v()` + `valign()` handles SOC grouping rows in AE tables
5.  `add_header_row()` creates column spanners for shift tables
6.  `save_as_rtf()` exports directly; for Word use `officer::read_docx()` + `print(target = ...)`

------------------------------------------------------------------------

## Next Steps

**Day 27:** `r2rtf` -- submission-ready RTF tables **Day 28:** `rtables` + `tern` -- structured clinical tables

------------------------------------------------------------------------

## Resources

-   flextable documentation: <https://davidgohel.github.io/flextable/>
-   officer CRAN: <https://cran.r-project.org/package=officer>
-   r4csr.org: <https://r4csr.org>

------------------------------------------------------------------------

**End of Day 26**
