---
title: "Day 23: ADCM and ADRS - Concomitant Meds and Oncology Response"
subtitle: "Week 4, Day 23: OCCDS period flags and RECIST 1.1 with admiralonco"


format:
  html:
    theme:
      light: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css

editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---
::: {style="margin-bottom:12px;"}
<a href="../index.html" class="resource-link"> Back to Roadmap</a>
:::

## Overview

Today covers two supporting ADaM datasets:

-   **ADCM** - Concomitant Medications (OCCDS) with pre/concomitant/prior period flags
-   **ADRS** - Oncology Response (OCCDS) with RECIST 1.1 Best Overall Response using `admiralonco`

> Note: `pharmaversesdtm` exports `rs_onco` (not `rs`) for oncology response data.

------------------------------------------------------------------------

## Setup

```{r setup}
library(admiral)
library(admiralonco)
library(pharmaversesdtm)
library(pharmaverseadam)
library(dplyr)
library(lubridate)
library(xportr)

# Load SDTM source data
cm     <- pharmaversesdtm::cm
rs_onco <- pharmaversesdtm::rs_onco   # oncology-specific RS domain

# Load ADSL for merging
adsl <- pharmaverseadam::adsl

cat("Loaded CM:", nrow(cm), "records\n")
cat("Loaded RS (oncology):", nrow(rs_onco), "records\n")
cat("Loaded ADSL:", nrow(adsl), "subjects\n")
```

------------------------------------------------------------------------

## ADCM - Concomitant Medications

### Step 1: Check CM Structure

```{r check-cm}
cm %>%
  dplyr::select(STUDYID, USUBJID, CMTRT, CMDECOD, CMSTDTC, CMENDTC) %>%
  head(5)
```

**Key CM Variables:**

-   CMTRT: Verbatim medication name
-   CMDECOD: Standardised medication name (WHODrug)
-   CMSTDTC / CMENDTC: Start / end dates (ISO 8601)

------------------------------------------------------------------------

### Step 2: Merge ADSL into CM

```{r adcm-merge}
adcm <- cm %>%
  admiral::derive_vars_merged(
    dataset_add = adsl,
    new_vars = exprs(TRTSDT, TRTEDT, TRT01A),
    by_vars = exprs(STUDYID, USUBJID)
  )

cat("ADCM after ADSL merge:", nrow(adcm), "records\n")
```

------------------------------------------------------------------------

### Step 3: Derive Start and End Dates

```{r adcm-dates}
adcm <- adcm %>%
  admiral::derive_vars_dt(
    new_vars_prefix = "AST",
    dtc = CMSTDTC,
    highest_imputation = "M"
  ) %>%
  admiral::derive_vars_dt(
    new_vars_prefix = "AEN",
    dtc = CMENDTC,
    highest_imputation = "M"
  )

cat("Dates derived\n")

adcm %>%
  dplyr::select(USUBJID, CMTRT, ASTDT, AENDT) %>%
  head(5)
```

------------------------------------------------------------------------

### Step 4: Derive Period Flags

```{r adcm-period-flags}
adcm <- adcm %>%
  dplyr::mutate(
    # Pre-treatment: started before TRTSDT and ongoing at TRTSDT
    PREFL = dplyr::case_when(
      !is.na(ASTDT) & ASTDT < TRTSDT &
        (is.na(AENDT) | AENDT >= TRTSDT) ~ "Y",
      TRUE ~ NA_character_
    ),
    # Concomitant: overlaps with treatment period
    CONFL = dplyr::case_when(
      !is.na(ASTDT) &
        ASTDT <= TRTEDT &
        (is.na(AENDT) | AENDT >= TRTSDT) ~ "Y",
      TRUE ~ NA_character_
    ),
    # Prior: ended strictly before treatment start
    PRVFL = dplyr::case_when(
      !is.na(AENDT) & AENDT < TRTSDT ~ "Y",
      TRUE ~ NA_character_
    )
  )

cat("Period flags derived\n")

adcm %>%
  dplyr::summarise(
    N_PREFL = sum(PREFL == "Y", na.rm = TRUE),
    N_CONFL = sum(CONFL == "Y", na.rm = TRUE),
    N_PRVFL = sum(PRVFL == "Y", na.rm = TRUE)
  )
```

**Period Flag Logic:**

| Flag  | Meaning                                  |
|-------|------------------------------------------|
| PREFL | Started before TRTSDT, ongoing at TRTSDT |
| CONFL | Overlapping with treatment window        |
| PRVFL | Ended strictly before TRTSDT             |

------------------------------------------------------------------------

### Step 5: Derive Analysis Sequence

```{r adcm-aseq}
adcm <- adcm %>%
  dplyr::arrange(USUBJID, ASTDT, CMSEQ) %>%
  dplyr::group_by(USUBJID) %>%
  dplyr::mutate(ASEQ = dplyr::row_number()) %>%
  dplyr::ungroup()

cat("ADCM records:", nrow(adcm), "\n")
```

------------------------------------------------------------------------

### Step 6: ADCM Validation

```{r adcm-validate}
cat("\n=== ADCM Validation ===\n\n")

check1 <- adcm %>% dplyr::filter(is.na(ASTDT))
cat("Check 1 - Missing ASTDT:", nrow(check1), "\n")

check2 <- adcm %>% dplyr::filter(!is.na(PREFL) & PREFL != "Y")
cat("Check 2 - Invalid PREFL values:", nrow(check2), "\n")

check3 <- adcm %>%
  dplyr::count(USUBJID, ASEQ) %>%
  dplyr::filter(n > 1)
cat("Check 3 - Duplicate ASEQ per subject:", nrow(check3), "\n")

cat("\n✓ ADCM validation complete\n")
```

------------------------------------------------------------------------

## ADRS - Oncology Tumor Response

### Step 7: Check RS Oncology Structure

```{r check-rs}
# pharmaversesdtm exports rs_onco (not rs)
names(rs_onco)

rs_onco %>%
  dplyr::select(USUBJID, RSTESTCD, RSSTRESC, VISIT, RSDTC) %>%
  head(5)
```

------------------------------------------------------------------------

### Step 8: Merge ADSL into RS

```{r adrs-base}
adrs <- rs_onco %>%
  admiral::derive_vars_merged(
    dataset_add = adsl,
    new_vars = exprs(RANDDT, TRTSDT, TRT01A),
    by_vars = exprs(STUDYID, USUBJID)
  ) %>%
  admiral::derive_vars_dt(
    new_vars_prefix = "A",
    dtc = RSDTC
  )

cat("ADRS base created:", nrow(adrs), "records\n")

adrs %>%
  dplyr::select(USUBJID, RSTESTCD, RSSTRESC, ADT, TRTSDT) %>%
  head(5)
```

------------------------------------------------------------------------

### Step 9: Derive AVAL from Response

```{r adrs-aval}
# admiralonco::aval_resp() maps RECIST 1.1 labels to numeric values
adrs <- adrs %>%
  dplyr::mutate(
    PARAMCD = RSTESTCD,
    PARAM   = RSTEST,
    AVALC   = RSSTRESC,
    AVAL    = admiralonco::aval_resp(RSSTRESC)
  )

cat("AVAL derived\n")

adrs %>%
  dplyr::count(AVALC, AVAL) %>%
  dplyr::arrange(AVAL)
```

**RECIST 1.1 Response Coding:**

| AVALC | AVAL |
|-------|------|
| CR    | 1    |
| PR    | 2    |
| SD    | 3    |
| PD    | 4    |
| NE    | 5    |

------------------------------------------------------------------------

### Step 10: Filter to Overall Visit Response

```{r adrs-ovr}
# Filter to overall response records and rename parameter
adrs_ovr <- adrs %>%
  dplyr::filter(RSTESTCD == "OVRLRESP") %>%
  dplyr::mutate(
    PARAMCD = "OVR",
    PARAM   = "Overall Response by Visit"
  )

cat("OVR records:", nrow(adrs_ovr), "\n")

adrs_ovr %>%
  dplyr::count(AVALC, AVAL)
```

------------------------------------------------------------------------

### Step 11: Derive Best Overall Response

```{r adrs-bor}
# BOR = best (lowest AVAL) response across all visits per subject
adrs_bor <- adrs_ovr %>%
  dplyr::group_by(STUDYID, USUBJID) %>%
  dplyr::slice_min(AVAL, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    PARAMCD = "BOR",
    PARAM   = "Best Overall Response"
  )

cat("BOR records:", nrow(adrs_bor), "\n")

adrs_bor %>%
  dplyr::count(AVALC, AVAL)
```

------------------------------------------------------------------------

### Step 12: Derive Responder Flag

```{r adrs-rsp}
# Responder = CR (AVAL=1) or PR (AVAL=2)
adrs_rsp <- adrs_bor %>%
  dplyr::mutate(
    PARAMCD = "RSP",
    PARAM   = "Responder (CR or PR)",
    AVALC   = dplyr::if_else(AVAL <= 2, "Y", "N"),
    AVAL    = dplyr::if_else(AVALC == "Y", 1, 0)
  )

cat("RSP records:", nrow(adrs_rsp), "\n")

adrs_rsp %>%
  dplyr::count(AVALC, AVAL)
```

------------------------------------------------------------------------

### Step 13: Combine ADRS Parameters

```{r adrs-combine}
adrs_final <- dplyr::bind_rows(adrs_ovr, adrs_bor, adrs_rsp) %>%
  dplyr::arrange(USUBJID, PARAMCD, ADT) %>%
  dplyr::group_by(USUBJID) %>%
  dplyr::mutate(ASEQ = dplyr::row_number()) %>%
  dplyr::ungroup()

cat("ADRS final:", nrow(adrs_final), "records\n")

adrs_final %>%
  dplyr::count(PARAMCD, PARAM)
```

------------------------------------------------------------------------

### Step 14: ADRS Validation

```{r adrs-validate}
cat("\n=== ADRS Validation ===\n\n")

# One BOR per subject
check1 <- adrs_final %>%
  dplyr::filter(PARAMCD == "BOR") %>%
  dplyr::count(USUBJID) %>%
  dplyr::filter(n > 1)
cat("Check 1 - Multiple BOR per subject:", nrow(check1), "\n")

# Valid AVAL range for OVR (1-5)
check2 <- adrs_final %>%
  dplyr::filter(PARAMCD == "OVR", !AVAL %in% 1:5)
cat("Check 2 - Invalid OVR AVAL:", nrow(check2), "\n")

# RSP only 0 or 1
check3 <- adrs_final %>%
  dplyr::filter(PARAMCD == "RSP", !AVAL %in% c(0, 1))
cat("Check 3 - Invalid RSP AVAL:", nrow(check3), "\n")

cat("\n✓ ADRS validation complete\n")
```

------------------------------------------------------------------------

## Summary

```{r summary}
cat("\n=== Day 23 Summary ===\n\n")

cat("ADCM:\n")
cat("  Records:", nrow(adcm), "\n")
cat("  Concomitant (CONFL=Y):", sum(adcm$CONFL == "Y", na.rm = TRUE), "\n")
cat("  Prior (PRVFL=Y):", sum(adcm$PRVFL == "Y", na.rm = TRUE), "\n\n")

cat("ADRS:\n")
adrs_final %>%
  dplyr::count(PARAMCD, PARAM)
```

------------------------------------------------------------------------

## Export

```{r export}
attr(adcm$ASTDT, "label") <- "Analysis Start Date"
attr(adcm$AENDT, "label") <- "Analysis End Date"
attr(adcm$PREFL, "label") <- "Pre-treatment Flag"
attr(adcm$CONFL, "label") <- "Concomitant Flag"
attr(adcm$PRVFL, "label") <- "Prior Flag"
attr(adcm$ASEQ,  "label") <- "Analysis Sequence Number"

attr(adrs_final$PARAMCD, "label") <- "Parameter Code"
attr(adrs_final$PARAM,   "label") <- "Parameter"
attr(adrs_final$AVAL,    "label") <- "Analysis Value"
attr(adrs_final$AVALC,   "label") <- "Analysis Value (C)"
attr(adrs_final$ADT,     "label") <- "Analysis Date"
attr(adrs_final$ASEQ,    "label") <- "Analysis Sequence Number"

xportr_write(adcm,       path = "adcm.xpt", domain = "ADCM")
xportr_write(adrs_final, path = "adrs.xpt", domain = "ADRS")

cat("\n✓ adcm.xpt exported\n")
cat("✓ adrs.xpt exported\n")
```

------------------------------------------------------------------------

## Key Takeaways

1.  `pharmaversesdtm` exports `rs_onco` for oncology response - not `rs`
2.  ADCM period flags describe timing relative to TRTSDT/TRTEDT
3.  `admiralonco::aval_resp()` converts RECIST labels to numeric AVAL
4.  BOR = `slice_min(AVAL)` per subject across OVR records
5.  RSP = 1 if BOR is CR or PR (AVAL ≤ 2), else 0

------------------------------------------------------------------------

## Next Steps

**Day 24:** ARD-first reporting with `cards` and `cardx` **Day 25:** `gtsummary` + `tfrmt` ARD-backed tables **Day 26:** `flextable` + `officer` Word/RTF output

------------------------------------------------------------------------

## Resources

-   admiral ADCM vignette: <https://pharmaverse.github.io/admiral/articles/occds.html>
-   admiralonco ADRS vignette: <https://pharmaverse.github.io/admiralonco/articles/adrs.html>
-   pharmaversesdtm datasets: <https://pharmaverse.github.io/pharmaversesdtm/>

------------------------------------------------------------------------

**End of Day 23**

___
