---
title: "Day 27: rtables, tern, and r2rtf - Structured Clinical Tables"
subtitle: "Declarative table layout and RTF output"

format:
  html:
    theme:
      light: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css
editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---
::: {style="margin-bottom:12px;"}
< a target="_blank" href="../index.html" class="resource-link"> Back to Roadmap</a>
:::

## Overview

`rtables` provides a declarative layout engine for clinical tables. `tern` adds pre-built clinical analysis functions on top of `rtables`. `r2rtf` produces RTF output suitable for clinical reporting.

------------------------------------------------------------------------

## Setup

```{r setup}
library(rtables)
library(tern)
library(r2rtf)
library(pharmaverseadam)
library(dplyr)
library(tidyr)

adsl <- pharmaverseadam::adsl
adae  <- pharmaverseadam::adae
adlb  <- pharmaverseadam::adlb

# Factor all grouping variables -- rtables requires identical levels per column
adsl_safe <- adsl |>
  dplyr::filter(SAFFL == "Y") |>
  dplyr::mutate(
    TRT01A = factor(TRT01A),
    SEX    = factor(SEX),
    RACE   = factor(RACE),
    AGEGR1 = factor(AGEGR1)
  )

adae_safe <- adae |>
  dplyr::filter(SAFFL == "Y", TRTEMFL == "Y") |>
  dplyr::mutate(
    TRT01A   = factor(TRT01A,   levels = levels(adsl_safe$TRT01A)),
    AEBODSYS = factor(AEBODSYS),
    AEDECOD  = factor(AEDECOD)
  )

adlb_safe <- adlb |>
  dplyr::filter(SAFFL == "Y", !is.na(AVAL)) |>
  dplyr::mutate(
    TRT01A = factor(TRT01A, levels = levels(adsl_safe$TRT01A))
  )

cat("ADSL safety N:", nrow(adsl_safe), "\n")
cat("Treatment arms:", levels(adsl_safe$TRT01A), "\n")
```

> **Key rules:** 1. **Factor** all categorical variables before `build_table()`. 2. Add **`split_fun = drop_split_levels`** to every `split_rows_by()` call. 3. **Pre-order factor levels** in dplyr for sorting -- `sort_at_path()` with `cont_n_allcols` requires content rows that `count_occurrences()` does not produce. 4. **`rtf_encode()` only accepts `doc_type = "table"` or `"figure"`** -- use `assemble_rtf()` to merge multiple RTF files into one document.

------------------------------------------------------------------------

## Part 1: rtables

### Step 1: Minimal rtables table

```{r rtables-concepts}
afun_mean_sd <- function(x, ...) {
  list(
    "Mean (SD)" = rtables::rcell(
      c(mean(x, na.rm = TRUE), sd(x, na.rm = TRUE)),
      format = "xx.x (xx.xx)"
    )
  )
}

lyt_simple <- rtables::basic_table() |>
  rtables::split_cols_by("TRT01A") |>
  rtables::add_colcounts() |>
  rtables::analyze("AGE", afun = afun_mean_sd, var_labels = "Age (years)")

tbl_simple <- rtables::build_table(lyt_simple, adsl_safe)
print(tbl_simple)
```

------------------------------------------------------------------------

### Step 2: Demographics table with rtables

```{r rtables-demo}
# Pre-format multi-value cells with sprintf() -- rcell() accepts plain character
afun_age <- function(x, ...) {
  rtables::in_rows(
    "Mean (SD)"       = rtables::rcell(
      sprintf("%.1f (%.2f)", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE))
    ),
    "Median (Q1, Q3)" = rtables::rcell(
      sprintf("%.1f (%.1f, %.1f)",
              median(x, na.rm = TRUE),
              stats::quantile(x, 0.25, na.rm = TRUE),
              stats::quantile(x, 0.75, na.rm = TRUE))
    ),
    "Min, Max"        = rtables::rcell(
      sprintf("%.1f, %.1f", min(x, na.rm = TRUE), max(x, na.rm = TRUE))
    )
  )
}

# Closure captures all factor levels so every column emits identical rows
make_afun_cat <- function(var_levels) {
  function(x, ...) {
    tab <- table(factor(x, levels = var_levels), useNA = "no")
    tot <- sum(tab)
    rtables::in_rows(
      .list  = lapply(var_levels, function(v) {
        rtables::rcell(
          sprintf("%d (%.1f%%)", tab[[v]], tab[[v]] / tot * 100)
        )
      }),
      .names = var_levels
    )
  }
}

sex_levels  <- levels(adsl_safe$SEX)
race_levels <- levels(adsl_safe$RACE)

lyt_demo <- rtables::basic_table(show_colcounts = TRUE) |>
  rtables::split_cols_by("TRT01A") |>
  rtables::add_overall_col("All Subjects") |>
  rtables::analyze("AGE",  afun = afun_age,  var_labels = "Age (years)") |>
  rtables::analyze("SEX",  afun = make_afun_cat(sex_levels),  var_labels = "Sex") |>
  rtables::analyze("RACE", afun = make_afun_cat(race_levels), var_labels = "Race")

tbl_demo_rt <- rtables::build_table(lyt_demo, adsl_safe)
print(tbl_demo_rt)
```

------------------------------------------------------------------------

## Part 2: tern

### Step 3: Demographics table with tern

```{r tern-demo}
# analyze_vars() is the correct exported tern function for variable summaries
lyt_tern_demo <- rtables::basic_table(show_colcounts = TRUE) |>
  rtables::split_cols_by("TRT01A") |>
  rtables::add_overall_col("All Subjects") |>
  tern::analyze_vars(
    vars       = c("AGE", "SEX", "RACE"),
    var_labels = c("Age (years)", "Sex", "Race")
  )

tbl_tern_demo <- rtables::build_table(lyt_tern_demo, adsl_safe)
print(tbl_tern_demo)
```

------------------------------------------------------------------------

### Step 4: AE frequency table with pre-sorted factor levels

```{r tern-ae}
# Pre-order AEBODSYS and AEDECOD by descending frequency via dplyr.
# This is the reliable alternative to sort_at_path(cont_n_allcols),
# which needs content rows that count_occurrences() does not produce.
soc_order <- adae_safe |>
  dplyr::distinct(USUBJID, AEBODSYS) |>
  dplyr::count(AEBODSYS, sort = TRUE) |>
  dplyr::pull(AEBODSYS) |>
  as.character()

pt_order <- adae_safe |>
  dplyr::distinct(USUBJID, AEBODSYS, AEDECOD) |>
  dplyr::count(AEBODSYS, AEDECOD, sort = TRUE) |>
  dplyr::pull(AEDECOD) |>
  as.character() |>
  unique()

adae_sorted <- adae_safe |>
  dplyr::mutate(
    AEBODSYS = factor(AEBODSYS, levels = soc_order),
    AEDECOD  = factor(AEDECOD,  levels = pt_order)
  )

lyt_ae <- rtables::basic_table(show_colcounts = TRUE) |>
  rtables::split_cols_by("TRT01A") |>
  rtables::add_overall_col("All Subjects") |>
  rtables::split_rows_by(
    "AEBODSYS",
    label_pos   = "topleft",
    split_label = "System Organ Class",
    split_fun   = drop_split_levels
  ) |>
  tern::count_occurrences(vars = "AEDECOD", .indent_mods = 1L)

tbl_ae <- rtables::build_table(
  lyt_ae,
  df            = adae_sorted,
  alt_counts_df = adsl_safe
)
tbl_ae_pruned <- rtables::prune_table(tbl_ae)
print(tbl_ae_pruned)
```

------------------------------------------------------------------------

### Step 5: Lab shift table with tern

```{r tern-lab}
adlb_alt <- adlb_safe |>
  dplyr::filter(PARAMCD == "ALT", ANL01FL == "Y",
                !is.na(BNRIND), !is.na(ANRIND)) |>
  dplyr::mutate(
    BNRIND = factor(BNRIND, levels = c("LOW", "NORMAL", "HIGH")),
    ANRIND = factor(ANRIND, levels = c("LOW", "NORMAL", "HIGH"))
  )

cat("ALT records for shift table:", nrow(adlb_alt), "\n")

# split_fun = drop_split_levels silently removes levels absent from data
lyt_shift <- rtables::basic_table(show_colcounts = TRUE) |>
  rtables::split_cols_by("TRT01A") |>
  rtables::add_overall_col("All Subjects") |>
  rtables::split_rows_by(
    "BNRIND",
    label_pos   = "topleft",
    split_label = "Baseline Reference Range",
    split_fun   = drop_split_levels
  ) |>
  tern::count_occurrences(vars = "ANRIND", .indent_mods = 1L)

tbl_shift <- rtables::build_table(
  lyt_shift,
  df            = adlb_alt,
  alt_counts_df = adsl_safe
)
print(tbl_shift)
```

------------------------------------------------------------------------

## Part 3: r2rtf

### Step 6: Prepare data frames for RTF

```{r r2rtf-data}
demo_df <- adsl_safe |>
  dplyr::group_by(TRT01A) |>
  dplyr::summarise(
    N        = dplyr::n(),
    age_mean = mean(AGE, na.rm = TRUE),
    age_sd   = sd(AGE, na.rm = TRUE),
    female_n = sum(SEX == "F", na.rm = TRUE),
    .groups  = "drop"
  ) |>
  dplyr::mutate(
    `Age, Mean (SD)` = sprintf("%.1f (%.2f)", age_mean, age_sd),
    `Female, n (%)`  = sprintf("%d (%.1f%%)", female_n, female_n / N * 100)
  ) |>
  dplyr::select(
    Treatment = TRT01A, N, `Age, Mean (SD)`, `Female, n (%)`
  )
print(demo_df)
```

------------------------------------------------------------------------

### Step 7: Write an RTF demographics table

```{r r2rtf-demo}
demo_df |>
  r2rtf::rtf_page(orientation = "portrait") |>
  r2rtf::rtf_title(
    title    = "Table 14.1.1",
    subtitle = "Demographic and Baseline Characteristics (Safety Population)"
  ) |>
  r2rtf::rtf_colheader(
    colheader     = "Treatment | N | Age, Mean (SD) | Female, n (%)",
    col_rel_width = c(3, 1, 2, 2)
  ) |>
  r2rtf::rtf_body(
    col_rel_width      = c(3, 1, 2, 2),
    text_justification = c("l", "c", "c", "c")
  ) |>
  r2rtf::rtf_footnote(footnote = "SD = standard deviation.") |>
  r2rtf::rtf_encode() |>
  r2rtf::write_rtf("table14_1_1_r2rtf.rtf")
cat("Exported: table14_1_1_r2rtf.rtf\n")
```

------------------------------------------------------------------------

### Step 8: Combine two RTF files with assemble_rtf()

```{r r2rtf-multipage}
# rtf_encode() only accepts doc_type = "table" or "figure".
# To produce a single combined RTF, write each table separately then
# use r2rtf::assemble_rtf() to merge them into one file.

arm_names <- levels(adsl_safe$TRT01A)

ae_df <- adae_safe |>
  dplyr::count(TRT01A, AEBODSYS, AEDECOD, name = "n") |>
  dplyr::left_join(
    adsl_safe |> dplyr::count(TRT01A, name = "N"),
    by = "TRT01A"
  ) |>
  dplyr::mutate(cell = sprintf("%d (%.1f%%)", n, n / N * 100)) |>
  dplyr::select(AEBODSYS, AEDECOD, TRT01A, cell) |>
  tidyr::pivot_wider(
    names_from  = TRT01A,
    values_from = cell,
    values_fill = "0 (0.0%)"
  ) |>
  dplyr::arrange(AEBODSYS, AEDECOD) |>
  dplyr::slice_head(n = 12)

col_header <- paste(
  c("System Organ Class", "Preferred Term", arm_names),
  collapse = " | "
)

# Write table 1 (portrait)
file1 <- tempfile(fileext = ".rtf")
demo_df |>
  r2rtf::rtf_page(orientation = "portrait") |>
  r2rtf::rtf_title(title = "Table 14.1.1", subtitle = "Demographics") |>
  r2rtf::rtf_colheader(
    colheader     = "Treatment | N | Age, Mean (SD) | Female, n (%)",
    col_rel_width = c(3, 1, 2, 2)
  ) |>
  r2rtf::rtf_body(
    col_rel_width      = c(3, 1, 2, 2),
    text_justification = c("l", "c", "c", "c")
  ) |>
  r2rtf::rtf_encode() |>
  r2rtf::write_rtf(file1)

# Write table 2 (landscape)
file2 <- tempfile(fileext = ".rtf")
ae_df |>
  r2rtf::rtf_page(orientation = "landscape") |>
  r2rtf::rtf_title(title = "Table 14.3.1",
                   subtitle = "TEAEs by SOC and Preferred Term") |>
  r2rtf::rtf_colheader(
    colheader     = col_header,
    col_rel_width = c(3, 3, rep(2, length(arm_names)))
  ) |>
  r2rtf::rtf_body(
    col_rel_width      = c(3, 3, rep(2, length(arm_names))),
    text_justification = c("l", "l", rep("c", length(arm_names)))
  ) |>
  r2rtf::rtf_encode() |>
  r2rtf::write_rtf(file2)

# Assemble into one combined RTF file
r2rtf::assemble_rtf(
  input  = c(file1, file2),
  output = "tlf_package_day27.rtf"
)
cat("Exported: tlf_package_day27.rtf\n")
```

------------------------------------------------------------------------

## Validation Checks

```{r validate}
cat("\n=== Day 27 Validation ===\n\n")
cat("Check 1 - rtables simple table:",    class(tbl_simple),    "\n")
cat("Check 2 - tern demo table:",         class(tbl_tern_demo), "\n")
cat("Check 3 - AE pruned rows:",          nrow(tbl_ae_pruned),  "\n")
cat("Check 4 - SOC order (top 3):",       soc_order[1:3],       "\n")
files_out <- c("table14_1_1_r2rtf.rtf", "tlf_package_day27.rtf")
cat("\nOutput files:\n")
for (f in files_out) {
  cat(sprintf("  %s: %s\n", f, if (file.exists(f)) "EXISTS" else "MISSING"))
}
cat("\nValidation complete\n")
```

------------------------------------------------------------------------

## Key Takeaways

1.  **Factor all categorical variables** before `build_table()` -- rtables requires identical levels per column
2.  **`split_fun = drop_split_levels`** on every `split_rows_by()` -- silently drops factor levels absent from data
3.  **Pre-order factor levels via dplyr** for sorted output -- `sort_at_path(cont_n_allcols)` requires content rows that `count_occurrences()` does not produce
4.  **`rcell()` with `sprintf()`** avoids the format-label validator for multi-value cells
5.  **`tern::analyze_vars()`** is the correct exported demographics function (not `summarize_vars()`)
6.  **`rtf_encode()` only accepts `doc_type = "table"` or `"figure"`** -- use `assemble_rtf(input, output)` to merge separate RTF files into one document

------------------------------------------------------------------------

## Resources

-   rtables: <https://insightsengineering.github.io/rtables/>
-   tern: <https://insightsengineering.github.io/tern/>
-   r2rtf: <https://merck.github.io/r2rtf/>
-   r4csr book: <https://r4csr.org>

------------------------------------------------------------------------

End of Day 27

< a target="_blank" href="https://visitorbadge.io/status?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse"><img src="https://api.visitorbadge.io/api/visitors?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse&label=Visits%20till%20date&countColor=%23263759&style=plastic&labelStyle=none" /></a>
