---
title: "Day 22: Demography Table with gtsummary + gt"
subtitle: "First TLF from ADSL"


format:
  html:
    theme:
      light: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css

editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---
::: {style="margin-bottom:12px;"}
<a href="../index.html" class="resource-link"> Back to Roadmap</a>
:::

## Welcome to Week 4: TLFs

Week 4 moves from **building ADaMs** to **producing TLFs**.

Today we build two classic tables using `pharmaverseadam::adsl`:

-   **Table 14.1.1** Demography and Baseline Characteristics
-   **Table 14.2.1** Disposition Summary

------------------------------------------------------------------------

## Setup

```{r setup}
library(pharmaverseadam)
library(dplyr)
library(gtsummary)
library(gt)

adsl <- pharmaverseadam::adsl

cat("Loaded ADSL:", nrow(adsl), "subjects\n")
cat("Variables:", ncol(adsl), "\n")
```

------------------------------------------------------------------------

## Step 1: Check Available Columns

```{r check-cols}
# Always check what exists before selecting
names(adsl)
```

------------------------------------------------------------------------

## Step 2: Define Analysis Population

```{r analysis-population}
adsl_saf <- adsl %>%
  dplyr::filter(SAFFL == "Y")

cat("Safety population:", nrow(adsl_saf), "subjects\n")

adsl_saf %>%
  dplyr::count(TRT01A)
```

------------------------------------------------------------------------

## Step 3: Select Variables for Table 1

```{r select-vars}
# Only columns confirmed in pharmaverseadam::adsl
adsl_tab <- adsl_saf %>%
  dplyr::select(TRT01A, AGE, SEX, RACE, ETHNIC, AGEGR1)

adsl_tab %>%
  head(5)
```

Variables used:

-   AGE (continuous)
-   SEX, RACE, ETHNIC, AGEGR1 (categorical)

------------------------------------------------------------------------

## Step 4: Build Summary Table with gtsummary

```{r build-gtsummary}
var_types <- list(
  AGE    ~ "continuous2",
  SEX    ~ "categorical",
  RACE   ~ "categorical",
  ETHNIC ~ "categorical",
  AGEGR1 ~ "categorical"
)

tbl_demo <- adsl_tab %>%
  gtsummary::tbl_summary(
    by = TRT01A,
    type = var_types,
    statistic = list(
      gtsummary::all_continuous()  ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
      gtsummary::all_categorical() ~ "{n} ({p}%)"
    ),
    digits  = gtsummary::all_continuous() ~ 1,
    missing = "no"
  ) %>%
  gtsummary::add_overall(last = TRUE) %>%
  gtsummary::add_n()

tbl_demo
```

**Features:**

-   Continuous: mean (SD) and median (Q1, Q3)
-   Categorical: n (%)
-   Columns: each treatment arm + overall
-   `add_n()` shows N per column header

------------------------------------------------------------------------

## Step 5: Convert to gt Table

```{r to-gt}
gt_demo <- tbl_demo %>%
  gtsummary::as_gt() %>%
  gt::tab_header(
    title    = "Table 14.1.1",
    subtitle = "Summary of Demographic and Baseline Characteristics (Safety Population)"
  ) %>%
  gt::tab_source_note(
    source_note = "Source: pharmaverseadam::adsl | Population: Safety (SAFFL=Y)"
  )

gt_demo
```

------------------------------------------------------------------------

## Step 6: Explore Disposition Variables

```{r disposition-explore}
# EOSSTT = End of Study Status (confirmed in adsl)
# DTHFL = Death Flag (confirmed in adsl)

adsl_saf %>%
  dplyr::count(EOSSTT)
```

------------------------------------------------------------------------

## Step 7: Disposition Summary Table

```{r disposition-gtsummary}
disp_tab <- adsl_saf %>%
  dplyr::select(TRT01A, EOSSTT, DTHFL)

tbl_disp <- disp_tab %>%
  gtsummary::tbl_summary(
    by = TRT01A,
    type = list(
      EOSSTT ~ "categorical",
      DTHFL  ~ "categorical"
    ),
    statistic = gtsummary::all_categorical() ~ "{n} ({p}%)",
    missing = "no"
  ) %>%
  gtsummary::add_overall(last = TRUE) %>%
  gtsummary::add_n()

tbl_disp
```

------------------------------------------------------------------------

## Step 8: Disposition gt Table

```{r disposition-gt}
gt_disp <- tbl_disp %>%
  gtsummary::as_gt() %>%
  gt::tab_header(
    title    = "Table 14.2.1",
    subtitle = "Subject Disposition Summary (Safety Population)"
  ) %>%
  gt::tab_source_note(
    source_note = "Source: pharmaverseadam::adsl | Population: Safety (SAFFL=Y)"
  )

gt_disp
```

------------------------------------------------------------------------

## Validation Checks

```{r validate}
cat("\n=== Table Validation ===\n\n")

# Check 1: Safety N
cat("Check 1 - Safety N:", nrow(adsl_saf), "\n")

# Check 2: All SAFFL subjects have TRT01A
check2 <- adsl_saf %>% dplyr::filter(is.na(TRT01A))
cat("Check 2 - Missing TRT01A:", nrow(check2), "\n")

# Check 3: EOSSTT completeness
check3 <- adsl_saf %>% dplyr::filter(is.na(EOSSTT))
cat("Check 3 - Missing EOSSTT:", nrow(check3), "\n")

cat("\n✓ Validation complete\n")
```

------------------------------------------------------------------------

## Summary

```{r summary}
cat("\n=== Day 22 Summary ===\n\n")
cat("Safety N:", nrow(adsl_saf), "\n\n")

cat("Treatment arms:\n")
adsl_saf %>% dplyr::count(TRT01A)

cat("\nEnd of study status:\n")
adsl_saf %>% dplyr::count(EOSSTT)
```

------------------------------------------------------------------------

## Export Reference (not run)

```{r export-tables, eval=FALSE}
# Export gtsummary table to Word/RTF (covered in a later day)
# library(flextable)
# library(officer)
#
# ft_demo <- gtsummary::as_flex_table(tbl_demo)
# doc <- officer::read_docx() %>%
#   officer::body_add_flextable(ft_demo)
# print(doc, target = "table14_1_1_demography.docx")
```

------------------------------------------------------------------------

## Key Takeaways

1.  Always run `names(adsl)` first to confirm which columns exist
2.  `pharmaverseadam::adsl` confirmed variables: AGE, SEX, RACE, ETHNIC, AGEGR1, EOSSTT, DTHFL, TRT01A
3.  `gtsummary::tbl_summary()` handles both continuous and categorical in one call
4.  `gtsummary::as_gt()` converts to gt for headers, footnotes, and final formatting
5.  No `DCSREAS`, `BMIBL`, or `adtte` in `pharmaverseadam` - always verify before use

------------------------------------------------------------------------

## Next Steps

**Day 23:** AE tables with `pharmaverseadam::adae` and `gt` **Day 24:** KM curves with `ggsurvfit` **Day 25:** ARD-first workflow with `cards`

------------------------------------------------------------------------

## Resources

-   gtsummary: <https://www.danieldsjoberg.com/gtsummary/>
-   gt clinical case study: <https://gt.rstudio.com/articles/case-study-clinical-tables.html>
-   pharmaverseadam datasets: <https://pharmaverse.github.io/pharmaverseadam/>

------------------------------------------------------------------------

**End of Day 22**

<a href="https://visitorbadge.io/status?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse"><img src="https://api.visitorbadge.io/api/visitors?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse&label=Visits%20till%20date&countColor=%23263759&style=plastic&labelStyle=none" /></a>
