---
title: "Day 28: Tplyr - Declarative Clinical Table Programming"
subtitle: "Grammar of clinical data summaries"

format:
  html:
    theme:
      light: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css
editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---
::: {style="margin-bottom:12px;"}
< a target="_blank" href="../index.html" class="resource-link"> Back to Roadmap</a>
:::

## Overview

`Tplyr` provides a grammar-of-tables approach to clinical data summaries. A table is built by stacking independent **layers**. When `build()` is called, all layers execute and assemble into a single data frame ready for downstream display or RTF export.

| Concept | Tplyr | rtables / tern |
|--------------------|------------------|-----------------------------------|
| Entry point | `tplyr_table()` | `basic_table()` |
| Analysis unit | layer (`add_layer()`) | `analyze()` / `split_rows_by()` |
| Execution | `build()` returns a data frame | `build_table()` returns an `rtable` |
| Sorting | dplyr post-`build()` | pre-ordered factors |
| Output | data frame -\> any renderer | `rtable` -\> `r2rtf` |

------------------------------------------------------------------------

## Setup

```{r setup}
library(Tplyr)
library(pharmaverseadam)
library(dplyr)
library(tidyr)
library(knitr)

adsl <- pharmaverseadam::adsl
adae  <- pharmaverseadam::adae
adlb  <- pharmaverseadam::adlb

adsl_safe <- adsl |>
  dplyr::filter(SAFFL == "Y") |>
  dplyr::mutate(
    TRT01A = factor(TRT01A),
    SEX    = factor(SEX),
    RACE   = factor(RACE)
  )

adae_safe <- adae |>
  dplyr::filter(SAFFL == "Y", TRTEMFL == "Y")

adlb_alt <- adlb |>
  dplyr::filter(SAFFL == "Y", PARAMCD == "ALT",
                ANL01FL == "Y", !is.na(BNRIND), !is.na(ANRIND)) |>
  dplyr::mutate(
    BNRIND = factor(BNRIND, levels = c('LOW', 'NORMAL', 'HIGH')),
    ANRIND = factor(ANRIND, levels = c('LOW', 'NORMAL', 'HIGH'))
  )

cat("ADSL safety N:", nrow(adsl_safe), "\n")
cat("Treatment arms:", levels(adsl_safe$TRT01A), "\n")
cat("ADAE TEAE records:", nrow(adae_safe), "\n")
cat("ADLB ALT records:", nrow(adlb_alt), "\n")
```

------------------------------------------------------------------------

## Part 1: Core Tplyr Concepts

### Step 1: Single-layer count table

```{r tplyr-basic}
t_sex <- tplyr_table(adsl_safe, TRT01A) |>
  add_total_group() |>
  add_layer(
    group_count(SEX, by = 'Sex') |>
      set_format_strings(f_str('xx (xx.x%)', n, pct))
  )
build(t_sex) |>
  dplyr::select(-dplyr::starts_with('ord')) |>
  knitr::kable()
```

------------------------------------------------------------------------

### Step 2: Descriptive statistics layer

```{r tplyr-desc-simple}
# Named entries in set_format_strings() become row labels.
# f_str() maps positional args to statistic keywords: mean, sd, median, q1, q3, min, max
t_age <- tplyr_table(adsl_safe, TRT01A) |>
  add_total_group() |>
  add_layer(
    group_desc(AGE, by = 'Age (years)') |>
      set_format_strings(
        'n'           = f_str('xx', n),
        'Mean (SD)'   = f_str('xx.x (xx.xx)', mean, sd),
        'Median'      = f_str('xx.x', median),
        'Q1, Q3'      = f_str('xx.x, xx.x', q1, q3),
        'Min, Max'    = f_str('xx.x, xx.x', min, max)
      )
  )
build(t_age) |>
  dplyr::select(-dplyr::starts_with('ord')) |>
  knitr::kable()
```

------------------------------------------------------------------------

## Part 2: Demographics Table

### Step 3: Multi-layer demographics table

```{r tplyr-demo}
t_demo <- tplyr_table(adsl_safe, TRT01A) |>
  add_total_group() |>
  add_layer(
    group_desc(AGE, by = 'Age (years)') |>
      set_format_strings(
        'n'         = f_str('xx', n),
        'Mean (SD)' = f_str('xx.x (xx.xx)', mean, sd),
        'Median'    = f_str('xx.x', median),
        'Min, Max'  = f_str('xx.x, xx.x', min, max)
      )
  ) |>
  add_layer(
    group_count(SEX, by = 'Sex') |>
      set_format_strings(f_str('xx (xx.x%)', n, pct))
  ) |>
  add_layer(
    group_count(RACE, by = 'Race') |>
      set_format_strings(f_str('xx (xx.x%)', n, pct))
  ) |>
  add_layer(
    group_count(AGEGR1, by = 'Age Group') |>
      set_format_strings(f_str('xx (xx.x%)', n, pct))
  )
demo_result <- build(t_demo) |>
  dplyr::select(-dplyr::starts_with('ord'))
knitr::kable(demo_result)
```

------------------------------------------------------------------------

## Part 3: AE Frequency Table

### Step 4: AE table with subject-level denominators

```{r tplyr-ae}
# Required trio for subject-level AE denominators:
#   set_pop_data() / set_pop_treat_var() / set_distinct_by(USUBJID)
# Sort by frequency AFTER build() using dplyr on the Total column.
t_ae <- tplyr_table(adae_safe, TRT01A) |>
  set_pop_data(adsl_safe) |>
  set_pop_treat_var(TRT01A) |>
  add_total_group() |>
  add_layer(
    group_count(AEDECOD, by = vars(AEBODSYS)) |>
      set_distinct_by(USUBJID) |>
      set_format_strings(f_str('xx (xx.x%)', distinct_n, distinct_pct))
  )
ae_result_raw <- build(t_ae) |>
  dplyr::select(-dplyr::starts_with('ord'))
total_col <- grep('Total', names(ae_result_raw), value = TRUE)[1]
ae_result <- ae_result_raw |>
  dplyr::mutate(
    .sort_n = suppressWarnings(
      as.integer(trimws(sub('\\s*\\(.*', '', .data[[total_col]])))
    )
  ) |>
  dplyr::arrange(row_label1, dplyr::desc(.sort_n)) |>
  dplyr::select(-.sort_n)
cat("AE table rows:", nrow(ae_result), "\n")
knitr::kable(head(ae_result, 20))
```

------------------------------------------------------------------------

### Step 5: Apply a frequency threshold

```{r tplyr-ae-filter}
arm_cols <- grep('^var1_', names(ae_result), value = TRUE)
arm_cols <- arm_cols[!grepl('Total', arm_cols, ignore.case = TRUE)]
arm1_col <- arm_cols[1]
cat("First arm column:", arm1_col, "\n")
ae_filtered <- ae_result |>
  dplyr::filter(
    !is.na(row_label2),
    suppressWarnings(
      as.integer(trimws(sub('\\s*\\(.*', '', .data[[arm1_col]]))) >= 5
    )
  )
cat("PT rows with n >= 5:", nrow(ae_filtered), "\n")
knitr::kable(ae_filtered)
```

------------------------------------------------------------------------

## Part 4: Shift Table

### Step 6: Lab shift table with group_shift()

```{r tplyr-shift}
# group_shift() uses vars(row = <baseline>, column = <post-baseline>).
# The keyword is 'column', not 'col'.
t_shift <- tplyr_table(adlb_alt, TRT01A) |>
  set_pop_data(adsl_safe) |>
  set_pop_treat_var(TRT01A) |>
  add_layer(
    group_shift(vars(row = BNRIND, column = ANRIND), by = 'ALT') |>
      set_format_strings(f_str('xx (xx.x%)', n, pct))
  )
shift_result <- build(t_shift) |>
  dplyr::select(-dplyr::starts_with('ord'))
knitr::kable(shift_result)
```

------------------------------------------------------------------------

## Part 5: Extracting Numeric Data

### Step 7: get_numeric_data() for traceability

```{r tplyr-numeric}
# get_numeric_data() returns a list of data frames with unformatted statistics.
# Column names depend on which statistics were requested.
# Always inspect with names() first -- never hard-code statistic column names.
t_age2 <- tplyr_table(adsl_safe, TRT01A) |>
  add_layer(
    group_desc(AGE) |>
      set_format_strings(
        'Mean (SD)' = f_str('xx.x (xx.xx)', mean, sd),
        'Median'    = f_str('xx.x', median)
      )
  )
built_age <- build(t_age2)
num_data  <- get_numeric_data(t_age2)
cat("Layers returned:", length(num_data), "\n")
cat("Layer 1 columns:", paste(names(num_data[[1]]), collapse = ", "), "\n")
# Display all columns; select() avoided because column names vary by statistics
knitr::kable(num_data[[1]], digits = 2)
```

------------------------------------------------------------------------

## Part 6: Column Nesting

### Step 8: cols argument for nested column splits

```{r tplyr-cols}
# cols adds a second level of column splitting (here: SEX within TRT01A)
t_nested <- tplyr_table(adsl_safe, TRT01A, cols = SEX) |>
  add_layer(
    group_count(AGEGR1, by = 'Age Group') |>
      set_format_strings(f_str('xx (xx.x%)', n, pct))
  )
build(t_nested) |>
  dplyr::select(-dplyr::starts_with('ord')) |>
  knitr::kable()
```

------------------------------------------------------------------------

## Validation Checks

```{r validate}
cat("\n=== Day 28 Validation ===\n\n")
cat("Check 1 - Demographics rows:", nrow(demo_result), "\n")
cat("Check 2 - AE table rows:", nrow(ae_result), "\n")
cat("Check 3 - AE rows after n>=5 filter:", nrow(ae_filtered), "\n")
cat("Check 4 - Shift table rows:", nrow(shift_result), "\n")
cat("Check 5 - Numeric data layers:", length(num_data), "\n")
cat("\nValidation complete\n")
```

------------------------------------------------------------------------

## Key Takeaways

1.  **`tplyr_table(data, treat_var)`** -- unquoted NSE; `add_total_group()` adds a Total column
2.  **Three layer types:** `group_desc()` continuous, `group_count()` categorical, `group_shift()` shift
3.  **`f_str('xx.x (xx.xx)', mean, sd)`** -- positional vars map to statistics; named entries become row labels
4.  **`group_shift()` uses `column=`** (not `col=`) in `vars(row = ..., column = ...)`
5.  **`set_pop_data()` + `set_pop_treat_var()` + `set_distinct_by(USUBJID)`** -- required for AE subject-level denominators
6.  **Never combine `set_order_count_method()` + `set_ordering_cols()`** -- sort with `dplyr::arrange()` after `build()` instead
7.  **`get_numeric_data()` column names vary** -- always inspect with `names()` first; never hard-code statistic column names
8.  **`build()` returns a plain data frame** -- filter, sort, join like any tibble; drop sort columns with `select(-starts_with('ord'))`

------------------------------------------------------------------------

## Resources

-   Tplyr documentation: <https://atorus-research.github.io/Tplyr/>
-   Tplyr cheatsheet: <https://atorus-research.github.io/Tplyr_cheatsheet.pdf>
-   r4csr book: <https://r4csr.org>

------------------------------------------------------------------------

End of Day 28

< a target="_blank" href="https://visitorbadge.io/status?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse"><img src="https://api.visitorbadge.io/api/visitors?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse&label=Visits%20till%20date&countColor=%23263759&style=plastic&labelStyle=none" /></a>
