---
title: "Day 25: gtsummary and tfrmt - ARD-Backed Production Tables"
subtitle: "Flexible TLGs from ARD and format-spec driven TLF libraries"


format:
  html:
    theme:
      light: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css

editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---
::: {style="margin-bottom:12px;"}
<a href="../index.html" class="resource-link"> Back to Roadmap</a>
:::

## Overview

Today covers two complementary approaches to clinical table production:

-   **`gtsummary`** - builds publication-quality tables directly from ADaM data, and can extract an ARD via `gather_ard()`
-   **`tfrmt`** - separates data preparation from display spec; one format spec applies to many tables

------------------------------------------------------------------------

## Setup

```{r setup}
library(gtsummary)
library(tfrmt)
library(pharmaverseadam)
library(dplyr)
library(gt)

# Used explicitly later
library(purrr)
library(tidyr)

# Safety population
adsl_safe <- pharmaverseadam::adsl |>
  dplyr::filter(SAFFL == "Y")

# Lab data for tfrmt section
adlb <- pharmaverseadam::adlb |>
  dplyr::filter(SAFFL == "Y", !is.na(AVAL))

cat("ADSL Safety N:", nrow(adsl_safe), "\n")
cat("ADLB records:", nrow(adlb), "\n")

cat("\nADLB columns (first 30):\n")
print(utils::head(names(adlb), 30))
```

------------------------------------------------------------------------

## Part 1: gtsummary

### Step 1: Demographics Table

```{r tbl-demo}
tbl_demo <- adsl_safe |>
  gtsummary::tbl_summary(
    by       = TRT01A,
    include  = c(AGE, AGEGR1, SEX, RACE, ETHNIC),
    type     = list(AGE ~ "continuous2"),
    statistic = list(
      gtsummary::all_continuous2() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
      gtsummary::all_categorical() ~ "{n} ({p}%)"
    ),
    digits  = gtsummary::all_continuous() ~ 1,
    missing = "no"
  ) |>
  gtsummary::add_overall(last = TRUE) |>
  gtsummary::add_n() |>
  gtsummary::modify_header(label = "**Characteristic**") |>
  gtsummary::bold_labels() |>
  gtsummary::add_stat_label()

tbl_demo
```

------------------------------------------------------------------------

### Step 2: Convert to gt and Add Headers

```{r tbl-demo-gt}
tbl_demo |>
  gtsummary::as_gt() |>
  gt::tab_header(
    title    = "Table 14.1.1",
    subtitle = "Demographic and Baseline Characteristics (Safety Population)"
  ) |>
  gt::tab_source_note(
    source_note = "Source: pharmaverseadam::adsl | Population: Safety (SAFFL=Y)"
  )
```

------------------------------------------------------------------------

### Step 3: Extract ARD from gtsummary

```{r gather-ard}
ard_from_gtsumm <- tbl_demo |>
  gtsummary::gather_ard() |>
  purrr::pluck("tbl_summary")

cat("ARD rows extracted from gtsummary:", nrow(ard_from_gtsumm), "\n")

ard_from_gtsumm |>
  dplyr::select(group1, group1_level, variable, stat_name, stat_label, stat) |>
  head(10)
```

------------------------------------------------------------------------

### Step 4: Lab Summary Table from ADLB

```{r tbl-lab}
# Summarise ALT at baseline per treatment
# (Do not assume AVALU exists in pharmaverseadam::adlb)
adlb_alt_bl <- adlb |>
  dplyr::filter(PARAMCD == "ALT", ABLFL == "Y") |>
  dplyr::select(TRT01A, AVAL)

cat("ALT baseline records:", nrow(adlb_alt_bl), "\n")

tbl_lab <- adlb_alt_bl |>
  gtsummary::tbl_summary(
    by       = TRT01A,
    include  = AVAL,
    type     = list(AVAL ~ "continuous2"),
    statistic = list(
      gtsummary::all_continuous2() ~ c(
        "{mean} ({sd})",
        "{median} ({p25}, {p75})",
        "{min}, {max}"
      )
    ),
    digits  = gtsummary::all_continuous() ~ 1,
    missing = "no",
    label   = list(AVAL ~ "ALT")
  ) |>
  gtsummary::add_overall(last = TRUE) |>
  gtsummary::add_n() |>
  gtsummary::bold_labels()

tbl_lab
```

------------------------------------------------------------------------

### Step 5: Logistic Regression Table

```{r tbl-regression}
# Logistic regression: SEX ~ AGE + TRT01A
adsl_model <- adsl_safe |>
  dplyr::mutate(SEX_BIN = dplyr::if_else(SEX == "M", 1L, 0L))

logit_mod <- glm(
  SEX_BIN ~ AGE + TRT01A,
  data   = adsl_model,
  family = binomial()
)

tbl_reg <- logit_mod |>
  gtsummary::tbl_regression(
    exponentiate = TRUE,
    label        = list(AGE ~ "Age (years)", TRT01A ~ "Treatment Arm")
  ) |>
  gtsummary::add_n() |>
  gtsummary::add_nevent() |>
  gtsummary::bold_p(t = 0.05) |>
  gtsummary::bold_labels()

tbl_reg
```

------------------------------------------------------------------------

### Step 6: Merge Tables with tbl_merge

```{r tbl-merge}
tbl_merged <- gtsummary::tbl_merge(
  tbls        = list(tbl_demo, tbl_lab),
  tab_spanner = c("**Demographics**", "**ALT at Baseline**")
)

tbl_merged
```

------------------------------------------------------------------------

## Part 2: tfrmt

### Step 7: Build a tfrmt Format Spec

```{r tfrmt-spec}
# tfrmt expects a long dataset with:
# - group columns, label column, column columns
# - param column (e.g., mean, sd)
# - value column (numeric)

my_fmt <- tfrmt::tfrmt(
  group  = vars(PARAMCD),
  label  = vars(VISIT),
  column = vars(TRT01A),
  param  = PARAM,
  value  = VALUE,
  body_plan = tfrmt::body_plan(
    tfrmt::frmt_structure(
      group_val = ".default",
      label_val = ".default",
      tfrmt::frmt_combine(
        "{mean} ({sd})",
        mean = tfrmt::frmt("xx.x"),
        sd   = tfrmt::frmt("xx.xx")
      )
    )
  )
)

my_fmt
```

------------------------------------------------------------------------

### Step 8: Prepare BDS Summary Data for tfrmt

```{r tfrmt-data}
# Summarise ADLB: mean and SD of AVAL by PARAMCD, AVISIT, TRT01A
adlb_summary <- adlb |>
  dplyr::filter(PARAMCD %in% c("ALT", "AST", "BILI")) |>
  dplyr::filter(!is.na(AVISIT), !is.na(AVAL)) |>
  dplyr::group_by(PARAMCD, AVISIT, TRT01A) |>
  dplyr::summarise(
    mean = mean(AVAL, na.rm = TRUE),
    sd   = sd(AVAL, na.rm = TRUE),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(
    cols      = c(mean, sd),
    names_to  = "PARAM",
    values_to = "VALUE"
  ) |>
  dplyr::rename(VISIT = AVISIT)

cat("tfrmt input rows:", nrow(adlb_summary), "\n")
head(adlb_summary, 10)
```

------------------------------------------------------------------------

### Step 9: Apply tfrmt Format Spec

```{r tfrmt-apply}
tfrmt_gt <- tfrmt::print_to_gt(my_fmt, .data = adlb_summary)

tfrmt_gt
```

------------------------------------------------------------------------

### Step 10: Mock Table from tfrmt Spec

```{r tfrmt-mock}
tfrmt::print_mock_gt(my_fmt, .data = adlb_summary)
```

------------------------------------------------------------------------

## Step 11: Export tfrmt table to HTML

```{r export-html}
tfrmt_gt |>
  gt::gtsave("tfrmt_lab_summary.html")

cat("✓ tfrmt_lab_summary.html exported\n")
```

------------------------------------------------------------------------

## Validation Checks

```{r validate}
cat("\n=== Day 25 Validation ===\n\n")

# Check 1: Safety N by arm
cat("Check 1 - Safety N by arm:\n")
print(adsl_safe |> dplyr::count(TRT01A))

# Check 2: ARD extracted from gtsummary has rows
cat("\nCheck 2 - ARD rows from gather_ard():", nrow(ard_from_gtsumm), "\n")

# Check 3: tfrmt data has all 3 parameters
cat("\nCheck 3 - PARAMCD in tfrmt data:\n")
print(adlb_summary |> dplyr::distinct(PARAMCD))

# Check 4: No NA VALUE in tfrmt input
cat("\nCheck 4 - NA VALUE rows:", nrow(adlb_summary |> dplyr::filter(is.na(VALUE))), "\n")

cat("\n✓ Validation complete\n")
```

------------------------------------------------------------------------

## Key Takeaways

1.  `tbl_summary(by = TRT01A)` is the workhorse for clinical demographics tables
2.  `gather_ard()` extracts a CDISC-compatible ARD from any `gtsummary` table
3.  `tfrmt` decouples format spec from data; `print_mock_gt()` helps you agree shells early
4.  Always validate dataset columns (e.g., do not assume `AVALU` exists)

------------------------------------------------------------------------

## Next Steps

**Day 26:** `flextable` + `officer` - Word/RTF clinical output

------------------------------------------------------------------------

## Resources

-   gtsummary documentation: <https://www.danieldsjoberg.com/gtsummary/>
-   tfrmt examples: <https://gsk-biostatistics.github.io/tfrmt/articles/examples.html>

------------------------------------------------------------------------

**End of Day 25**

<a href="https://visitorbadge.io/status?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse"><img src="https://api.visitorbadge.io/api/visitors?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse&label=Visits%20till%20date&countColor=%23263759&style=plastic&labelStyle=none" /></a>
