---
title: "Day 17: ADSL Part 2 - Population Flags & Demographics"
subtitle: "Week 3, Day 17: SAFFL, ITTFL, demographic groupings, and baseline measurements"


format:
  html:
    theme:
      light: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css

editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---
::: {style="margin-bottom:12px;"}
< a target="_blank" href="../index.html" class="resource-link"> Back to Roadmap</a>
:::

## Welcome to ADSL Part 2

Today we **complete ADSL** by adding population flags, demographic groupings, and baseline measurements.

**What we'll add:**

-   Population flags (SAFFL, ITTFL)
-   Demographics groupings (AGEGR1, SEXN, RACEN)
-   Baseline measurements (HEIGHT, WEIGHT, BMI)

------------------------------------------------------------------------

## Setup and Load Day 16 ADSL

```{r setup}
library(admiral)
library(pharmaversesdtm)
library(dplyr)
library(lubridate)
library(stringr)
library(xportr)

# Load SDTM domains
dm <- pharmaversesdtm::dm
ex <- pharmaversesdtm::ex
ds <- pharmaversesdtm::ds
vs <- pharmaversesdtm::vs

cat("Loaded SDTM domains\n")
```

### Rebuild ADSL Part 1 from Day 16

```{r rebuild-adsl-part1}
# Start with DM
adsl <- dm %>%
  select(-DOMAIN)

# Treatment variables
adsl <- adsl %>%
  mutate(
    TRT01P = ARM,
    TRT01A = ACTARM
  )

# Prepare EX with dates
ex_ext <- ex %>%
  derive_vars_dtm(dtc = EXSTDTC, new_vars_prefix = "EXST") %>%
  derive_vars_dtm(dtc = EXENDTC, new_vars_prefix = "EXEN", time_imputation = "last")

# First dose date
adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) & !is.na(EXSTDTM),
    new_vars = exprs(TRTSDTM = EXSTDTM),
    order = exprs(EXSTDTM, EXSEQ),
    mode = "first",
    by_vars = exprs(STUDYID, USUBJID)
  )

# Last dose date
adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) & !is.na(EXENDTM),
    new_vars = exprs(TRTEDTM = EXENDTM),
    order = exprs(EXENDTM, EXSEQ),
    mode = "last",
    by_vars = exprs(STUDYID, USUBJID)
  )

# Convert to dates
adsl <- adsl %>%
  derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM, TRTEDTM))

# Treatment duration
adsl <- adsl %>%
  derive_var_trtdurd()

cat("\nADSL Part 1 rebuilt:", nrow(adsl), "subjects\n")
```

------------------------------------------------------------------------

## Step 1: Derive Safety Population Flag (SAFFL)

**Safety Population:** Subjects who received at least one dose of study treatment.

```{r safety-flag}
# Derive SAFFL
adsl <- adsl %>%
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars = exprs(STUDYID, USUBJID),
    new_var = SAFFL,
    condition = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))
  )

cat("\nSafety population flag derived\n")

# Distribution
adsl %>%
  count(SAFFL, TRT01A)
```

**Logic:**

-   SAFFL = "Y" if subject has any EX record with EXDOSE \> 0
-   Used for safety analyses (AEs, labs, vitals)

------------------------------------------------------------------------

## Step 2: Derive ITT Population Flag (ITTFL)

**Intent-to-Treat Population:** Randomized subjects.

```{r itt-flag}
# Derive ITTFL
adsl <- adsl %>%
  derive_var_merged_exist_flag(
    dataset_add = ds,
    by_vars = exprs(STUDYID, USUBJID),
    new_var = ITTFL,
    condition = DSDECOD == "RANDOMIZED"
  )

cat("\nITT population flag derived\n")

# Distribution
adsl %>%
  count(ITTFL, SAFFL)
```

**Logic:**

-   ITTFL = "Y" if randomization event exists in DS
-   Used for efficacy analyses

------------------------------------------------------------------------

## Step 3: Create Age Groupings

```{r age-groups}
# Derive age group categories
adsl <- adsl %>%
  mutate(
    AGEGR1 = case_when(
      AGE < 65 ~ "<65",
      AGE >= 65 & AGE < 75 ~ "65-74",
      AGE >= 75 ~ ">=75",
      TRUE ~ NA_character_
    ),
    AGEGR1N = case_when(
      AGE < 65 ~ 1,
      AGE >= 65 & AGE < 75 ~ 2,
      AGE >= 75 ~ 3,
      TRUE ~ NA_real_
    )
  )

cat("\nAge groupings created\n")

# Distribution
adsl %>%
  count(AGEGR1, AGEGR1N)
```

**Age Groups:**

-   \<65: Younger adults (N=1)

-   65-74: Older adults (N=2)

-   

    > =75: Elderly (N=3)

------------------------------------------------------------------------

## Step 4: Derive Numeric Demographics

```{r numeric-demographics}
# Numeric versions of categorical variables
adsl <- adsl %>%
  mutate(
    SEXN = case_when(
      SEX == "F" ~ 1,
      SEX == "M" ~ 2,
      TRUE ~ NA_real_
    ),
    RACEN = case_when(
      RACE == "AMERICAN INDIAN OR ALASKA NATIVE" ~ 1,
      RACE == "ASIAN" ~ 2,
      RACE == "BLACK OR AFRICAN AMERICAN" ~ 3,
      RACE == "NATIVE HAWAIIAN OR OTHER PACIFIC ISLANDER" ~ 4,
      RACE == "WHITE" ~ 5,
      RACE == "MULTIPLE" ~ 6,
      TRUE ~ NA_real_
    )
  )

cat("\nNumeric demographics derived\n")

# Check mappings
adsl %>%
  count(SEX, SEXN)

adsl %>%
  count(RACE, RACEN) %>%
  head(10)
```

**Why numeric versions?**

-   Enable table sorting
-   Required for some stats procedures
-   Standard across studies

------------------------------------------------------------------------

## Step 5: Prepare VS for Baseline Merging

```{r prepare-vs}
# Filter VS for baseline
vs_bl <- vs %>%
  filter(VISIT %in% c("BASELINE", "SCREENING") | VISITNUM == 0) %>%
  derive_vars_dt(dtc = VSDTC, new_vars_prefix = "VS")

# Available baseline measurements
vs_bl %>%
  count(VSTESTCD)
```

------------------------------------------------------------------------

## Step 6: Derive Baseline Height

```{r baseline-height}
# Merge baseline height
adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = vs_bl,
    by_vars = exprs(STUDYID, USUBJID),
    filter_add = VSTESTCD == "HEIGHT" & !is.na(VSSTRESN),
    new_vars = exprs(HEIGHTBL = VSSTRESN),
    order = exprs(VSDT),
    mode = "last"
  )

cat("\nBaseline height merged\n")

# Summary
adsl %>%
  filter(!is.na(HEIGHTBL)) %>%
  summarise(
    N = n(),
    Mean = round(mean(HEIGHTBL), 1),
    SD = round(sd(HEIGHTBL), 1),
    Median = median(HEIGHTBL),
    Min = min(HEIGHTBL),
    Max = max(HEIGHTBL)
  )
```

**Logic:** Take last height at or before baseline

------------------------------------------------------------------------

## Step 7: Derive Baseline Weight

```{r baseline-weight}
# Merge baseline weight
adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = vs_bl,
    by_vars = exprs(STUDYID, USUBJID),
    filter_add = VSTESTCD == "WEIGHT" & !is.na(VSSTRESN),
    new_vars = exprs(WEIGHTBL = VSSTRESN),
    order = exprs(VSDT),
    mode = "last"
  )

cat("\nBaseline weight merged\n")

# Summary
adsl %>%
  filter(!is.na(WEIGHTBL)) %>%
  summarise(
    N = n(),
    Mean = round(mean(WEIGHTBL), 1),
    SD = round(sd(WEIGHTBL), 1),
    Median = median(WEIGHTBL),
    Min = min(WEIGHTBL),
    Max = max(WEIGHTBL)
  )
```

------------------------------------------------------------------------

## Step 8: Calculate Baseline BMI

```{r baseline-bmi}
# Calculate BMI
adsl <- adsl %>%
  mutate(
    BMIBL = case_when(
      !is.na(HEIGHTBL) & !is.na(WEIGHTBL) ~ round(WEIGHTBL / (HEIGHTBL / 100)^2, 1),
      TRUE ~ NA_real_
    )
  )

cat("\nBaseline BMI calculated\n")

# Summary
adsl %>%
  filter(!is.na(BMIBL)) %>%
  summarise(
    N = n(),
    Mean = round(mean(BMIBL), 1),
    SD = round(sd(BMIBL), 1),
    Median = median(BMIBL),
    Min = min(BMIBL),
    Max = max(BMIBL)
  )
```

**Formula:** BMI = Weight (kg) / (Height (m))²

------------------------------------------------------------------------

## Step 9: Create BMI Groups

```{r bmi-groups}
# BMI groupings
adsl <- adsl %>%
  mutate(
    BMIBLGR1 = case_when(
      BMIBL < 18.5 ~ "<18.5",
      BMIBL >= 18.5 & BMIBL < 25 ~ "18.5-<25",
      BMIBL >= 25 & BMIBL < 30 ~ "25-<30",
      BMIBL >= 30 ~ ">=30",
      TRUE ~ NA_character_
    )
  )

# Distribution
adsl %>%
  count(BMIBLGR1)
```

**BMI Categories:**

-   \<18.5: Underweight

-   18.5-\<25: Normal

-   25-\<30: Overweight

-   

    > =30: Obese

------------------------------------------------------------------------

## Validation Checks

```{r validate}
cat("\n=== ADSL Validation ===\n\n")

# Check 1: SAFFL consistency
check1 <- adsl %>%
  filter(SAFFL == "Y" & is.na(TRTSDT))
cat("Check 1 - SAFFL='Y' but no TRTSDT:", nrow(check1), "\n")

# Check 2: Age groups
check2 <- adsl %>%
  filter(!is.na(AGEGR1) & is.na(AGEGR1N))
cat("Check 2 - AGEGR1 without AGEGR1N:", nrow(check2), "\n")

# Check 3: BMI calculation
check3 <- adsl %>%
  filter(!is.na(BMIBL)) %>%
  mutate(
    BMI_check = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1),
    Match = abs(BMIBL - BMI_check) < 0.1
  ) %>%
  filter(!Match)
cat("Check 3 - BMI mismatch:", nrow(check3), "\n")

# Check 4: Numeric demographics
check4 <- adsl %>%
  filter(!is.na(SEX) & is.na(SEXN))
cat("Check 4 - SEX without SEXN:", nrow(check4), "\n")

cat("\n✓ Validation complete\n")
```

------------------------------------------------------------------------

## Complete ADSL Summary

```{r final-summary}
cat("\n=== Complete ADSL ===\n\n")
cat("Total subjects:", nrow(adsl), "\n")
cat("Variables:", ncol(adsl), "\n\n")

# Population flags
cat("Population Flags:\n")
adsl %>%
  summarise(
    SAFFL_Y = sum(SAFFL == "Y", na.rm = TRUE),
    ITTFL_Y = sum(ITTFL == "Y", na.rm = TRUE)
  )

# Demographics by treatment (Safety population)
cat("\nDemographics by Treatment (Safety Population):\n")
adsl %>%
  filter(SAFFL == "Y") %>%
  group_by(TRT01A) %>%
  summarise(
    N = n(),
    Age_Mean = round(mean(AGE), 1),
    Age_SD = round(sd(AGE), 1),
    Female_N = sum(SEX == "F"),
    Female_Pct = round(100 * sum(SEX == "F") / n(), 1),
    BMI_Mean = round(mean(BMIBL, na.rm = TRUE), 1),
    .groups = "drop"
  )
```

------------------------------------------------------------------------

## Export Complete ADSL

```{r export}
# Apply variable labels
attr(adsl$SAFFL, "label") <- "Safety Population Flag"
attr(adsl$ITTFL, "label") <- "Intent-to-Treat Population Flag"
attr(adsl$AGEGR1, "label") <- "Age Group 1"
attr(adsl$AGEGR1N, "label") <- "Age Group 1 (N)"
attr(adsl$SEXN, "label") <- "Sex (N)"
attr(adsl$RACEN, "label") <- "Race (N)"
attr(adsl$HEIGHTBL, "label") <- "Baseline Height (cm)"
attr(adsl$WEIGHTBL, "label") <- "Baseline Weight (kg)"
attr(adsl$BMIBL, "label") <- "Baseline Body Mass Index"
attr(adsl$BMIBLGR1, "label") <- "Baseline BMI Group 1"

# Export as XPT
xportr_write(adsl, path = "adsl.xpt", domain = "ADSL")

cat("\n✓ ADSL exported to: adsl.xpt\n")
```

------------------------------------------------------------------------

## Key Takeaways

**Population Flags:**

1.  SAFFL = received treatment (from EX)
2.  ITTFL = randomized (from DS)
3.  Both derived with `derive_var_merged_exist_flag()`

**Demographics:**

4.  Always create numeric versions (AGEGR1N, SEXN, RACEN)
5.  Numeric enables sorting and statistics
6.  Character for display in tables

**Baseline Measurements:**

7.  Merge from VS using `derive_vars_merged()`
8.  Take last baseline measurement
9.  Calculate BMI from HEIGHT and WEIGHT

**Validation:**

10. Check flag consistency (SAFFL="Y" must have TRTSDT)
11. Verify calculations (BMI formula)
12. Ensure char/numeric pairs match

------------------------------------------------------------------------

## Next Steps

**Day 18:** ADAE (Adverse Events) - OCCDS structure\
**Day 19:** ADLB (Labs) - BDS with baseline\
**Day 20:** ADVS (Vitals) - BDS with visits\
**Day 21:** ADTTE (Time-to-Event)

ADSL is complete - ready for all other ADaMs!

------------------------------------------------------------------------

## Resources

**Admiral:**

-   ADSL Vignette: <https://pharmaverse.github.io/admiral/articles/adsl.html>
-   derive_var_merged_exist_flag: <https://pharmaverse.github.io/admiral/reference/derive_var_merged_exist_flag.html>

**CDISC:**

-   ADaM IG v1.3: <https://www.cdisc.org/standards/foundational/adam>

**Pharmaverse:**

-   pharmaverse.org: <https://pharmaverse.org>

------------------------------------------------------------------------

**End of Day 17**

*Tomorrow: ADAE (Adverse Events)*

< a target="_blank" href="https://visitorbadge.io/status?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse"><img src="https://api.visitorbadge.io/api/visitors?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse&label=Visits%20till%20date&countColor=%23263759&style=plastic&labelStyle=none" /></a>
