---
title: "Day 15: ADaM Architecture & Admiral Core Engine"
subtitle: "Week 3, Day 15: Understanding ADaM Structures, Admiral Philosophy, and Core Derivation Patterns"


format:
  html:
    theme:
      light: cosmo

    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css
    
editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Welcome to Week 3: ADaM Deep Dive & Admiral Mastery

**Week 3 Overview:** After mastering SDTM creation and validation in Weeks 1-2, we now transition to **ADaM (Analysis Data Model)** - the analysis-ready datasets that feed directly into regulatory Tables, Listings, and Figures (TLFs).

This week is **intensive** - we'll use the `{admiral}` package to build every major ADaM dataset type. Admiral was co-developed by multiple pharmaceutical companies as a **tidyverse-based, modular, non-black-box toolbox** for ADaM creation, and is now maintained by 50+ pharma company programmers across the pharmaverse community.

**Day 15 is foundational** - today we learn the **architecture** before building datasets tomorrow.

------------------------------------------------------------------------

## 1 Learning Objectives

By the end of Day 15, you will be able to:

-   **Understand the three ADaM data structures:** ADSL (subject-level), BDS (Basic Data Structure), and OCCDS (Occurrence Data Structure)
-   **Explain Admiral's design philosophy:** Modular, transparent, tidyverse-native, and cross-company collaborative
-   **Master Admiral's four core derivation function categories:** `derive_vars_*`, `derive_var_*`, `derive_param_*`, and `derive_extreme_*`
-   **Map each of 8 key ADaM datasets** to their source SDTM domain, ADaM structure type, and key Admiral functions
-   **Navigate Admiral's function taxonomy** to quickly find the right function for any derivation task
-   **Understand Admiral's input-output contract** and how functions compose sequentially

**Clinical Context:** Today is conceptual but critical. Understanding ADaM architecture prevents hours of debugging later. Knowing which Admiral functions to use for which patterns makes you 10x more efficient in Days 16-30.

------------------------------------------------------------------------

## 2 What is ADaM? (Deep Dive)

### ADaM in the Regulatory Workflow

```         
Clinical Trial → EDC Data → SDTM (Tabulation) → ADaM (Analysis) → TLFs → Regulatory Submission
                                                      ↑
                                              Admiral creates this
```

**ADaM (Analysis Data Model)** is the CDISC standard for **analysis-ready datasets** used in regulatory submissions to FDA, EMA, and other health authorities.

### Why ADaM Exists

**SDTM limitations for analysis:** - One record per observation (not per analysis unit) - No derived variables (change from baseline, treatment-emergent flags) - No baseline or analysis visit structure - Not optimized for statistical procedures

**ADaM solves this by:** - Structuring data for analysis (one row per subject/parameter/timepoint) - Adding derived variables (flags, categories, calculated values) - Ensuring traceability back to SDTM - Enabling direct input to SAS PROC or R statistical functions

### ADaM Fundamental Principles (CDISC ADaM IG)

1.  **Analysis-ready:** Data can be directly input to statistical procedures
2.  **Self-contained:** All variables needed for analysis are in one dataset
3.  **Traceable:** Every value traces back to SDTM source
4.  **One record per analysis unit:** Structure depends on analysis type
5.  **Standardized naming:** Consistent variable names across studies

**Clinical Example:**

**In SDTM (LB - Laboratory):**

```         
USUBJID  LBTESTCD  LBSTRESN  LBDTC        VISIT
001      ALT       45        2024-01-15   Baseline
001      ALT       52        2024-02-15   Week 4
001      ALT       48        2024-03-15   Week 8
```

**In ADaM (ADLB - BDS structure):**

```         
USUBJID  PARAMCD  AVISIT    AVAL  ABLFL  BASE  CHG  ANRIND
001      ALT      Baseline  45    Y      45    NA   NORMAL
001      ALT      Week 4    52    N      45    7    NORMAL
001      ALT      Week 8    48    N      45    3    NORMAL
```

**Analysis benefit:** - Baseline flag (ABLFL) clearly identifies reference measurement - Change from baseline (CHG) pre-calculated - Reference range indicator (ANRIND) ready for shift tables - Structure enables `lm(CHG ~ AVISIT + BASE, data = adlb_filtered)`

------------------------------------------------------------------------

## 3 The Three ADaM Data Structures

CDISC defines **three primary ADaM structures**. Understanding which structure to use is the first decision in any ADaM creation.

### 3.1 ADSL: Subject-Level Analysis Dataset

**Structure:** **One record per subject**

**Purpose:** - Demographics and baseline characteristics - Treatment assignment (planned and actual) - Population flags (Safety, ITT, Per-Protocol) - Study dates and disposition - **Foundation for all other ADaM datasets**

**Key Variables:** - Identifiers: `USUBJID`, `SUBJID`, `SITEID` - Demographics: `AGE`, `SEX`, `RACE`, `ETHNIC` - Treatment: `TRT01P`, `TRT01A`, `TRTSDT`, `TRTEDT` - Flags: `SAFFL`, `ITTFL`, `COMPLFL` - Disposition: `EOSSTT`, `DCSREAS`

**SDTM Sources:** - Primary: `DM` (Demographics) - Secondary: `EX` (Exposure for treatment dates), `DS` (Disposition), `VS`/`LB` (baseline values)

**Clinical Use:** - Demographic tables (Table 14.1.1 in submissions) - Population definitions for all analyses - Subgroup definitions (age groups, sex, baseline severity)

**Example Record:**

```         
USUBJID: "01-701-1015"
AGE: 75, SEX: "F", RACE: "WHITE"
TRT01P: "Xanomeline High Dose", TRT01A: "Xanomeline High Dose"
TRTSDT: 2024-01-20, TRTEDT: 2024-07-15, TRTDURD: 178
SAFFL: "Y", ITTFL: "Y", COMPLFL: "Y"
EOSSTT: "COMPLETED"
```

### 3.2 BDS: Basic Data Structure

**Structure:** **One record per subject per parameter per analysis timepoint**

**Purpose:** - Repeated measures over time (labs, vital signs, efficacy scales) - Baseline identification and change from baseline - Reference range and shift analysis

**Key Variables:** - `PARAMCD`/`PARAM`: Parameter identifier (e.g., "ALT", "DIABP") - `AVAL`/`AVALC`: Analysis value (numeric/character) - `AVISIT`/`AVISITN`: Analysis visit - `ABLFL`: Baseline flag (Y for baseline record) - `BASE`: Baseline value (carried forward to all post-baseline) - `CHG`: Change from baseline (`AVAL - BASE`) - `ANRIND`: Analysis reference range indicator (LOW/NORMAL/HIGH)

**SDTM Sources:** - `LB` → `ADLB` (Lab results) - `VS` → `ADVS` (Vital signs) - `QS` → `ADQS` (Questionnaires) - `EG` → `ADEG` (ECG results)

**BDS Formula:**

```         
Number of Records = N subjects × N parameters × N timepoints
```

**Example: ADLB**

```         
USUBJID  PARAMCD  AVISIT    AVAL  ABLFL  BASE  CHG   PCHG   ANRIND
001      ALT      Baseline  45    Y      45    NA    NA     NORMAL
001      ALT      Week 4    52    N      45    7     15.6   NORMAL
001      ALT      Week 8    48    N      45    3     6.7    NORMAL
001      ALKPH    Baseline  78    Y      78    NA    NA     NORMAL
001      ALKPH    Week 4    82    N      78    4     5.1    NORMAL
```

**Clinical Use:** - Change from baseline analysis - Shift tables (baseline NORMAL → Week 4 HIGH) - Time-course plots - Repeated measures ANCOVA

### 3.3 OCCDS: Occurrence Data Structure

**Structure:** **One record per subject per event occurrence**

**Purpose:** - Adverse events - Medications - Medical history events - Any event that can occur 0, 1, or multiple times per subject

**Key Variables:** - `ASTDT`/`AENDT`: Analysis start/end dates - `ADURN`: Analysis duration - `TRTEMFL`: Treatment-emergent flag - `AOCCFL`: First occurrence flag - `ANL01FL`: Primary analysis flag - Event-specific variables (e.g., severity, causality for AE)

**SDTM Sources:** - `AE` → `ADAE` (Adverse events) - `CM` → `ADCM` (Concomitant medications) - `MH` → `ADMH` (Medical history)

**OCCDS Formula:**

```         
Number of Records = Variable (0 to N events per subject)
```

**Example: ADAE**

```         
USUBJID  AESEQ  AETERM      ASTDT       AENDT       TRTEMFL  AOCCFL  ASEV
001      1      Headache    2024-01-22  2024-01-23  Y        Y       MILD
001      2      Nausea      2024-02-10  2024-02-12  Y        N       MODERATE
001      3      Dizziness   2024-03-05  2024-03-05  Y        N       MILD
002      1      Headache    2024-01-18  2024-01-19  N        N       MILD
```

**Clinical Use:** - Adverse event frequency tables - Time-to-first-event analysis - Treatment-emergent vs pre-existing events - Causality and severity analysis

------------------------------------------------------------------------

## 4 Admiral Philosophy & Design Principles

### 4.1 Why Admiral Was Created

**Pre-Admiral Era (2010-2020):** - Every pharma company had proprietary SAS macro libraries - Black-box functions (couldn't see or modify code) - Duplicated effort (50 companies solving the same problems) - Difficult to validate (macros hidden in compiled code) - Knowledge siloed within companies

**Admiral Vision (2020-present):** - **Open-source:** Code is fully transparent and inspectable - **Cross-company collaboration:** Multiple pharmaceutical companies co-develop - **Tidyverse-native:** Leverages dplyr, tidyr, lubridate - familiar to R users - **Modular:** Small, composable functions instead of monolithic macros - **Non-black-box:** Every derivation is explicit in your script - **Validated:** Extensive test coverage and regulatory use

### 4.2 Admiral Core Design Principles

#### Principle 1: Modularity

**Bad (monolithic):**

``` r
create_adsl(dm, ex, ds, ae, lb)  # What does this do? Can't see inside!
```

**Good (modular Admiral):**

``` r
adsl <- dm %>%
  derive_vars_merged(dataset_add = ex, ...) %>%  # Clear: merging EX
  derive_var_trtdurd() %>%                       # Clear: deriving duration
  derive_var_merged_exist_flag(dataset_add = ex, new_var = SAFFL, ...)
```

Every step is visible and modifiable.

#### Principle 2: Tidyverse Integration

Admiral functions work seamlessly with `%>%` pipes and dplyr verbs:

``` r
adsl <- dm %>%
  filter(ARMCD != "SCRNFAIL") %>%          # dplyr
  derive_vars_merged(ex, ...) %>%          # admiral
  mutate(AGEGR1 = case_when(...)) %>%      # dplyr
  derive_var_trtdurd()                     # admiral
```

#### Principle 3: Transparency

**Every Admiral function:** - Returns a modified data frame (never a hidden state) - Has explicit input and output - Can be inspected mid-pipeline with `glimpse()` or `View()`

``` r
adsl_step1 <- dm %>% derive_vars_merged(ex, ...)
glimpse(adsl_step1)  # Inspect intermediate result
adsl_step2 <- adsl_step1 %>% derive_var_trtdurd()
```

#### Principle 4: Reusability

Admiral functions are **parameterized** - same function, different arguments:

``` r
# Derive first dose date
derive_vars_merged(dataset_add = ex, mode = "first", ...)

# Derive last dose date (same function, different mode)
derive_vars_merged(dataset_add = ex, mode = "last", ...)
```

#### Principle 5: Traceability

Admiral encourages inline documentation:

``` r
adsl <- dm %>%
  # Treatment start date: first dose from EX
  derive_vars_merged(
    dataset_add = ex,
    by_vars = exprs(STUDYID, USUBJID),
    new_vars = exprs(TRTSDTM = EXSTDTM),
    mode = "first",
    filter_add = EXDOSE > 0
  )
```

### 4.3 Admiral vs Traditional SAS Macros

| Aspect | SAS Macros (Legacy) | Admiral (Modern) |
|------------------------|------------------------|------------------------|
| **Source code** | Hidden in compiled `.sas7bcat` | Fully open on GitHub |
| **Collaboration** | Company-specific | Cross-company |
| **Debugging** | Difficult (macro expansion errors) | Easy (standard R debugging) |
| **Testing** | Manual | Automated with `testthat` |
| **Community** | Internal only | 1000+ GitHub stars, Slack |
| **Learning curve** | Steep (macro language) | Moderate (tidyverse knowledge) |
| **Validation** | Company validates internally | Shared validation evidence |

------------------------------------------------------------------------

## 5 Admiral's Four Core Function Categories

Admiral has **170+ functions**, but they fall into **4 main categories**. Learning these categories is like learning grammar - it lets you construct any derivation.

### 5.1 Category 1: `derive_vars_*` - Add Multiple Variables

**Pattern:** `derive_vars_*(dataset, dataset_add, by_vars, new_vars, ...)`

**Purpose:** Add multiple variables to a dataset by merging from another dataset

**Key Functions:**

#### `derive_vars_merged()`

Merge variables from another dataset (most versatile function in Admiral)

``` r
# Pseudo-code example
adsl <- dm %>%
  derive_vars_merged(
    dataset_add = ex,
    by_vars = exprs(USUBJID),
    new_vars = exprs(TRTSDTM = EXSTDTM),
    mode = "first",       # or "last"
    order = exprs(EXSTDTM)
  )
```

**Use cases:** - Treatment dates from EX - Disposition reasons from DS - Baseline values from LB/VS

#### `derive_vars_dt()` and `derive_vars_dtm()`

Convert character dates to Date or POSIXct

``` r
adsl <- adsl %>%
  derive_vars_dt(dtc = RFSTDTC, new_vars_prefix = "RFST")
  # Creates RFSTDT
```

**Use cases:** - All date conversions in ADSL, ADAE, ADLB, etc. - Handles partial dates with imputation

#### `derive_vars_merged_lookup()`

Add variables from a lookup table (e.g., parameter metadata)

``` r
adlb <- lb %>%
  derive_vars_merged_lookup(
    dataset_add = param_lookup,
    by_vars = exprs(PARAMCD),
    new_vars = exprs(PARAM, PARAMN)
  )
```

**Use cases:** - Parameter labels in BDS datasets - Decode values from reference tables

### 5.2 Category 2: `derive_var_*` - Add One Variable

**Pattern:** `derive_var_*(dataset, ...)`

**Purpose:** Add a single derived variable to a dataset

**Key Functions:**

#### `derive_var_trtdurd()`

Treatment duration (days)

``` r
adsl <- adsl %>%
  derive_var_trtdurd()
  # Creates TRTDURD = TRTEDT - TRTSDT + 1
```

#### `derive_var_merged_exist_flag()`

Flag if a condition exists in another dataset

``` r
adsl <- adsl %>%
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars = exprs(USUBJID),
    new_var = SAFFL,
    condition = EXDOSE > 0
  )
  # SAFFL = "Y" if subject has any EX record with dose > 0
```

**Use cases:** - Population flags (SAFFL, ITTFL) - Event flags (had SAE, had dose interruption)

#### `derive_var_extreme_flag()`

Flag first/last/min/max record within a group

``` r
adlb <- adlb %>%
  derive_var_extreme_flag(
    by_vars = exprs(USUBJID, PARAMCD),
    order = exprs(ADT),
    new_var = ABLFL,
    mode = "last",
    filter = ADT < TRTSDT
  )
  # Flags last pre-treatment lab as baseline
```

**Use cases:** - Baseline flags (ABLFL) - First occurrence flags (AOCCFL)

### 5.3 Category 3: `derive_param_*` - Add Computed Parameters

**Pattern:** `derive_param_*(dataset, ...)`

**Purpose:** Add **new rows** with computed parameters (BDS-specific)

**Key Functions:**

#### `derive_param_computed()`

Create a parameter computed from other parameters

``` r
advs <- advs %>%
  derive_param_computed(
    by_vars = exprs(USUBJID, AVISIT),
    parameters = c("SYSBP", "DIABP"),
    set_values_to = exprs(
      PARAMCD = "MAP",
      PARAM = "Mean Arterial Pressure",
      AVAL = (SYSBP + 2 * DIABP) / 3
    )
  )
  # Adds new rows with PARAMCD = "MAP"
```

**Use cases:** - Derived parameters (MAP from BP, BMI from height/weight) - Ratios (LDL/HDL)

#### `derive_param_exist_flag()`

Create a parameter indicating if an event occurred

``` r
adlb <- adlb %>%
  derive_param_exist_flag(
    condition = ANRIND == "HIGH",
    set_values_to = exprs(
      PARAMCD = "ANYHIGH",
      PARAM = "Any High Lab Value"
    )
  )
  # Adds new rows flagging subjects with any high lab
```

### 5.4 Category 4: `derive_extreme_records()` and Utilities

**Pattern:** `derive_extreme_records(dataset, dataset_add, ...)`

**Purpose:** Compute records (e.g., LOCF, average across visits)

**Key Functions:**

#### `derive_extreme_records()`

Create records by selecting extremes (first, last, min, max)

``` r
adlb_locf <- adlb %>%
  derive_extreme_records(
    dataset_add = adlb,
    by_vars = exprs(USUBJID, PARAMCD, AVISITN),
    order = exprs(ADT),
    mode = "last",
    filter_add = !is.na(AVAL)
  )
  # LOCF imputation: carry forward last observed value
```

**Use cases:** - LOCF/WOCF imputation - Summary records (average of multiple readings)

#### Higher-Order Functions (HOFs)

**`call_derivation()`** Apply the same derivation with different parameters

``` r
adsl <- call_derivation(
  dataset = adsl,
  derivation = derive_var_extreme_flag,
  variable_params = list(
    params(new_var = AOCCFL, filter = !is.na(ASTDT)),
    params(new_var = AOCCSFL, filter = AESER == "Y")
  ),
  by_vars = exprs(USUBJID)
)
# Derives two flags with one call
```

**`restrict_derivation()`** Apply derivation only to subset

``` r
adae <- restrict_derivation(
  dataset = adae,
  derivation = derive_var_extreme_flag,
  args = params(...),
  filter = TRTEMFL == "Y"
)
# Only applies to treatment-emergent AEs
```

**`slice_derivation()`** Apply different derivations to different slices

``` r
adlb <- slice_derivation(
  dataset = adlb,
  derivation = derive_var_base,
  args = params(by_vars = exprs(USUBJID, PARAMCD)),
  slices = list(
    slice(filter = PARAMCD %in% c("ALT", "AST"), args = params(source_var = AVAL)),
    slice(filter = PARAMCD == "BILI", args = params(source_var = AVALC))
  )
)
```

------------------------------------------------------------------------

## 6 Mapping ADaM Datasets to Admiral Patterns

This table is your **rosetta stone** for ADaM creation. Memorize this mapping.

| ADaM Dataset | SDTM Source(s) | Structure | Key Admiral Functions | Primary Use |
|---------------|---------------|---------------|---------------|---------------|
| **ADSL** | DM, EX, DS, AE, VS, LB | Subject-level (1 row/subject) | `derive_vars_merged()`, `derive_var_merged_exist_flag()`, `derive_var_trtdurd()` | Demographics, population flags, treatment dates |
| **ADAE** | AE, ADSL | OCCDS (1 row/event) | `derive_vars_merged()` (ADSL), `derive_var_extreme_flag()` (TRTEMFL, AOCCFL), `derive_vars_dt()` | Safety tables, AE listings, time-to-first-AE |
| **ADLB** | LB, ADSL | BDS (1 row/subject/param/visit) | `derive_var_extreme_flag()` (ABLFL), `derive_var_base()`, `derive_var_chg()`, `derive_var_pchg()` | Lab result tables, shift tables, change analysis |
| **ADVS** | VS, ADSL | BDS | `derive_var_extreme_flag()` (ABLFL), `derive_var_base()`, `derive_var_chg()`, `derive_vars_joined()` (visit mapping) | Vital signs tables, plots |
| **ADTTE** | DM, DS, AE, ADSL | BDS-like (1 row/subject/endpoint) | `event_source()`, `censor_source()`, `derive_param_tte()` | Kaplan-Meier, Cox regression, time-to-event |
| **ADCM** | CM, ADSL | OCCDS | `derive_vars_merged()` (ADSL), `derive_vars_dt()`, period flags (CMPREFL, CMCONFL) | Concomitant meds tables |
| **ADEX** | EX, ADSL | BDS (1 row/subject/param/dose) | `derive_vars_merged()` (ADSL), `derive_param_exposure()` (cumulative dose, intensity) | Dose intensity, compliance |
| **ADRS** | RS, TR, TU, ADSL | BDS (oncology-specific) | `admiral.onco` package: `derive_param_response()`, `derive_param_confirmed_resp()` | Tumor response, RECIST criteria |

------------------------------------------------------------------------

## 7 Hands-On: Exploring Admiral Function Taxonomy

Let's load Admiral and explore its structure.

```{r setup}
# Load packages
library(admiral)
library(dplyr)
library(pharmaversesdtm)  # For example SDTM data
library(tibble)

# Check Admiral version
packageVersion("admiral")
```

### 7.1 Listing All Admiral Functions

```{r admiral-functions}
# Get all Admiral functions
admiral_functions <- ls("package:admiral")

# Filter by category
cat("derive_vars_* functions (add multiple variables):\n")
admiral_functions[grepl("^derive_vars_", admiral_functions)] %>% head(10)

cat("\nderive_var_* functions (add one variable):\n")
admiral_functions[grepl("^derive_var_[^s]", admiral_functions)] %>% head(10)

cat("\nderive_param_* functions (add computed parameters):\n")
admiral_functions[grepl("^derive_param_", admiral_functions)] %>% head(10)

cat("\nderive_extreme_* functions:\n")
admiral_functions[grepl("^derive_extreme_", admiral_functions)]

cat("\nHigher-order functions (HOFs):\n")
admiral_functions[admiral_functions %in% c("call_derivation", "restrict_derivation", "slice_derivation")]
```

### 7.2 Examining Function Arguments

```{r function-args}
# Example: What arguments does derive_vars_merged() take?
cat("derive_vars_merged() arguments:\n")
cat(paste(names(formals(derive_vars_merged)), collapse = ", "))

cat("\n\nderive_var_extreme_flag() arguments:\n")
cat(paste(names(formals(derive_var_extreme_flag)), collapse = ", "))
```

### 7.3 Finding the Right Function

**Scenario:** "I need to flag the first adverse event per subject."

**Thought process:** 1. First occurrence → need a **flag** → `derive_var_*` 2. First/extreme → `derive_var_extreme_flag()` 3. Check help: `?derive_var_extreme_flag`

```{r find-function, eval=FALSE}
# Search Admiral help
?derive_var_extreme_flag

# Or browse all Admiral vignettes
browseVignettes("admiral")
```

------------------------------------------------------------------------

## 8 Deliverable: ADaM Dataset Mapping Matrix

Your deliverable today is a **mapping matrix** documenting the architecture of your future ADaM package.

```{r deliverable}
# Create mapping matrix
adam_mapping <- tribble(
  ~ADAM_Dataset, ~SDTM_Sources, ~Structure_Type, ~Admiral_Function_1, ~Admiral_Function_2, ~Admiral_Function_3, ~Clinical_Purpose,

  "ADSL", 
  "DM (primary), EX, DS, AE, VS, LB", 
  "Subject-level", 
  "derive_vars_merged()", 
  "derive_var_merged_exist_flag()", 
  "derive_var_trtdurd()",
  "Demographics tables, population flags, treatment assignments",

  "ADAE",
  "AE (primary), ADSL",
  "OCCDS",
  "derive_vars_merged() [ADSL]",
  "derive_var_extreme_flag() [TRTEMFL]",
  "derive_vars_dt() [dates]",
  "Adverse event frequency tables, AE listings, safety summaries",

  "ADLB",
  "LB (primary), ADSL",
  "BDS",
  "derive_var_extreme_flag() [ABLFL]",
  "derive_var_base()",
  "derive_var_chg()",
  "Lab result tables, shift tables, change from baseline analysis",

  "ADVS",
  "VS (primary), ADSL",
  "BDS",
  "derive_var_extreme_flag() [ABLFL]",
  "derive_var_base()",
  "derive_vars_joined() [visit mapping]",
  "Vital signs tables, BP plots, vital sign changes",

  "ADTTE",
  "DS, AE, ADSL",
  "BDS-like (time-to-event)",
  "event_source()",
  "censor_source()",
  "derive_param_tte()",
  "Kaplan-Meier curves, Cox regression, survival analysis",

  "ADCM",
  "CM (primary), ADSL",
  "OCCDS",
  "derive_vars_merged() [ADSL]",
  "derive_vars_dt() [dates]",
  "Period flags (CMPREFL, CMCONFL)",
  "Concomitant medications tables, prior/concomitant distinction",

  "ADEX",
  "EX (primary), ADSL",
  "BDS (exposure metrics)",
  "derive_vars_merged() [ADSL]",
  "derive_param_exposure() [cumulative dose]",
  "derive_var_duration()",
  "Dose intensity tables, treatment compliance, exposure summaries",

  "ADRS",
  "RS, TR, TU, ADSL",
  "BDS (oncology)",
  "admiral.onco::derive_param_response()",
  "admiral.onco::derive_param_confirmed_resp()",
  "admiral.onco::derive_param_clinbenefit()",
  "Tumor response tables, RECIST assessments, best overall response"
)

# Display mapping
adam_mapping %>%
  select(ADAM_Dataset, SDTM_Sources, Structure_Type, Clinical_Purpose) %>%
  knitr::kable(caption = "ADaM Dataset Architecture Overview")

# Display function mappings
adam_mapping %>%
  select(ADAM_Dataset, Admiral_Function_1, Admiral_Function_2, Admiral_Function_3) %>%
  knitr::kable(caption = "Key Admiral Functions by ADaM Dataset")
```

### Save Mapping for Reference

```{r save-mapping}
# Save as CSV for future reference
write.csv(adam_mapping, "adam_dataset_mapping.csv", row.names = FALSE)

cat("✓ Mapping matrix saved to adam_dataset_mapping.csv\n")
cat("\nThis matrix will guide your ADaM builds in Days 16-30!\n")
```

------------------------------------------------------------------------

## 9 Admiral Input-Output Contract

Understanding Admiral's **input-output contract** prevents common errors.

### Rule 1: Functions Return Modified Data Frames

``` r
# Input: data frame
# Output: modified data frame (never modifies in place)

adsl_new <- adsl_old %>%
  derive_var_trtdurd()

# adsl_old is unchanged
# adsl_new has TRTDURD column added
```

### Rule 2: Functions Compose with Pipes

``` r
adsl <- dm %>%
  derive_vars_dt(...) %>%
  derive_vars_merged(...) %>%
  derive_var_trtdurd() %>%
  mutate(...)
```

**Each function:** - Takes a data frame as first argument - Returns a modified data frame - Can be inspected mid-pipeline

### Rule 3: Required Variables Must Exist

``` r
# This will error if TRTSDT or TRTEDT don't exist
adsl <- adsl %>%
  derive_var_trtdurd()

# Check required variables
required_vars <- exprs(TRTSDT, TRTEDT)
assert_data_frame(adsl, required_vars = required_vars)
```

### Rule 4: Functions Don't Modify Unexpected Variables

Admiral functions only add new variables or modify explicitly named variables. They never silently change existing variables.

``` r
# derive_var_trtdurd() only adds TRTDURD
# It never modifies TRTSDT, TRTEDT, or any other existing variable
```

------------------------------------------------------------------------

## 10 Navigating Admiral Documentation

### 10.1 Admiral Website

**URL:** <https://pharmaverse.github.io/admiral/>

**Key sections:** - **Get Started:** Quick introduction - **Vignettes:** Topic-based guides (ADSL, ADAE, ADLB, dates, etc.) - **Reference:** Alphabetical function list with examples - **Articles:** Advanced topics (Higher-Order Functions, metadata, programming strategy)

### 10.2 Function Help

``` r
# In R console
?derive_vars_merged
?derive_var_extreme_flag

# Or click "Help" tab in RStudio and search
```

### 10.3 Vignettes (Most Important!)

```{r browse-vignettes, eval=FALSE}
# List all vignettes
browseVignettes("admiral")

# Key vignettes to read:
# - "Creating ADSL"
# - "Creating a BDS Finding ADaM"
# - "Date and Time Imputation"
# - "Higher Order Functions"
```

### 10.4 GitHub Issues & Discussions

**URL:** <https://github.com/pharmaverse/admiral>

-   **Issues:** Bug reports and feature requests
-   **Discussions:** Q&A, best practices
-   Search before asking - your question may be answered

### 10.5 Pharmaverse Slack

-   **#admiral channel:** Active community support
-   Join at [pharmaverse.org](https://pharmaverse.org)

------------------------------------------------------------------------

## 11 Common Admiral Patterns (Cheat Sheet)

Memorize these patterns - they appear in 90% of ADaM derivations.

### Pattern 1: Merge Variables from Another Dataset

``` r
adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = ex,
    by_vars = exprs(USUBJID),
    new_vars = exprs(TRTSDTM = EXSTDTM),
    mode = "first",  # or "last"
    order = exprs(EXSTDTM),
    filter_add = EXDOSE > 0
  )
```

**Use for:** Treatment dates, disposition reasons, any variable from another SDTM domain

### Pattern 2: Flag if Condition Exists

``` r
adsl <- adsl %>%
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars = exprs(USUBJID),
    new_var = SAFFL,
    condition = EXDOSE > 0
  )
```

**Use for:** Population flags (SAFFL, ITTFL), event occurrence flags

### Pattern 3: Flag First/Last/Extreme Record

``` r
adlb <- adlb %>%
  derive_var_extreme_flag(
    by_vars = exprs(USUBJID, PARAMCD),
    order = exprs(ADT),
    new_var = ABLFL,
    mode = "last",
    filter = ADT < TRTSDT
  )
```

**Use for:** Baseline flags, first occurrence flags, min/max flags

### Pattern 4: Add Baseline Value

``` r
adlb <- adlb %>%
  derive_var_base(
    by_vars = exprs(USUBJID, PARAMCD),
    source_var = AVAL,
    new_var = BASE,
    filter = ABLFL == "Y"
  )
```

**Use for:** Carrying baseline forward in BDS datasets

### Pattern 5: Compute Change from Baseline

``` r
adlb <- adlb %>%
  derive_var_chg() %>%
  derive_var_pchg()
```

**Use for:** CHG, PCHG in all BDS datasets

### Pattern 6: Date Conversion with Imputation

``` r
adsl <- adsl %>%
  derive_vars_dt(
    dtc = RFSTDTC,
    new_vars_prefix = "RFST",
    highest_imputation = "M"
  )
```

**Use for:** All date conversions in all ADaM datasets

------------------------------------------------------------------------

## 12 ADaM Creation Workflow (Conceptual)

**Standard sequence for any ADaM dataset:**

```         
1. Start with primary SDTM domain
   ↓
2. Merge ADSL (for demographics, treatment, flags)
   ↓
3. Derive dates (convert DTC → DT/DTM)
   ↓
4. Derive analysis values (AVAL, AVALC)
   ↓
5. Derive baseline flags (ABLFL for BDS)
   ↓
6. Derive baseline values (BASE for BDS)
   ↓
7. Derive change from baseline (CHG, PCHG for BDS)
   ↓
8. Derive analysis flags (TRTEMFL, ANL01FL, etc.)
   ↓
9. Derive computed parameters (for BDS)
   ↓
10. Add metadata (variable labels, formats)
   ↓
11. Export to XPT
```

**Admiral makes each step explicit in your code.**

------------------------------------------------------------------------

## 13 Key Takeaways

### Technical Takeaways

1.  **Three ADaM structures:** ADSL (subject-level), BDS (parameters × visits), OCCDS (events)
2.  **Admiral is modular:** Small, composable functions that chain with `%>%`
3.  **Four function categories:** `derive_vars_*` (multi-var), `derive_var_*` (single-var), `derive_param_*` (computed parameters), `derive_extreme_*` (record computations)
4.  **Functions compose:** Each takes and returns a data frame
5.  **Transparency:** Every derivation is explicit in your script
6.  **Reusability:** Same functions work across datasets with different arguments

### Strategic Takeaways

7.  **ADSL is the foundation:** Build ADSL first, then all other ADaMs merge from it
8.  **Structure determines function choice:** BDS needs baseline derivations, OCCDS needs occurrence flags
9.  **Admiral documentation is extensive:** Vignettes are your best friend
10. **Community support:** Pharmaverse Slack, GitHub discussions
11. **Cross-company patterns:** What works at one company works at another

### Prepare for Tomorrow

12. **Day 16:** Hands-on ADSL creation (treatment dates, population flags)
13. **Day 17:** ADSL demographics and baseline values
14. **Days 18-24:** Build every ADaM type (ADAE, ADLB, ADVS, ADTTE, ADCM, ADEX)

------------------------------------------------------------------------

## 14 Additional Resources

### Official Admiral Resources

-   **Admiral Website:** <https://pharmaverse.github.io/admiral/>
-   **Admiral GitHub:** <https://github.com/pharmaverse/admiral>
-   **Pharmaverse Website:** <https://pharmaverse.org>
-   **Admiral Get Started:** <https://pharmaverse.github.io/admiral/articles/admiral.html>

### Key Vignettes (Read These!)

-   **Creating ADSL:** Full ADSL workflow
-   **Creating a BDS Finding ADaM:** ADLB/ADVS patterns
-   **Occurrence Datasets:** ADAE/ADCM patterns
-   **Higher Order Functions:** Advanced patterns
-   **Date and Time Imputation:** Complete date handling guide

### CDISC Standards

-   **ADaM IG v1.3:** <https://www.cdisc.org/standards/foundational/adam>
-   **ADaM Examples:** <https://github.com/cdisc-org/sdtm-adam-pilot-project>

### Community

-   **Pharmaverse Slack:** `#admiral` channel
-   **Pharmaverse Blog:** <https://pharmaverse.github.io/blog/>
-   **Pharmaverse Examples:** <https://pharmaverse.github.io/examples/>

------------------------------------------------------------------------

## 15 Homework (Optional)

Prepare for Day 16 (hands-on ADSL) by:

1.  **Read Admiral ADSL vignette:** <https://pharmaverse.github.io/admiral/articles/adsl.html>
2.  **Explore `pharmaversesdtm` datasets:** Load DM, EX, DS and examine their structure
3.  **Review your ADaM mapping matrix:** Understand how ADSL sources from multiple SDTM domains
4.  **Browse Admiral function reference:** Get familiar with function names and categories

**Tomorrow we build ADSL from scratch!**

------------------------------------------------------------------------

**End of Day 15**

*Today you mastered ADaM architecture and Admiral philosophy.*\
*Tomorrow you'll put it into practice building your first ADaM dataset: ADSL!*
