---
title: "Day 18: ADAE - Adverse Events Analysis Dataset"
subtitle: "Week 3, Day 18: OCCDS structure with treatment emergent flags"


format:
  html:
    theme:
      light: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css

editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---
::: {style="margin-bottom:12px;"}
<a href="../index.html" class="resource-link"> Back to Roadmap</a>
:::

## Welcome to ADAE

ADAE is the backbone of safety reporting. Every safety table flows from this dataset.

**Today's focus:** Build ADAE with treatment emergent flags and occurrence indicators.

------------------------------------------------------------------------

## Setup

```{r setup}
library(admiral)
library(pharmaversesdtm)
library(pharmaverseadam)
library(dplyr)
library(lubridate)
library(stringr)
library(xportr)

# Load SDTM domains
ae <- pharmaversesdtm::ae
ex <- pharmaversesdtm::ex

# Load pre-built ADSL
adsl <- pharmaverseadam::adsl

cat("Loaded AE:", nrow(ae), "records\n")
cat("Loaded ADSL:", nrow(adsl), "subjects\n")
```

**Why pharmaverseadam::adsl?**

-   Pre-built reference ADSL
-   Has all required variables
-   Saves rebuild time

------------------------------------------------------------------------

## Step 1: Merge ADSL to AE

```{r merge-adsl}
# Merge treatment dates to AE
adae <- ae %>%
  admiral::derive_vars_merged(
    dataset_add = adsl,
    new_vars = exprs(TRTSDT, TRTEDT, TRT01A),
    by_vars = exprs(STUDYID, USUBJID)
  )

cat("\nADAE initialized:", nrow(adae), "records\n")

# Check
adae %>%
  select(USUBJID, AETERM, TRTSDT, TRTEDT) %>%
  head(5)
```

------------------------------------------------------------------------

## Step 2: Derive AE Start Date

```{r derive-astdt}
# Derive AE start date
adae <- adae %>%
  admiral::derive_vars_dt(
    new_vars_prefix = "AST",
    dtc = AESTDTC,
    highest_imputation = "M",
    flag_imputation = "auto"
  )

cat("\nAE start date derived\n")

# Check
adae %>%
  select(USUBJID, AESEQ, AESTDTC, ASTDT, ASTDTF) %>%
  head(5)
```

**Imputation:**

-   Partial dates like "2024-03" → "2024-03-01"
-   ASTDTF flag shows if imputed

------------------------------------------------------------------------

## Step 3: Derive AE End Date

```{r derive-aendt}
# Derive AE end date
adae <- adae %>%
  admiral::derive_vars_dt(
    new_vars_prefix = "AEN",
    dtc = AEENDTC,
    highest_imputation = "M",
    date_imputation = "last",
    flag_imputation = "auto"
  )

cat("\nAE end date derived\n")

# Check
adae %>%
  select(USUBJID, AESEQ, AEENDTC, AENDT, AENDTF) %>%
  head(5)
```

**Difference from start:**

-   `date_imputation = "last"` → "2024-03" becomes "2024-03-31"

------------------------------------------------------------------------

## Step 4: Derive Study Days

```{r derive-study-days}
# Calculate relative days
adae <- adae %>%
  admiral::derive_vars_dy(
    reference_date = TRTSDT,
    source_vars = exprs(ASTDT, AENDT)
  )

cat("\nStudy days derived\n")

# Check
adae %>%
  select(USUBJID, TRTSDT, ASTDT, ASTDY, AENDT, AENDY) %>%
  filter(!is.na(ASTDY)) %>%
  head(5)
```

**Study day convention:**

-   Day 1 = first treatment day
-   No Day 0

------------------------------------------------------------------------

## Step 5: Derive Duration

```{r derive-duration}
# Calculate duration
adae <- adae %>%
  mutate(
    ADURN = case_when(
      !is.na(ASTDT) & !is.na(AENDT) ~ as.numeric(AENDT - ASTDT) + 1,
      TRUE ~ NA_real_
    )
  )

cat("\nDuration calculated\n")

# Summary
adae %>%
  filter(!is.na(ADURN)) %>%
  summarise(
    N = n(),
    Mean = round(mean(ADURN), 1),
    Median = median(ADURN),
    Min = min(ADURN),
    Max = max(ADURN)
  )
```

**Formula:** AENDT - ASTDT + 1

------------------------------------------------------------------------

## Step 6: Derive Treatment Emergent Flag

```{r derive-trtemfl}
# Flag AEs that occurred during treatment
adae <- adae %>%
  mutate(
    TRTEMFL = case_when(
      !is.na(ASTDT) & !is.na(TRTSDT) & ASTDT >= TRTSDT ~ "Y",
      TRUE ~ NA_character_
    )
  )

cat("\nTreatment emergent flag derived\n")

# Distribution
adae %>%
  count(TRTEMFL)
```

**Logic:**

-   AE started on or after treatment start
-   Used in all safety tables

------------------------------------------------------------------------

## Step 7: Derive Severity Numeric

```{r derive-severity}
# Create numeric severity
adae <- adae %>%
  mutate(
    ASEV = AESEV,
    ASEVN = case_when(
      AESEV == "MILD" ~ 1,
      AESEV == "MODERATE" ~ 2,
      AESEV == "SEVERE" ~ 3,
      TRUE ~ NA_real_
    )
  )

cat("\nSeverity numeric created\n")

# Distribution
adae %>%
  count(ASEV, ASEVN)
```

------------------------------------------------------------------------

## Step 8: Derive First Occurrence Flag

```{r derive-aoccfl}
# Flag first AE per subject
adae <- adae %>%
  arrange(USUBJID, ASTDT, AESEQ) %>%
  group_by(USUBJID) %>%
  mutate(
    AOCCFL = if_else(row_number() == 1, "Y", NA_character_)
  ) %>%
  ungroup()

cat("\nFirst occurrence flag derived\n")

# Check
adae %>%
  filter(AOCCFL == "Y") %>%
  select(USUBJID, AETERM, ASTDT, AOCCFL) %>%
  head(5)
```

**Logic:**

-   First AE record per subject
-   Used for "subjects with at least one AE"

------------------------------------------------------------------------

## Step 9: Derive Serious AE Flag

```{r derive-serious}
# Copy serious flag
adae <- adae %>%
  mutate(
    ASER = AESER,
    ASERN = if_else(AESER == "Y", 1, 0)
  )

cat("\nSerious flag created\n")

# Distribution
adae %>%
  count(ASER, ASERN)
```

------------------------------------------------------------------------

## Step 10: Derive Analysis Flag

```{r derive-anl01fl}
# Primary analysis flag
adae <- adae %>%
  mutate(
    ANL01FL = if_else(TRTEMFL == "Y", "Y", NA_character_)
  )

cat("\nAnalysis flag created\n")

# Distribution
adae %>%
  count(ANL01FL, TRTEMFL)
```

**ANL01FL:**

-   Primary analysis records
-   Typically = TRTEMFL

------------------------------------------------------------------------

## Validation

```{r validate}
cat("\n=== ADAE Validation ===\n\n")

# Check 1: Start <= End
check1 <- adae %>%
  filter(!is.na(ASTDT) & !is.na(AENDT) & ASTDT > AENDT)
cat("Check 1 - ASTDT > AENDT:", nrow(check1), "\n")

# Check 2: Duration
check2 <- adae %>%
  filter(!is.na(ADURN)) %>%
  mutate(
    Expected = as.numeric(AENDT - ASTDT) + 1,
    Match = ADURN == Expected
  ) %>%
  filter(!Match)
cat("Check 2 - ADURN mismatch:", nrow(check2), "\n")

# Check 3: One AOCCFL per subject
check3 <- adae %>%
  filter(AOCCFL == "Y") %>%
  count(USUBJID) %>%
  filter(n > 1)
cat("Check 3 - Multiple AOCCFL:", nrow(check3), "\n")

cat("\n✓ Validation complete\n")
```

------------------------------------------------------------------------

## Summary

```{r summary}
cat("\n=== ADAE Summary ===\n\n")
cat("Total records:", nrow(adae), "\n")
cat("Variables:", ncol(adae), "\n\n")

# By treatment
adae %>%
  filter(TRTEMFL == "Y") %>%
  group_by(TRT01A) %>%
  summarise(
    N_AEs = n(),
    N_Subjects = n_distinct(USUBJID),
    N_Serious = sum(ASER == "Y", na.rm = TRUE),
    .groups = "drop"
  )

# By severity
adae %>%
  filter(TRTEMFL == "Y") %>%
  count(ASEV)
```

------------------------------------------------------------------------

## Export

```{r export}
# Add labels
attr(adae$ASTDT, "label") <- "Analysis Start Date"
attr(adae$ASTDY, "label") <- "Analysis Start Relative Day"
attr(adae$AENDT, "label") <- "Analysis End Date"
attr(adae$AENDY, "label") <- "Analysis End Relative Day"
attr(adae$ADURN, "label") <- "Analysis Duration (Days)"
attr(adae$TRTEMFL, "label") <- "Treatment Emergent Flag"
attr(adae$ASEV, "label") <- "Analysis Severity"
attr(adae$ASEVN, "label") <- "Analysis Severity (N)"
attr(adae$AOCCFL, "label") <- "First Occurrence Flag"
attr(adae$ANL01FL, "label") <- "Analysis Record Flag 01"

# Export - xportr is already loaded via library()
xportr_write(adae, path = "adae.xpt", domain = "ADAE")

cat("\n✓ ADAE exported to: adae.xpt\n")
```

------------------------------------------------------------------------

## Key Takeaways

1.  Use pharmaverseadam::adsl for reference ADSL
2.  Impute start date "first", end date "last"
3.  TRTEMFL = AE on or after treatment start
4.  AOCCFL = first AE per subject (arrange + group_by + row_number)
5.  Always validate dates and flags

------------------------------------------------------------------------

## Next Steps

**Day 19:** ADLB - Lab Analysis Dataset\
**Day 20:** ADVS - Vitals Analysis Dataset\
**Day 21:** ADTTE - Time-to-Event

------------------------------------------------------------------------

## Resources

-   OCCDS Vignette: <https://pharmaverse.github.io/admiral/articles/occds.html>
-   ADAE Example: <https://pharmaverse.github.io/examples/adam/adae.html>
-   pharmaverseadam: <https://pharmaverse.github.io/pharmaverseadam/>
-   xportr package: <https://atorus-research.github.io/xportr/>

------------------------------------------------------------------------

**End of Day 18**

<a href="https://visitorbadge.io/status?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse"><img src="https://api.visitorbadge.io/api/visitors?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse&label=Visits%20till%20date&countColor=%23263759&style=plastic&labelStyle=none" /></a>
