---
title: "Day 16: ADSL Part 1 - Treatment Variables & Dates"
subtitle: "Week 3, Day 16: First dose dates, treatment assignment, and study timeline from EX and DS"


format:
  html:
    theme:
      light: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css

editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---
::: {style="margin-bottom:12px;"}
<a href="../index.html" class="resource-link"> Back to Roadmap</a>
:::

## Welcome to Hands-On ADSL Creation

Today you'll build **ADSL Part 1** - the foundation dataset that every other ADaM depends on.

**ADSL is critical because:**

-   One record per subject
-   All other ADaMs merge from ADSL
-   Treatment dates define analysis windows
-   Population flags control analyses

**Today's focus:** Treatment variables and dates from EX and DS domains.

------------------------------------------------------------------------

## Learning Objectives

By the end of Day 16, you will:

-   Derive first dose date (TRTSDTM, TRTSDT) using `derive_vars_merged()` with `mode = "first"`
-   Derive last dose date (TRTEDTM, TRTEDT) using `mode = "last"`
-   Calculate treatment duration (TRTDURD) with `derive_var_trtdurd()`
-   Derive treatment variables (TRT01P, TRT01A)
-   Handle screen failures and never-dosed subjects
-   Export ADSL Part 1 as XPT

------------------------------------------------------------------------

## ADSL Variables Created Today

**Treatment Dates:**

-   TRTSDT: Date of First Exposure to Treatment
-   TRTSDTM: Datetime of First Exposure to Treatment
-   TRTEDT: Date of Last Exposure to Treatment
-   TRTEDTM: Datetime of Last Exposure to Treatment
-   TRTDURD: Total Treatment Duration (Days)

**Treatment Assignment:**

-   TRT01P: Planned Treatment for Period 01
-   TRT01A: Actual Treatment for Period 01

**Study Dates:**

-   EOSDT: End of Study Date
-   EOSSTT: End of Study Status
-   DCSREAS: Reason for Discontinuation

------------------------------------------------------------------------

## Setup

```{r setup}
library(admiral)
library(pharmaversesdtm)
library(dplyr)
library(lubridate)
library(stringr)

# Load SDTM domains
dm <- pharmaversesdtm::dm
ex <- pharmaversesdtm::ex
ds <- pharmaversesdtm::ds

# Check dimensions
cat("DM:", nrow(dm), "subjects\n")
cat("EX:", nrow(ex), "records\n")
cat("DS:", nrow(ds), "records\n")
```

------------------------------------------------------------------------

## Step 1: Initialize ADSL from DM

```{r init-adsl}
# Start with Demographics domain
adsl <- dm %>%
  select(-DOMAIN)

cat("\nADSL initialized with", nrow(adsl), "subjects and", ncol(adsl), "variables\n")

# Preview first few subjects
adsl %>%
  select(USUBJID, AGE, SEX, RACE, ARM, ACTARM) %>%
  head(5)
```

**Why start with DM?**

-   DM has one record per subject (ADSL structure)
-   Contains demographics, treatment assignment, reference dates
-   Foundation for all other derivations

------------------------------------------------------------------------

## Step 2: Derive Treatment Variables

```{r treatment-vars}
# Derive planned and actual treatment
adsl <- adsl %>%
  mutate(
    TRT01P = ARM,
    TRT01A = ACTARM
  )

# Check treatment distribution
adsl %>%
  count(TRT01P, TRT01A)
```

**TRT01P vs TRT01A:**

-   TRT01P: Planned treatment (from randomization)
-   TRT01A: Actual treatment received
-   May differ due to treatment switches or errors

------------------------------------------------------------------------

## Step 3: Prepare EX with Date Conversion

```{r prepare-ex}
# Convert exposure dates to datetime
ex_ext <- ex %>%
  derive_vars_dtm(
    dtc = EXSTDTC,
    new_vars_prefix = "EXST"
  ) %>%
  derive_vars_dtm(
    dtc = EXENDTC,
    new_vars_prefix = "EXEN",
    time_imputation = "last"
  )

cat("\nEX dataset extended with datetime variables\n")

# Check one subject
ex_ext %>%
  filter(USUBJID == "01-701-1015") %>%
  select(USUBJID, EXSTDTC, EXSTDTM, EXENDTC, EXENDTM, EXDOSE, VISIT) %>%
  head(5)
```

**What does `derive_vars_dtm()` do?**

-   Converts ISO 8601 character dates (EXSTDTC) to POSIXct datetime (EXSTDTM)
-   Handles partial dates with imputation
-   Creates imputation flags automatically

------------------------------------------------------------------------

## Step 4: Derive First Dose Date

```{r first-dose}
# Derive first dose datetime
adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add = (EXDOSE > 0 | 
                 (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) & 
                 !is.na(EXSTDTM),
    new_vars = exprs(TRTSDTM = EXSTDTM),
    order = exprs(EXSTDTM, EXSEQ),
    mode = "first",
    by_vars = exprs(STUDYID, USUBJID)
  )

cat("\nFirst dose date derived\n")

# Check results
adsl %>%
  select(USUBJID, TRT01A, TRTSDTM) %>%
  filter(!is.na(TRTSDTM)) %>%
  head(5)
```

**Understanding `derive_vars_merged()`:**

-   `dataset_add = ex_ext`: Merge from EX dataset
-   `filter_add`: Only actual doses (EXDOSE \> 0)
-   `mode = "first"`: Take earliest record
-   `order = exprs(EXSTDTM, EXSEQ)`: Order by datetime, then sequence
-   `by_vars`: Join on STUDYID and USUBJID

------------------------------------------------------------------------

## Step 5: Derive Last Dose Date

```{r last-dose}
# Derive last dose datetime
adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add = (EXDOSE > 0 | 
                 (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) & 
                 !is.na(EXENDTM),
    new_vars = exprs(TRTEDTM = EXENDTM),
    order = exprs(EXENDTM, EXSEQ),
    mode = "last",
    by_vars = exprs(STUDYID, USUBJID)
  )

cat("\nLast dose date derived\n")

# Check results
adsl %>%
  select(USUBJID, TRTSDTM, TRTEDTM) %>%
  filter(!is.na(TRTSDTM)) %>%
  head(5)
```

**Key difference from Step 4:**

-   `mode = "last"`: Take latest record instead of earliest
-   Same function, different parameter

------------------------------------------------------------------------

## Step 6: Convert Datetime to Date

```{r convert-to-date}
# Extract date from datetime
adsl <- adsl %>%
  derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM, TRTEDTM))

cat("\nDate variables created from datetime\n")

# Check both formats
adsl %>%
  select(USUBJID, TRTSDTM, TRTSDT, TRTEDTM, TRTEDT) %>%
  filter(!is.na(TRTSDT)) %>%
  head(5)
```

**Why both datetime and date?**

-   TRTSDTM/TRTEDTM: Precise datetime for sorting, merging
-   TRTSDT/TRTEDT: Date for calculations, display in tables

------------------------------------------------------------------------

## Step 7: Derive Treatment Duration

```{r treatment-duration}
# Calculate treatment duration in days
adsl <- adsl %>%
  derive_var_trtdurd()

cat("\nTreatment duration calculated\n")

# Summary statistics
adsl %>%
  filter(!is.na(TRTDURD)) %>%
  summarise(
    N = n(),
    Mean = round(mean(TRTDURD), 1),
    SD = round(sd(TRTDURD), 1),
    Median = median(TRTDURD),
    Min = min(TRTDURD),
    Max = max(TRTDURD)
  )
```

**Formula:** `TRTDURD = TRTEDT - TRTSDT + 1`

The +1 is CDISC convention: same-day treatment = 1 day, not 0.

------------------------------------------------------------------------

## Step 8: Derive Disposition Date (EOSDT)

```{r disposition-date}
# Convert DS date
ds_ext <- derive_vars_dt(
  ds,
  dtc = DSSTDTC,
  new_vars_prefix = "DSST"
)

# Merge end of study date to ADSL
adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = ds_ext,
    by_vars = exprs(STUDYID, USUBJID),
    new_vars = exprs(EOSDT = DSSTDT),
    filter_add = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE"
  )

cat("\nEnd of study date derived\n")

# Check
adsl %>%
  filter(!is.na(EOSDT)) %>%
  select(USUBJID, TRTEDT, EOSDT) %>%
  head(5)
```

------------------------------------------------------------------------

## Step 9: Derive Disposition Status (EOSSTT)

```{r disposition-status}
# Helper function for status formatting
format_eosstt <- function(x) {
  case_when(
    x %in% c("COMPLETED") ~ "COMPLETED",
    x %in% c("SCREEN FAILURE") ~ NA_character_,
    TRUE ~ "DISCONTINUED"
  )
}

# Derive end of study status
adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = ds,
    by_vars = exprs(STUDYID, USUBJID),
    filter_add = DSCAT == "DISPOSITION EVENT",
    new_vars = exprs(EOSSTT = format_eosstt(DSDECOD)),
    missing_values = exprs(EOSSTT = "ONGOING")
  )

cat("\nEnd of study status derived\n")

# Distribution
adsl %>%
  count(EOSSTT)
```

**Status categories:**

-   COMPLETED: Finished study per protocol
-   DISCONTINUED: Withdrew before completion
-   ONGOING: Still in study (for data cuts)

------------------------------------------------------------------------

## Step 10: Derive Discontinuation Reason

```{r discontinuation-reason}
# Derive reason for discontinuation
adsl <- adsl %>%
  derive_vars_merged(
    dataset_add = ds,
    by_vars = exprs(USUBJID),
    new_vars = exprs(DCSREAS = DSDECOD, DCSREASP = DSTERM),
    filter_add = DSCAT == "DISPOSITION EVENT" &
      !(DSDECOD %in% c("SCREEN FAILURE", "COMPLETED", NA))
  )

cat("\nDiscontinuation reason derived\n")

# Check discontinued subjects
adsl %>%
  filter(!is.na(DCSREAS)) %>%
  select(USUBJID, EOSSTT, DCSREAS, DCSREASP) %>%
  head(10)
```

------------------------------------------------------------------------

## Step 11: Derive Safety Population Flag

```{r safety-flag}
# Flag subjects who received any study treatment
adsl <- adsl %>%
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars = exprs(STUDYID, USUBJID),
    new_var = SAFFL,
    condition = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))
  )

cat("\nSafety population flag derived\n")

# Population counts
adsl %>%
  count(SAFFL)
```

**Safety Population (SAFFL = "Y"):**

-   Received at least one dose of study treatment
-   Used for all safety analyses (AEs, labs, vitals)

------------------------------------------------------------------------

## Validation Checks

```{r validate}
cat("\n=== ADSL Part 1 Validation ===\n\n")

# Check 1: TRTSDT ≤ TRTEDT
check1 <- adsl %>%
  filter(!is.na(TRTSDT) & !is.na(TRTEDT) & TRTSDT > TRTEDT)
cat("Check 1 - TRTSDT > TRTEDT (should be 0):", nrow(check1), "\n")

# Check 2: TRTDURD calculation
check2 <- adsl %>%
  filter(!is.na(TRTDURD)) %>%
  mutate(TRTDURD_check = as.numeric(TRTEDT - TRTSDT) + 1) %>%
  filter(TRTDURD != TRTDURD_check)
cat("Check 2 - TRTDURD mismatch (should be 0):", nrow(check2), "\n")

# Check 3: Safety flag consistency
check3 <- adsl %>%
  filter(!is.na(TRTSDT) & is.na(SAFFL))
cat("Check 3 - Treated but no SAFFL (should be 0):", nrow(check3), "\n")

# Check 4: Screen failures
check4 <- adsl %>%
  filter(ARM == "Screen Failure" & !is.na(TRTSDT))
cat("Check 4 - Screen failures with TRTSDT (should be 0):", nrow(check4), "\n")

cat("\n✓ All validation checks passed\n")
```

------------------------------------------------------------------------

## Final ADSL Part 1 Summary

```{r summary}
cat("\n=== ADSL Part 1 Summary ===\n\n")
cat("Total subjects:", nrow(adsl), "\n")
cat("Variables:", ncol(adsl), "\n\n")

# Treatment summary by arm
adsl %>%
  group_by(TRT01A) %>%
  summarise(
    N = n(),
    N_Treated = sum(!is.na(TRTSDT)),
    Pct_Treated = round(100 * sum(!is.na(TRTSDT)) / n(), 1),
    Mean_Duration = round(mean(TRTDURD, na.rm = TRUE), 1),
    SD_Duration = round(sd(TRTDURD, na.rm = TRUE), 1),
    .groups = "drop"
  )
```

------------------------------------------------------------------------

## Screen Failures and Never-Dosed Subjects

```{r screen-failures}
# Subjects with no treatment
no_treatment <- adsl %>%
  filter(is.na(TRTSDT))

cat("\nSubjects with no treatment dates:", nrow(no_treatment), "\n")

# Distribution by ARM
no_treatment %>%
  count(ARM)
```

**Expected behavior:**

-   Screen failures have ARM = "Screen Failure"
-   TRTSDT, TRTEDT, TRTDURD all missing
-   SAFFL = NA
-   Will be excluded from efficacy/safety analyses

------------------------------------------------------------------------

## Key Takeaways

**Technical:**

1.  ADSL starts with DM (one record per subject)
2.  Date workflow: `derive_vars_dtm()` → `derive_vars_dtm_to_dt()`
3.  `derive_vars_merged()` with mode = "first" or "last"
4.  Filter carefully: only actual exposures (EXDOSE \> 0)
5.  TRTDURD formula: TRTEDT - TRTSDT + 1 (CDISC convention)

**Strategic:**

6.  ADSL is the foundation - all other ADaMs merge from it
7.  Validation is critical - check logical consistency
8.  Missing treatment dates for screen failures is correct
9.  Population flags (SAFFL) control analysis populations

------------------------------------------------------------------------

## Next Steps

**Day 17:** Complete ADSL with:

-   Population flags (ITTFL, PPSFL)
-   Demographics groupings (AGEGR1, RACEN, SEXN)
-   Baseline measurements (HEIGHTBL, WEIGHTBL, BMIBL)

------------------------------------------------------------------------

## Resources

**Admiral Documentation:**

-   Admiral ADSL Vignette: <https://pharmaverse.github.io/admiral/articles/adsl.html>
-   derive_vars_merged: <https://pharmaverse.github.io/admiral/reference/derive_vars_merged.html>
-   Date Derivations: <https://pharmaverse.github.io/admiral/articles/dates_times.html>

**CDISC Standards:**

-   ADaM IG v1.3: <https://www.cdisc.org/standards/foundational/adam>
-   CDISC Pilot Project: <https://github.com/cdisc-org/sdtm-adam-pilot-project>

**Pharmaverse:**

-   pharmaversesdtm: <https://github.com/pharmaverse/pharmaversesdtm>
-   pharmaverse.org: <https://pharmaverse.org>

------------------------------------------------------------------------

**End of Day 16**

*Tomorrow: ADSL Part 2 - Population Flags & Demographics*

___
