---
title: "Day 19: ADLB - Lab Analysis Dataset (BDS)"
subtitle: "Week 3, Day 19: BDS structure with baseline, change, and toxicity grading"


format:
  html:
    theme:
      light: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css

editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---
::: {style="margin-bottom:12px;"}
<a href="../index.html" class="resource-link"> Back to Roadmap</a>
:::

## Welcome to ADLB

ADLB is your first **BDS (Basic Data Structure)** dataset. Unlike ADSL (one row per subject) and ADAE (one row per event), BDS has **one row per subject per parameter per timepoint**.

**Today's focus:** Build ADLB with baseline flags, change from baseline, and toxicity grades.

------------------------------------------------------------------------

## Setup

```{r setup}
library(admiral)
library(pharmaversesdtm)
library(pharmaverseadam)
library(dplyr)
library(lubridate)
library(stringr)
library(xportr)

# Load SDTM and ADaM
lb <- pharmaversesdtm::lb
vs <- pharmaversesdtm::vs
adsl <- pharmaverseadam::adsl

cat("Loaded LB:", nrow(lb), "records\n")
cat("Loaded ADSL:", nrow(adsl), "subjects\n")
```

------------------------------------------------------------------------

## Step 1: Merge ADSL Variables

```{r merge-adsl}
# Start with LB and merge treatment info
adlb <- lb %>%
  admiral::derive_vars_merged(
    dataset_add = adsl,
    new_vars = exprs(TRTSDT, TRTEDT, TRT01A),
    by_vars = exprs(STUDYID, USUBJID)
  )

cat("\nADLB initialized:", nrow(adlb), "records\n")

# Check
adlb %>%
  select(USUBJID, LBTESTCD, LBDTC, TRTSDT) %>%
  head(5)
```

------------------------------------------------------------------------

## Step 2: Derive Analysis Date

```{r derive-adt}
# Derive analysis date from LBDTC
adlb <- adlb %>%
  admiral::derive_vars_dt(
    new_vars_prefix = "A",
    dtc = LBDTC
  )

cat("\nAnalysis date derived\n")

# Check
adlb %>%
  select(USUBJID, LBTESTCD, LBDTC, ADT) %>%
  head(5)
```

------------------------------------------------------------------------

## Step 3: Derive Study Day

```{r derive-ady}
# Calculate relative day
adlb <- adlb %>%
  admiral::derive_vars_dy(
    reference_date = TRTSDT,
    source_vars = exprs(ADT)
  )

cat("\nStudy day derived\n")

# Check
adlb %>%
  select(USUBJID, LBTESTCD, TRTSDT, ADT, ADY) %>%
  filter(!is.na(ADY)) %>%
  head(5)
```

------------------------------------------------------------------------

## Step 4: Create PARAM and AVAL

```{r create-param-aval}
# Create BDS variables
adlb <- adlb %>%
  mutate(
    PARAMCD = LBTESTCD,
    PARAM = LBTEST,
    AVAL = LBSTRESN,
    AVALU = LBSTRESU
  )

cat("\nPARAM and AVAL created\n")

# Check parameters
adlb %>%
  count(PARAMCD, PARAM) %>%
  head(10)
```

**BDS Key Variables:**

-   PARAMCD: Parameter code (ALT, AST, etc.)
-   PARAM: Parameter description
-   AVAL: Analysis value (numeric)
-   AVALU: Analysis value unit

------------------------------------------------------------------------

## Step 5: Derive Baseline Flag

```{r derive-ablfl}
# Flag last value before treatment start
adlb <- adlb %>%
  arrange(USUBJID, PARAMCD, ADT) %>%
  group_by(USUBJID, PARAMCD) %>%
  mutate(
    ABLFL = case_when(
      !is.na(AVAL) & ADT <= TRTSDT & row_number() == max(which(ADT <= TRTSDT)) ~ "Y",
      TRUE ~ NA_character_
    )
  ) %>%
  ungroup()

cat("\nBaseline flag derived\n")

# Check
adlb %>%
  filter(ABLFL == "Y") %>%
  select(USUBJID, PARAMCD, ADT, TRTSDT, AVAL, ABLFL) %>%
  head(5)
```

**ABLFL Logic:**

-   Last non-missing value on or before TRTSDT
-   One ABLFL="Y" per subject per parameter

------------------------------------------------------------------------

## Step 6: Derive Baseline Value

```{r derive-base}
# Merge baseline value to all records
adlb <- adlb %>%
  group_by(USUBJID, PARAMCD) %>%
  mutate(
    BASE = AVAL[ABLFL == "Y"][1]
  ) %>%
  ungroup()

cat("\nBaseline value derived\n")

# Check
adlb %>%
  select(USUBJID, PARAMCD, ADT, AVAL, ABLFL, BASE) %>%
  head(10)
```

------------------------------------------------------------------------

## Step 7: Derive Change from Baseline

```{r derive-chg}
# Calculate change and percent change
adlb <- adlb %>%
  mutate(
    CHG = case_when(
      !is.na(AVAL) & !is.na(BASE) ~ AVAL - BASE,
      TRUE ~ NA_real_
    ),
    PCHG = case_when(
      !is.na(CHG) & BASE != 0 ~ (CHG / BASE) * 100,
      TRUE ~ NA_real_
    )
  )

cat("\nChange from baseline calculated\n")

# Summary
adlb %>%
  filter(!is.na(CHG)) %>%
  group_by(PARAMCD) %>%
  summarise(
    N = n(),
    Mean_CHG = round(mean(CHG), 2),
    .groups = "drop"
  ) %>%
  head(5)
```

**Change Variables:**

-   CHG = AVAL - BASE
-   PCHG = (CHG / BASE) \* 100

------------------------------------------------------------------------

## Step 8: Derive Analysis Visit

```{r derive-avisit}
# Map VISIT to analysis visit
adlb <- adlb %>%
  mutate(
    AVISIT = case_when(
      VISIT == "SCREENING" ~ "Baseline",
      VISIT == "BASELINE" ~ "Baseline",
      TRUE ~ VISIT
    ),
    AVISITN = case_when(
      AVISIT == "Baseline" ~ 0,
      VISIT == "WEEK 2" ~ 2,
      VISIT == "WEEK 4" ~ 4,
      VISIT == "WEEK 8" ~ 8,
      TRUE ~ as.numeric(VISITNUM)
    )
  )

cat("\nAnalysis visit derived\n")

# Distribution
adlb %>%
  count(AVISIT, AVISITN) %>%
  head(5)
```

------------------------------------------------------------------------

## Step 9: Derive Reference Range Indicators

```{r derive-anrind}
# Create reference range indicators
adlb <- adlb %>%
  mutate(
    ANRIND = case_when(
      !is.na(AVAL) & !is.na(LBSTNRLO) & AVAL < LBSTNRLO ~ "LOW",
      !is.na(AVAL) & !is.na(LBSTNRHI) & AVAL > LBSTNRHI ~ "HIGH",
      !is.na(AVAL) ~ "NORMAL",
      TRUE ~ NA_character_
    ),
    BNRIND = case_when(
      !is.na(BASE) & !is.na(LBSTNRLO) & BASE < LBSTNRLO ~ "LOW",
      !is.na(BASE) & !is.na(LBSTNRHI) & BASE > LBSTNRHI ~ "HIGH",
      !is.na(BASE) ~ "NORMAL",
      TRUE ~ NA_character_
    )
  )

cat("\nReference range indicators derived\n")

# Distribution
adlb %>%
  count(PARAMCD, ANRIND) %>%
  head(10)
```

**Reference Range:**

-   ANRIND: Analysis reference range (LOW/NORMAL/HIGH)
-   BNRIND: Baseline reference range

------------------------------------------------------------------------

## Step 10: Derive Toxicity Grades

```{r derive-atoxgr}
# Simple toxicity grading for ALT as example
adlb <- adlb %>%
  mutate(
    ATOXGR = case_when(
      PARAMCD == "ALT" & !is.na(AVAL) & !is.na(LBSTNRHI) ~ case_when(
        AVAL > 20 * LBSTNRHI ~ "4",
        AVAL > 5 * LBSTNRHI ~ "3",
        AVAL > 3 * LBSTNRHI ~ "2",
        AVAL > LBSTNRHI ~ "1",
        TRUE ~ "0"
      ),
      TRUE ~ NA_character_
    ),
    BTOXGR = case_when(
      PARAMCD == "ALT" & !is.na(BASE) & !is.na(LBSTNRHI) ~ case_when(
        BASE > 20 * LBSTNRHI ~ "4",
        BASE > 5 * LBSTNRHI ~ "3",
        BASE > 3 * LBSTNRHI ~ "2",
        BASE > LBSTNRHI ~ "1",
        TRUE ~ "0"
      ),
      TRUE ~ NA_character_
    )
  )

cat("\nToxicity grades derived (ALT only)\n")

# Check ALT toxicity
adlb %>%
  filter(PARAMCD == "ALT", !is.na(ATOXGR)) %>%
  count(ATOXGR)
```

**NCI CTCAE Grading (ALT Example):**

-   Grade 0: Normal
-   Grade 1: \>ULN - 3×ULN
-   Grade 2: \>3×ULN - 5×ULN
-   Grade 3: \>5×ULN - 20×ULN
-   Grade 4: \>20×ULN

------------------------------------------------------------------------

## Step 11: Derive Analysis Sequence

```{r derive-aseq}
# Create sequence number
adlb <- adlb %>%
  arrange(USUBJID, PARAMCD, ADT) %>%
  group_by(USUBJID) %>%
  mutate(ASEQ = row_number()) %>%
  ungroup()

cat("\nAnalysis sequence derived\n")
```

------------------------------------------------------------------------

## Validation

```{r validate}
cat("\n=== ADLB Validation ===\n\n")

# Check 1: One ABLFL per subject-parameter
check1 <- adlb %>%
  filter(ABLFL == "Y") %>%
  count(USUBJID, PARAMCD) %>%
  filter(n > 1)
cat("Check 1 - Multiple ABLFL per subject-PARAMCD:", nrow(check1), "\n")

# Check 2: CHG is NA at baseline
check2 <- adlb %>%
  filter(ABLFL == "Y", !is.na(CHG))
cat("Check 2 - CHG not NA at baseline:", nrow(check2), "\n")

# Check 3: BASE consistency
check3 <- adlb %>%
  filter(!is.na(BASE)) %>%
  group_by(USUBJID, PARAMCD) %>%
  summarise(n_distinct_base = n_distinct(BASE), .groups = "drop") %>%
  filter(n_distinct_base > 1)
cat("Check 3 - Inconsistent BASE within subject-PARAMCD:", nrow(check3), "\n")

cat("\n✓ Validation complete\n")
```

------------------------------------------------------------------------

## Summary

```{r summary}
cat("\n=== ADLB Summary ===\n\n")
cat("Total records:", nrow(adlb), "\n")
cat("Variables:", ncol(adlb), "\n\n")

# By parameter
adlb %>%
  group_by(PARAMCD) %>%
  summarise(
    N_Records = n(),
    N_Subjects = n_distinct(USUBJID),
    N_Baseline = sum(ABLFL == "Y", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  head(10)

# By visit
adlb %>%
  count(AVISIT, AVISITN) %>%
  arrange(AVISITN)
```

------------------------------------------------------------------------

## Export

```{r export}
# Add labels
attr(adlb$PARAMCD, "label") <- "Parameter Code"
attr(adlb$PARAM, "label") <- "Parameter"
attr(adlb$AVAL, "label") <- "Analysis Value"
attr(adlb$AVALU, "label") <- "Analysis Value Unit"
attr(adlb$ADT, "label") <- "Analysis Date"
attr(adlb$ADY, "label") <- "Analysis Relative Day"
attr(adlb$AVISIT, "label") <- "Analysis Visit"
attr(adlb$AVISITN, "label") <- "Analysis Visit (N)"
attr(adlb$ABLFL, "label") <- "Baseline Record Flag"
attr(adlb$BASE, "label") <- "Baseline Value"
attr(adlb$CHG, "label") <- "Change from Baseline"
attr(adlb$PCHG, "label") <- "Percent Change from Baseline"
attr(adlb$ANRIND, "label") <- "Analysis Reference Range Indicator"
attr(adlb$BNRIND, "label") <- "Baseline Reference Range Indicator"
attr(adlb$ATOXGR, "label") <- "Analysis Toxicity Grade"
attr(adlb$ASEQ, "label") <- "Analysis Sequence Number"

# Export
xportr_write(adlb, path = "adlb.xpt", domain = "ADLB")

cat("\n✓ ADLB exported to: adlb.xpt\n")
```

------------------------------------------------------------------------

## Key Takeaways

1.  BDS structure: One row per subject-parameter-timepoint
2.  ABLFL = last value on or before TRTSDT (one per subject-PARAMCD)
3.  BASE is constant across all records for same subject-PARAMCD
4.  CHG = AVAL - BASE, must be NA at baseline rows
5.  ANRIND compares to reference range (LOW/NORMAL/HIGH)
6.  Toxicity grades based on multiples of ULN (Upper Limit of Normal)

------------------------------------------------------------------------

## Next Steps

**Day 20:** ADVS - Vitals Analysis Dataset\
**Day 21:** ADTTE - Time-to-Event Analysis\
**Day 22:** ADCM and ADRS

------------------------------------------------------------------------

## Resources

-   BDS Structure: <https://www.cdisc.org/standards/foundational/adam>
-   Admiral BDS Vignette: <https://pharmaverse.github.io/admiral/articles/bds_finding.html>
-   NCI CTCAE: <https://ctep.cancer.gov/protocoldevelopment/electronic_applications/ctc.htm>

------------------------------------------------------------------------

**End of Day 19**

___
