---
title: "Day 21: ADTTE - Time-to-Event Analysis Dataset"
subtitle: "Week 3, Day 21: Survival analysis structure with events and censoring"


format:
  html:
    theme:
      light: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css

editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---
::: {style="margin-bottom:12px;"}
< a target="_blank" href="../index.html" class="resource-link"> Back to Roadmap</a>
:::

## Welcome to ADTTE

ADTTE is the dataset for **survival analysis** - time-to-event endpoints like overall survival, progression-free survival, and time to first AE.

**Today's focus:** Build ADTTE with event/censor indicators and time-to-event calculations.

---

## Setup

```{r setup}
library(admiral)
library(pharmaversesdtm)
library(pharmaverseadam)
library(dplyr)
library(lubridate)
library(stringr)
library(xportr)

# Load SDTM and ADaM
dm <- pharmaversesdtm::dm
ae <- pharmaversesdtm::ae
adsl <- pharmaverseadam::adsl

cat("Loaded DM:", nrow(dm), "subjects\n")
cat("Loaded AE:", nrow(ae), "AE records\n")
cat("Loaded ADSL:", nrow(adsl), "subjects\n")
```

---

## Step 1: Start with ADSL

```{r start-adsl}
# ADTTE starts with one row per subject from ADSL
adtte <- adsl %>%
  select(STUDYID, USUBJID, TRTSDT, TRT01A, AGE, SEX)

cat("\nADTTE initialized:", nrow(adtte), "subjects\n")

# Check
adtte %>%
  select(USUBJID, TRTSDT, TRT01A) %>%
  head(5)
```

**ADTTE Structure:**

- One row per subject per parameter
- PARAMCD identifies the endpoint (OS, TTAE, etc.)
- Each parameter has event/censor logic

---

## Step 2: Prepare AE Data

```{r prepare-ae}
# Get first AE date per subject
ae_first <- ae %>%
  admiral::derive_vars_dt(
    new_vars_prefix = "AE",
    dtc = AESTDTC
  ) %>%
  filter(!is.na(AEDT)) %>%
  group_by(STUDYID, USUBJID) %>%
  summarise(
    FIRST_AEDT = min(AEDT),
    .groups = "drop"
  )

cat("\nFirst AE dates prepared\n")

# Check
ae_first %>%
  head(5)
```

---

## Step 3: Merge AE Dates

```{r merge-ae}
# Merge first AE date to ADTTE
adtte <- adtte %>%
  left_join(
    ae_first,
    by = c("STUDYID", "USUBJID")
  )

cat("\nAE dates merged\n")

# Check
adtte %>%
  select(USUBJID, TRTSDT, FIRST_AEDT) %>%
  head(10)
```

---

## Step 4: Create TTAE Parameter

```{r create-ttae}
# Time to First AE parameter
adtte_ttae <- adtte %>%
  mutate(
    PARAMCD = "TTAE",
    PARAM = "Time to First Adverse Event",
    # Event if AE occurred after treatment start
    CNSR = case_when(
      !is.na(FIRST_AEDT) & FIRST_AEDT >= TRTSDT ~ 0,  # Event
      TRUE ~ 1  # Censored
    ),
    # Analysis date
    ADT = case_when(
      CNSR == 0 ~ FIRST_AEDT,  # Event date
      TRUE ~ TRTSDT + 365  # Censor at 1 year (example)
    ),
    # Time to event in days
    AVAL = case_when(
      !is.na(ADT) & !is.na(TRTSDT) ~ as.numeric(ADT - TRTSDT) + 1,
      TRUE ~ NA_real_
    ),
    AVALU = "DAYS"
  )

cat("\nTTAE parameter created\n")

# Summary
adtte_ttae %>%
  count(CNSR) %>%
  mutate(
    Label = case_when(
      CNSR == 0 ~ "Event",
      CNSR == 1 ~ "Censored"
    )
  )
```

**CNSR Convention:**

- CNSR = 0: Event occurred
- CNSR = 1: Censored (no event)

---

## Step 5: Add Event Description

```{r add-event-desc}
# Add event description
adtte_ttae <- adtte_ttae %>%
  mutate(
    EVNTDESC = case_when(
      CNSR == 0 ~ "Adverse Event",
      TRUE ~ NA_character_
    ),
    CNSDTDSC = case_when(
      CNSR == 1 ~ "No Event by Study End",
      TRUE ~ NA_character_
    )
  )

cat("\nEvent descriptions added\n")

# Check
adtte_ttae %>%
  select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC, CNSDTDSC) %>%
  head(10)
```

---

## Step 6: Create Serious AE Parameter

```{r create-ttserae}
# Get first serious AE
ae_serious <- ae %>%
  admiral::derive_vars_dt(
    new_vars_prefix = "AE",
    dtc = AESTDTC
  ) %>%
  filter(AESER == "Y", !is.na(AEDT)) %>%
  group_by(STUDYID, USUBJID) %>%
  summarise(
    FIRST_SERAEDT = min(AEDT),
    .groups = "drop"
  )

# Merge to base
adtte_base <- adtte %>%
  left_join(ae_serious, by = c("STUDYID", "USUBJID"))

# Create parameter
adtte_ttserae <- adtte_base %>%
  mutate(
    PARAMCD = "TTSERAE",
    PARAM = "Time to First Serious Adverse Event",
    CNSR = case_when(
      !is.na(FIRST_SERAEDT) & FIRST_SERAEDT >= TRTSDT ~ 0,
      TRUE ~ 1
    ),
    ADT = case_when(
      CNSR == 0 ~ FIRST_SERAEDT,
      TRUE ~ TRTSDT + 365
    ),
    AVAL = case_when(
      !is.na(ADT) & !is.na(TRTSDT) ~ as.numeric(ADT - TRTSDT) + 1,
      TRUE ~ NA_real_
    ),
    AVALU = "DAYS",
    EVNTDESC = case_when(
      CNSR == 0 ~ "Serious Adverse Event",
      TRUE ~ NA_character_
    ),
    CNSDTDSC = case_when(
      CNSR == 1 ~ "No Serious Event by Study End",
      TRUE ~ NA_character_
    )
  )

cat("\nTTSERAE parameter created\n")

# Summary
adtte_ttserae %>%
  count(CNSR) %>%
  mutate(
    Label = case_when(
      CNSR == 0 ~ "Event",
      CNSR == 1 ~ "Censored"
    )
  )
```

---

## Step 7: Combine Parameters

```{r combine-params}
# Combine TTAE and TTSERAE
adtte_final <- bind_rows(adtte_ttae, adtte_ttserae)

cat("\nParameters combined\n")
cat("Total records:", nrow(adtte_final), "\n")

# Distribution
adtte_final %>%
  count(PARAMCD, PARAM)
```

---

## Step 8: Derive Analysis Sequence

```{r derive-aseq}
# Create sequence number
adtte_final <- adtte_final %>%
  arrange(USUBJID, PARAMCD) %>%
  group_by(USUBJID) %>%
  mutate(ASEQ = row_number()) %>%
  ungroup()

cat("\nAnalysis sequence derived\n")
```

---

## Step 9: Derive Study Day

```{r derive-ady}
# Calculate study day for analysis date
adtte_final <- adtte_final %>%
  mutate(
    ADY = case_when(
      !is.na(ADT) & !is.na(TRTSDT) ~ as.numeric(ADT - TRTSDT) + 1,
      TRUE ~ NA_real_
    )
  )

cat("\nStudy day derived\n")

# Check
adtte_final %>%
  select(USUBJID, PARAMCD, ADT, TRTSDT, ADY, AVAL) %>%
  head(10)
```

---

## Validation

```{r validate}
cat("\n=== ADTTE Validation ===\n\n")

# Check 1: CNSR values
check1 <- adtte_final %>%
  filter(!CNSR %in% c(0, 1))
cat("Check 1 - Invalid CNSR values:", nrow(check1), "\n")

# Check 2: AVAL consistency
check2 <- adtte_final %>%
  filter(!is.na(AVAL) & !is.na(ADY) & abs(AVAL - ADY) > 0.1)
cat("Check 2 - AVAL != ADY:", nrow(check2), "\n")

# Check 3: Event has EVNTDESC
check3 <- adtte_final %>%
  filter(CNSR == 0, is.na(EVNTDESC))
cat("Check 3 - Event missing EVNTDESC:", nrow(check3), "\n")

# Check 4: Censored has CNSDTDSC
check4 <- adtte_final %>%
  filter(CNSR == 1, is.na(CNSDTDSC))
cat("Check 4 - Censored missing CNSDTDSC:", nrow(check4), "\n")

cat("\n✓ Validation complete\n")
```

---

## Summary

```{r summary}
cat("\n=== ADTTE Summary ===\n\n")
cat("Total records:", nrow(adtte_final), "\n")
cat("Parameters:", n_distinct(adtte_final$PARAMCD), "\n")
cat("Subjects:", n_distinct(adtte_final$USUBJID), "\n\n")

# By parameter
adtte_final %>%
  group_by(PARAMCD, PARAM) %>%
  summarise(
    N_Subjects = n(),
    N_Events = sum(CNSR == 0),
    N_Censored = sum(CNSR == 1),
    Median_AVAL = median(AVAL, na.rm = TRUE),
    .groups = "drop"
  )

# AVAL distribution
adtte_final %>%
  group_by(PARAMCD) %>%
  summarise(
    N = sum(!is.na(AVAL)),
    Mean = round(mean(AVAL, na.rm = TRUE), 1),
    Median = round(median(AVAL, na.rm = TRUE), 1),
    Min = min(AVAL, na.rm = TRUE),
    Max = max(AVAL, na.rm = TRUE),
    .groups = "drop"
  )
```

---

## Export

```{r export}
# Add labels
attr(adtte_final$PARAMCD, "label") <- "Parameter Code"
attr(adtte_final$PARAM, "label") <- "Parameter"
attr(adtte_final$AVAL, "label") <- "Analysis Value"
attr(adtte_final$AVALU, "label") <- "Analysis Value Unit"
attr(adtte_final$ADT, "label") <- "Analysis Date"
attr(adtte_final$ADY, "label") <- "Analysis Relative Day"
attr(adtte_final$CNSR, "label") <- "Censor"
attr(adtte_final$EVNTDESC, "label") <- "Event Description"
attr(adtte_final$CNSDTDSC, "label") <- "Censored Description"
attr(adtte_final$ASEQ, "label") <- "Analysis Sequence Number"

# Export
xportr_write(adtte_final, path = "adtte.xpt", domain = "ADTTE")

cat("\n✓ ADTTE exported to: adtte.xpt\n")
```

---

## Key Takeaways

1. ADTTE has one row per subject per parameter
2. CNSR = 0 (event), CNSR = 1 (censored)
3. AVAL = time to event in days (ADT - TRTSDT + 1)
4. Events need EVNTDESC, censored need CNSDTDSC
5. Multiple parameters: TTAE, TTSERAE, etc.
6. Simple dplyr approach without complex admiral TTE functions

---

## Next Steps

**Day 22:** Tables and Listings with gt  
**Day 23:** Figures with ggplot2  
**Day 24:** Complete TLF Package

---

## Resources

- ADTTE Structure: <https://www.cdisc.org/standards/foundational/adam>
- Admiral TTE Vignette: <https://pharmaverse.github.io/admiral/articles/bds_tte.html>
- Survival Analysis in R: <https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html>

---

**End of Day 21**

**Week 3 Complete! You've built:**
- ADSL (Days 16-17)
- ADAE (Day 18)
- ADLB (Day 19)
- ADVS (Day 20)
- ADTTE (Day 21)

Ready for TLF production in Week 4!

< a target="_blank" href="https://visitorbadge.io/status?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse"><img src="https://api.visitorbadge.io/api/visitors?path=https%3A%2F%2Fineelhere.github.io%2F30-days-of-pharmaverse&label=Visits%20till%20date&countColor=%23263759&style=plastic&labelStyle=none" /></a>
