---
title: "Day 20: ADVS - Vitals Analysis Dataset (BDS)"
subtitle: "Week 3, Day 20: BDS structure with visit windows and multiple readings"


format:
  html:
    theme:
      light: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    number-sections: true
    smooth-scroll: true
    highlight-style: github
    page-layout: full
    css: ../styles/site.css

editor: visual
execute:
  echo: true
  eval: true
  warning: false
  message: false
---
::: {style="margin-bottom:12px;"}
<a href="../index.html" class="resource-link"> Back to Roadmap</a>
:::

## Welcome to ADVS

ADVS follows the same **BDS structure** as ADLB, but with additional complexity: visit windows and multiple readings per timepoint.

**Today's focus:** Build ADVS with visit mapping, baseline flags, and change from baseline.

------------------------------------------------------------------------

## Setup

```{r setup}
library(admiral)
library(pharmaversesdtm)
library(pharmaverseadam)
library(dplyr)
library(lubridate)
library(stringr)
library(xportr)

# Load SDTM and ADaM
vs <- pharmaversesdtm::vs
adsl <- pharmaverseadam::adsl

cat("Loaded VS:", nrow(vs), "records\n")
cat("Loaded ADSL:", nrow(adsl), "subjects\n")
```

------------------------------------------------------------------------

## Step 1: Merge ADSL Variables

```{r merge-adsl}
# Start with VS and merge treatment info
advs <- vs %>%
  admiral::derive_vars_merged(
    dataset_add = adsl,
    new_vars = exprs(TRTSDT, TRT01A),
    by_vars = exprs(STUDYID, USUBJID)
  )

cat("\nADVS initialized:", nrow(advs), "records\n")

# Check
advs %>%
  select(USUBJID, VSTESTCD, VSDTC, TRTSDT) %>%
  head(5)
```

------------------------------------------------------------------------

## Step 2: Derive Analysis Date

```{r derive-adt}
# Derive analysis date from VSDTC
advs <- advs %>%
  admiral::derive_vars_dt(
    new_vars_prefix = "A",
    dtc = VSDTC
  )

cat("\nAnalysis date derived\n")

# Check
advs %>%
  select(USUBJID, VSTESTCD, VSDTC, ADT) %>%
  head(5)
```

------------------------------------------------------------------------

## Step 3: Derive Study Day

```{r derive-ady}
# Calculate relative day
advs <- advs %>%
  admiral::derive_vars_dy(
    reference_date = TRTSDT,
    source_vars = exprs(ADT)
  )

cat("\nStudy day derived\n")

# Check
advs %>%
  select(USUBJID, VSTESTCD, TRTSDT, ADT, ADY) %>%
  filter(!is.na(ADY)) %>%
  head(5)
```

------------------------------------------------------------------------

## Step 4: Create PARAM and AVAL

```{r create-param-aval}
# Create BDS variables
advs <- advs %>%
  mutate(
    PARAMCD = VSTESTCD,
    PARAM = VSTEST,
    AVAL = VSSTRESN,
    AVALU = VSSTRESU
  )

cat("\nPARAM and AVAL created\n")

# Check parameters
advs %>%
  count(PARAMCD, PARAM) %>%
  head(10)
```

------------------------------------------------------------------------

## Step 5: Map Analysis Visits

```{r map-avisit}
# Map VISIT to analysis visit
advs <- advs %>%
  mutate(
    AVISIT = case_when(
      VISIT == "SCREENING" ~ "Baseline",
      VISIT == "BASELINE" ~ "Baseline",
      VISIT == "WEEK 2" ~ "Week 2",
      VISIT == "WEEK 4" ~ "Week 4",
      VISIT == "WEEK 8" ~ "Week 8",
      TRUE ~ VISIT
    ),
    AVISITN = case_when(
      AVISIT == "Baseline" ~ 0,
      AVISIT == "Week 2" ~ 2,
      AVISIT == "Week 4" ~ 4,
      AVISIT == "Week 8" ~ 8,
      TRUE ~ as.numeric(VISITNUM)
    )
  )

cat("\nAnalysis visit mapped\n")

# Distribution
advs %>%
  count(AVISIT, AVISITN) %>%
  arrange(AVISITN)
```

**Visit Mapping:**

-   AVISIT: Analysis visit label
-   AVISITN: Analysis visit number for sorting

------------------------------------------------------------------------

## Step 6: Handle Position and Timepoint

```{r derive-position}
# Map position and timepoint
advs <- advs %>%
  mutate(
    ATPT = VSTPT,
    ATPTN = VSTPTNUM
  )

cat("\nPosition and timepoint mapped\n")

# Check
advs %>%
  count(VSPOS, VSTPT, ATPTN) %>%
  head(5)
```

**Multiple Readings:**

-   ATPT: Analysis timepoint (e.g., "PRE-DOSE", "1H POST-DOSE")
-   ATPTN: Timepoint number for ordering

------------------------------------------------------------------------

## Step 7: Derive Baseline Flag

```{r derive-ablfl}
# Flag last value before treatment start
advs <- advs %>%
  arrange(USUBJID, PARAMCD, ADT, ATPTN) %>%
  group_by(USUBJID, PARAMCD) %>%
  mutate(
    ABLFL = case_when(
      !is.na(AVAL) & !is.na(TRTSDT) & ADT <= TRTSDT & 
        row_number() == max(which(ADT <= TRTSDT)) ~ "Y",
      TRUE ~ NA_character_
    )
  ) %>%
  ungroup()

cat("\nBaseline flag derived\n")

# Check
advs %>%
  filter(ABLFL == "Y") %>%
  select(USUBJID, PARAMCD, ADT, TRTSDT, AVAL, ABLFL) %>%
  head(5)
```

------------------------------------------------------------------------

## Step 8: Derive Baseline Value

```{r derive-base}
# Merge baseline value to all records
advs <- advs %>%
  group_by(USUBJID, PARAMCD) %>%
  mutate(
    BASE = AVAL[ABLFL == "Y"][1]
  ) %>%
  ungroup()

cat("\nBaseline value derived\n")

# Check
advs %>%
  select(USUBJID, PARAMCD, AVISIT, AVAL, ABLFL, BASE) %>%
  head(10)
```

------------------------------------------------------------------------

## Step 9: Derive Change from Baseline

```{r derive-chg}
# Calculate change and percent change
advs <- advs %>%
  mutate(
    CHG = case_when(
      !is.na(AVAL) & !is.na(BASE) ~ AVAL - BASE,
      TRUE ~ NA_real_
    ),
    PCHG = case_when(
      !is.na(CHG) & BASE != 0 ~ (CHG / BASE) * 100,
      TRUE ~ NA_real_
    )
  )

cat("\nChange from baseline calculated\n")

# Summary by parameter
advs %>%
  filter(!is.na(CHG)) %>%
  group_by(PARAMCD) %>%
  summarise(
    N = n(),
    Mean_CHG = round(mean(CHG), 1),
    SD_CHG = round(sd(CHG), 1),
    .groups = "drop"
  ) %>%
  head(5)
```

------------------------------------------------------------------------

## Step 10: Derive Reference Range Indicators

```{r derive-anrind}
# Create reference range indicators for BP and HR
advs <- advs %>%
  mutate(
    ANRIND = case_when(
      PARAMCD == "SYSBP" & !is.na(AVAL) ~ case_when(
        AVAL < 90 ~ "LOW",
        AVAL > 140 ~ "HIGH",
        TRUE ~ "NORMAL"
      ),
      PARAMCD == "DIABP" & !is.na(AVAL) ~ case_when(
        AVAL < 60 ~ "LOW",
        AVAL > 90 ~ "HIGH",
        TRUE ~ "NORMAL"
      ),
      PARAMCD == "PULSE" & !is.na(AVAL) ~ case_when(
        AVAL < 60 ~ "LOW",
        AVAL > 100 ~ "HIGH",
        TRUE ~ "NORMAL"
      ),
      TRUE ~ NA_character_
    ),
    BNRIND = case_when(
      PARAMCD == "SYSBP" & !is.na(BASE) ~ case_when(
        BASE < 90 ~ "LOW",
        BASE > 140 ~ "HIGH",
        TRUE ~ "NORMAL"
      ),
      PARAMCD == "DIABP" & !is.na(BASE) ~ case_when(
        BASE < 60 ~ "LOW",
        BASE > 90 ~ "HIGH",
        TRUE ~ "NORMAL"
      ),
      PARAMCD == "PULSE" & !is.na(BASE) ~ case_when(
        BASE < 60 ~ "LOW",
        BASE > 100 ~ "HIGH",
        TRUE ~ "NORMAL"
      ),
      TRUE ~ NA_character_
    )
  )

cat("\nReference range indicators derived\n")

# Distribution
advs %>%
  count(PARAMCD, ANRIND) %>%
  head(10)
```

**Clinical Ranges (Simplified):**

-   Systolic BP: \<90 LOW, 90-140 NORMAL, \>140 HIGH
-   Diastolic BP: \<60 LOW, 60-90 NORMAL, \>90 HIGH
-   Pulse: \<60 LOW, 60-100 NORMAL, \>100 HIGH

------------------------------------------------------------------------

## Step 11: Derive Analysis Sequence

```{r derive-aseq}
# Create sequence number
advs <- advs %>%
  arrange(USUBJID, PARAMCD, AVISITN, ATPTN, ADT) %>%
  group_by(USUBJID) %>%
  mutate(ASEQ = row_number()) %>%
  ungroup()

cat("\nAnalysis sequence derived\n")
```

------------------------------------------------------------------------

## Validation

```{r validate}
cat("\n=== ADVS Validation ===\n\n")

# Check 1: One ABLFL per subject-parameter
check1 <- advs %>%
  filter(ABLFL == "Y") %>%
  count(USUBJID, PARAMCD) %>%
  filter(n > 1)
cat("Check 1 - Multiple ABLFL per subject-PARAMCD:", nrow(check1), "\n")

# Check 2: CHG is NA at baseline
check2 <- advs %>%
  filter(ABLFL == "Y", !is.na(CHG))
cat("Check 2 - CHG not NA at baseline:", nrow(check2), "\n")

# Check 3: BASE consistency
check3 <- advs %>%
  filter(!is.na(BASE)) %>%
  group_by(USUBJID, PARAMCD) %>%
  summarise(n_distinct_base = n_distinct(BASE), .groups = "drop") %>%
  filter(n_distinct_base > 1)
cat("Check 3 - Inconsistent BASE within subject-PARAMCD:", nrow(check3), "\n")

# Check 4: Visit ordering
check4 <- advs %>%
  filter(!is.na(AVISITN)) %>%
  group_by(AVISIT) %>%
  summarise(n_distinct_avisitn = n_distinct(AVISITN), .groups = "drop") %>%
  filter(n_distinct_avisitn > 1)
cat("Check 4 - Multiple AVISITN per AVISIT:", nrow(check4), "\n")

cat("\n✓ Validation complete\n")
```

------------------------------------------------------------------------

## Summary

```{r summary}
cat("\n=== ADVS Summary ===\n\n")
cat("Total records:", nrow(advs), "\n")
cat("Variables:", ncol(advs), "\n\n")

# By parameter
advs %>%
  group_by(PARAMCD, PARAM) %>%
  summarise(
    N_Records = n(),
    N_Subjects = n_distinct(USUBJID),
    N_Baseline = sum(ABLFL == "Y", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  head(10)

# By visit
advs %>%
  count(AVISIT, AVISITN) %>%
  arrange(AVISITN)
```

------------------------------------------------------------------------

## Export

```{r export}
# Add labels
attr(advs$PARAMCD, "label") <- "Parameter Code"
attr(advs$PARAM, "label") <- "Parameter"
attr(advs$AVAL, "label") <- "Analysis Value"
attr(advs$AVALU, "label") <- "Analysis Value Unit"
attr(advs$ADT, "label") <- "Analysis Date"
attr(advs$ADY, "label") <- "Analysis Relative Day"
attr(advs$AVISIT, "label") <- "Analysis Visit"
attr(advs$AVISITN, "label") <- "Analysis Visit (N)"
attr(advs$ATPT, "label") <- "Analysis Timepoint"
attr(advs$ATPTN, "label") <- "Analysis Timepoint (N)"
attr(advs$ABLFL, "label") <- "Baseline Record Flag"
attr(advs$BASE, "label") <- "Baseline Value"
attr(advs$CHG, "label") <- "Change from Baseline"
attr(advs$PCHG, "label") <- "Percent Change from Baseline"
attr(advs$ANRIND, "label") <- "Analysis Reference Range Indicator"
attr(advs$BNRIND, "label") <- "Baseline Reference Range Indicator"
attr(advs$ASEQ, "label") <- "Analysis Sequence Number"

# Export
xportr_write(advs, path = "advs.xpt", domain = "ADVS")

cat("\n✓ ADVS exported to: advs.xpt\n")
```

------------------------------------------------------------------------

## Key Takeaways

1.  ADVS follows same BDS structure as ADLB
2.  Visit mapping: VISIT → AVISIT/AVISITN
3.  Multiple readings handled via ATPT/ATPTN (timepoint)
4.  ABLFL logic same as ADLB (last value ≤ TRTSDT)
5.  Clinical ranges differ by vital sign parameter
6.  Position (SITTING/STANDING) captured in original VS

------------------------------------------------------------------------

## Next Steps

**Day 21:** ADTTE - Time-to-Event Analysis\
**Day 22:** ADCM and ADRS\
**Day 23:** TLF Production

------------------------------------------------------------------------

## Resources

-   BDS Structure: <https://www.cdisc.org/standards/foundational/adam>
-   Admiral BDS Vignette: <https://pharmaverse.github.io/admiral/articles/bds_finding.html>
-   ADVS Examples: <https://pharmaverse.github.io/examples/adam/advs.html>

------------------------------------------------------------------------

**End of Day 20**

___
